### 2021-04-28 citation_fetching.Rmd
fetching citation info

---
title: "Fetch Citation Info"
author: "Mark Ziemann"
date: "28/04/2021"
output: html_document
---

## Intro

Here we are going to fetch the number of citations as well as the publication date for every PMCID.

The code below fetches the number of times that A PMC has been cited.

The output format is

1. PMCID

2. PMID

3. Number of cites

```{r,reutils2}

get_cites <- function(PMCID) {
  mylink <- elink(uid=PMCID , dbFrom="pmc", dbTo = "pubmed")
  PMID <- mylink[1]
  CITES <- elink(uid = PMID , dbFrom = "pubmed"  , linkname = "pubmed_pubmed_citedin" )
  N_CITES <- length(CITES[1])
  return(c(PMCID,PMID,N_CITES))
}

PMCID=3155154

get_cites(PMCID = PMCID)

```

### 2021-04-29 citation_fetching.Rmd
script works

@@
 ---
-title: "Fetch Citation Info"
+title: "Fetch PMC Citation Info"
 author: "Mark Ziemann"
-date: "28/04/2021"
-output: html_document
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    toc: true
+    toc_float: true
+    fig_width: 7
+    fig_height: 7
+theme: cosmo
 ---
@@
 
+library("reutils")
+
 get_cites <- function(PMCID) {
@@
 ```
+
+## Try the following
+
+Now to get a bunch of these.
+
+```{r,reutils3}
+
+PMCIDS <- c("PMC6317205", "PMC6317219", "PMC6318854", "PMC6318897", "PMC6322516", "PMC6323737", "PMC6323741", "PMC6325784", "PMC6325795", "PMC6325889", "PMC6326702", "PMC6327033",
+            "PMC6327913", "PMC6328174", "PMC6328465", "PMC6328499", "PMC6329100", "PMC6329100", "PMC6329125", "PMC6329125", "PMC6329193", "PMC6329348")
+
+PMCIDS <- as.numeric(gsub("PMC","",PMCIDS))
+
+res <- sapply(X = PMCIDS, FUN = get_cites)
+
+res
+```
+
+Looks like it works :)
+
+## Session information
+
+```{r,sessioninfo}
+
+sessionInfo()
+
+```

### 2021-04-29 citation_fetching.Rmd
include publication date

@@
 
-```{r,reutils2}
+4. Pubdate
+
+```{r,library}
 
 library("reutils")
+library("XML")
+library("kableExtra")
+
+```
+
+Here are the functions.
+
+```{r,functions}
+
+get_pub_date <- function(PMCID) {
+  res2 <- esummary(uid=PMCID, db = "pmc")
+  res3 <- content(x=res2,as="text")
+  res5 <- xmlToList(res3)
+  d <- res5$DocumentSummarySet$DocumentSummary$EPubDate
+  d <- gsub(" Jan ","-1-",d)
+  d <- gsub(" Feb ","-2-",d)
+  d <- gsub(" Mar ","-3-",d)
+  d <- gsub(" Apr ","-4-",d)
+  d <- gsub(" May ","-5-",d)
+  d <- gsub(" Jun ","-6-",d)
+  d <- gsub(" Jul ","-7-",d)
+  d <- gsub(" Aug ","-8-",d)
+  d <- gsub(" Sep ","-9-",d)
+  d <- gsub(" Oct ","-10-",d)
+  d <- gsub(" Nov ","-11-",d)
+  d <- gsub(" Dec ","-12-",d)
+  return(as.Date(d))
+}
 
@@
   N_CITES <- length(CITES[1])
-  return(c(PMCID,PMID,N_CITES))
+  PUBDATE <- get_pub_date(PMCID)
+  return(c(PMCID,PMID,N_CITES,PUBDATE))
 }
 
+# test the function
 PMCID=3155154
-
-get_cites(PMCID = PMCID)
+res <- get_cites(PMCID = PMCID)
+res
 
@@
 
-## Try the following
+## Get cites  and pu
 
@@
 
-res <- sapply(X = PMCIDS, FUN = get_cites)
+res <- lapply(X = PMCIDS, FUN = get_cites)
+mx <- do.call(rbind,res)
+mx <- apply(X = mx, MARGIN = 2, FUN = as.numeric)
+df <- as.data.frame(mx)
+colnames(df) <- c("PMCID","PMID","NumCites","Pubdate")
 
-res
+df %>% kbl %>% kable_paper("hover", full_width = F)
+
+plot(df$Pubdate,df$NumCites)
+mylm <- lm(df$NumCites ~ df$Pubdate)
+
+mylm
+
 ```

### 2021-07-15 cleaning_aid.Rmd
started cleaning

---
title: "Cleaning Data"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Intro

here is some background info blah blah

```{r,load data}

x <- read.table("PMC 2019 - analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
head(x)
colnames(x)

# omics type
omics <- x$Omics.type
omics_split <- strsplit(omics,", ")
omics <- unlist(omics_split)
table(omics)

# organism
org <- x$Organism
org_split <- strsplit(org,", ")
org <- unlist(org_split)
table(org)

```

### 2021-07-19 cleaning_aid.Rmd
First changes
codes for gene set library

@@
 colnames(x)
+dim(x)
 
@@
 
+# Gene set library
+GSL <-x$Gene.set.library
+GSL_split <- strsplit(GSL,", ")
+GSL <- unlist(GSL_split)
+table(GSL)
+
 ```

### 2021-07-28 cleaning_aid.Rmd
codes for 'app used'
Gene set version cleaned
gene set version cleaned
fix name change

@@
 
-x <- read.table("PMC 2019 - analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
+x <- read.table("PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
 head(x)
@@
 
+# App used
+App <-x$App.used
+App_split <- strsplit(App,", ")
+App <- unlist(App_split)
+table(App)
+
+# Gene set version
+GSV <-x$GS.version
+table(GSV)
+
 ```

### 2021-07-28 cleaning_aid.Rmd
cleaning_aid.Rmd

@@
 
+```{r,begin}
+
+knitr::opts_chunk$set(fig.width=7, fig.height=7)
+
+```
+
 ```{r,load data}
@@
 
+```
+
+```{r,columns1}
+
+par(mar=c(5,12,3,1))
+
 # omics type
@@
 omics <- unlist(omics_split)
-table(omics)
+res <- table(omics)
+res <- res[order(res)]
+res
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
+        main = "Omics type", xlim=c(0,600))
+grid()
+
+```
+
+```{r,columns2}
 
@@
 table(App)
+```
 
+```{r,columnsX}
+par(mar=c(5,7,3,1))
 # Gene set version
 GSV <-x$GS.version
-table(GSV)
+res <- table(GSV)
+res
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Gene set version defined", xlim=c(0,1600))
+grid()
 

### 2021-07-30 cleaning_aid.Rmd
bar charts section 1

--- 2021-07-28 cleaning_aid.Rmd
@@
 
-here is some background info blah blah
+here is some background info
+
+```{r,begin}
+
+knitr::opts_chunk$set(fig.width=7, fig.height=7)
+
+```
 
@@
 
+```
+
+```{r,columns1}
+
+par(mar=c(5,12,3,1))
 # omics type
@@
 omics <- unlist(omics_split)
-table(omics)
+res <- table(omics)
+res <- res[order(res)]
+res
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
+        main = "Omics type", xlim=c(0,600))
+grid()
+
+```
+
+```{r,columns2}
+par(mar=c(5,12,3,1))
 
@@
 org <- unlist(org_split)
-table(org)
+res <- table(org)
+res <- res[order(res)]
+res
+barplot(tail(res,30),horiz=TRUE,las=1,cex.names = 0.5, xlab ="no. analyses",
+        main = "Organism", xlim=c(0,1000))
+grid()
 
+```
+
+```{r,columns3}
+par(mar=c(5,12,3,1))
 # Gene set library
@@
 GSL <- unlist(GSL_split)
-table(GSL)
+res<- table(GSL)
+res <- res[order(res)]
+res
+barplot(tail(res,30),horiz=TRUE,las=1,cex.names = 0.5, xlab ="no. analyses",
+        main = "Gene set library", xlim=c(0,500))
+grid()
+
+```
+
+```{r,columns4}
+par(mar=c(5,7,3,1))
+
+# Gene set version
+GSV <-x$GS.version
+res <- table(GSV)
+res <- res[order(res)]
+res
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Gene set version", xlim=c(0,1600))
+grid()
+
+```
+```{r,columns5}
+par(mar=c(5,7,3,1))
+
+# Statistical test used
+ST <-x$Statistical.test.used
+res <- table(ST)
+res <- res[order(res)]
+res
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.5, xlab="no. analyses",
+        main = "Statistical test used", xlim=c(0,500))
+grid()
+
+```
+
+```{r,columns6}
+par(mar=c(5,7,3,1))
+
+# FDR correction
+FDR <-x$FDR.Correction
+res <- table(FDR)
+res <- res[order(res)]
+res
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "FDR correction", xlim=c(0,500))
+grid()
+
+```
+
+```{r,columns7}
+par(mar=c(5,12,3,1))
 
@@
 App <- unlist(App_split)
-table(App)
-
-# Gene set version
-GSV <-x$GS.version
-table(GSV)
+res <- table(App)
+res <- res[order(res)]
+res
+barplot(tail(res,30),horiz=TRUE,las=1,cex.names = 0.5, xlab ="no. analyses",
+        main = "App used", xlim=c(0,200))
+grid()
 
 ```
+
+## Session Information
+
+```{r,sessioninfo}
+sessionInfo()
+```

### 2021-07-30 analysis.Rmd

---
title: "PMC Survey Data Analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

## Intro

Here we are performing an analysis of the ways enrichment analysis has been done in a sample of 1500 PMC articles from 2019.

```{r,begin}

knitr::opts_chunk$set(fig.width=7, fig.height=5)

library("wordcloud")
library("RColorBrewer")
library("wordcloud")

```

## Overview of included and excluded analyses

```{r,exclude1}

x <- read.table("PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
head(x)
colnames(x)
dim(x)

exclude <- subset(x,x$GS.version=="EXCLUDE")
nrow(exclude)
length(unique(exclude$Pubmed.Central.ID))

x <- subset(x,x$GS.version!="EXCLUDE")
nrow(x)
length(unique(x$Pubmed.Central.ID))

```

## Omics type

```{r,omics1}

omics <- x$Omics.type
omics_split <- strsplit(omics,", ")
omics <- unlist(omics_split)
res <- table(omics)
res <- res[order(res)]
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Gene expression array","RNA array",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Omics type", xlim=c(0,600))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Omics type", xlim=c(0,600))
grid()

```

## Organism

```{r,org1}

org <- x$Organism
org_split <- strsplit(org,", ")
org <- unlist(org_split)
res <- table(org)
res <- res[order(res)]
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Homo sapiens","human",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"), scale=c(4,.5))

par(mar=c(5,12,3,1))
names(res) <- gsub("human","Homo sapiens",names(res))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Organism", xlim=c(0,1200))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Organism", xlim=c(0,1200))
grid()

```

## Gene set library

```{r,gsl1}

GSL <-x$Gene.set.library
GSL_split <- strsplit(GSL,", ")
GSL <- unlist(GSL_split)
res <- table(GSL)
res <- res[order(res)]
which(names(res)=="Not stated")/sum(res)*100
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Homo sapiens","human",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,1000))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,1200))
grid()

```

## Gene set version

```{r,gsv1}

GSV <-x$GS.version
res <- table(GSV)
res
res[1]/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set version defined", xlim=c(0,1600))
grid()

```

## Statistical test used

```{r,test1}

test <-x$Statistical.test.used
test <- strsplit(test,", ")
test <- unlist(test)
res <- table(test)
res <- res[order(res)]
res[which(names(res)=="Not stated")] / sum(res) * 100
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Test used", xlim=c(0,1200))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Test used", xlim=c(0,1200))
grid()

```

## FDR Correction

```{r,fdr1}

fdr <-x$FDR.Correction
fdr <- strsplit(fdr,", ")
fdr <- unlist(fdr)
res <- table(fdr)
res <- res[order(res)]
sum(res[which(names(res)!="Yes")])/sum(res)*100
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "FDR", xlim=c(0,800))
grid()

```

## App used

```{r,app1}

App <-x$App.used
App_split <- strsplit(App,", ")
App <- unlist(App_split)
res <- table(App)
res <- res[order(res)]
res[which(names(res)=="Not stated")]/sum(res)*100
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "App used", xlim=c(0,400))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "App used", xlim=c(0,1200))
grid()

```

## App version

```{r,appv1}

APV <-x$App.Version
res <- table(APV)
res
res[1]/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App version defined", xlim=c(0,1600))
grid()

```

## Code available

```{r,code1}

code <-x$Code.availability
res <- table(code)
res
res[1]/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Code availability", xlim=c(0,300))
grid()

```

## Background gene set

```{r,background1}

BG <-x$Background.gene.set
res <- table(BG)
res
sum(res[which(names(res)!="Yes")])/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Background list defined", xlim=c(0,1200))
grid()

```

## Gene lists provided

```{r,genelists1}

GL <-x$Gene.lists.provided
res <- table(GL)
res
sum(res[which(names(res)!="Yes")])/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene lists provided", xlim=c(0,1200))
grid()
colnames(x)

```

## Assumptions violated

```{r,assumptions1}

ok <- nrow(subset(x,Assumptions.violated=="No"))
ok
bad <- nrow(subset(x,Assumptions.violated!="No"))
bad
ok/sum(bad,ok)*100

ass <-x$Assumptions.violated
ass <- strsplit(ass,", ")
ass <- unlist(ass)
res <- table(ass)
res <- res[order(res)]
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Assumptions violated", xlim=c(0,1400))
grid()

```

## Session information

```{r,session}

sessionInfo()

```

### 2021-08-04 analysis.Rmd
correcting analysis

@@
 length(unique(x$Pubmed.Central.ID))
+
+```
+
+## Journal
+
+```{r,Journal1}
+
+journal <- x$Journal
+journal_split <- strsplit(journal,", ")
+journal <- unlist(journal_split)
+res <- table(journal)
+res <- res[order(res)]
+res
+
+par(mar=c(1,1,1,1))
+#names(res) <- gsub("Gene expression array","RNA array",names(res))
+wordcloud(words = names(res), freq = res, min.freq = 1, 
+    max.words=200, random.order=FALSE, rot.per=0.35, 
+    colors=brewer.pal(8, "Dark2"))
+
+par(mar=c(5,12,3,1))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
+        main = "Journal", xlim=c(0,120))
+grid()
+
+other <- sum(res[1:(nrow(res)-10)])
+res2 <- c(other,tail(res,9))
+names(res2)[1] <- "Other"
+par(mar=c(5,12,3,1))
+barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
+        main = "Journal", xlim=c(0,1200))
+grid()
 

### 2021-08-05 analysis.Rmd
corrected pmc sheet
minor updates

@@
 res <- res[order(res)]
+length(res)
 res
@@
 res <- res[order(res)]
+length(res)
 which(names(res)=="Not stated")/sum(res)*100
@@
 barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Gene set library", xlim=c(0,1200))
+        main = "Gene set library", xlim=c(0,500))
 grid()
@@
 barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "App used", xlim=c(0,1200))
+        main = "App used", xlim=c(0,500))
 grid()

### 2021-08-05 analysis.Rmd
jif

@@
 library("RColorBrewer")
-library("wordcloud")
+library("wordcloud2")
 
@@
 
+## Journal rank
+
+```{r,journal rank}
+
+jmetrics <- read.table("scimagojr_2020.csv",sep=";",header=TRUE,dec = ",")
+jmetrics$Title <- toupper(jmetrics$Title)
+head(jmetrics)
+dim(jmetrics)
+
+nlmcat <- readLines("nlmcatalog_result.txt")
+journaltitle <- nlmcat[grep("Title\\(s\\):",nlmcat)]
+journaltitle <- gsub("Title\\(s\\): ","",journaltitle)
+journaltitle <- gsub("\\.$","",journaltitle)
+journaltitle <- toupper(journaltitle) 
+head(journaltitle)
+head(nlmcat)
+journalabbrev <- nlmcat[grep("Title Abbreviation:",nlmcat)]
+journalabbrev <- sapply(strsplit(journalabbrev,":"),"[[",2)
+journalabbrev <- gsub(" $","",journalabbrev)
+journalabbrev <- gsub("^ ","",journalabbrev)
+head(journalabbrev)
+jdf <- data.frame(journalabbrev,journaltitle)
+head(jdf)
+
+mjdf <- merge(jdf,jmetrics,by.x="journaltitle",by.y="Title")
+head(mjdf)
+dim(mjdf)
+
+```
+journalabbrev <- medline[grep("Title(s):",medline)]
+
+jdf <- data.frame(journaltitle,journalabbrev)
+
+m <- merge(jmetrics,jdf,by.x="Title",by.y="journaltitle")
+dim(m)
+mm <- m[,c("journalabbrev","SJR")]
+dim(mm)
+
+dim(x)
+xm <- merge(x,mm,by.x="Journal",by.y="journalabbrev")
+dim(xm)
+tail(xm)
+
+dim(x)
+length(which(x$Journal %in% mm$journalabbrev))
+
+x[which(! x$Journal %in% mm$journalabbrev),"Journal"]
+
+mm$journalabbrev
+head(x$Journal)
+
+```
+
 ## Session information

### 2021-08-05 analysis.Rmd
citation analysis

@@
 library("wordcloud2")
+library("reutils")
+library("XML")
+library("kableExtra")
 
@@
 
-```
-journalabbrev <- medline[grep("Title(s):",medline)]
-
-jdf <- data.frame(journaltitle,journalabbrev)
-
-m <- merge(jmetrics,jdf,by.x="Title",by.y="journaltitle")
-dim(m)
-mm <- m[,c("journalabbrev","SJR")]
-dim(mm)
-
-dim(x)
-xm <- merge(x,mm,by.x="Journal",by.y="journalabbrev")
+xm <- merge(x,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)
 dim(xm)
@@
 
-dim(x)
-length(which(x$Journal %in% mm$journalabbrev))
+```
 
-x[which(! x$Journal %in% mm$journalabbrev),"Journal"]
+## Article citations
 
-mm$journalabbrev
-head(x$Journal)
+```{r,cites}
+
+get_pub_date <- function(PMCID) {
+  res2 <- esummary(uid=PMCID, db = "pmc")
+  res3 <- content(x=res2,as="text")
+  res5 <- xmlToList(res3)
+  d <- res5$DocumentSummarySet$DocumentSummary$EPubDate
+  d <- gsub(" Jan ","-1-",d)
+  d <- gsub(" Feb ","-2-",d)
+  d <- gsub(" Mar ","-3-",d)
+  d <- gsub(" Apr ","-4-",d)
+  d <- gsub(" May ","-5-",d)
+  d <- gsub(" Jun ","-6-",d)
+  d <- gsub(" Jul ","-7-",d)
+  d <- gsub(" Aug ","-8-",d)
+  d <- gsub(" Sep ","-9-",d)
+  d <- gsub(" Oct ","-10-",d)
+  d <- gsub(" Nov ","-11-",d)
+  d <- gsub(" Dec ","-12-",d)
+  return(as.Date(d))
+}
+
+get_cites <- function(PMCID) {
+  mylink <- elink(uid=PMCID , dbFrom="pmc", dbTo = "pubmed")
+  PMID <- mylink[1]
+  CITES <- elink(uid = PMID , dbFrom = "pubmed"  , linkname = "pubmed_pubmed_citedin" )
+  N_CITES <- length(CITES[1])
+  PUBDATE <- get_pub_date(PMCID)
+  return(c(PMCID,PMID,N_CITES,PUBDATE))
+}
+
+PMCIDS <- as.numeric(gsub("PMC","",x$Pubmed.Central.ID))
+
+# getcites needs to be fixed so that it can handle all the results
+PMCIDS <- head(PMCIDS,100)
+res <- lapply(X = PMCIDS, FUN = get_cites)
+mx <- do.call(rbind,res)
+mx <- apply(X = mx, MARGIN = 2, FUN = as.numeric)
+df <- as.data.frame(mx)
+colnames(df) <- c("PMCID","PMID","NumCites","Pubdate")
+
+df %>% kbl %>% kable_paper("hover", full_width = F)
+
+plot(df$Pubdate,df$NumCites)
+mylm <- lm(df$NumCites ~ df$Pubdate)
+
+mylm
 

### 2021-08-08 analysis.Rmd
scoring scheme

@@
 
+## Score
+
+Here I am proposing a scoring scheme.
+If there is information missing from the article, or basic mistakes are made, then a point is deducted.
+If the article goes over and above the basic reproducibility, then they are awarded a point.
+
+Gene set library origin not stated = -1
+
+Gene set library version not stated = -1
+
+Stat test not stated = -1
+
+No stat test conducted = -1
+
+No FDR correction conducted = -1
+
+App used not stated = -1
+
+App version not stated = -1
+
+Background list not defined = -1
+
+Inappropriate background list used = -1
+
+Code availability = +1
+
+Gene lists provided = +1
+
+```{r,score}
+
+head(x)
+x1 <- x[1,]
+x1
+
+score <- function(r){
+  r[is.na(r)] <- 0
+  SCORE=0
+  # gene set lib
+  if(r[7]=="Not stated"){
+    SCORE=SCORE-1
+  }
+  # GS version
+  if(r[8]=="No"){
+    SCORE=SCORE-1
+  }
+  # stat test
+  if(r[9]=="No test"){
+    SCORE=SCORE-1
+  }
+  if(r[9]=="Not stated"){
+    SCORE=SCORE-1
+  }
+  # FDR
+  if(r[10]!="Yes"){
+    SCORE=SCORE-1
+  }
+  # app used
+  if(r[11]=="Not stated"){
+    SCORE=SCORE-1
+  }
+  # App version
+  if(r[12]=="No"){
+    SCORE=SCORE-1
+  }
+  # Code availability
+  if(r[13]=="Yes"){
+    SCORE=SCORE+1
+  }
+  # Background
+  if(r[14]=="Not stated"){
+    SCORE=SCORE-1
+  }
+  if(r[14]=="No"){
+    SCORE=SCORE-1
+  }
+  if(r[14]=="Stated, but incorrect"){
+    SCORE=SCORE-1
+  }
+  # gene list
+  # Code availability
+  if(r[16]=="Yes"){
+    SCORE=SCORE+1
+  }
+  return(SCORE)
+}
+
+scores <- apply(X = x,MARGIN = 1, FUN = score)
+hist(scores)
+x$scores <- scores
+
+```
+
 ## Journal rank
@@
 
+## Analysis scores vs Journal rank
+
+```{r,scoresvsjrank}
+
+colnames(xm)
+
+tail(xm$Journal)
+
+mylm1 <- lm (xm$scores ~ xm$SJR)
+plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis quality with journal metrics")
+abline(mylm1,col="red")
+
+cor.test(xm$scores,xm$SJR,method="pearson")
+
+cor.test(xm$scores,xm$SJR,method="spearman")
+```
+## Best and worst journals
+
+Minimum of 5 analyses.
+
+```{r,goodbadugly}
+
+colnames(xm)
+
+xmm <-xm[,c(1,17)]
+tail(xmm)
+tab <- table(xmm$Journal)
+tab <- tab[which(tab>5)]
+xmm <- xmm[which(xmm$Journal %in% names(tab)),]
+head(xmm)
+dim(xmm)
+jres <- aggregate(. ~ Journal, xmm ,mean)
+jres <- jres[order(jres$scores),]
+head(jres)
+dim(jres)
+
+par(mar=c(5,12,3,1))
+barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="scores",
+        main = "Best journals",xlim=c(-4,0))
+grid()
+
+par(mar=c(5,12,3,1))
+barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="scores",
+        main = "Best journals",xlim=c(-5,0))
+grid()
+
+```
+
+## Mean analysis scores vs Journal rank
+
+```{r,scoresvsjrank}
+
+head(jres)
+head(mjdf)
+
+jjres <- merge(jres,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)
+
+head(jjres)
+
+mylm2 <- lm (jjres$scores ~ jjres$SJR)
+plot(jjres$SJR,jjres$scores,xlab="SJR",ylab="score",main="Association of mean analysis quality with journal metrics")
+abline(mylm2,col="red")
+
+cor.test(jjres$scores,jjres$SJR,method="pearson")
+
+cor.test(jjres$scores,jjres$SJR,method="spearman")
+
+```
+
 ## Article citations
@@
   N_CITES <- length(CITES[1])
-  PUBDATE <- get_pub_date(PMCID)
-  return(c(PMCID,PMID,N_CITES,PUBDATE))
+  return(c(PMCID,PMID,N_CITES))
 }
 
+```
+
+```{r,runcites}
+
+if (file.exists("cites_data.rds") ) {
+  res <- readRDS("cites_data.rds")
+} else {
   PMCIDS <- as.numeric(gsub("PMC","",x$Pubmed.Central.ID))
-
-# getcites needs to be fixed so that it can handle all the results
-PMCIDS <- head(PMCIDS,100)
   res <- lapply(X = PMCIDS, FUN = get_cites)
+  saveRDS(res,file="cites_data.rds")
+}
+res
 mx <- do.call(rbind,res)
@@
 mylm
+df$Pubmed.Central.ID <- gsub("^","PMC",df$PMCID)
+head(df)
+
+dfm <- merge(df,xm,by.x=,by.y="Pubmed.Central.ID")
+colnames(dfm)
+
+mylm2 <- lm (dfm$scores ~ dfm$NumCites)
+plot(dfm$NumCites,dfm$scores,xlab="Num cites",ylab="score",main="Association of citations with journal metrics")
+abline(mylm2,col="red")
 

### 2021-08-09 QC analysis.Rmd
QC analysis

--- 2021-08-05 analysis.Rmd
@@
 ---
-title: "PMC Survey Data Analysis"
+  title: "PMC Survey QC Analysis"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
   ## Intro
-
-Here we are performing an analysis of the ways enrichment analysis has been done in a sample of 1500 PMC articles from 2019.
+  Here we are performing an analysis of 200 articles which is randomly selected from 1500 PMC articles.
 
@@
 
-x <- read.table("PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
+x <- read.table("QC-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
 head(x)
@@
 ```
-
 ## Journal
@@
 ```
-
-## Journal rank
-
-```{r,journal rank}
-
-jmetrics <- read.table("scimagojr_2020.csv",sep=";",header=TRUE,dec = ",")
-jmetrics$Title <- toupper(jmetrics$Title)
-head(jmetrics)
-dim(jmetrics)
-
-nlmcat <- readLines("nlmcatalog_result.txt")
-journaltitle <- nlmcat[grep("Title\\(s\\):",nlmcat)]
-journaltitle <- gsub("Title\\(s\\): ","",journaltitle)
-journaltitle <- gsub("\\.$","",journaltitle)
-journaltitle <- toupper(journaltitle) 
-head(journaltitle)
-head(nlmcat)
-journalabbrev <- nlmcat[grep("Title Abbreviation:",nlmcat)]
-journalabbrev <- sapply(strsplit(journalabbrev,":"),"[[",2)
-journalabbrev <- gsub(" $","",journalabbrev)
-journalabbrev <- gsub("^ ","",journalabbrev)
-head(journalabbrev)
-jdf <- data.frame(journalabbrev,journaltitle)
-head(jdf)
-
-mjdf <- merge(jdf,jmetrics,by.x="journaltitle",by.y="Title")
-head(mjdf)
-dim(mjdf)
-
-```
-journalabbrev <- medline[grep("Title(s):",medline)]
-
-jdf <- data.frame(journaltitle,journalabbrev)
-
-m <- merge(jmetrics,jdf,by.x="Title",by.y="journaltitle")
-dim(m)
-mm <- m[,c("journalabbrev","SJR")]
-dim(mm)
-
-dim(x)
-xm <- merge(x,mm,by.x="Journal",by.y="journalabbrev")
-dim(xm)
-tail(xm)
-
-dim(x)
-length(which(x$Journal %in% mm$journalabbrev))
-
-x[which(! x$Journal %in% mm$journalabbrev),"Journal"]
-
-mm$journalabbrev
-head(x$Journal)
-
-```
-
-## Session information
-
-```{r,session}
-
-sessionInfo()
-
-```

### 2021-08-09 analysis.Rmd
update cites

@@
 
-```{r,scoresvsjrank}
+```{r,scoresvsjrank1}
 
@@
 
-```{r,scoresvsjrank}
+```{r,scoresvsjrank2}
 
@@
 
+Get the number of citations per article.
+
 ```{r,cites}
 
-get_pub_date <- function(PMCID) {
-  res2 <- esummary(uid=PMCID, db = "pmc")
-  res3 <- content(x=res2,as="text")
-  res5 <- xmlToList(res3)
-  d <- res5$DocumentSummarySet$DocumentSummary$EPubDate
-  d <- gsub(" Jan ","-1-",d)
-  d <- gsub(" Feb ","-2-",d)
-  d <- gsub(" Mar ","-3-",d)
-  d <- gsub(" Apr ","-4-",d)
-  d <- gsub(" May ","-5-",d)
-  d <- gsub(" Jun ","-6-",d)
-  d <- gsub(" Jul ","-7-",d)
-  d <- gsub(" Aug ","-8-",d)
-  d <- gsub(" Sep ","-9-",d)
-  d <- gsub(" Oct ","-10-",d)
-  d <- gsub(" Nov ","-11-",d)
-  d <- gsub(" Dec ","-12-",d)
-  return(as.Date(d))
+if (file.exists("cites_data.rds") ) {
+  cites <- readRDS("cites_data.rds")
+} else {
+  PMCIDS <- as.numeric(gsub("PMC","",x$Pubmed.Central.ID ))
+  mylink <- elink(uid=PMCIDS , dbFrom="pmc", dbTo = "pubmed")
+  PMID <- mylink[1]
+  #writeLines(PMID,"pubmedids.txt")
+  res <- sapply(X=PMID,FUN=function(x){
+      Sys.sleep(1)
+      CITES <- elink(uid = x , dbFrom = "pubmed"  , linkname = "pubmed_pubmed_citedin" ) 
+      CITES=length(CITES[1])
+      return(CITES)
+   })
+  mylink2 <- elink(uid=PMID , dbFrom="pubmed", dbTo = "pmc")
+  cites <- data.frame(mylink2[1],PMID,res)
+  colnames(cites) <- c("PMCID","PMID","cites")
+  saveRDS(cites,file="cites_data.rds")
 }
 
-get_cites <- function(PMCID) {
-  mylink <- elink(uid=PMCID , dbFrom="pmc", dbTo = "pubmed")
-  PMID <- mylink[1]
-  CITES <- elink(uid = PMID , dbFrom = "pubmed"  , linkname = "pubmed_pubmed_citedin" )
-  N_CITES <- length(CITES[1])
-  return(c(PMCID,PMID,N_CITES))
-}
+cites$PMCID <- gsub("^","PMC",cites$PMCID)
+cites$PMID <- gsub("^","PMID",cites$PMID)
+head(cites)
+xm <- merge(x,cites,by.x="Pubmed.Central.ID",by.y="PMCID")
+head(xm)
+dim(xm)
+```
+Plot the number of citations against analysis scores
 
-```
+```{r,citesplot}
 
-```{r,runcites}
-
-if (file.exists("cites_data.rds") ) {
-  res <- readRDS("cites_data.rds")
-} else {
-  PMCIDS <- as.numeric(gsub("PMC","",x$Pubmed.Central.ID))
-  res <- lapply(X = PMCIDS, FUN = get_cites)
-  saveRDS(res,file="cites_data.rds")
-}
-res
-mx <- do.call(rbind,res)
-mx <- apply(X = mx, MARGIN = 2, FUN = as.numeric)
-df <- as.data.frame(mx)
-colnames(df) <- c("PMCID","PMID","NumCites","Pubdate")
-
-df %>% kbl %>% kable_paper("hover", full_width = F)
-
-plot(df$Pubdate,df$NumCites)
-mylm <- lm(df$NumCites ~ df$Pubdate)
-
-mylm
-df$Pubmed.Central.ID <- gsub("^","PMC",df$PMCID)
-head(df)
-
-dfm <- merge(df,xm,by.x=,by.y="Pubmed.Central.ID")
-colnames(dfm)
-
-mylm2 <- lm (dfm$scores ~ dfm$NumCites)
-plot(dfm$NumCites,dfm$scores,xlab="Num cites",ylab="score",main="Association of citations with journal metrics")
+colnames(xm)
+mylm2 <- lm (xm$scores ~ xm$cites)
+plot(xm$cites,xm$scores,xlab="Num cites",ylab="score",main="Association of citations with journal metrics")
 abline(mylm2,col="red")
 
+cor.test(xm$scores , xm$cites)
 ```

### 2021-08-09 analysis.Rmd
minor improvements

@@
 ---
+
+Source: https://github.com/markziemann/SurveyEnrichmentMethods
 
@@
 library("kableExtra")
+library("vioplot")
 
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Journal", xlim=c(0,120))
+        main = "Journal", xlim=c(0,100))
 grid()
@@
 
-head(x)
-x1 <- x[1,]
-x1
-
 score <- function(r){
@@
   # gene list
-  # Code availability
   if(r[16]=="Yes"){
@@
 jmetrics$Title <- toupper(jmetrics$Title)
-head(jmetrics)
+head(jmetrics,3)
 dim(jmetrics)
@@
 journaltitle <- toupper(journaltitle) 
-head(journaltitle)
-head(nlmcat)
 journalabbrev <- nlmcat[grep("Title Abbreviation:",nlmcat)]
@@
 journalabbrev <- gsub("^ ","",journalabbrev)
-head(journalabbrev)
 jdf <- data.frame(journalabbrev,journaltitle)
-head(jdf)
 
 mjdf <- merge(jdf,jmetrics,by.x="journaltitle",by.y="Title")
-head(mjdf)
-dim(mjdf)
 
 xm <- merge(x,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)
-dim(xm)
-tail(xm)
 
@@
 
-colnames(xm)
-
 tail(xm$Journal)
@@
 mylm1 <- lm (xm$scores ~ xm$SJR)
-plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis quality with journal metrics")
+plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
 abline(mylm1,col="red")
@@
 cor.test(xm$scores,xm$SJR,method="spearman")
+
 ```
@@
 
-colnames(xm)
-
 xmm <-xm[,c(1,17)]
-tail(xmm)
 tab <- table(xmm$Journal)
@@
 xmm <- xmm[which(xmm$Journal %in% names(tab)),]
-head(xmm)
 dim(xmm)
 jres <- aggregate(. ~ Journal, xmm ,mean)
+jsd <- aggregate(. ~ Journal, xmm ,sd)
+jres$sd <- jsd$scores
 jres <- jres[order(jres$scores),]
+dim(jres)
 head(jres)
-dim(jres)
 
 par(mar=c(5,12,3,1))
-barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="scores",
-        main = "Best journals",xlim=c(-4,0))
+barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Highest scoring journals",xlim=c(-4,0))
 grid()
@@
 par(mar=c(5,12,3,1))
-barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="scores",
-        main = "Best journals",xlim=c(-5,0))
+barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Lowest scoring journals",xlim=c(-5,0))
 grid()
+
+j <- unique(xmm$Journal)
+jscores <- sapply(j,function(jj) {
+ xmm[which(xmm$Journal == jj ),2]  
+})
+
+jscores <- jscores[order(unlist(lapply(jscores,mean)))]
+
+par(mar=c(5,12,3,1))
+vioplot(tail(jscores,20),horizontal = TRUE,las=1,cex.axis=0.75,main="Highest scoring journals")
+
+par(mar=c(5,12,3,1))
+vioplot(head(jscores,20),horizontal = TRUE,las=1,cex.axis=0.75,main="Lowest scoring journals")
 
@@
 
-head(jres)
-head(mjdf)
-
 jjres <- merge(jres,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)
 
-head(jjres)
-
 mylm2 <- lm (jjres$scores ~ jjres$SJR)
-plot(jjres$SJR,jjres$scores,xlab="SJR",ylab="score",main="Association of mean analysis quality with journal metrics")
+plot(jjres$SJR,jjres$scores,xlab="SJR",ylab="score",main="Association of mean analysis score with journal metrics")
 abline(mylm2,col="red")
@@
 cites$PMID <- gsub("^","PMID",cites$PMID)
-head(cites)
 xm <- merge(x,cites,by.x="Pubmed.Central.ID",by.y="PMCID")
-head(xm)
 dim(xm)
+
 ```
+
 Plot the number of citations against analysis scores
@@
 
-colnames(xm)
 mylm2 <- lm (xm$scores ~ xm$cites)
-plot(xm$cites,xm$scores,xlab="Num cites",ylab="score",main="Association of citations with journal metrics")
+plot(xm$cites,xm$scores,xlab="Num cites",ylab="score",main="Association of analysis scores with number of citations")
 abline(mylm2,col="red")
 
-cor.test(xm$scores , xm$cites)
+cor.test(xm$scores , xm$cites, method="pearson")
+
+cor.test(xm$scores , xm$cites, method="spearman")
+
+mylm2 <- lm (xm$scores ~ log2(xm$cites))
+plot(log2(xm$cites),xm$scores,xlab="log2(citations)",ylab="score",main="Association of analysis scores with log2(citations)")
+abline(mylm2,col="red")
+
+cor.test(xm$scores , xm$cites, method="pearson")
+
+cor.test(xm$scores , xm$cites, method="spearman")
 ```

### 2021-08-10 analysis.Rmd
working on the manuscript

@@
 
+png("images/journals1.png")
+par(mar=c(5,8,3,1))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Journal", xlim=c(0,1200))
+grid()
+dev.off()
 ```
@@
 grid()
+
+png("images/omics1.png")
+par(mar=c(5,10,3,1))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Omics type", xlim=c(0,600))
+grid()
+dev.off()
 
@@
 
+png("images/organisms1.png")
+par(mar=c(5,10,3,1))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Organism", xlim=c(0,1200))
+grid()
+dev.off()
+
 ```
@@
 barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Gene set library", xlim=c(0,500))
+        main = "Gene set library", xlim=c(0,1000))
 grid()
 
+png("images/genesetlib1.png")
+par(mar=c(5,12,3,1))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Gene set library", xlim=c(0,1000))
+grid()
+dev.off()
 ```
@@
 
+png("images/genesetvers1.png",width = 260, height = 260)
+par(mar=c(5,7,3,1))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Gene set version defined", xlim=c(0,1600))
+grid()
+dev.off()
 ```
@@
 
+png("images/stattest1.png")
+par(mar=c(5,10,3,1))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Test used", xlim=c(0,1200))
+grid()
+dev.off()
 ```
@@
 
+png("images/fdr1.png",width = 260, height = 260)
+par(mar=c(5,7,3,1))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "FDR", xlim=c(0,800))
+grid()
+dev.off()
+
 ```
@@
 res[which(names(res)=="Not stated")]/sum(res)*100
+length(res)
 res
@@
 
+png("images/app1.png")
+par(mar=c(5,12,3,1))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "App used", xlim=c(0,500))
+grid()
+dev.off()
 ```

### 2021-08-10 manuscript.Rmd
working on the manuscript

---
title: "A survey of rigour and reporting standards for gene set enrichment tests"
author: "Kaumadi Wijesooriya, Sameer A Jadaan, Tanuveer Kaur, Kaushalya L Perera, Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichmentMethods

Target Journal: Briefings in Bioinformatics (JIF=9.0)

* Affiliations *

1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences

2. Sameer to provide affiliation information

## Abstract

Gene set enrichment tests (aka functional enrichment analysis) are amongst the most frequently used methods in computational biology.
Despite this popularity, there are concerns that these methods are being applied incorrectly and the results of some peer-reviewed publications are unreliable.
To ascertain the frequency of these errors, we performed a screen of ~1300-1500 articles describing gene set test results.
We find that XX% of articles using over-representation tests implemented the reference gene list incorrectly or did not describe this in the methods.
Failure to perform correction for multiple tests was identified in XX% of studies.
Many studies lacked detail in the methods section about the tools and gene sets used.
Only XX% of studies achieved the satisfactory standard of reporting and rigour.
This highlights the poor state of enrichment test reporting and rigour standards in the recent literature.
We provide a set of minimum standards that should act as a checklist for researchers and peer-reviewers.

## Introduction

Since the turn of the millennium, high throughput omics techniques like gene expression microarrays and next generation sequencing have brought with them a deluge of data. They have revealed how genes are regulated in development and many different diseases.
These experiments involve the measurement of thousands of genes simultaneously, and can identify hundreds or even thousands of genes’ expression being associated with a disease.
Interpreting such data is extraordinarily challenging, as the sheer number of associations can be difficult to investigate in a gene-by-gene manner.
Instead, many different tools have been developed in an effort to summarise gene profiles into simplified functional categories.
These functional categories typically represent signaling or biochemical pathways, curated from information present in the literature, hence the name functional enrichment.

Widely used functional enrichment tools can be classified into two main categories; (i) overrepresentation analysis (ORA) and (ii) functional class scoring (FCS).
In ORA, differentially expressed genes (DEGs) meeting a significance and/or fold change threshold are queried against curated gene sets.
A statistical test is performed to ascertain whether the number of DEGs in a particular gene set is higher than that expected by random chance.
ORA tools can be stand alone software packages or web services, and they use one or more statistical tests (Fisher’s Exact test, hypergeometric test, binomial test, χ2 test, etc) (Draghici et al, 2003; Hosack et al, 2003).
FCS tools involve giving each detected gene a differential expression score and then evaluating whether the scores are more positive or negative than expected by chance for each set of genes.
The popular Gene Set Enrichment Analysis (GSEA) tool uses permutation approaches to establish whether a gene set is associated with higher or lower scores, either by permuting sample labels or by permuting genes in the differential expression profile (Subramanian et al, 2005).
From the user’s perspective, ORA is easier to conduct because it is as simple as pasting a list of gene names into a text box on a website, on the other hand FCS tools are more difficult to use but are more sensitive at detecting subtle associations (Kaspi and Ziemann 2020).

Although these are powerful tools to summarise complex genomics data, there are concerns that they are not being correctly used.
Timmons et al (2015) highlight instances where a failure to account for sampling bias in ORA leads to biased results and likely invalidated conclusions in published work.
Many times as a peer reviewer, we have had to request authors include details about what reference or background gene list was used to account for sampling bias.
In other cases we have requested authors to correct their p-values to account for the possibility of false positives when performing hundreds to thousands of tests in parallel.
The purpose of this work is to determine how frequent errors like lack of a background gene set, lack of adjusting for multiple comparisons and lack of essential methodological details required for reproducibility.
By performing a screen of published articles we will identify the most frequent error modes, and assist in developing a set of minimum standards for functional enrichment analysis (MSFEA).

## Methods

We have collated a list of 2941 articles in PubMed Central published in 2019 that have “enrichment analysis” and related keyword terms. We sampled 1500 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.

* Journal name

* Type of omics data

* Gene set library used, and whether a version was reported

* Statistical test used

* Whether p-values were corrected for multiple comparisons

* Software package used, and whether a version was reported

* Whether a background gene set was used

* Code availability

* Whether gene profile is provided in the supplement

* Whether assumptions are correctly applied

Some articles presented the results of >1 enrichment analysis, so additional rows were added to accommodate them. 
These data were entered into a Google Spreadsheet by a team of five researchers.
For quality control, 200 randomly selected assessed articles were cross-checked by two senior team members to ensure consistency across the five team members.
Data were then cleaned of spelling mistakes and other irregularities by loading them into R and tabulating the common occurrences, then fixing the original spreadsheet.
Cleaned data were then loaded into R v4.1 for downstream analysis. 
We rated each analysis with a simple approach that deducted points for missing methodological details and awarding points for including extra information (Table 1).

| 1 point deducted | 1 point awarded |
| --- | --- |
| Gene set library origin not stated | Code made available |
| Gene set library version not stated | Gene lists provided |
| Stat test not stated | |
| No stat test conducted | |
| No FDR correction conducted | |
| App used not stated | |
| App version not stated | |
| Background list not defined | |
| Inappropriate background list used | |

Scimago Journal Ranks were downloaded from the Scimago website (https://www.scimagojr.com/journalrank.php) and used to rank journals by their citation metrics.
Using NCBI’s Eutils API, we collected the number of citations each article accrued since publication.
We used Spearman and Pearson correlation tests to determine any association with the analysis scores we generated.
All data analyses were conducted in R v4.1 and the analysis scripts are available at GitHub (https://github.com/markziemann/SurveyEnrichmentMethods).

## Results

### Dataset overview

A search of PubMed Central showed 2941 articles published in 2019 with the keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
From these, 1500 articles were screened for methodological errors.
We excluded 133 articles from the screen because they did not present any enrichment analysis. 
Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
As some articles included more than one enrichment analysis, the final dataset included 1624 analyses from 1363 articles.
There were articles from 330 journals in the screen, with *Scientific Reports*, *Oncology Letters* and *International Journal of Molecular Science* being the biggest contributors (Figure 1).
In the analyses, there were 35 different omics types, with RNA-seq and gene expression microarray being the most popular(Figure 2).
There were 175 different species under study, but human was the most common with 1104 analyses (Figure 3).
We recorded the use of 101 different gene set libraries, with GO and KEGG being the most frequently used (Figure 4).
There were 85 studies where the gene set libraries used were not defined in the article.
Only 108 analyses reported the version of the gene set library used (Figure 5).
There were XX different statistical tests used, and the most common reported tests were Fisher's Exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Figure 6).
Only 754 of 1594 studies described performing FDR or other p-value correction for multiple testing (Figure 7).
There were 145 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 174 analyses (XX%) did not state what tool was used (Figure 8).

![Figure 1. Representation of journals in the data set.](images/journals1.png "Journals")

![Figure 2. Representation of different omics types in the data set.](images/omics1.png "Omics")

![Figure 3. Representation of different organisms in the data set.](images/organisms1.png "Organisms")

![Figure 4. Representation of different gene set libraries in the data set.](images/genesetlib1.png "Geneset libraries")

![Figure 5. Proportion of analyses that reported version information for gene sets used.](images/genesetvers1.png "Geneset version")

![Figure 6. Representation of different statistical tests in the data set.](images/stattest1.png "Geneset libraries")

![Figure 7. Recognition of p-value correction for multiple testing in the data set.](images/fdr1.png "FDR")

![Figure 8. Representation of different software tools for enrichment analysis in the data set.](images/app1.png "Software used")

app1.png

## Discussion

## References

Draghici, S., Khatri, P., Martins, R. P., Ostermeier, G. C., & Krawetz, S. A. (2003). Global functional profiling of gene expression. Genomics, 81(2), 98–104.

Hosack, D. A., Dennis, G., Jr, Sherman, B. T., Lane, H. C., & Lempicki, R. A. (2003). Identifying biological themes within lists of genes with EASE. Genome Biology, 4(10), R70.

Huang, D. W., Sherman, B. T., & Lempicki, R. A. (2009). Bioinformatics enrichment tools: paths toward the comprehensive functional analysis of large gene lists. Nucleic Acids Research, 37(1), 1–13.

Kaspi, A., & Ziemann, M. (2020). Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data. BMC Genomics, 21(1), 447.

Subramanian, A., Tamayo, P., Mootha, V. K., Mukherjee, S., Ebert, B. L., Gillette, M. A., … Mesirov, J. P. (2005). Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. Proceedings of the National Academy of Sciences of the United States of America, 102(43), 15545–15550.

Timmons, J. A., Szkop, K. J., & Gallagher, I. J. (2015). Multiple sources of bias confound functional enrichment analysis of global -omics data. Genome Biology, 16(1), 186.

Xie, C., Jauhari, S., & Mora, A. (2021). Popularity and performance of bioinformatics software: the case of gene set analysis. BMC Bioinformatics, 22(1), 191.

### 2021-08-10 analysis.Rmd
many improvements

@@
 res[which(names(res)=="Not stated")]/sum(res)*100
+length(unique(subset(x,App.used=="Not stated")$Journal))
+unique(subset(x,App.used=="Not stated")$Journal)
 length(res)
@@
 dev.off()
+
+length(unique(subset(x,App.used=="Not stated")$Journal))
+
 ```
@@
 grid()
+
+png("images/appvers1.png",width = 260, height = 260)
+par(mar=c(5,7,3,1))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "App version defined", xlim=c(0,1600))
+grid()
+dev.off()
 
@@
 
+png("images/code1.png",width = 260, height = 260)
+par(mar=c(5,7,3,1))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Code availability", xlim=c(0,300))
+grid()
+dev.off()
+
 ```
@@
 
+png("images/bg1.png")
+par(mar=c(5,12,3,1))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Background list defined", xlim=c(0,1200))
+grid()
+dev.off()
 ```
@@
 colnames(x)
+
+png("images/genelists1.png",width = 260, height = 260)
+par(mar=c(5,7,3,1))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Gene lists provided", xlim=c(0,1200))
+grid()
+dev.off()
 
@@
 grid()
+
+png("images/assumptions1.png")
+par(mar=c(5,12,3,1))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Assumptions violated", xlim=c(0,1400))
+grid()
+dev.off()
 
@@
 
+mean(x$scores)
+sd(x$scores)
+median(x$scores)
+
+png("images/hist1.png")
+par(mar=c(5,5,3,1))
+hist(scores)
+grid()
+dev.off()
+
 ```
@@
 
+png("images/score_sjr1.png")
+par(mar=c(5,5,3,1))
+plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
+abline(mylm1,col="red")
+grid()
+dev.off()
+
 ```
@@
 
+png("images/score_highest1.png")
+par(mar=c(5,12,3,1))
+barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Highest scoring journals",xlim=c(-4,0))
+grid()
+dev.off()
+
 par(mar=c(5,12,3,1))
@@
 grid()
+
+png("images/score_lowest1.png")
+par(mar=c(5,12,3,1))
+barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Lowest scoring journals",xlim=c(-5,0))
+grid()
+dev.off()
 
@@
 mylm2 <- lm (jjres$scores ~ jjres$SJR)
-plot(jjres$SJR,jjres$scores,xlab="SJR",ylab="score",main="Association of mean analysis score with journal metrics")
+plot(jjres$SJR,jjres$scores,xlab="SJR",ylab="mean score",main="Association of mean analysis score with journal metrics")
 abline(mylm2,col="red")
+
+png("images/mean_sjr1.png")
+par(mar=c(5,5,3,1))
+plot(jjres$SJR,jjres$scores,xlab="SJR",ylab="mean score",main="Association of mean analysis score with journal metrics")
+abline(mylm2,col="red")
+dev.off()
 
@@
 
+png("images/cites1.png")
+par(mar=c(5,5,3,1))
+plot(log2(xm$cites),xm$scores,xlab="log2(citations)",ylab="score",main="Association of analysis scores with log2(citations)")
+abline(mylm2,col="red")
+dev.off()
+
 cor.test(xm$scores , xm$cites, method="pearson")

### 2021-08-10 manuscript.Rmd
many improvements

@@
 
-* Affiliations *
+**Affiliations**
 
@@
 
-Gene set enrichment tests (aka functional enrichment analysis) are amongst the most frequently used methods in computational biology.
+Gene set enrichment tests (a.k.a. functional enrichment analysis) are amongst the most frequently used methods in computational biology.
 Despite this popularity, there are concerns that these methods are being applied incorrectly and the results of some peer-reviewed publications are unreliable.
-To ascertain the frequency of these errors, we performed a screen of ~1300-1500 articles describing gene set test results.
-We find that XX% of articles using over-representation tests implemented the reference gene list incorrectly or did not describe this in the methods.
-Failure to perform correction for multiple tests was identified in XX% of studies.
+To ascertain the frequency of these errors, we performed a screen of 1363 open access research articles describing gene set test results.
+We find that 95% of analyses using over-representation tests did not implement an appropriate reference gene list or did not describe this in the methods.
+Failure to perform p-value correction for multiple tests was identified in 53% of analyses
 Many studies lacked detail in the methods section about the tools and gene sets used.
-Only XX% of studies achieved the satisfactory standard of reporting and rigour.
-This highlights the poor state of enrichment test reporting and rigour standards in the recent literature.
+Only 10% of studies achieved the satisfactory standard of reporting and rigour.
+This highlights the poor state of enrichment test reporting and rigour standards in the contemporary literature.
 We provide a set of minimum standards that should act as a checklist for researchers and peer-reviewers.
@@
 Timmons et al (2015) highlight instances where a failure to account for sampling bias in ORA leads to biased results and likely invalidated conclusions in published work.
-Many times as a peer reviewer, we have had to request authors include details about what reference or background gene list was used to account for sampling bias.
+Many times as peer reviewers, we have had to request authors include details about what reference or background gene list was used to account for sampling bias.
 In other cases we have requested authors to correct their p-values to account for the possibility of false positives when performing hundreds to thousands of tests in parallel.
-The purpose of this work is to determine how frequent errors like lack of a background gene set, lack of adjusting for multiple comparisons and lack of essential methodological details required for reproducibility.
-By performing a screen of published articles we will identify the most frequent error modes, and assist in developing a set of minimum standards for functional enrichment analysis (MSFEA).
+The purpose of this work is to determine how frequent methodological issues in functional enrichment are, including lack of a background gene set, lack of adjustment for multiple comparisons and lack of essential methodological details.
+By performing a screen of published articles we will identify the most frequent error modes, and use this to assist in developing a set of minimum standards for functional enrichment analysis (MSFEA).
 
@@
 A search of PubMed Central showed 2941 articles published in 2019 with the keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
-From these, 1500 articles were screened for methodological errors.
+From these, 1500 articles were screened for methodological errors; data supplied in Supplementary Table 1.
 We excluded 133 articles from the screen because they did not present any enrichment analysis. 
@@
 There were articles from 330 journals in the screen, with *Scientific Reports*, *Oncology Letters* and *International Journal of Molecular Science* being the biggest contributors (Figure 1).
-In the analyses, there were 35 different omics types, with RNA-seq and gene expression microarray being the most popular(Figure 2).
+There were 35 different omics types, with RNA-seq and gene expression microarray being the most popular (Figure 2).
 There were 175 different species under study, but human was the most common with 1104 analyses (Figure 3).
@@
 There were 145 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 174 analyses (XX%) did not state what tool was used (Figure 8).
+The version of the software used was provided in only 421 of 1611 analyses (26%) (Figure 9).
+Of the 293 analyses which used a scripted computer language, only 10 provided links to the code used for enrichment analysis (3.4%) (Figure 10).
+For analyses using ORA methods, we studied what background gene set was used (Figure 11). 
+This revealed that in most cases, the background list was not stated or it was clear from the article methods section that no background list was used.
+In a few cases, background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
+In only 67/1460 of cases (4.6%), the appropriate background list was described in the article.
+For 579 of 1618 analyses, the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Figure 12).
+Next, we quantified the frequency of methodological issues that would undermine the conclusions (Figure 13).
+Lack of appropriate background was the most common issue (1361 cases), followed by FDR (761 cases), then inference without test (61 cases), inference without data shown (32 cases) and misinterpreted FDR values (3 cases).
+Only 160 analyses did not exhibit any of these major methodological issues.
 
@@
 
-app1.png
+![Figure 9. Proportion of analyses that reported version information for software used.](images/appvers1.png "Software version")
+
+![Figure 10. Proportion of scripted analyses that provided software code.](images/code1.png "Code available")
+
+![Figure 11. Background gene set usage and reporting in the dataset.](images/bg1.png "Background gene list used")
+
+![Figure 12. Proportion of scripted analyses that provided gene profiles.](images/genelists1.png "Gene lists provided")
+
+![Figure 13. Proportion of analyses with different methodological problems.](images/assumptions1.png "Assumptions violated")
+
+### Analysis scores
+
+We then scored each analysis based on the inclusion or lack of methodological issues and included details.
+The median score was -4, with a mean of -3.5 and standard deviation of 1.3 (Figure 14).
+Next, we assessed whether these analysis scores associate with journal citation metrics (Figure 15).
+There was a slight positive correlation using the Pearson method but it was not statistically significant (r=0.04, p=0.16).
+Next we wanted to know which journals had the highest and lowest scores.
+Only journals with five or more analyses were included.
+The best scoring journals were *Transl Psychiatry*, *Metabolites* and *J Exp Clin Cancer Res* (Figure 16), while the poorest were *Mol Med Rep*, *Cancer Med* and *World J Gastroenterol* (Figure 17), although we note that there was a wide variation between articles of the same journal.
+Then, we assessed for an association between mean analysis score and the SJR (Figure 17).
+Again, there was a slightly positive association that wasn't statistically significant (Pearson r=0.09, p=0.55).
+Next we assessed whether there was any association between analysis scores and the number of citations received by the articles.
+After log transforming the citation data, there was no association between citations and analysis scores (Pearson r=0.002, p=0.94).
+
+![Figure 14. Distribution of analysis scores.](images/hist1.png "Analysis scores")
+
+![Figure 15. Association of analysis scores with journal rank.](images/score_sjr1.png "Analysis scores versus journal rank")
+
+![Figure 16. Journals with highest analysis scores.](images/score_highest1.png "Highest analysis scores")
+
+![Figure 17. Journals with Lowest analysis scores.](images/score_lowest1.png "Lowest analysis scores")
+
+![Figure 18. Association between SJR and mean analysis score for journals with >=5 analyses.](images/mean_sjr1.png "Mean score vs SJR")
 
@@
 
+While there have been several articles evaluating the performance of functional enrichment tools, there has been relatively little interest in how existing tools are (mis)used. 
+A recent study looked at the the popularity of frequently used tools and their performance (Xie et al, 2021), but this assumes that the tools were properly used.
+In this sample of open access research articles, there appears to be a bias toward tools that are easy to use.
+ORA tools that only require pasting lists of gene identifiers into a webpage (ie: DAVID, IPA, KOBAS and PANTHER) are collectively more popular than other solutions like GSEA (a standalone graphical user interface software for FCS) or any command line tool.
+This is despite the fact that ORA tools perform worse than FCS and new generation pathway topology tools (Kaspi & Ziemann 2020; Maleki et al, 2020; Xie et al, 2021).
+
+The lack of methodological details regarding enrichment analysis in some articles was disturbing.
+To illustrate this, many articles mention prominently in the abstract the results of the functional enrichment analysis yet not detail in their methods sections any detail whatsoever about what type of enrichment analysis was actually conducted.
+This is not limited to a few obscure journals; there were 88 journals that published articles where the tool used for enrichment analysis was not defined.
+
+Surprisingly, the analysis scores we generated did not correlate with a journal level bibliographic metric, nor accrued citations of individual articles.
+This suggests no association between methodological rigour and "impact" in this sample.
+
+Still there are some limitations that need to be recognised.
+Many open access articles examined here are from journals not considered to be highly prestigious, and the distribution of analysis scores might be different for closed access articles.
+The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in more prestigious journals contain such details in the abstract at lower rates.
+
+Nevertheless, these results are a wake up call for reproducibility and highlight the urgent need for minimum standards for functional enrichment analysis (MSFEA). 
+To address this, we provide a set of minimum standards for functional enrichment analysis, as well as additional recommendations to achieve gold standard reliability and reproducibility (Box 1).
+
+<hr>
+
+### Essential minimum standards for functional enrichment analysis
+
+The following guidelines are selected to avoid the most severe methodological issues while requiring minimal effort to address.
+
+1. Report the origin of the gene sets and the version used or date downloaded.
+These databases are regularly upgraded so this is important for reproducibility.
+
+2. Report the tool used and its software version.
+As these tools are regularly updated, this will aid reproducibility.
+
+3. If making claims about the regulation of a gene set, then results of a statistical test must be shown (including measures of significance and effect size).
+"Data not shown" is not acceptable when it is possible to attach supplementary files.
+
+4. Report which statistical test was used.
+This will help long term reproducibility, for example if in future the tool is no longer available.
+This is especially relevant for tools that report the results of different statistical tests.
+
+5. Always report FDR or q-values (p-values that have been corrected for multiple tests).
+This will reduce the chance of false positives when performing many tests simultaneously.
+Bonferonni, FDR and FWER are examples.
+
+6. If ORA is used, it is vital to use a background list consisting of all detected genes.
+Do not use tools that cannot accept a background list.
+
+7. Criteria and parameters matter.
+If performing ORA, then the inclusion thresholds need to be mentioned.
+Moreover it needs to be mentioned whether up- and down-regulated genes are combined or separated into different tests.
+If using GSEA, parameters around gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels).
+
+### Additional recommendations for enhanced reproducibility
+
+The following guidelines are directed at going over and above the minimum standards towards a gold standard of reliability and reproducibility.
+
+8. Preference FCS and pathway topology tools over ORA tools.
+ORA tools have a lower sensitivity. 
+In practice, they are rarely used properly.
+FCS tests distinguish between up- and down-regulated gene sets by default.
+
+9. Include the gene profile data (including any background lists) in the supplementary data in TSV or CSV formats.
+Excel spreadsheets are not recommended for genomics work (Ziemann et al, 2016)
+
+10.
+Scripted analysis workflows are preferred over analysis involving graphical user interfaces because they can attain a higher degree of computational reproducibility, so consider investing in upgrading your analysis.
+Similarly, free and open source software tools are preferred over proprietary software.
+
+11. If using a scripted analysis, provide code by depositing it to a repository like GitHub or Zenodo.
+Link the data to the code, so anyone can download and reproduce your analysis (Peng 2011).
+
+<hr>
+
+## Conclusions
+
+We echo the sentiments of Timmons et al (2015) that functional enrichment analysis results should be interpreted with extreme caution, and that they shouldn't be considered proof of biological plausibility nor validity of omics data analysis.
+Functional enrichment analysis is best used as a hypothesis generating procedure to inform subsequent biochemical or signaling experimentation.
+
 ## References
 
+References
 Draghici, S., Khatri, P., Martins, R. P., Ostermeier, G. C., & Krawetz, S. A. (2003). Global functional profiling of gene expression. Genomics, 81(2), 98–104.
@@
 Kaspi, A., & Ziemann, M. (2020). Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data. BMC Genomics, 21(1), 447.
+
+Maleki, F., Ovens, K., Hogan, D. J., & Kusalik, A. J. (2020). Gene set analysis: Challenges, opportunities, and future research. Frontiers in Genetics, 11, 654.
+
+Peng, R. D. (2011). Reproducible research in computational science. Science (New York, N.Y.), 334(6060), 1226–1227.
 
@@
 Xie, C., Jauhari, S., & Mora, A. (2021). Popularity and performance of bioinformatics software: the case of gene set analysis. BMC Bioinformatics, 22(1), 191.
+
+Ziemann, M., Eren, Y., & El-Osta, A. (2016). Gene name errors are widespread in the scientific literature. Genome Biology, 17(1), 177.

### 2021-08-11 manuscript.Rmd
minor update to ms

@@
 
-6. If ORA is used, it is vital to use a background list consisting of all detected genes.
-Do not use tools that cannot accept a background list.
+6. If ORA is used, it is vital to use a background list consisting of all detected genes, and report this in the methods section.
+Avoid tools that cannot accept a background list.
 
@@
 If performing ORA, then the inclusion thresholds need to be mentioned.
-Moreover it needs to be mentioned whether up- and down-regulated genes are combined or separated into different tests.
-If using GSEA, parameters around gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels).
+If using GSEA or another FCS tool, parameters around ranking, gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels).
+
+8. If using ORA, it needs to be mentioned whether up- and down-regulated genes are separated into different tests as recommended by Hong et al (2014) or combined.
 
@@
 
-8. Preference FCS and pathway topology tools over ORA tools.
+9. Preference FCS and pathway topology tools over ORA tools.
 ORA tools have a lower sensitivity. 
@@
 
-9. Include the gene profile data (including any background lists) in the supplementary data in TSV or CSV formats.
+10. Include the gene profile data (including any background lists) in the supplementary data in TSV or CSV formats.
 Excel spreadsheets are not recommended for genomics work (Ziemann et al, 2016)
 
-10.
+11.
 Scripted analysis workflows are preferred over analysis involving graphical user interfaces because they can attain a higher degree of computational reproducibility, so consider investing in upgrading your analysis.
@@
 
-11. If using a scripted analysis, provide code by depositing it to a repository like GitHub or Zenodo.
+12. If using a scripted analysis, provide code by depositing it to a repository like GitHub or Zenodo.
 Link the data to the code, so anyone can download and reproduce your analysis (Peng 2011).

### 2021-08-11 example.Rmd
added an example analysis on why BG is so important!

---
title: "PMC Survey Data Analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichmentMethods

## Intro

Here we are performing an analysis of some gene expression data to demonstrate the difference between ORA and FCS methods and to highlight the differences caused by improper background gene set use.

```{r,begin}

library("getDEE2") 
library("DESeq2")
library("clusterProfiler")
library("mitch")
library("kableExtra")
library("eulerr")

```

## Get expression data

I'm using some RNA-seq data looking at the effect of hyperglycemia on hepatocytes.

```{r,getdata}

mdat<-getDEE2Metadata("hsapiens")

#SRP128998
#samplesheet<-mdat[which(mdat$GEO_series=="GSE109140"),]
samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
samplesheet$HG<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))

# samplesheets
s1 <- subset(samplesheet,VPA==0)

w<-getDEE2("hsapiens",samplesheet$SRR_accession,metadata=mdat,legacy = TRUE)
x<-Tx2Gene(w)
x<-x$Tx2Gene

# save the genetable for later
gt<-w$GeneInfo[,1,drop=FALSE]
gt$accession<-rownames(gt)
head(gt)

# counts 
x1<-x[,which(colnames(x) %in% s1$SRR_accession)]

# filter out lowly expressed genes
x1<-x1[which(rowSums(x1)/ncol(x1)>=(10)),]

#run DESeq2 for LG vs HG
y <- DESeqDataSetFromMatrix(countData = round(x1), colData = s1, design = ~ HG)
y <- DESeq(y)
LGvHG <- results(y)
LGvHG<-as.data.frame(LGvHG[order(LGvHG$pvalue),])
rownames(LGvHG)<-sapply(strsplit(rownames(LGvHG),"\\."),"[[",1)
head(LGvHG)

```

## Gene sets from Reactome

```{r,reactome}

download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
unzip("ReactomePathways.gmt.zip")
genesets<-gmt_import("ReactomePathways.gmt")

```

## FCS with Mitch

```{r,mitch1}
library(kableExtra)

m <- mitch_import(LGvHG,DEtype = "DEseq2", geneTable = gt)
head(m)
mres <- mitch_calc(m,genesets = genesets)
head(mres$enrichment_result,50)
m_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
m_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
length(m_up)
length(m_dn)

```

## ORA with clusterprofiler

```{r,cp1}

genesets2 <- read.gmt("ReactomePathways.gmt")
head(genesets2)

head(LGvHG)

LGvHG_up <- rownames(subset(LGvHG,log2FoldChange>0,padj<0.05))
LGvHG_up <- unique(gt[which(rownames(gt) %in% LGvHG_up),1])
head(LGvHG_up)

LGvHG_dn <- rownames(subset(LGvHG,log2FoldChange<0,padj<0.05))
LGvHG_dn <- unique(gt[which(rownames(gt) %in% LGvHG_dn),1])
head(LGvHG_dn)

LGvHG_bg <- rownames(LGvHG)
LGvHG_bg <- unique(gt[which(rownames(gt) %in% LGvHG_bg),1])
length(LGvHG_bg)
head(LGvHG_bg)

c_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg, TERM2GENE = genesets2)
c_up <- head(c_up,1000)
head(c_up,20)
c_up <- rownames(subset(c_up, p.adjust < 0.05))

c_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, TERM2GENE = genesets2)
c_dn <- head(c_dn,1000)
head(c_up,20)
c_dn <- rownames(subset(c_dn, p.adjust < 0.05))

```

## ORA with clusterprofiler with whole genome background list

```{r,cp1}

LGvHG_bg <- w$GeneInfo$GeneSymbol
length(LGvHG_bg)
head(LGvHG_bg)

f_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg, TERM2GENE = genesets2)
f_up <- head(f_up,1000)
head(f_up,20)
f_up <- rownames(subset(f_up, p.adjust < 0.05))

f_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, TERM2GENE = genesets2)
f_dn <- head(f_dn,1000)
head(f_up,20)
f_dn <- rownames(subset(f_dn, p.adjust < 0.05))

```
## Venn diagram comparison

```{r,venn1}

v1 <- list("FCS up"=m_up, "FCS dn"=m_dn,
           "ORA up"=c_up,"ORA dn"=c_dn)

plot(euler(v1),quantities = TRUE)

v2 <- list("ORA up"=c_up,"ORA dn"=c_dn, 
           "ORA* up"=f_up,"ORA* dn"=f_dn )

plot(euler(v2),quantities = TRUE)

v3 <- list("FCS up"=m_up, "FCS dn"=m_dn,
           "ORA up"=c_up,"ORA dn"=c_dn, 
           "ORA* up"=f_up,"ORA* dn"=f_dn )

plot(euler(v3),quantities = TRUE)

```

## Session information

```{r,session}

sessionInfo()

```

### 2021-08-11 manuscript.Rmd
refs
added an example analysis on why BG is so important!

@@
 
-Since the turn of the millennium, high throughput omics techniques like gene expression microarrays and next generation sequencing have brought with them a deluge of data.
+Since the turn of the millennium, high throughput omics techniques like microarrays and next generation sequencing have brought with them a deluge of data.
 They have revealed how genes are regulated in development and many different diseases.
-These experiments involve the measurement of thousands of genes simultaneously, and can identify hundreds or even thousands of genes’ expression being associated with a disease.
+These experiments involve the measurement of thousands of genes simultaneously, and can identify hundreds or even thousands of significant associations.
 Interpreting such data is extraordinarily challenging, as the sheer number of associations can be difficult to investigate in a gene-by-gene manner.
@@
 
-Widely used functional enrichment tools can be classified into two main categories; (i) overrepresentation analysis (ORA) and (ii) functional class scoring (FCS).
+Widely used functional enrichment tools can be classified into two main categories; (i) over-representation analysis (ORA) and (ii) functional class scoring (FCS), and the most common application is in differential expression analysis.
 In ORA, differentially expressed genes (DEGs) meeting a significance and/or fold change threshold are queried against curated gene sets.
@@
 The popular Gene Set Enrichment Analysis (GSEA) tool uses permutation approaches to establish whether a gene set is associated with higher or lower scores, either by permuting sample labels or by permuting genes in the differential expression profile (Subramanian et al, 2005).
-From the user’s perspective, ORA is easier to conduct because it is as simple as pasting a list of gene names into a text box on a website, on the other hand FCS tools are more difficult to use but are more sensitive at detecting subtle associations (Kaspi and Ziemann 2020).
+From a user's perspective, ORA is easier to conduct because it is as simple as pasting a list of gene names into a text box on a website.
+On the other hand FCS tools are more difficult to use but are more sensitive at detecting subtle associations (Kaspi and Ziemann 2020; Maleki et al, 2020; Xie et al, 2021).
 
 Although these are powerful tools to summarise complex genomics data, there are concerns that they are not being correctly used.
-Timmons et al (2015) highlight instances where a failure to account for sampling bias in ORA leads to biased results and likely invalidated conclusions in published work.
+Timmons et al (2015) highlight instances where a failure to account for sampling bias in ORA leads to biased results and invalidated conclusions in published work.
 Many times as peer reviewers, we have had to request authors include details about what reference or background gene list was used to account for sampling bias.
-In other cases we have requested authors to correct their p-values to account for the possibility of false positives when performing hundreds to thousands of tests in parallel.
+In other cases we have requested authors to correct their p-values to account for the possibility of false positives when performing hundreds to thousands of gene set tests in parallel.
 The purpose of this work is to determine how frequent methodological issues in functional enrichment are, including lack of a background gene set, lack of adjustment for multiple comparisons and lack of essential methodological details.
-By performing a screen of published articles we will identify the most frequent error modes, and use this to assist in developing a set of minimum standards for functional enrichment analysis (MSFEA).
+By performing a screen of published articles we will identify the most frequent error modes, and use this to inform us to develop a set of minimum standards for functional enrichment analysis (MSFEA).
 
@@
 
-References
 Draghici, S., Khatri, P., Martins, R. P., Ostermeier, G. C., & Krawetz, S. A. (2003). Global functional profiling of gene expression. Genomics, 81(2), 98–104.
+
+Hong, G., Zhang, W., Li, H., Shen, X., & Guo, Z. (2014). Separate enrichment analysis of pathways for up- and downregulated genes. Journal of the Royal Society, Interface, 11(92), 20130950.
 

### 2021-08-11 example.Rmd
added stuff

@@
 ---
-title: "PMC Survey Data Analysis"
+title: "Example gene set analysis"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
+suppressPackageStartupMessages({
 library("getDEE2") 
@@
 library("eulerr")
+})
 
@@
 
-```{r,getdata}
+```{r,getdata,fig.width=7,fig.height=7}
 
 mdat<-getDEE2Metadata("hsapiens")
-
-#SRP128998
-#samplesheet<-mdat[which(mdat$GEO_series=="GSE109140"),]
 samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
@@
 samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
-
-# samplesheets
 s1 <- subset(samplesheet,VPA==0)
@@
 
+# MDS plot
+plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$HG, main="MDS")
+
 #run DESeq2 for LG vs HG
@@
 
+```{r,deplots,fig.width=7,fig.height=7}
+
+maplot <- function(de,contrast_name) {
+  sig <-subset(de, padj < 0.05 )
+  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
+  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
+  GENESUP <- length(up)
+  GENESDN <- length(dn)
+  DET=nrow(de)
+  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
+  ns <-subset(de, padj > 0.05 )
+  plot(log2(de$baseMean),de$log2FoldChange, 
+       xlab="log2 basemean", ylab="log2 foldchange",
+       pch=19, cex=0.5, col="dark gray",
+       main=contrast_name, cex.main=0.7)
+  points(log2(sig$baseMean),sig$log2FoldChange,
+         pch=19, cex=0.5, col="red")
+  mtext(SUBHEADER,cex = 0.7)
+}
+
+make_volcano <- function(de,name) {
+    sig <- subset(de,padj<0.05)
+    N_SIG=nrow(sig)
+    N_UP=nrow(subset(sig,log2FoldChange>0))
+    N_DN=nrow(subset(sig,log2FoldChange<0))
+    DET=nrow(de)
+    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
+    plot(de$log2FoldChange,-log10(de$padj),cex=0.5,pch=19,col="darkgray",
+        main=name, xlab="log2 FC", ylab="-log10 pval", xlim=c(-6,6))
+    mtext(HEADER)
+    grid()
+    points(sig$log2FoldChange,-log10(sig$padj),cex=0.5,pch=19,col="red")
+}
+
+maplot(LGvHG,"LG v HG")
+make_volcano(LGvHG,"LG v HG")
+
+```
+
 ## Gene sets from Reactome
@@
 
+if (! file.exists("ReactomePathways.gmt")) {
   download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
@@
   unzip("ReactomePathways.gmt.zip")
+}
 genesets<-gmt_import("ReactomePathways.gmt")
@@
 ```{r,mitch1}
-library(kableExtra)
 
@@
 mres <- mitch_calc(m,genesets = genesets)
-head(mres$enrichment_result,50)
+head(mres$enrichment_result) 
 m_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
@@
 head(genesets2)
-
-head(LGvHG)
 
@@
 c_up <- head(c_up,1000)
-head(c_up,20)
+head(c_up)
 c_up <- rownames(subset(c_up, p.adjust < 0.05))
@@
 c_dn <- head(c_dn,1000)
-head(c_up,20)
+head(c_dn)
 c_dn <- rownames(subset(c_dn, p.adjust < 0.05))
+
+```
+
+## ORA with clusterprofiler combining up and down
+
+```{r,cp2}
+
+LGvHG_de <- rownames(subset(LGvHG,padj<0.05))
+LGvHG_de <- unique(gt[which(rownames(gt) %in% LGvHG_de),1])
+head(LGvHG_de)
+
+d_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, TERM2GENE = genesets2)
+d_de <- head(d_de,1000)
+d_de <- rownames(subset(d_de, p.adjust < 0.05))
+head(d_de)
 
@@
 
-```{r,cp1}
+```{r,cp3}
 
@@
 f_up <- head(f_up,1000)
-head(f_up,20)
+head(f_up)
 f_up <- rownames(subset(f_up, p.adjust < 0.05))
@@
 f_dn <- head(f_dn,1000)
-head(f_up,20)
+head(f_dn)
 f_dn <- rownames(subset(f_dn, p.adjust < 0.05))
@@
 ```
+
+## ORA combined with clusterprofiler with whole genome background list
+
+```{r,cp4}
+
+e_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, TERM2GENE = genesets2)
+e_de <- head(e_de,1000)
+head(e_de)
+e_de <- rownames(subset(e_de, p.adjust < 0.05))
+
+```
+
 ## Venn diagram comparison
@@
 
+par(cex.main=0.5)
+
+par(mar=c(2,2,2,2))
+
+v0 <- list("ORA up"=c_up,"ORA dn"=c_dn,
+           "ORA comb" = d_de)
+
+plot(euler(v0),quantities = TRUE,main="effect of combining up and down regulated genes")
+
 v1 <- list("FCS up"=m_up, "FCS dn"=m_dn,
@@
 
-plot(euler(v1),quantities = TRUE)
+plot(euler(v1),quantities = TRUE, main="FCS compared to ORA")
 
@@
 
-plot(euler(v2),quantities = TRUE)
+plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")
 
-v3 <- list("FCS up"=m_up, "FCS dn"=m_dn,
-           "ORA up"=c_up,"ORA dn"=c_dn, 
-           "ORA* up"=f_up,"ORA* dn"=f_dn )
+vx <- list("ORA up"=c_up,"ORA dn"=c_dn,
+           "ORA comb" = d_de, "ORA* comb" = e_de)
 
-plot(euler(v3),quantities = TRUE)
+plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
+
+v3 <- list("ORA up"=c_up,"ORA dn"=c_dn, 
+           "ORA* up"=f_up,"ORA* dn"=f_dn ,
+           "FCS up"=m_up, "FCS dn"=m_dn)
+
+plot(euler(v3),quantities = TRUE, main="all sets")
+
+```
+
+## Jaccard calculation
+
+```{r,jacc1}
+
+# ORA vs ORA combined
+dc <- length(intersect(d_de, c(c_up,c_dn))) / length(union(d_de, c(c_up,c_dn)))
+dc
+
+# FCS vs ORA
+cm <- length(intersect(c(c_up,c_dn), c(m_up,m_dn))) / length(union(c(c_up,c_dn), c(m_up,m_dn)))
+cm
+
+m_up <- gsub("^","up ",m_up)
+m_dn <- gsub("^","dn ",m_dn)
+m_de <- union(m_up,m_dn)
+
+c_up <- gsub("^","up ",c_up)
+c_dn <- gsub("^","dn ",c_dn)
+c_de <- union(c_up,c_dn)
+
+f_up <- gsub("^","up ",f_up)
+f_dn <- gsub("^","dn ",f_dn)
+f_de <- union(f_up,f_dn)
+
+# ORA vs ORA*
+cf <- length(intersect(c_de, f_de )) / length(union(c_de, f_de))
+cf
+
+# FCS vs ORA*
+mf <- length(intersect(m_de, f_de )) / length(union(m_de, f_de))
+mf
+
+dat <- c("ORA vs ORA comb"=dc,"FCS vs ORA"=cm,"ORA vs ORA*"=cf,"FCS vs ORA*"=mf)
+
+dat
+
+barplot(dat,ylab="jaccard metric")
 

### 2021-08-11 manuscript.Rmd
added stuff

@@
 
+As we screened these articles, we observed other types of details lacking.
+For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles.
+For ORA, in some instances it was unclear whether up- and down-regulated gene lists were combined before analysis.
+For differential expression analysis, combining up- and down-regulated gene lists for ORA is considered poor practice (Hong et al, 2014).
+Our own tests indicate that combining up- and down-regulated gene lists prior to ORA dramatically reduces the statistical strength when the proportion of differentially expressed genes is high.
+
 Surprisingly, the analysis scores we generated did not correlate with a journal level bibliographic metric, nor accrued citations of individual articles.

### 2021-08-13 example.Rmd
example RNA-seq ORA

@@
 x1<-x1[which(rowSums(x1)/ncol(x1)>=(10)),]
+nrow(x)
+nrow(x1)
 
@@
 
-c_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg, TERM2GENE = genesets2)
+c_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
 c_up <- head(c_up,1000)
@@
 
-c_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, TERM2GENE = genesets2)
+c_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
 c_dn <- head(c_dn,1000)
@@
 
-d_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, TERM2GENE = genesets2)
+d_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
 d_de <- head(d_de,1000)
@@
 
-f_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg, TERM2GENE = genesets2)
+f_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
 f_up <- head(f_up,1000)
@@
 
-f_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, TERM2GENE = genesets2)
+f_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, maxGSSize = 5000, TERM2GENE = genesets2)
 f_dn <- head(f_dn,1000)
@@
 
-e_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, TERM2GENE = genesets2)
+e_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, maxGSSize = 5000, TERM2GENE = genesets2)
 e_de <- head(e_de,1000)
@@
 
+png("images/fcs_ora1.png")
+plot(euler(v1),quantities = TRUE, main="FCS vs ORA")
+dev.off()
+
+png("images/orabg1.png")
+plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")
+dev.off()
+
+png("images/oracomb1.png")
+plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
+dev.off()
+
 ```
@@
 dc
+
+# ORA vs ORA* combined
+ec <- length(intersect(e_de, c(c_up,c_dn))) / length(union(e_de, c(c_up,c_dn)))
+ec
 
@@
 
-dat <- c("ORA vs ORA comb"=dc,"FCS vs ORA"=cm,"ORA vs ORA*"=cf,"FCS vs ORA*"=mf)
+dat <- c("FCS vs ORA"=cm,"ORA vs ORA*"=cf,"FCS vs ORA*"=mf, "ORA vs ORA comb"=dc, "ORA vs ORA* comb"=ec)
 

### 2021-08-13 manuscript.Rmd
example RNA-seq ORA

@@
 ---
-title: "A survey of rigour and reporting standards for gene set enrichment tests"
+title: "Guidelines for reliable and reproducible functional enrichment analysis"
 author: "Kaumadi Wijesooriya, Sameer A Jadaan, Tanuveer Kaur, Kaushalya L Perera, Mark Ziemann"
@@
 In other cases we have requested authors to correct their p-values to account for the possibility of false positives when performing hundreds to thousands of gene set tests in parallel.
-The purpose of this work is to determine how frequent methodological issues in functional enrichment are, including lack of a background gene set, lack of adjustment for multiple comparisons and lack of essential methodological details.
+The purpose of this work is to demonstrate the effect of misusing enrichment tools and determine how frequent methodological issues in functional enrichment are, specifically 
+(i) lack of a background gene set, 
+(ii) lack of adjustment for multiple comparisons and 
+(ii) lack of essential methodological details.
 By performing a screen of published articles we will identify the most frequent error modes, and use this to inform us to develop a set of minimum standards for functional enrichment analysis (MSFEA).
@@
 ## Methods
+
+### An example gene set analysis
+
+To demonstrate the effect of misusing gene set analysis, a publicly available RNA-seq dataset (accession SRP128998) was downloaded from DEE2 (Ziemann et al, 2019).
+This data consists of hepatocytes grown in culture with and without glucose first described by Felisbino et al (2021) and here we are using the comparison of cells grown in low glucose control condition compared to high glucose as an example.
+Transcript level counts were aggregated to genes using the getDEE2 R package. 
+Next, genes with an average of 10 reads per sample were discarded.
+Differential expression statistical analysis was conducted with DESeq2 (Love et al, 2014).
+For gene set analysis, Reactome gene sets (Jassal et al, 2020) were downloaded (URL: https://reactome.org/download/current/ReactomePathways.gmt.zip , accessed 13th August 2021).
+FCS was performed using the mitch R package, which uses a rank-ANOVA statistical test (Kaspi and Ziemann 2020).
+ORA analysis was performed using the clusterProfiler R package enricher function that implements a hypergeometric test (Yu et al, 2012).
+For ORA, two types of background gene sets were used (i) all detected genes, or (ii) all genes in the annotation set.
+For genes and gene sets, a false discovery rate adjusted p-value (FDR) of 0.05 was considered significant.
+Analyses were conducted in R version 4.1.
+
+### Survey of enrichment analysis
 
@@
 
-### Dataset overview
+### An example of gene set analysis misuse
+
+In order to demonstrate the effect of gene set test misuse, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes.
+Out of 39297 genes in the annotation set, 15635 were above the detection threshold.
+Statistical analysis revealed 3472 differentially expressed genes with 1560 up-regulated and 1912 down-regulated (FDR<0.05) due to glucose exposure.
+FCS revealed 95 up and 316 downregulated pathways (FDR<0.05).
+ORA with a background gene list consisting of detected genes revealed 55 upregulated and 238 downregulated gene sets (FDR<0.05).
+There was a strong overlap between FCS and ORA results with a Jaccard statistic of 0.66 (Figure 1A).
+We then performed ORA using the whole genome gene list (indicated as ORA*), which resulted in 147 up and 530 down regulated gene sets (FDR<0.05),
+The overlap of ORA with ORA* was relatively smaller, with a Jaccard statistic of 0.41 (Figure 1B).
+Interestingly, 26 gene sets were simultaneously up and downregulated with this approach.
+Then we performed ORA analysis after combining up and downregulated genes as sometimes observed in the literature (indicated as ORA comb) with the standard background and again with a whole genome background (indicated as ORA* comb).
+ORA comb revealed only 22 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.04 (Figure 1C). 
+ORA* comb yielded 64 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.08.
+These results demonstrate that commonly observed practices of combining the up and downregulated gene lists and using the whole genome background gene list severely distort ORA results.
+
+![Figure 1A. Comparison of FCS and ORA methods.](images/fcs_ora1.png "FCS vs ORA")
+
+![Figure 1B. ORA with standard and whole genome background.](images/orabg1.png "ORA with different background sets")
+
+![Figure 1C. Effect of combining up and downregulated genes on ORA results.](images/oracomb1.png "ORA combining up and downregulated genes")
+
+### A screen of functional enrichment analyses in open access journals
 

### 2021-08-16 analysis.Rmd
incremental

@@
         main = "Journal", xlim=c(0,1200))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2+50, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Omics type", xlim=c(0,600))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Organism", xlim=c(0,1200))
+        main = "Organism", xlim=c(0,1300))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Organism", xlim=c(0,1200))
+        main = "Organism", xlim=c(0,1300))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Gene set library", xlim=c(0,1000))
+        main = "Gene set library", xlim=c(0,1100))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Gene set library", xlim=c(0,1000))
+        main = "Gene set library", xlim=c(0,1100))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Gene set version defined", xlim=c(0,1600))
+text(y = (1:length(res)*1.2) - 0.6 , x = res-65, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Test used", xlim=c(0,1200))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
 par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "FDR", xlim=c(0,800))
+barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "FDR", xlim=c(0,900))
+text(y = (1:length(res)*1.2)-0.7  , x = res-30, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "FDR", xlim=c(0,800))
+        main = "FDR", xlim=c(0,900))
+text(y = (1:length(res)*1.2)-0.7  , x = res-30, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "App used", xlim=c(0,500))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "App version defined", xlim=c(0,1600))
+text(y = (1:length(res)*1.2) - 0.6 , x = res-70, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Code availability", xlim=c(0,300))
+        main = "Code availability", xlim=c(0,330))
+text(y = (1:length(res)*1.2) - 0.6 , x = res+15, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Background list defined", xlim=c(0,1200))
+        main = "Background list defined", xlim=c(0,1300))
+text(y = (1:length(res)*1.2)-0.7  , x = res+60, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Gene lists provided", xlim=c(0,1200))
+text(y = (1:length(res)*1.2) - 0.6 , x = res-70, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Assumptions violated", xlim=c(0,1400))
+        main = "Assumptions violated", xlim=c(0,1500))
+text(y = (1:length(res)*1.2)-0.75  , x = res+60, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
   # gene set lib
-  if(r[7]=="Not stated"){
+  if(r["Gene.set.library"]=="Not stated"){
     SCORE=SCORE-1
@@
   # GS version
-  if(r[8]=="No"){
+  if(r["GS.version"]=="No"){
     SCORE=SCORE-1
@@
   # stat test
-  if(r[9]=="No test"){
+  if(r["Statistical.test.used"]=="No test"){
     SCORE=SCORE-1
   }
-  if(r[9]=="Not stated"){
+  if(r["Statistical.test.used"]=="Not stated"){
     SCORE=SCORE-1
@@
   # FDR
-  if(r[10]!="Yes"){
+  if(r["FDR.Correction"]!="Yes"){
     SCORE=SCORE-1
@@
   # app used
-  if(r[11]=="Not stated"){
+  if(r["App.used"]=="Not stated"){
     SCORE=SCORE-1
@@
   # App version
-  if(r[12]=="No"){
+  if(r["App.Version"]=="No"){
     SCORE=SCORE-1
@@
   # Code availability
-  if(r[13]=="Yes"){
+  if(r["Code.availability"]=="Yes"){
     SCORE=SCORE+1
@@
   # Background
-  if(r[14]=="Not stated"){
+  if(r["Background.gene.set"]=="Not stated"){
     SCORE=SCORE-1
   }
-  if(r[14]=="No"){
+  if(r["Background.gene.set"]=="No"){
     SCORE=SCORE-1
   }
-  if(r[14]=="Stated, but incorrect"){
+  if(r["Background.gene.set"]=="Stated, but incorrect"){
     SCORE=SCORE-1
@@
   # gene list
-  if(r[16]=="Yes"){
+  if(r["Gene.lists.provided"]=="Yes"){
     SCORE=SCORE+1
@@
 x$scores <- scores
+length(which(scores>0))
 
@@
 
-xmm <-xm[,c(1,17)]
+xmm <-xm[,c("Journal","scores")]
 tab <- table(xmm$Journal)

### 2021-08-16 manuscript.Rmd
incremental

@@
 FCS tools involve giving each detected gene a differential expression score and then evaluating whether the scores are more positive or negative than expected by chance for each set of genes.
-The popular Gene Set Enrichment Analysis (GSEA) tool uses permutation approaches to establish whether a gene set is associated with higher or lower scores, either by permuting sample labels or by permuting genes in the differential expression profile (Subramanian et al, 2005).
+The popular Gene Set Enrichment Analysis (GSEA) tool uses permutation approaches to establish whether a gene set is significantly associated with higher or lower scores, either by permuting sample labels or by permuting genes in the differential expression profile (Subramanian et al, 2005).
 From a user's perspective, ORA is easier to conduct because it is as simple as pasting a list of gene names into a text box on a website.
@@
 
-To demonstrate the effect of misusing gene set analysis, a publicly available RNA-seq dataset (accession SRP128998) was downloaded from DEE2 (Ziemann et al, 2019).
-This data consists of hepatocytes grown in culture with and without glucose first described by Felisbino et al (2021) and here we are using the comparison of cells grown in low glucose control condition compared to high glucose as an example.
-Transcript level counts were aggregated to genes using the getDEE2 R package. 
+To demonstrate the effect of misusing gene set analysis, a publicly available RNA-seq dataset (accession SRP128998) was downloaded from DEE2 on 13th August 2021 (Ziemann et al, 2019).
+This data consists of immortalised human hepatocytes cultured in standard or high glucose, first described by Felisbino et al (2021).
+Transcript level counts were aggregated to genes using the getDEE2 R package v1.2.0. 
 Next, genes with an average of 10 reads per sample were discarded.
-Differential expression statistical analysis was conducted with DESeq2 (Love et al, 2014).
+Differential expression statistical analysis was conducted with DESeq2 v1.32.0 (Love et al, 2014) to identify genes altered by high glucose exposure.
 For gene set analysis, Reactome gene sets (Jassal et al, 2020) were downloaded (URL: https://reactome.org/download/current/ReactomePathways.gmt.zip , accessed 13th August 2021).
-FCS was performed using the mitch R package, which uses a rank-ANOVA statistical test (Kaspi and Ziemann 2020).
-ORA analysis was performed using the clusterProfiler R package enricher function that implements a hypergeometric test (Yu et al, 2012).
+FCS was performed using the mitch R package v1.4.0 with default settings, which uses a rank-ANOVA statistical test (Kaspi and Ziemann 2020).
+ORA analysis was performed using the clusterProfiler R package (v4.0.2) enricher function that implements a hypergeometric test (Yu et al, 2012).
 For ORA, two types of background gene sets were used (i) all detected genes, or (ii) all genes in the annotation set.
 For genes and gene sets, a false discovery rate adjusted p-value (FDR) of 0.05 was considered significant.
-Analyses were conducted in R version 4.1.
+Analyses were conducted in R version 4.1.0.
 
@@
 
-* Whether gene profile is provided in the supplement
+* Whether gene profile was provided in the supplement
 
-* Whether assumptions are correctly applied
+* Whether assumptions was correctly applied
 
+We excluded articles describing novel enrichment analysis techniques/tools, review articles and conference abstracts.
 Some articles presented the results of >1 enrichment analysis, so additional rows were added to accommodate them. 
@@
 In order to demonstrate the effect of gene set test misuse, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes.
-Out of 39297 genes in the annotation set, 15635 were above the detection threshold.
-Statistical analysis revealed 3472 differentially expressed genes with 1560 up-regulated and 1912 down-regulated (FDR<0.05) due to glucose exposure.
+Out of 39,297 genes in the annotation set, 15,635 were above the detection threshold (>10 reads per sample on average).
+Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to glucose exposure.
 FCS revealed 95 up and 316 downregulated pathways (FDR<0.05).
@@
 There was a strong overlap between FCS and ORA results with a Jaccard statistic of 0.66 (Figure 1A).
-We then performed ORA using the whole genome gene list (indicated as ORA*), which resulted in 147 up and 530 down regulated gene sets (FDR<0.05),
-The overlap of ORA with ORA* was relatively smaller, with a Jaccard statistic of 0.41 (Figure 1B).
+We then performed ORA using the whole genome gene list (indicated as ORA\*), which resulted in 147 up and 530 down regulated gene sets (FDR<0.05),
+The overlap of ORA with ORA\* was relatively smaller, with a Jaccard statistic of 0.41 (Figure 1B).
 Interestingly, 26 gene sets were simultaneously up and downregulated with this approach.
-Then we performed ORA analysis after combining up and downregulated genes as sometimes observed in the literature (indicated as ORA comb) with the standard background and again with a whole genome background (indicated as ORA* comb).
+Then we performed ORA analysis after combining up and downregulated genes as sometimes observed in the literature (indicated as ORA comb) with the standard background and again with a whole genome background (indicated as ORA\* comb).
 ORA comb revealed only 22 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.04 (Figure 1C). 
-ORA* comb yielded 64 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.08.
+ORA\* comb yielded 64 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.08.
 These results demonstrate that commonly observed practices of combining the up and downregulated gene lists and using the whole genome background gene list severely distort ORA results.
@@
 As some articles included more than one enrichment analysis, the final dataset included 1624 analyses from 1363 articles.
-There were articles from 330 journals in the screen, with *Scientific Reports*, *Oncology Letters* and *International Journal of Molecular Science* being the biggest contributors (Figure 1).
-There were 35 different omics types, with RNA-seq and gene expression microarray being the most popular (Figure 2).
-There were 175 different species under study, but human was the most common with 1104 analyses (Figure 3).
-We recorded the use of 101 different gene set libraries, with GO and KEGG being the most frequently used (Figure 4).
+There were articles from 330 journals in the screen, with *Scientific Reports*, *Oncology Letters* and *International Journal of Molecular Science* being the biggest contributors (Figure S1A).
+There were 35 different omics types, with RNA-seq and gene expression microarray being the most popular (Figure S1B).
+There were 175 different species under study, but human was the most common with 1104 analyses (Figure S1C).
+We recorded the use of 101 different gene set libraries, with GO and KEGG being the most frequently used (Figure 2A).
 There were 85 studies where the gene set libraries used were not defined in the article.
-Only 108 analyses reported the version of the gene set library used (Figure 5).
-There were XX different statistical tests used, and the most common reported tests were Fisher's Exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Figure 6).
-Only 754 of 1594 studies described performing FDR or other p-value correction for multiple testing (Figure 7).
-There were 145 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 174 analyses (XX%) did not state what tool was used (Figure 8).
-The version of the software used was provided in only 421 of 1611 analyses (26%) (Figure 9).
-Of the 293 analyses which used a scripted computer language, only 10 provided links to the code used for enrichment analysis (3.4%) (Figure 10).
-For analyses using ORA methods, we studied what background gene set was used (Figure 11). 
+Only 108 analyses reported the version of the gene set library used (Figure 2B).
+There were XX different statistical tests used, and the most common reported tests were Fisher's Exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Figure 2C).
+Only 754 of 1594 studies described performing FDR or other p-value correction for multiple testing (Figure 2D).
+There were 145 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 174 analyses (XX%) did not state what tool was used (Figure 2E).
+The version of the software used was provided in only 421 of 1611 analyses (26%) (Figure 2F).
+For analyses using ORA methods, we studied what background gene set was used (Figure 2G). 
 This revealed that in most cases, the background list was not stated or it was clear from the article methods section that no background list was used.
@@
 In only 67/1460 of cases (4.6%), the appropriate background list was described in the article.
-For 579 of 1618 analyses, the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Figure 12).
-Next, we quantified the frequency of methodological issues that would undermine the conclusions (Figure 13).
+Of the 293 analyses which used a scripted computer language, only 10 provided links to the code used for enrichment analysis (3.4%) (Figure 2H).
+For 579 of 1618 analyses, the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Figure 2I).
+Next, we quantified the frequency of methodological issues that would undermine the conclusions (Figure 2J).
 Lack of appropriate background was the most common issue (1361 cases), followed by FDR (761 cases), then inference without test (61 cases), inference without data shown (32 cases) and misinterpreted FDR values (3 cases).
 Only 160 analyses did not exhibit any of these major methodological issues.
+Taken together, these data suggest methodological and reporting deficiencies are widespread in published functional enrichment analyses.
 
-![Figure 1. Representation of journals in the data set.](images/journals1.png "Journals")
+![Figure S1a. Representation of journals in the data set.](images/journals1.png "Journals")
 
-![Figure 2. Representation of different omics types in the data set.](images/omics1.png "Omics")
+![Figure S1B. Representation of different omics types in the data set.](images/omics1.png "Omics")
 
-![Figure 3. Representation of different organisms in the data set.](images/organisms1.png "Organisms")
+![Figure S1C. Representation of different organisms in the data set.](images/organisms1.png "Organisms")
 
-![Figure 4. Representation of different gene set libraries in the data set.](images/genesetlib1.png "Geneset libraries")
+![Figure 2A. Representation of different gene set libraries in the data set.](images/genesetlib1.png "Geneset libraries")
 
-![Figure 5. Proportion of analyses that reported version information for gene sets used.](images/genesetvers1.png "Geneset version")
+![Figure 2B. Proportion of analyses that reported version information for gene sets used.](images/genesetvers1.png "Geneset version")
 
-![Figure 6. Representation of different statistical tests in the data set.](images/stattest1.png "Geneset libraries")
+![Figure 2C. Representation of different statistical tests in the data set.](images/stattest1.png "Geneset libraries")
 
-![Figure 7. Recognition of p-value correction for multiple testing in the data set.](images/fdr1.png "FDR")
+![Figure 2D. Recognition of p-value correction for multiple testing in the data set.](images/fdr1.png "FDR")
 
-![Figure 8. Representation of different software tools for enrichment analysis in the data set.](images/app1.png "Software used")
+![Figure 2E. Representation of different software tools for enrichment analysis in the data set.](images/app1.png "Software used")
 
-![Figure 9. Proportion of analyses that reported version information for software used.](images/appvers1.png "Software version")
+![Figure 2F. Proportion of analyses that reported version information for software used.](images/appvers1.png "Software version")
 
-![Figure 10. Proportion of scripted analyses that provided software code.](images/code1.png "Code available")
+![Figure 2G. Background gene set usage and reporting in the dataset.](images/bg1.png "Background gene list used")
 
-![Figure 11. Background gene set usage and reporting in the dataset.](images/bg1.png "Background gene list used")
+![Figure 2H. Proportion of scripted analyses that provided software code.](images/code1.png "Code available")
 
-![Figure 12. Proportion of scripted analyses that provided gene profiles.](images/genelists1.png "Gene lists provided")
+![Figure 2I. Proportion of scripted analyses that provided gene profiles.](images/genelists1.png "Gene lists provided")
 
-![Figure 13. Proportion of analyses with different methodological problems.](images/assumptions1.png "Assumptions violated")
+![Figure 2J. Proportion of analyses with different methodological problems.](images/assumptions1.png "Assumptions violated")
 
@@
 
-![Figure 14. Distribution of analysis scores.](images/hist1.png "Analysis scores")
+![Figure 3A. Distribution of analysis scores.](images/hist1.png "Analysis scores")
 
-![Figure 15. Association of analysis scores with journal rank.](images/score_sjr1.png "Analysis scores versus journal rank")
+![Figure 3B. Association of analysis scores with journal rank.](images/score_sjr1.png "Analysis scores versus journal rank")
 
-![Figure 16. Journals with highest analysis scores.](images/score_highest1.png "Highest analysis scores")
+![Figure 3C. Journals with highest analysis scores.](images/score_highest1.png "Highest analysis scores")
 
-![Figure 17. Journals with Lowest analysis scores.](images/score_lowest1.png "Lowest analysis scores")
+![Figure 3D. Journals with Lowest analysis scores.](images/score_lowest1.png "Lowest analysis scores")
 
-![Figure 18. Association between SJR and mean analysis score for journals with >=5 analyses.](images/mean_sjr1.png "Mean score vs SJR")
+![Figure 3E. Association between SJR and mean analysis score for journals with >=5 analyses.](images/mean_sjr1.png "Mean score vs SJR")
 
@@
 The lack of methodological details regarding enrichment analysis in some articles was disturbing.
-To illustrate this, many articles mention prominently in the abstract the results of the functional enrichment analysis yet not detail in their methods sections any detail whatsoever about what type of enrichment analysis was actually conducted.
+To illustrate this, many articles mention prominently in the abstract the functional enrichment results but do not detail in their methods sections anything about what type of enrichment analysis was actually conducted.
 This is not limited to a few obscure journals; there were 88 journals that published articles where the tool used for enrichment analysis was not defined.
@@
 For ORA, in some instances it was unclear whether up- and down-regulated gene lists were combined before analysis.
-For differential expression analysis, combining up- and down-regulated gene lists for ORA is considered poor practice (Hong et al, 2014).
-Our own tests indicate that combining up- and down-regulated gene lists prior to ORA dramatically reduces the statistical strength when the proportion of differentially expressed genes is high.
+When applying ORA to differential expression, combining up- and down-regulated gene lists is considered poor practice (Hong et al, 2014).
+Our own tests indicate that combining up- and down-regulated gene lists prior to ORA results in a dramatic loss of sensitivity when the proportion of differentially expressed genes is high.
+Using an inappropriate background list resulted in substantially altered results (Figure 1B).
+In the example here, using the inappropriate background resulted in 2.3 fold more differentially regulated gene sets.
+Thus if there are few or no significant results using a correct background list, there may be a temptation may be to use an inappropriate background.
+Results here indicate most ORA articles do not report using a background gene list, suggesting that inappropriate ORA use is common.
 
+We scored each of the analyses included in the screen which revealed only 30 articles scored a satisfactory score of zero or higher.
 Surprisingly, the analysis scores we generated did not correlate with a journal level bibliographic metric, nor accrued citations of individual articles.
-This suggests no association between methodological rigour and "impact" in this sample.
+This suggests no association between methodological rigour and traditional research impact metrics in this sample.
 
@@
 
-Nevertheless, these results are a wake up call for reproducibility and highlight the urgent need for minimum standards for functional enrichment analysis (MSFEA). 
+Nevertheless, these results are a wake up call for reproducibility and highlight the urgent need for minimum standards for functional enrichment analysis. 
 To address this, we provide a set of minimum standards for functional enrichment analysis, as well as additional recommendations to achieve gold standard reliability and reproducibility (Box 1).
@@
 This will reduce the chance of false positives when performing many tests simultaneously.
-Bonferonni, FDR and FWER are examples.
+Report what method was used for p-value adjustment; Bonferonni, FDR and FWER are examples.
 
 6. If ORA is used, it is vital to use a background list consisting of all detected genes, and report this in the methods section.
-Avoid tools that cannot accept a background list.
+Avoid tools that don't accept a background list.
 
-7. Criteria and parameters matter.
+7. Report selection criteria and parameters.
 If performing ORA, then the inclusion thresholds need to be mentioned.
 If using GSEA or another FCS tool, parameters around ranking, gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels).
+Any parameters that are vary from the defaults should be reported.
 
-8. If using ORA, it needs to be mentioned whether up- and down-regulated genes are separated into different tests as recommended by Hong et al (2014) or combined.
+8. If using ORA for gene expression analysis, refrain from combining up- and down-regulated genes before analysis.
 
@@
 In practice, they are rarely used properly.
-FCS tests distinguish between up- and down-regulated gene sets by default.
 

### 2021-08-16 manuscript.Rmd
affiliation

@@
 title: "Guidelines for reliable and reproducible functional enrichment analysis"
-author: "Kaumadi Wijesooriya, Sameer A Jadaan, Tanuveer Kaur, Kaushalya L Perera, Mark Ziemann"
+author: "Kaumadi Wijesooriya1, Sameer A Jadaan2, Tanuveer Kaur1, Kaushalya L Perera1, Mark Ziemann1"
 date: "`r Sys.Date()`"
@@
 
-1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences
+1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.
 
-2. Sameer to provide affiliation information
+2. College of Health and Medical Technology, Middle Technical University, Baghdad, Iraq.
 

### 2021-08-18 QC analysis.Rmd
fix qc
start comparision

@@
 ## Intro
+
 Here we are performing an analysis of 200 articles which is randomly selected from 1500 PMC articles.
@@
 ```
+
+## Cross check
+
+```{R,crosscheck}
+y <- read.table("PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
+myqc<-x[1,,drop=TRUE]
+head(myqc)
+mypmc<-myqc[[1]]
+str(myqc)
+
+ydat<-y[which(y$Pubmed.Central.ID==mypmc),,drop=TRUE]
+str(ydat)
+
+```

### 2021-08-18 analysis.Rmd
PMC numbers included
masnuscript update

@@
 
+dir.create("images")
 png("images/journals1.png")
@@
         main = "Journal", xlim=c(0,1200))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2+50, label = res2, pos = 3, cex = 1, col = "black")
+text(y = (1:length(res2)*1.2)-0.8  , x = res2+50, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Omics type", xlim=c(0,600))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
+text(y = (1:length(res2)*1.2)-0.8  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Organism", xlim=c(0,1300))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
+text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Gene set library", xlim=c(0,1100))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
+text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Gene set version defined", xlim=c(0,1600))
-text(y = (1:length(res)*1.2) - 0.6 , x = res-65, label = res, pos = 3, cex = 1, col = "black")
+        main = "Gene set version defined", xlim=c(0,1900))
+text(y = (1:length(res)*1.2) - 0.6 , x = res+200, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Test used", xlim=c(0,1200))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
+text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "FDR", xlim=c(0,900))
-text(y = (1:length(res)*1.2)-0.7  , x = res-30, label = res, pos = 3, cex = 1, col = "black")
+        main = "FDR", xlim=c(0,1000))
+text(y = (1:length(res)*1.2)-0.7  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "App used", xlim=c(0,500))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
+text(y = (1:length(res2)*1.2)-0.8  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "App version defined", xlim=c(0,1600))
-text(y = (1:length(res)*1.2) - 0.6 , x = res-70, label = res, pos = 3, cex = 1, col = "black")
+text(y = (1:length(res)*1.2) - 0.6 , x = res-200, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Code availability", xlim=c(0,330))
-text(y = (1:length(res)*1.2) - 0.6 , x = res+15, label = res, pos = 3, cex = 1, col = "black")
+        main = "Code availability", xlim=c(0,350))
+text(y = (1:length(res)*1.2) - 0.6 , x = res+30, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Background list defined", xlim=c(0,1300))
-text(y = (1:length(res)*1.2)-0.7  , x = res+60, label = res, pos = 3, cex = 1, col = "black")
+        main = "Background list defined", xlim=c(0,1400))
+text(y = (1:length(res)*1.2)-0.7  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
         main = "Gene lists provided", xlim=c(0,1200))
-text(y = (1:length(res)*1.2) - 0.6 , x = res-70, label = res, pos = 3, cex = 1, col = "black")
+text(y = (1:length(res)*1.2) - 0.6 , x = res-120, label = res, pos = 3, cex = 1, col = "black")
 grid()
@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Assumptions violated", xlim=c(0,1500))
-text(y = (1:length(res)*1.2)-0.75  , x = res+60, label = res, pos = 3, cex = 1, col = "black")
+        main = "Assumptions violated", xlim=c(0,100))
+text(y = (1:length(res)*1.2)-0.75  , x = res+100, label = res, pos = 3, cex = 1, col = "black")
 grid()

### 2021-08-18 manuscript.Rmd
masnuscript update

@@
 
-Gene set enrichment tests (a.k.a. functional enrichment analysis) are amongst the most frequently used methods in computational biology.
+Gene set enrichment tests (a.k.a. functional enrichment analysis) are among the most frequently used methods in computational biology.
 Despite this popularity, there are concerns that these methods are being applied incorrectly and the results of some peer-reviewed publications are unreliable.
-To ascertain the frequency of these errors, we performed a screen of 1363 open access research articles describing gene set test results.
+These methodological problems include the use of inappropriate background gene lists, lack of false discovery rate correction and combining up-and downregulated gene sets before enrichment analysis.
+An example analysis of public RNA-seq reveals that inappropriate background gene list and combining up-and downregulated gene sets alters enrichment results sufficiently to .
+To ascertain the frequency of these errors in the literature, we performed a screen of 1363 open access research articles describing functional enrichment results.
 We find that 95% of analyses using over-representation tests did not implement an appropriate reference gene list or did not describe this in the methods.
-Failure to perform p-value correction for multiple tests was identified in 53% of analyses
+Failure to perform p-value correction for multiple tests was identified in 53% of analyses.
 Many studies lacked detail in the methods section about the tools and gene sets used.
@@
 (ii) lack of essential methodological details.
-By performing a screen of published articles we will identify the most frequent error modes, and use this to inform us to develop a set of minimum standards for functional enrichment analysis (MSFEA).
+By performing a screen of published articles we will identify the most frequent error modes, and use this to inform us to develop a set of minimum standards for functional enrichment analysis.
 
@@
 
-We have collated a list of 2941 articles in PubMed Central published in 2019 that have “enrichment analysis” and related keyword terms.
+We have collated a list of 2941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
 We sampled 1500 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.
@@
 As some articles included more than one enrichment analysis, the final dataset included 1624 analyses from 1363 articles.
+A flow diagram of the survey is provided in Figure 2.
+
+```{r,diagram1,echo=FALSE}
+
+library(DiagrammeR)
+grViz("digraph flowchart {
+      # node definitions with substituted label text
+      node [fontname = Helvetica, shape = rectangle]        
+      tab1 [label = '@@1']
+      tab2 [label = '@@2']
+      tab3 [label = '@@3']
+      tab4 [label = '@@4']
+      tab5 [label = '@@5']
+      tab6 [label = '@@6']
+      tab7 [label = '@@7']
+      tab8 [label = '@@8']
+      tab9 [label = '@@9']
+      tab10 [label = '@@10']
+      tab11 [label = '@@11']
+      tab12 [label = '@@12']
+      tab13 [label = '@@13']
+
+      # edge definitions with the node IDs
+      tab1 -> tab2  -> tab4 -> tab5 -> tab6 -> tab7 -> tab11;
+      tab7 -> tab12 ;
+      tab7 -> tab13 ;
+      tab2-> tab3 ;
+      tab4-> tab8 -> tab9 -> tab10 -> tab13;
+      }
+      [1]: 'PMC keyword search (2019): 2941 results'
+      [2]: '1500 PMC articles selected'
+      [3]: '133 articles excluded'
+      [4]: '1363 articles analyzed by team of five researchers'
+      [5]: '1624 enrichment analyses described'
+      [6]: 'data entry to spreadsheet'
+      [7]: 'data analysis in R'
+      [8]: 'Subset of 231 analyses cross-checked by two senior researchers'
+      [9]: 'data entry to spreadsheet'
+      [10]: 'data analysis in R'
+      [11]: 'Summary charts'
+      [12]: 'Integration with journal and article metrics'
+      [13]: 'Data validation/crosschecking'
+      ")
+
+```
+Figure 2. A summary of the survey of functional enrichment analyses.
+
 There were articles from 330 journals in the screen, with *Scientific Reports*, *Oncology Letters* and *International Journal of Molecular Science* being the biggest contributors (Figure S1A).
@@
 There were 175 different species under study, but human was the most common with 1104 analyses (Figure S1C).
-We recorded the use of 101 different gene set libraries, with GO and KEGG being the most frequently used (Figure 2A).
+We recorded the use of 101 different gene set libraries, with GO and KEGG being the most frequently used (Figure 3A).
 There were 85 studies where the gene set libraries used were not defined in the article.
-Only 108 analyses reported the version of the gene set library used (Figure 2B).
-There were XX different statistical tests used, and the most common reported tests were Fisher's Exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Figure 2C).
-Only 754 of 1594 studies described performing FDR or other p-value correction for multiple testing (Figure 2D).
-There were 145 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 174 analyses (XX%) did not state what tool was used (Figure 2E).
-The version of the software used was provided in only 421 of 1611 analyses (26%) (Figure 2F).
-For analyses using ORA methods, we studied what background gene set was used (Figure 2G). 
+Only 108 analyses reported the version of the gene set library used (Figure 3B).
+There were XX different statistical tests used, and the most common reported tests were Fisher's Exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Figure 3C).
+Only 754 of 1594 studies (XX%) described correction of p-values for multiple testing (Figure 3D).
+There were 145 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 174 analyses (XX%) did not state what tool was used (Figure 3E).
+The version of the software used was provided in only 421 of 1611 analyses (26%) (Figure 3F).
+For analyses using ORA methods, we studied what background gene set was used (Figure 3G). 
 This revealed that in most cases, the background list was not stated or it was clear from the article methods section that no background list was used.
@@
 In only 67/1460 of cases (4.6%), the appropriate background list was described in the article.
-Of the 293 analyses which used a scripted computer language, only 10 provided links to the code used for enrichment analysis (3.4%) (Figure 2H).
-For 579 of 1618 analyses, the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Figure 2I).
-Next, we quantified the frequency of methodological issues that would undermine the conclusions (Figure 2J).
+Of the 293 analyses which used a scripted computer language, only 10 provided links to the code used for enrichment analysis (3.4%) (Figure 3H).
+For 579 of 1618 analyses, the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Figure 3I).
+Next, we quantified the frequency of methodological issues that would undermine the conclusions (Figure 3J).
 Lack of appropriate background was the most common issue (1361 cases), followed by FDR (761 cases), then inference without test (61 cases), inference without data shown (32 cases) and misinterpreted FDR values (3 cases).
@@
 
-![Figure 2A. Representation of different gene set libraries in the data set.](images/genesetlib1.png "Geneset libraries")
+![Figure 3A. Representation of different gene set libraries in the data set.](images/genesetlib1.png "Geneset libraries")
 
-![Figure 2B. Proportion of analyses that reported version information for gene sets used.](images/genesetvers1.png "Geneset version")
+![Figure 3B. Proportion of analyses that reported version information for gene sets used.](images/genesetvers1.png "Geneset version")
 
-![Figure 2C. Representation of different statistical tests in the data set.](images/stattest1.png "Geneset libraries")
+![Figure 3C. Representation of different statistical tests in the data set.](images/stattest1.png "Geneset libraries")
 
-![Figure 2D. Recognition of p-value correction for multiple testing in the data set.](images/fdr1.png "FDR")
+![Figure 3D. Recognition of p-value correction for multiple testing in the data set.](images/fdr1.png "FDR")
 
-![Figure 2E. Representation of different software tools for enrichment analysis in the data set.](images/app1.png "Software used")
+![Figure 3E. Representation of different software tools for enrichment analysis in the data set.](images/app1.png "Software used")
 
-![Figure 2F. Proportion of analyses that reported version information for software used.](images/appvers1.png "Software version")
+![Figure 3F. Proportion of analyses that reported version information for software used.](images/appvers1.png "Software version")
 
-![Figure 2G. Background gene set usage and reporting in the dataset.](images/bg1.png "Background gene list used")
+![Figure 3G. Background gene set usage and reporting in the dataset.](images/bg1.png "Background gene list used")
 
-![Figure 2H. Proportion of scripted analyses that provided software code.](images/code1.png "Code available")
+![Figure 3H. Proportion of scripted analyses that provided software code.](images/code1.png "Code available")
 
-![Figure 2I. Proportion of scripted analyses that provided gene profiles.](images/genelists1.png "Gene lists provided")
+![Figure 3I. Proportion of scripted analyses that provided gene profiles.](images/genelists1.png "Gene lists provided")
 
-![Figure 2J. Proportion of analyses with different methodological problems.](images/assumptions1.png "Assumptions violated")
+![Figure 3J. Proportion of analyses with different methodological problems.](images/assumptions1.png "Assumptions violated")
 
@@
 We then scored each analysis based on the inclusion or lack of methodological issues and included details.
-The median score was -4, with a mean of -3.5 and standard deviation of 1.3 (Figure 14).
-Next, we assessed whether these analysis scores associate with journal citation metrics (Figure 15).
+The median score was -4, with a mean of -3.5 and standard deviation of 1.3 (Figure 4A).
+Next, we assessed whether these analysis scores associate with journal citation metrics (Figure 4B).
 There was a slight positive correlation using the Pearson method but it was not statistically significant (r=0.04, p=0.16).
@@
 Only journals with five or more analyses were included.
-The best scoring journals were *Transl Psychiatry*, *Metabolites* and *J Exp Clin Cancer Res* (Figure 16), while the poorest were *Mol Med Rep*, *Cancer Med* and *World J Gastroenterol* (Figure 17), although we note that there was a wide variation between articles of the same journal.
-Then, we assessed for an association between mean analysis score and the SJR (Figure 17).
+The best scoring journals were *Transl Psychiatry*, *Metabolites* and *J Exp Clin Cancer Res* (Figure 4C), while the poorest were *Mol Med Rep*, *Cancer Med* and *World J Gastroenterol* (Figure 4D), although we note that there was a wide variation between articles of the same journal.
+Then, we assessed for an association between mean analysis score and the SJR (Figure 4E).
 Again, there was a slightly positive association that wasn't statistically significant (Pearson r=0.09, p=0.55).
@@
 
-![Figure 3A. Distribution of analysis scores.](images/hist1.png "Analysis scores")
+![Figure 4A. Distribution of analysis scores.](images/hist1.png "Analysis scores")
 
-![Figure 3B. Association of analysis scores with journal rank.](images/score_sjr1.png "Analysis scores versus journal rank")
+![Figure 4B. Association of analysis scores with journal rank.](images/score_sjr1.png "Analysis scores versus journal rank")
 
-![Figure 3C. Journals with highest analysis scores.](images/score_highest1.png "Highest analysis scores")
+![Figure 4C. Journals with highest analysis scores.](images/score_highest1.png "Highest analysis scores")
 
-![Figure 3D. Journals with Lowest analysis scores.](images/score_lowest1.png "Lowest analysis scores")
+![Figure 4D. Journals with Lowest analysis scores.](images/score_lowest1.png "Lowest analysis scores")
 
-![Figure 3E. Association between SJR and mean analysis score for journals with >=5 analyses.](images/mean_sjr1.png "Mean score vs SJR")
+![Figure 4E. Association between SJR and mean analysis score for journals with >=5 analyses.](images/mean_sjr1.png "Mean score vs SJR")
 

### 2021-08-18 QC analysis.Rmd
working on QC

@@
 library("kableExtra")
+library("Biobase")
+library("vioplot")
 
@@
 
-```{R,crosscheck}
+```{R,crosscheck,fig.height=8,fig.width=8}
+
 y <- read.table("PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
-myqc<-x[1,,drop=TRUE]
-head(myqc)
-mypmc<-myqc[[1]]
-str(myqc)
+y <- y[which(isUnique(y$Pubmed.Central.ID)),]
+x <- x[which(isUnique(x$Pubmed.Central.ID)),]
+yunique <- y[which(isUnique(y$Pubmed.Central.ID)),"Pubmed.Central.ID"]
+xunique <- x[which(isUnique(x$Pubmed.Central.ID)),"Pubmed.Central.ID"]
+xyunique <- intersect(yunique,xunique)
+head(xyunique)
 
-ydat<-y[which(y$Pubmed.Central.ID==mypmc),,drop=TRUE]
-str(ydat)
+res <- sapply(xyunique,function(myid){
+  myqc <- x[which(x$Pubmed.Central.ID==myid),,drop=TRUE]
+  ydat <- y[which(y$Pubmed.Central.ID==myid),,drop=TRUE]
+  qcres <- sapply(5:16, function(m) {
+    mqc <- myqc[[m]]
+    mqc <- unlist(strsplit(mqc,", "))
+    myy <- ydat[[m]]
+    myy <- unlist(strsplit(myy,", "))
+  length(intersect(mqc,myy)) / length(union(mqc,myy))
+  })
+  return(qcres)
+})
+
+rownames(res) <- colnames(x)[5:16]
+categoryMeans <- rowMeans(res)
+categoryMeans
+
+tres <- t(res)
+head(tres)
+par(mar=c(5,12,3,1))
+vioplot(tres,horizontal=TRUE,las=1,xlab="mean Jaccard score")
+par(mar=c(5,12,3,1))
+barplot(categoryMeans,horiz=TRUE,las=1,xlim=c(0,1),xlab="mean Jaccard score")
+
+articleMeans <- colMeans(res)
+hist(articleMeans)
 
 ```
+
+```{r,qcfunc}
+
+ppl <- sapply(xyunique,function(myid){
+  myqc <- x[which(x$Pubmed.Central.ID==myid),,drop=TRUE]
+  ydat <- y[which(y$Pubmed.Central.ID==myid),,drop=TRUE]
+  px <- myqc[[3]]
+  py <- ydat[[3]]
+  return(c(py,px))
+})
+
+tppl <- t(ppl)
+
+tppl_comb <- paste(tppl[,1],tppl[,2])
+tppl_combs <- unique(tppl_comb)
+cres <- lapply(tppl_combs,function(comb){
+ combres <- tres[which(tppl_comb %in% comb),]
+})
+
+names(cres) <- tppl_combs
+str(cres)
+
+cres2 <- lapply(cres,mean)
+cres2
+par(mar=c(5,12,3,1))
+barplot(unlist(cres2),horiz=TRUE,las=1,xlim=c(0,1),xlab="mean Jaccard score")
+
+```
+
+## Session information
+
+```{r,session}
+
+sessionInfo()
+
+```

### 2021-08-18 manuscript.Rmd
working on QC

@@
 title: "Guidelines for reliable and reproducible functional enrichment analysis"
-author: "Kaumadi Wijesooriya1, Sameer A Jadaan2, Tanuveer Kaur1, Kaushalya L Perera1, Mark Ziemann1"
+author: "Kaumadi Wijesooriya, Sameer A Jadaan, Tanuveer Kaur, Kaushalya L Perera, Mark Ziemann"
 date: "`r Sys.Date()`"
 output:
+  rmdformats::robobook:
   html_document:
@@
 
+Kaumadi Wijesooriya<sup>1</sup>, Sameer A Jadaan<sup>2</sup>, Tanuveer Kaur<sup>1</sup>, Kaushalya L Perera<sup>1</sup>, Mark Ziemann<sup>1</sup>
+
 **Affiliations**
@@
 These methodological problems include the use of inappropriate background gene lists, lack of false discovery rate correction and combining up-and downregulated gene sets before enrichment analysis.
-An example analysis of public RNA-seq reveals that inappropriate background gene list and combining up-and downregulated gene sets alters enrichment results sufficiently to .
+An example analysis of public RNA-seq reveals that these methodological errors alter enrichment results dramatically.
 To ascertain the frequency of these errors in the literature, we performed a screen of 1363 open access research articles describing functional enrichment results.
@@
 Many studies lacked detail in the methods section about the tools and gene sets used.
-Only 10% of studies achieved the satisfactory standard of reporting and rigour.
-This highlights the poor state of enrichment test reporting and rigour standards in the contemporary literature.
+Only 10% of studies achieved an overall satisfactory standard, which highlights the poor state of functional enrichment rigour and reporting in the contemporary literature.
 We provide a set of minimum standards that should act as a checklist for researchers and peer-reviewers.
@@
 
-Since the turn of the millennium, high throughput omics techniques like microarrays and next generation sequencing have brought with them a deluge of data.
-They have revealed how genes are regulated in development and many different diseases.
-These experiments involve the measurement of thousands of genes simultaneously, and can identify hundreds or even thousands of significant associations.
+Since the turn of the millennium, high throughput "omics" techniques like microarrays and next generation sequencing have brought with them a deluge of data.
+These experiments involve the measurement of thousands of genes simultaneously, and can identify hundreds or even thousands of significant associations with developmental stages or diseases.
 Interpreting such data is extraordinarily challenging, as the sheer number of associations can be difficult to investigate in a gene-by-gene manner.
-Instead, many different tools have been developed in an effort to summarise gene profiles into simplified functional categories.
+Instead, many tools have been developed in an effort to summarise gene profiles into simplified functional categories.
 These functional categories typically represent signaling or biochemical pathways, curated from information present in the literature, hence the name functional enrichment.
 
-Widely used functional enrichment tools can be classified into two main categories; (i) over-representation analysis (ORA) and (ii) functional class scoring (FCS), and the most common application is in differential expression analysis.
-In ORA, differentially expressed genes (DEGs) meeting a significance and/or fold change threshold are queried against curated gene sets.
-A statistical test is performed to ascertain whether the number of DEGs in a particular gene set is higher than that expected by random chance.
-ORA tools can be stand alone software packages or web services, and they use one or more statistical tests (Fisher’s Exact test, hypergeometric test, binomial test, χ2 test, etc) (Draghici et al, 2003; Hosack et al, 2003).
+Widely used functional enrichment tools can be classified into two main categories; (i) over-representation analysis (ORA) and (ii) functional class scoring (FCS), and the most common application is in differential gene expression analysis.
+In ORA, differentially expressed genes (DEGs) meeting a significance and/or fold change threshold are queried against curated pathways (gene sets).
+A statistical test is performed to ascertain whether the number of DEGs belonging to a particular pathway (gene set) is higher than that expected by random chance.
+ORA tools can be stand alone software packages or web services, and they use one or more statistical tests (eg: Fisher’s Exact test, hypergeometric test, binomial test) (Draghici et al, 2003; Hosack et al, 2003).
 FCS tools involve giving each detected gene a differential expression score and then evaluating whether the scores are more positive or negative than expected by chance for each set of genes.
@@
 Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
-As some articles included more than one enrichment analysis, the final dataset included 1624 analyses from 1363 articles.
+As some articles included more than one enrichment analysis, the final dataset included 1624 analyses from 1363 articles (Supplementary Table 1).
 A flow diagram of the survey is provided in Figure 2.
@@
 In the example here, using the inappropriate background resulted in 2.3 fold more differentially regulated gene sets.
-Thus if there are few or no significant results using a correct background list, there may be a temptation may be to use an inappropriate background.
+Thus if there are few or no significant results using a correct background list, there may be a temptation to use an inappropriate background.
 Results here indicate most ORA articles do not report using a background gene list, suggesting that inappropriate ORA use is common.

### 2021-08-19 QC_analysis.Rmd
updated QC

--- 2021-08-18 QC analysis.Rmd
@@
 head(tres)
+
 par(mar=c(5,12,3,1))
@@
 hist(articleMeans)
+
+# PMCs that need to be redone
+tres[which(rowMeans(tres)<1),]  %>% kbl() %>% kable_paper("hover", full_width = F)
 

### 2021-08-25 QC_analysis.Rmd
minor updates

@@
     mqc <- unlist(strsplit(mqc,", "))
+    mqc <- gsub("Not stated","No",mqc)
     myy <- ydat[[m]]
     myy <- unlist(strsplit(myy,", "))
+    myy <- gsub("Not stated","No",myy)
   length(intersect(mqc,myy)) / length(union(mqc,myy))

### 2021-08-25 example2.Rmd
minor updates

--- 2021-08-13 example.Rmd
@@
 ---
-title: "Example gene set analysis"
+title: "Example2 gene set analysis: The case of Azacytidine"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
-I'm using some RNA-seq data looking at the effect of hyperglycemia on hepatocytes.
+I'm using some RNA-seq data looking at the effect of Axacytidine on AML3 cells.
 
@@
 mdat<-getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP038101",mdat$SRP_accession),]
 samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$HG<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
-samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
-s1 <- subset(samplesheet,VPA==0)
+samplesheet$Aza<-as.factor(c(1,1,1,0,0,0))
 
@@
 # counts 
-x1<-x[,which(colnames(x) %in% s1$SRR_accession)]
+x1<-x[,which(colnames(x) %in% samplesheet$SRR_accession)]
 
@@
 # MDS plot
-plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$HG, main="MDS")
+plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=samplesheet$Aza, main="MDS")
 
 #run DESeq2 for LG vs HG
-y <- DESeqDataSetFromMatrix(countData = round(x1), colData = s1, design = ~ HG)
+y <- DESeqDataSetFromMatrix(countData = round(x1), colData = samplesheet, design = ~ Aza)
 y <- DESeq(y)
-LGvHG <- results(y)
-LGvHG<-as.data.frame(LGvHG[order(LGvHG$pvalue),])
-rownames(LGvHG)<-sapply(strsplit(rownames(LGvHG),"\\."),"[[",1)
-head(LGvHG)
+CTLvAZA <- results(y)
+CTLvAZA<-as.data.frame(CTLvAZA[order(CTLvAZA$pvalue),])
+rownames(CTLvAZA)<-sapply(strsplit(rownames(CTLvAZA),"\\."),"[[",1)
+head(CTLvAZA)
 
@@
 
-maplot(LGvHG,"LG v HG")
-make_volcano(LGvHG,"LG v HG")
+maplot(CTLvAZA,"Ctl v Aza")
+make_volcano(CTLvAZA,"Ctl v Aza")
 
@@
 
-m <- mitch_import(LGvHG,DEtype = "DEseq2", geneTable = gt)
+m <- mitch_import(CTLvAZA,DEtype = "DEseq2", geneTable = gt)
 head(m)
@@
 
-LGvHG_up <- rownames(subset(LGvHG,log2FoldChange>0,padj<0.05))
-LGvHG_up <- unique(gt[which(rownames(gt) %in% LGvHG_up),1])
-head(LGvHG_up)
+CTLvAZA_up <- rownames(subset(CTLvAZA,log2FoldChange>0,padj<0.05))
+CTLvAZA_up <- unique(gt[which(rownames(gt) %in% CTLvAZA_up),1])
+head(CTLvAZA_up)
 
-LGvHG_dn <- rownames(subset(LGvHG,log2FoldChange<0,padj<0.05))
-LGvHG_dn <- unique(gt[which(rownames(gt) %in% LGvHG_dn),1])
-head(LGvHG_dn)
+CTLvAZA_dn <- rownames(subset(CTLvAZA,log2FoldChange<0,padj<0.05))
+CTLvAZA_dn <- unique(gt[which(rownames(gt) %in% CTLvAZA_dn),1])
+head(CTLvAZA_dn)
 
-LGvHG_bg <- rownames(LGvHG)
-LGvHG_bg <- unique(gt[which(rownames(gt) %in% LGvHG_bg),1])
-length(LGvHG_bg)
-head(LGvHG_bg)
+CTLvAZA_bg <- rownames(CTLvAZA)
+CTLvAZA_bg <- unique(gt[which(rownames(gt) %in% CTLvAZA_bg),1])
+length(CTLvAZA_bg)
+head(CTLvAZA_bg)
 
-c_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
+c_up <- enricher(gene = CTLvAZA_up, universe = CTLvAZA_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
 c_up <- head(c_up,1000)
@@
 
-c_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
+c_dn <- enricher(gene = CTLvAZA_dn, universe = CTLvAZA_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
 c_dn <- head(c_dn,1000)
@@
 
-LGvHG_de <- rownames(subset(LGvHG,padj<0.05))
-LGvHG_de <- unique(gt[which(rownames(gt) %in% LGvHG_de),1])
-head(LGvHG_de)
+CTLvAZA_de <- rownames(subset(CTLvAZA,padj<0.05))
+CTLvAZA_de <- unique(gt[which(rownames(gt) %in% CTLvAZA_de),1])
+head(CTLvAZA_de)
 
-d_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
+d_de <- enricher(gene = CTLvAZA_de, universe = CTLvAZA_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
 d_de <- head(d_de,1000)
@@
 
-LGvHG_bg <- w$GeneInfo$GeneSymbol
-length(LGvHG_bg)
-head(LGvHG_bg)
+CTLvAZA_bg <- w$GeneInfo$GeneSymbol
+length(CTLvAZA_bg)
+head(CTLvAZA_bg)
 
-f_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
+f_up <- enricher(gene = CTLvAZA_up, universe = CTLvAZA_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
 f_up <- head(f_up,1000)
@@
 
-f_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, maxGSSize = 5000, TERM2GENE = genesets2)
+f_dn <- enricher(gene = CTLvAZA_dn, universe = CTLvAZA_bg, maxGSSize = 5000, TERM2GENE = genesets2)
 f_dn <- head(f_dn,1000)
@@
 
-e_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, maxGSSize = 5000, TERM2GENE = genesets2)
+e_de <- enricher(gene = CTLvAZA_de, universe = CTLvAZA_bg, maxGSSize = 5000, TERM2GENE = genesets2)
 e_de <- head(e_de,1000)
@@
 
-png("images/fcs_ora1.png")
+png("images/fcs_ora2.png")
 plot(euler(v1),quantities = TRUE, main="FCS vs ORA")
@@
 
-png("images/orabg1.png")
+png("images/orabg2.png")
 plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")
@@
 
-png("images/oracomb1.png")
+png("images/oracomb2.png")
 plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")

### 2021-08-29 analysis.Rmd
Assed QC2 data with resolved discrepancies. Added ORA combined vs separate and GSEA parameters.

@@
 barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Assumptions violated", xlim=c(0,100))
-text(y = (1:length(res)*1.2)-0.75  , x = res+100, label = res, pos = 3, cex = 1, col = "black")
+        main = "Assumptions violated", xlim=c(0,1500))
+text(y = (1:length(res)*1.2)-0.75  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
 grid()

### 2021-08-29 manuscript.Rmd
Assed QC2 data with resolved discrepancies. Added ORA combined vs separate and GSEA parameters.

@@
 
+Felisbino, M.
+B., Ziemann, M., Khurana, I., Okabe, J., Al-Hasani, K., Maxwell, S., … El-Osta, A.
+(2021).
+Valproic acid influences the expression of genes implicated with hyperglycaemia-induced complement and coagulation pathways.
+Scientific Reports, 11(1), 2163.
+
 Hong, G., Zhang, W., Li, H., Shen, X., & Guo, Z. (2014). Separate enrichment analysis of pathways for up- and downregulated genes. Journal of the Royal Society, Interface, 11(92), 20130950.
@@
 
+Jassal, B., Matthews, L., Viteri, G., Gong, C., Lorente, P., Fabregat, A., … D’Eustachio, P. (2020). The reactome pathway knowledgebase. Nucleic Acids Research, 48(D1), D498–D503.
+
 Kaspi, A., & Ziemann, M. (2020). Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data. BMC Genomics, 21(1), 447.
+
+Love, M. I., Huber, W., & Anders, S. (2014). Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biology, 15(12), 550.
 
@@
 
+Yu, G., Wang, L.-G., Han, Y., & He, Q.-Y. (2012). clusterProfiler: an R package for comparing biological themes among gene clusters. Omics: A Journal of Integrative Biology, 16(5), 284–287.
+
 Ziemann, M., Eren, Y., & El-Osta, A. (2016). Gene name errors are widespread in the scientific literature. Genome Biology, 17(1), 177.
+
+Ziemann, M., Kaspi, A., & El-Osta, A. (2019). Digital expression explorer 2: a repository of uniformly processed RNA sequencing data. GigaScience, 8(4). doi:10.1093/gigascience/giz022

### 2021-08-30 QC_analysis2.Rmd
major manuscript changes

---
title: "PMC Survey QC Analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

## Intro

Here we are performing an analysis of 200 articles which is randomly selected from 1500 PMC articles.

```{r,begin}

knitr::opts_chunk$set(fig.width=7, fig.height=5)

library("wordcloud")
library("RColorBrewer")
library("wordcloud2")
library("reutils")
library("XML")
library("kableExtra")
library("Biobase")
library("vioplot")

```

## Overview of included and excluded analyses

```{r,exclude1}

x <- read.table("data/QC-analysis2.tsv",header=TRUE,fill=TRUE,sep="\t")
head(x)
colnames(x)
dim(x)

exclude <- subset(x,x$GS.version=="EXCLUDE")
nrow(exclude)
length(unique(exclude$Pubmed.Central.ID))

x <- subset(x,x$GS.version!="EXCLUDE")
nrow(x)
length(unique(x$Pubmed.Central.ID))

```
## Journal

```{r,Journal1}

journal <- x$Journal
journal_split <- strsplit(journal,", ")
journal <- unlist(journal_split)
res <- table(journal)
res <- res[order(res)]
length(res)
res

par(mar=c(1,1,1,1))
#names(res) <- gsub("Gene expression array","RNA array",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Journal", xlim=c(0,17))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,5))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Journal", xlim=c(0,165))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 7, label = res2, pos = 3, cex = 1, col = "black")

dir.create("images")
png("images/journals1.png",width=400,height=300)
par(mar=c(5,12,3,3))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Journal", xlim=c(0,170))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 10, label = res2, pos = 3, cex = 1, col = "black")
dev.off()
```

## Omics type

```{r,omics1}

omics <- x$Omics.type
omics_split <- strsplit(omics,", ")
omics <- unlist(omics_split)
res <- table(omics)
res <- res[order(res)]
length(res)
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Gene expression array","RNA array",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Omics type", xlim=c(0,100))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,5))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Omics type", xlim=c(0,100))
text(y = (1:length(res2)*1.2)-1.2  , x = res2+5, label = res2, pos = 3, cex = 1, col = "black")

png("images/omics1.png",width=400,height=300)
par(mar=c(5,12,3,3))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Omics type", xlim=c(0,100))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 6, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

```

## Organism

```{r,org1}

org <- x$Organism
org_split <- strsplit(org,", ")
org <- unlist(org_split)
res <- table(org)
res <- res[order(res)]
length(res)
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Homo sapiens","human",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"), scale=c(4,.5))

par(mar=c(5,12,3,5))
names(res) <- gsub("human","Homo sapiens",names(res))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Organism", xlim=c(0,200))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,5))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Organism", xlim=c(0,200))
text(y = (1:length(res2)*1.2)-1.2  , x = res2+7, label = res2, pos = 3, cex = 1, col = "black")

png("images/organisms1.png",width=400,height=300)
par(mar=c(5,12,3,3))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Organism", xlim=c(0,200))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 15, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

```

## Gene set library

```{r,gsl1}

GSL <-x$Gene.set.library
GSL_split <- strsplit(GSL,", ")
GSL <- unlist(GSL_split)
res <- table(GSL)
res <- res[order(res)]
length(res)
which(names(res)=="Not stated")/sum(res)*100
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Homo sapiens","human",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,140))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,5))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,140))
text(y = (1:length(res2)*1.2)-1.2  , x = res2+7, label = res2, pos = 3, cex = 1, col = "black")

png("images/genesetlib1.png",width=400,height=300)
par(mar=c(5,12,3,3))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,150))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 15, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

```

## Gene set version

```{r,gsv1}

GSV <-x$GS.version
res <- table(GSV)
res
res[1]/sum(res)*100

par(mar=c(14,12,3,10))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set version defined", xlim=c(0,250))
text(y = (1:length(res)*1.2) - 0.75 , x = res+15, label = res, pos = 3, cex = 1, col = "black")

png("images/genesetvers1.png",width=300,height=150)
par(mar=c(5,5,3,3))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set version defined", xlim=c(0,250))
text(y = (1:length(res)*1.2)-1.2  , x = res + 20, label = res, pos = 3, cex = 1, col = "black")
dev.off()

```

## Statistical test used

```{r,test1}

test <-x$Statistical.test.used
test <- strsplit(test,", ")
test <- unlist(test)
res <- table(test)
res <- res[order(res)]
res[which(names(res)=="Not stated")] / sum(res) * 100
length(res)
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Test used", xlim=c(0,150))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,5))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Test used", xlim=c(0,150))
text(y = (1:length(res2)*1.2)-1.1  , x = res2+10, label = res2, pos = 3, cex = 1, col = "black")

png("images/stattest1.png",width=400,height=300)
par(mar=c(5,12,3,3))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Test used", xlim=c(0,160))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 15, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

```

## FDR Correction

```{r,fdr1}

fdr <-x$FDR.Correction
fdr <- strsplit(fdr,", ")
fdr <- unlist(fdr)
res <- table(fdr)
res <- res[order(res)]
sum(res[which(names(res)!="Yes")])/sum(res)*100
res

par(mar=c(1,1,1,5))
wordcloud(words = names(res), freq = res, min.freq = 1, 
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))

par(mar=c(10,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "FDR correction performed", xlim=c(0,150))
text(y = (1:length(res)*1.2) - 0.8 , x = res+10, label = res, pos = 3, cex = 1, col = "black")

png("images/fdr1.png",width=300,height=200)
par(mar=c(5,8,3,3))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "FDR correction performed", xlim=c(0,160))
text(y = (1:length(res)*1.2)-1.2  , x = res + 15, label = res, pos = 3, cex = 1, col = "black")
dev.off()
```

## App used

```{r,app1}

App <-x$App.used
App_split <- strsplit(App,", ")
App <- unlist(App_split)
res <- table(App)
res <- res[order(res)]
res[which(names(res)=="Not stated")]/sum(res)*100
length(res)
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "App used", xlim=c(0,60))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,5))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App used", xlim=c(0,80))
text(y = (1:length(res2)*1.2)-1.1  , x = res2+5, label = res2, pos = 3, cex = 1, col = "black")

png("images/app1.png",width=400,height=300)
par(mar=c(5,12,3,3))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App used", xlim=c(0,80))
text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 5, label = res2, pos = 3, cex = 1, col = "black")
dev.off()

```

## App version

```{r,appv1}

APV <-x$App.Version
res <- table(APV)
res
res[1]/sum(res)*100

par(mar=c(14,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App version defined", xlim=c(0,200))
text(y = (1:length(res)*1.2) - 0.75 , x = res+15, label = res, pos = 3, cex = 1, col = "black")

png("images/appvers1.png",width=300,height=150)
par(mar=c(5,5,3,3))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App version defined", xlim=c(0,200))
text(y = (1:length(res)*1.2)-1.2  , x = res + 15, label = res, pos = 3, cex = 1, col = "black")
dev.off()

```

## Code available

```{r,code1}

code <-x$Code.availability
res <- table(code)
res
res[1]/sum(res)*100

par(mar=c(14,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Code availability", xlim=c(0,50))
text(y = (1:length(res)*1.2) - 0.75 , x = res+2, label = res, pos = 3, cex = 1, col = "black")

png("images/code1.png",width=300,height=150)
par(mar=c(5,5,3,3))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Code availability", xlim=c(0,50))
text(y = (1:length(res)*1.2)-1.2  , x = res + 3, label = res, pos = 3, cex = 1, col = "black")
dev.off()

```

## Background gene set

```{r,background1}

BG <-x$Background.gene.set
BG <- gsub("Yes","Yes, correct",BG)
res <- table(BG)
res
sum(res[which(names(res)!="Yes")])/sum(res)*100

par(mar=c(10,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Background list specification", xlim=c(0,200))
text(y = (1:length(res)*1.2) - 0.85 , x = res+10, label = res, pos = 3, cex = 1, col = "black")

png("images/bg1.png",width=300,height=200)
par(mar=c(5,9,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Background list specified", xlim=c(0,220))
text(y = (1:length(res)*1.2)-1.1  , x = res + 18, label = res, pos = 3, cex = 1, col = "black")
dev.off()

```

## Gene lists provided

```{r,genelists1}

GL <-x$Gene.lists.provided
res <- table(GL)
res
sum(res[which(names(res)!="Yes")])/sum(res)*100

par(mar=c(14,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene lists provided", xlim=c(0,160))
text(y = (1:length(res)*1.2) - 0.75 , x = res+8, label = res, pos = 3, cex = 1, col = "black")

png("images/genelists1.png",width=300,height=150)
par(mar=c(5,5,3,3))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene lists provided", xlim=c(0,160))
text(y = (1:length(res)*1.2)-1.2  , x = res + 9, label = res, pos = 3, cex = 1, col = "black")
dev.off()

```

## Methodological flaws

```{r,assumptions1}

ok <- nrow(subset(x,Assumptions.violated=="No"))
ok
bad <- nrow(subset(x,Assumptions.violated!="No"))
bad
ok/sum(bad,ok)*100

ass <-x$Assumptions.violated
ass <- gsub("^No$","None",ass)
ass <- strsplit(ass,", ")
ass <- unlist(ass)
res <- table(ass)
res <- res[order(res)]
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))

par(mar=c(8,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Methodological flaws", xlim=c(0,200))
text(y = (1:length(res)*1.2) - 0.9 , x = res+8, label = res, pos = 3, cex = 1, col = "black")

png("images/assumptions1.png",width=400,height=250)
par(mar=c(5,12,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Methodological flaws", xlim=c(0,220))
text(y = (1:length(res)*1.2)-1.1  , x = res + 18, label = res, pos = 3, cex = 1, col = "black")
dev.off()

```

## Separated or combined

```{r,combined}

comb <-x$Separated.up.and.down
res <- table(comb)
res
sum(res[which(names(res)!="Separated")])/sum(res)*100
res <- res[which(names(res)!="-")]

par(mar=c(10,12,3,5))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Separated or combined ORA (differential expression)", xlim=c(0,90))
grid()
text(y = (1:length(res)*1.2) - 0.9 , x = res+5, label = res, pos = 3, cex = 1, col = "black")

png("images/comb1.png",width=300,height=200)
par(mar=c(5,8,3,2))
barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Separated or combined ORA", xlim=c(0,100))
text(y = (1:length(res)*1.2)-1.1  , x = res + 5, label = res, pos = 3, cex = 1, col = "black")
dev.off()

```

## GSEA parameters

```{r,gseaparam}

param <-x$Parameters
param <- param[which(param!="-")]
param <- gsub(" ","",param)
str(param)

rankspec <- sapply(strsplit(param,","),"[[",1)
rankspec <- gsub("rank=","",rankspec)
rankspec <- table(rankspec)
rankspec

algospec <- sapply(strsplit(param,","),"[[",2)
algospec <- gsub("algo=","",algospec)
algospec <- table(algospec)
algospec

weightspec <- sapply(strsplit(param,","),"[[",3)
weightspec <- gsub("weight=","",weightspec)
weightspec <- table(weightspec)
weightspec

par(mar=c(10,12,3,4))
params <- rbind(rankspec,algospec,weightspec)
rownames(params) <- c("rank method", "test type", "weight type")
colnames(params) <- c("not stated", "stated")
barplot(t(params),horiz=TRUE,las=1,xlim=c(0,30),legend = colnames(params),
        main="GSEA parameter reporting", xlab="no. analyses",
        args.legend = list(x = "topleft", inset = c(0.05, 0.1)))
text(y = (1:nrow(params)*1.2) - 0.7 , x = params[,1]-3, label = params[,1], pos = 3, cex = 1, col = "white")
text(y = (1:nrow(params)*1.2) - 0.7 , x = params[,1]+2, label = params[,2], pos = 3, cex = 1, col = "black")

png("images/gseaparam1.png",width=350,height=250)
par(mar=c(5,8,3,2))
params <- rbind(rankspec,algospec,weightspec)
rownames(params) <- c("rank method", "test type", "weight type")
colnames(params) <- c("not stated", "stated")
barplot(t(params),horiz=TRUE,las=1,xlim=c(0,30),legend = colnames(params),
        main="GSEA parameter reporting", xlab="no. analyses",
        args.legend = list(x = "topleft", inset = c(0.05, 0.1)))
text(y = (1:nrow(params)*1.2) - 0.7 , x = params[,1]-3, label = params[,1], pos = 3, cex = 1, col = "white")
text(y = (1:nrow(params)*1.2) - 0.7 , x = params[,1]+2, label = params[,2], pos = 3, cex = 1, col = "black")
dev.off()

```

## Session information

```{r,session}

sessionInfo()

```

### 2021-08-30 manuscript.Rmd
major manuscript changes

@@
 We have collated a list of 2941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
-We sampled 1500 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.
+We sampled 200 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.
 
@@
 These data were entered into a Google Spreadsheet by a team of five researchers.
+These articles were cross checked by another team member and any discrepancies were resolved.
 For quality control, 200 randomly selected assessed articles were cross-checked by two senior team members to ensure consistency across the five team members.
-Data were then cleaned of spelling mistakes and other irregularities by loading them into R and tabulating the common occurrences, then fixing the original spreadsheet.
 Cleaned data were then loaded into R v4.1 for downstream analysis.
+
 We rated each analysis with a simple approach that deducted points for missing methodological details and awarding points for including extra information (Table 1).
@@
       tab8 [label = '@@8']
-      tab9 [label = '@@9']
-      tab10 [label = '@@10']
-      tab11 [label = '@@11']
-      tab12 [label = '@@12']
-      tab13 [label = '@@13']
 
       # edge definitions with the node IDs
-      tab1 -> tab2  -> tab4 -> tab5 -> tab6 -> tab7 -> tab11;
-      tab7 -> tab12 ;
-      tab7 -> tab13 ;
-      tab2-> tab3 ;
-      tab4-> tab8 -> tab9 -> tab10 -> tab13;
+      tab1 -> tab2  -> tab4 -> tab5 -> tab6 -> tab7;
+      tab2 -> tab3 ;
+      tab6 -> tab8 ;
       }
       [1]: 'PMC keyword search (2019): 2941 results'
-      [2]: '1500 PMC articles selected'
-      [3]: '133 articles excluded'
-      [4]: '1363 articles analyzed by team of five researchers'
-      [5]: '1624 enrichment analyses described'
-      [6]: 'data entry to spreadsheet'
-      [7]: 'data analysis in R'
-      [8]: 'Subset of 231 analyses cross-checked by two senior researchers'
-      [9]: 'data entry to spreadsheet'
-      [10]: 'data analysis in R'
-      [11]: 'Summary charts'
-      [12]: 'Integration with journal and article metrics'
-      [13]: 'Data validation/crosschecking'
+      [2]: '200 PMC articles selected'
+      [3]: '14 articles excluded'
+      [4]: '235 analyses screened and cross-checked'
+      [5]: 'data entry to spreadsheet'
+      [6]: 'data analysis in R'
+      [7]: 'Summary charts'
+      [8]: 'Integration with journal and article metrics'
 ")
@@
 ```
-Figure 2. A summary of the survey of functional enrichment analyses.
+*Figure 2. A summary of the survey of functional enrichment analyses.*
 
-There were articles from 330 journals in the screen, with *Scientific Reports*, *Oncology Letters* and *International Journal of Molecular Science* being the biggest contributors (Figure S1A).
-There were 35 different omics types, with RNA-seq and gene expression microarray being the most popular (Figure S1B).
-There were 175 different species under study, but human was the most common with 1104 analyses (Figure S1C).
-We recorded the use of 101 different gene set libraries, with GO and KEGG being the most frequently used (Figure 3A).
-There were 85 studies where the gene set libraries used were not defined in the article.
-Only 108 analyses reported the version of the gene set library used (Figure 3B).
-There were XX different statistical tests used, and the most common reported tests were Fisher's Exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Figure 3C).
-Only 754 of 1594 studies (XX%) described correction of p-values for multiple testing (Figure 3D).
-There were 145 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 174 analyses (XX%) did not state what tool was used (Figure 3E).
-The version of the software used was provided in only 421 of 1611 analyses (26%) (Figure 3F).
+There were articles from 100 journals in the sample, with *PLoS One*, *Scientific Reports*, and *PeerJ* being the biggest contributors (Figure S1A).
+There were 18 different omics types, with gene expression array and RNA-seq being the most popular (Figure S1B).
+There were 31 different species under study, but human was the most common with 157 analyses (Figure S1C).
+We recorded the use of 26 different gene set libraries, with GO and KEGG being the most frequently used (Figure 3A).
+There were 14 analyses where the gene set libraries used were not defined in the article.
+Only 18 analyses reported the version of the gene set library used (Figure 3B).
+There were 12 different statistical tests used, and the most common reported tests were Fisher's Exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Figure 3C).
+Out of the 225 analyses that performed a statistical test, only 119 (53%) described correction of p-values for multiple testing (Figure 3D).
+There were 50 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 15 analyses (6.4%) did not state what tool was used (Figure 3E).
+The version of the software used was provided in only 68 of 235 analyses (29%) (Figure 3F).
+
 For analyses using ORA methods, we studied what background gene set was used (Figure 3G). 
@@
 In a few cases, background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
-In only 67/1460 of cases (4.6%), the appropriate background list was described in the article.
-Of the 293 analyses which used a scripted computer language, only 10 provided links to the code used for enrichment analysis (3.4%) (Figure 3H).
-For 579 of 1618 analyses, the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Figure 3I).
-Next, we quantified the frequency of methodological issues that would undermine the conclusions (Figure 3J).
-Lack of appropriate background was the most common issue (1361 cases), followed by FDR (761 cases), then inference without test (61 cases), inference without data shown (32 cases) and misinterpreted FDR values (3 cases).
-Only 160 analyses did not exhibit any of these major methodological issues.
+In only 8/197 of cases (4.1%), the appropriate background list was described in the article.
+
+Of the 47 analyses which used a scripted computer language, only 3 provided links to the code used for enrichment analysis (6.4%) (Figure 3H).
+For 93 of 235 analyses (40%), the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Figure 3I).
+
+Next, we quantified the frequency of methodological issues and reporting that would undermine the conclusions (Figure 3J).
+Lack of appropriate background was the most common issue (179 cases), followed by FDR (94 cases), then lack of data shown (13), inference without test (11 cases), and misinterpreted FDR values (2 cases).
+Only 35 analyses (15%) did not exhibit any of these major methodological issues.
 Taken together, these data suggest methodological and reporting deficiencies are widespread in published functional enrichment analyses.

### 2021-08-30 QC_analysis2.Rmd
working on the results

@@
 dir.create("images")
-png("images/journals1.png",width=400,height=300)
+png("images/journals2.png",width=400,height=300)
 par(mar=c(5,12,3,3))
@@
 
-png("images/omics1.png",width=400,height=300)
+png("images/omics2.png",width=400,height=300)
 par(mar=c(5,12,3,3))
@@
 
-png("images/organisms1.png",width=400,height=300)
+png("images/organisms2.png",width=400,height=300)
 par(mar=c(5,12,3,3))
@@
 
-png("images/genesetlib1.png",width=400,height=300)
+png("images/genesetlib2.png",width=400,height=300)
 par(mar=c(5,12,3,3))
@@
 
-png("images/genesetvers1.png",width=300,height=150)
+png("images/genesetvers2.png",width=300,height=150)
 par(mar=c(5,5,3,3))
@@
 
-png("images/stattest1.png",width=400,height=300)
+png("images/stattest2.png",width=400,height=300)
 par(mar=c(5,12,3,3))
@@
 
-png("images/fdr1.png",width=300,height=200)
+png("images/fdr2.png",width=300,height=200)
 par(mar=c(5,8,3,3))
@@
 
-png("images/app1.png",width=400,height=300)
+png("images/app2.png",width=400,height=300)
 par(mar=c(5,12,3,3))
@@
 
-png("images/appvers1.png",width=300,height=150)
+png("images/appvers2.png",width=300,height=150)
 par(mar=c(5,5,3,3))
@@
 
-png("images/code1.png",width=300,height=150)
+png("images/code2.png",width=300,height=150)
 par(mar=c(5,5,3,3))
@@
 
-png("images/bg1.png",width=300,height=200)
+png("images/bg2.png",width=300,height=200)
 par(mar=c(5,9,3,2))
@@
 
-png("images/genelists1.png",width=300,height=150)
+png("images/genelists2.png",width=300,height=150)
 par(mar=c(5,5,3,3))
@@
 
-png("images/assumptions1.png",width=400,height=250)
+png("images/assumptions2.png",width=400,height=250)
 par(mar=c(5,12,3,2))
@@
 
-png("images/comb1.png",width=300,height=200)
+png("images/comb2.png",width=300,height=200)
 par(mar=c(5,8,3,2))
@@
 
-png("images/gseaparam1.png",width=350,height=250)
+png("images/gseaparam2.png",width=350,height=250)
 par(mar=c(5,8,3,2))
@@
 
+## Score
+
+Here I am proposing a scoring scheme.
+If there is information missing from the article, or basic mistakes are made, then a point is deducted.
+If the article goes over and above the basic reproducibility, then they are awarded a point.
+
+Gene set library origin not stated = -1
+
+Gene set library version not stated = -1
+
+Stat test not stated = -1
+
+No stat test conducted = -1
+
+No FDR correction conducted = -1
+
+App used not stated = -1
+
+App version not stated = -1
+
+Background list not defined = -1
+
+Inappropriate background list used = -1
+
+Code availability = +1
+
+Gene lists provided = +1
+
+```{r,score}
+
+score <- function(r){
+  r[is.na(r)] <- 0
+  SCORE=0
+  # gene set lib
+  if(r["Gene.set.library"]=="Not stated"){
+    SCORE=SCORE-1
+  }
+  # GS version
+  if(r["GS.version"]=="No"){
+    SCORE=SCORE-1
+  }
+  # stat test
+  if(r["Statistical.test.used"]=="No test"){
+    SCORE=SCORE-1
+  }
+  if(r["Statistical.test.used"]=="Not stated"){
+    SCORE=SCORE-1
+  }
+  # FDR
+  if(r["FDR.Correction"]!="Yes"){
+    SCORE=SCORE-1
+  }
+  # app used
+  if(r["App.used"]=="Not stated"){
+    SCORE=SCORE-1
+  }
+  # App version
+  if(r["App.Version"]=="No"){
+    SCORE=SCORE-1
+  }
+  # Code availability
+  if(r["Code.availability"]=="Yes"){
+    SCORE=SCORE+1
+  }
+  # Background
+  if(r["Background.gene.set"]=="Not stated"){
+    SCORE=SCORE-1
+  }
+  if(r["Background.gene.set"]=="No"){
+    SCORE=SCORE-1
+  }
+  if(r["Background.gene.set"]=="Stated, but incorrect"){
+    SCORE=SCORE-1
+  }
+  # gene list
+  if(r["Gene.lists.provided"]=="Yes"){
+    SCORE=SCORE+1
+  }
+  return(SCORE)
+}
+
+scores <- apply(X = x,MARGIN = 1, FUN = score)
+barplot(table(scores), xlab="analysis score",ylab="frequency")
+x$scores <- scores
+length(which(scores>0))
+
+mean(x$scores)
+sd(x$scores)
+median(x$scores)
+
+```
+
+## Journal rank
+
+```{r,journal rank}
+
+jmetrics <- read.table("scimagojr_2020.csv",sep=";",header=TRUE,dec = ",")
+jmetrics$Title <- toupper(jmetrics$Title)
+head(jmetrics,3)
+dim(jmetrics)
+
+nlmcat <- readLines("nlmcatalog_result.txt")
+journaltitle <- nlmcat[grep("Title\\(s\\):",nlmcat)]
+journaltitle <- gsub("Title\\(s\\): ","",journaltitle)
+journaltitle <- gsub("\\.$","",journaltitle)
+journaltitle <- toupper(journaltitle) 
+journalabbrev <- nlmcat[grep("Title Abbreviation:",nlmcat)]
+journalabbrev <- sapply(strsplit(journalabbrev,":"),"[[",2)
+journalabbrev <- gsub(" $","",journalabbrev)
+journalabbrev <- gsub("^ ","",journalabbrev)
+jdf <- data.frame(journalabbrev,journaltitle)
+
+mjdf <- merge(jdf,jmetrics,by.x="journaltitle",by.y="Title")
+
+xm <- merge(x,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)
+
+```
+
+## Analysis scores vs Journal rank
+
+```{r,scoresvsjrank1}
+
+tail(xm$Journal)
+
+mylm1 <- lm (xm$scores ~ xm$SJR)
+plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
+abline(mylm1,col="red")
+
+cor.test(xm$scores,xm$SJR,method="pearson")
+
+cor.test(xm$scores,xm$SJR,method="spearman")
+
+png("images/score_sjr1.png")
+par(mar=c(5,5,3,1))
+plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
+abline(mylm1,col="red")
+grid()
+dev.off()
+
+```
+
+## Best and worst journals
+
+Minimum of 5 analyses.
+
+```{r,goodbadugly}
+
+xmm <-xm[,c("Journal","scores")]
+tab <- table(xmm$Journal)
+tab <- tab[which(tab>3)]
+xmm <- xmm[which(xmm$Journal %in% names(tab)),]
+dim(xmm)
+jres <- aggregate(. ~ Journal, xmm ,mean)
+jsd <- aggregate(. ~ Journal, xmm ,sd)
+jres$sd <- jsd$scores
+jres <- jres[order(jres$scores),]
+dim(jres)
+head(jres)
+
+par(mar=c(5,12,3,1))
+barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Highest scoring journals",xlim=c(-4,0))
+grid()
+
+png("images/score_highest1.png")
+par(mar=c(5,12,3,1))
+barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Highest scoring journals",xlim=c(-4,0))
+grid()
+dev.off()
+
+par(mar=c(5,12,3,1))
+barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Lowest scoring journals",xlim=c(-5,0))
+grid()
+
+png("images/score_lowest1.png")
+par(mar=c(5,12,3,1))
+barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Lowest scoring journals",xlim=c(-5,0))
+grid()
+dev.off()
+
+j <- unique(xmm$Journal)
+jscores <- sapply(j,function(jj) {
+ xmm[which(xmm$Journal == jj ),2]  
+})
+
+jscores <- jscores[order(unlist(lapply(jscores,mean)))]
+
+par(mar=c(5,12,3,1))
+vioplot(tail(jscores,20),horizontal = TRUE,las=1,cex.axis=0.75,main="Highest scoring journals")
+
+par(mar=c(5,12,3,1))
+vioplot(head(jscores,20),horizontal = TRUE,las=1,cex.axis=0.75,main="Lowest scoring journals")
+
+```
+
 ## Session information

### 2021-08-30 analysis.Rmd
new separated/combined ora and GSEA params
working on the results

@@
 
+Here we are also using the QC2 data where possible.
+
 ```{r,exclude1}
 
-x <- read.table("PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
+x <- read.table("data/PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
+y <- read.table("data/QC-analysis2.tsv",header=TRUE,fill=TRUE,sep="\t")
+y <- y[,1:16]
+x <- x[which(! x$Pubmed.Central.ID %in% y$Pubmed.Central.ID),]
+x <- rbind(x,y)
+
 head(x)
@@
 length(unique(x$Pubmed.Central.ID))
+
+```
+
+## Filter for studies included in QC2
+
+```{r,xy}
 
@@
 scores <- apply(X = x,MARGIN = 1, FUN = score)
-hist(scores)
+barplot(table(scores), xlab="analysis score",ylab="frequency")
 x$scores <- scores
@@
 median(x$scores)
+length(scores)
+which(scores>0)
 
-png("images/hist1.png")
+png("images/hist1.png",width=300,height=200)
 par(mar=c(5,5,3,1))
-hist(scores)
-grid()
+barplot(table(scores), xlab="analysis score",ylab="frequency")
 dev.off()
@@
 
+hist(xm$SJR,xlab="SJR",main="SJR distribution")
+
 mylm1 <- lm (xm$scores ~ xm$SJR)
-plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
+plot(xm$SJR,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2), pch=19,
+  xlab="SJR",ylab="score",main="Analysis scores vs. journal metrics")
 abline(mylm1,col="red")
@@
 cor.test(xm$scores,xm$SJR,method="pearson")
-
 cor.test(xm$scores,xm$SJR,method="spearman")
 
-png("images/score_sjr1.png")
+png("images/score_sjr1.png",width=350,height=300)
 par(mar=c(5,5,3,1))
-plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
+plot(xm$SJR,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2), pch=19,
+  xlab="SJR",ylab="score",main="Analysis scores vs. journal metrics")
 abline(mylm1,col="red")
@@
 ```
+
 ## Best and worst journals
@@
 
-```{r,goodbadugly}
+```{r,goodbadugly,fig.height=5,fig.width=6}
 
@@
 tab <- table(xmm$Journal)
-tab <- tab[which(tab>5)]
+tab <- tab[which(tab>=5)]
 xmm <- xmm[which(xmm$Journal %in% names(tab)),]
@@
 barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
-        main = "Highest scoring journals",xlim=c(-4,0))
+        main = "Highest scoring",xlim=c(-4,0))
 grid()
 
-png("images/score_highest1.png")
+png("images/score_highest1.png",width=300,height=350)
 par(mar=c(5,12,3,1))
 barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
-        main = "Highest scoring journals",xlim=c(-4,0))
+        main = "Highest scoring",xlim=c(-4,0))
 grid()
@@
 
-png("images/score_lowest1.png")
+png("images/score_lowest1.png",width=300,height=350)
 par(mar=c(5,12,3,1))
 barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
-        main = "Lowest scoring journals",xlim=c(-5,0))
+        main = "Lowest scoring",xlim=c(-5,0))
 grid()
@@
 
-```{r,scoresvsjrank2}
+```{r,scoresvsjrank2,fig.height=5,fig.width=6}
 
@@
 mylm2 <- lm (jjres$scores ~ jjres$SJR)
-plot(jjres$SJR,jjres$scores,xlab="SJR",ylab="mean score",main="Association of mean analysis score with journal metrics")
+plot(jjres$SJR,jjres$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+  xlab="SJR",ylab="mean score",main="Mean analysis score vs. journal metrics")
 abline(mylm2,col="red")
 
-png("images/mean_sjr1.png")
+png("images/mean_sjr1.png",width=350,height=300)
 par(mar=c(5,5,3,1))
-plot(jjres$SJR,jjres$scores,xlab="SJR",ylab="mean score",main="Association of mean analysis score with journal metrics")
+plot(jjres$SJR,jjres$scores,  col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+  xlab="SJR",ylab="mean score",main="Mean analysis score vs. journal metrics")
 abline(mylm2,col="red")
@@
 mylm2 <- lm (xm$scores ~ xm$cites)
-plot(xm$cites,xm$scores,xlab="Num cites",ylab="score",main="Association of analysis scores with number of citations")
+plot(xm$cites,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+  xlab="Num cites",ylab="score",main="Association of analysis scores with number of citations")
 abline(mylm2,col="red")
@@
 mylm2 <- lm (xm$scores ~ log2(xm$cites))
-plot(log2(xm$cites),xm$scores,xlab="log2(citations)",ylab="score",main="Association of analysis scores with log2(citations)")
+plot(log2(xm$cites),xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
 abline(mylm2,col="red")
 
-png("images/cites1.png")
+png("images/cites1.png",width=350,height=300)
 par(mar=c(5,5,3,1))
-plot(log2(xm$cites),xm$scores,xlab="log2(citations)",ylab="score",main="Association of analysis scores with log2(citations)")
+plot(log2(xm$cites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
 abline(mylm2,col="red")

### 2021-08-30 manuscript.Rmd
working on the results

@@
 These articles were cross checked by another team member and any discrepancies were resolved.
-For quality control, 200 randomly selected assessed articles were cross-checked by two senior team members to ensure consistency across the five team members.
+For analyses using GSEA, we scanned the articles to identify whether key methodological steps were described, including (i) the gene weighting parameter, (ii) test type, ie: permuting sample labels or genes, and (iii) method used for ranking genes.
+For ORA of differential gene expression analyses, we inspected the articles to see evidence that researchers were conducting enrichment of up and down regulated genes separately or combined.
 Cleaned data were then loaded into R v4.1 for downstream analysis.
 
+For assessment of enrichment analysis quality with journal metrics and citations, we required a larger set of studies so we sampled and analysed a further 1300 articles from PMC.
+Results from this sample were not double-checked and so may have a small number of inaccuracies.
 We rated each analysis with a simple approach that deducted points for missing methodological details and awarding points for including extra information (Table 1).
@@
 Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
-As some articles included more than one enrichment analysis, the final dataset included 1624 analyses from 1363 articles (Supplementary Table 1).
+As some articles included more than one enrichment analysis, the dataset included 235 analyses from 186 articles (Supplementary Table 1).
 A flow diagram of the survey is provided in Figure 2.
 
-```{r,diagram1,echo=FALSE}
+```{r,diagram1,echo=FALSE,fig.width=8,fig.height=4}
 
@@
       tab8 [label = '@@8']
+      tab9 [label = '@@9']
+      tab10 [label = '@@10']
+      tab11 [label = '@@11']
+      tab12 [label = '@@12']
+      tab13 [label = '@@13']
 
@@
       tab2 -> tab3 ;
-      tab6 -> tab8 ;
+      tab1 -> tab9 -> tab10;
+      tab5 -> tab13   ;
+      tab9 -> tab11 -> tab12 -> tab13 -> tab8 ; 
       }
@@
       [6]: 'data analysis in R'
-      [7]: 'Summary charts'
-      [8]: 'Integration with journal and article metrics'
-")
+      [7]: 'summary charts'
+      [8]: 'integration with journal and article metrics'
+      [9]: 'additional 1300 PMC articles selected'
+      [10]: '118 articles excluded'
+      [11]: '1395 articles screened'
+      [12]: 'data entry to spreadsheet'
+      [13]: 'data analysis in R'
+",
+width = 700)
 
@@
 Only 35 analyses (15%) did not exhibit any of these major methodological issues.
+
+During the course of this survey, we also noticed some studies that grouped up- and down-regulated gene lists together prior to ORA, a practice we were not expecting.
+To assess this in a systematic way we assessed differential gene expression studies to determine how common it was to combine up- and down-regulated gene lists prior to ORA (Figure 3K).
+From 109 analyses, just 28 studies performed stated they performed separate analysis of up- and down-regulated gene sets.
+
+We also looked at studies performing GSEA, and whether three important analytical choices were described in the methods.
+These are (i) the gene weighting parameter, (ii) test type, ie: permuting sample labels or genes, and (iii) method used for ranking genes.
+These parameters were not stated in more than half of the analyses.
+
 Taken together, these data suggest methodological and reporting deficiencies are widespread in published functional enrichment analyses.
 
-![Figure S1a. Representation of journals in the data set.](images/journals1.png "Journals")
+![Figure S1a. Representation of journals in the data set.](images/journals2.png "Journals")
 
-![Figure S1B. Representation of different omics types in the data set.](images/omics1.png "Omics")
+![Figure S1B. Representation of different omics types in the data set.](images/omics2.png "Omics")
 
-![Figure S1C. Representation of different organisms in the data set.](images/organisms1.png "Organisms")
+![Figure S1C. Representation of different organisms in the data set.](images/organisms2.png "Organisms")
 
-![Figure 3A. Representation of different gene set libraries in the data set.](images/genesetlib1.png "Geneset libraries")
+![Figure 3A. Representation of different gene set libraries in the data set.](images/genesetlib2.png "Geneset libraries")
 
-![Figure 3B. Proportion of analyses that reported version information for gene sets used.](images/genesetvers1.png "Geneset version")
+![Figure 3B. Proportion of analyses that reported version information for gene sets used.](images/genesetvers2.png "Geneset version")
 
-![Figure 3C. Representation of different statistical tests in the data set.](images/stattest1.png "Geneset libraries")
+![Figure 3C. Representation of different statistical tests in the data set.](images/stattest2.png "Geneset libraries")
 
-![Figure 3D. Recognition of p-value correction for multiple testing in the data set.](images/fdr1.png "FDR")
+![Figure 3D. Recognition of p-value correction for multiple testing in the data set.](images/fdr2.png "FDR")
 
-![Figure 3E. Representation of different software tools for enrichment analysis in the data set.](images/app1.png "Software used")
+![Figure 3E. Representation of different software tools for enrichment analysis in the data set.](images/app2.png "Software used")
 
-![Figure 3F. Proportion of analyses that reported version information for software used.](images/appvers1.png "Software version")
+![Figure 3F. Proportion of analyses that reported version information for software used.](images/appvers2.png "Software version")
 
-![Figure 3G. Background gene set usage and reporting in the dataset.](images/bg1.png "Background gene list used")
+![Figure 3G. Background gene set usage and reporting in the dataset.](images/bg2.png "Background gene list used")
 
-![Figure 3H. Proportion of scripted analyses that provided software code.](images/code1.png "Code available")
+![Figure 3H. Proportion of scripted analyses that provided software code.](images/code2.png "Code available")
 
-![Figure 3I. Proportion of scripted analyses that provided gene profiles.](images/genelists1.png "Gene lists provided")
+![Figure 3I. Proportion of scripted analyses that provided gene profiles.](images/genelists2.png "Gene lists provided")
 
-![Figure 3J. Proportion of analyses with different methodological problems.](images/assumptions1.png "Assumptions violated")
+![Figure 3J. Proportion of analyses with different methodological problems.](images/assumptions2.png "Assumptions violated")
+
+![Figure 3K. Representation of journals in the data set.](images/comb2.png "Journals")
+
+![Figure 3L. Reporting of GSEA parameters.](images/gseaparam2.png "Journals")
 
@@
 
+In order to make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
+We therefore analysed a further 1300 articles, bringing the total number of analyses described up to 1365. 
+This larger sample did not undergo double-checking so may contain a small number of errors.
+
 We then scored each analysis based on the inclusion or lack of methodological issues and included details.
-The median score was -4, with a mean of -3.5 and standard deviation of 1.3 (Figure 4A).
+The median score was -4, with a mean of -3.5 and standard deviation of 1.4 (Figure 4A).
 Next, we assessed whether these analysis scores associate with journal citation metrics (Figure 4B).
-There was a slight positive correlation using the Pearson method but it was not statistically significant (r=0.04, p=0.16).
+There was no statistically significant association of analysis score and SJR (Pearson r=0.03, p=0.23).
 Next we wanted to know which journals had the highest and lowest scores.
 Only journals with five or more analyses were included.
-The best scoring journals were *Transl Psychiatry*, *Metabolites* and *J Exp Clin Cancer Res* (Figure 4C), while the poorest were *Mol Med Rep*, *Cancer Med* and *World J Gastroenterol* (Figure 4D), although we note that there was a wide variation between articles of the same journal.
-Then, we assessed for an association between mean analysis score and the SJR (Figure 4E).
-Again, there was a slightly positive association that wasn't statistically significant (Pearson r=0.09, p=0.55).
+The best scoring journals were *Transl Psychiatry*, *Metabolites* and *J Exp Clin Cancer Res* (Figure 4C), while the poorest were *RNA Biol*, *Mol Med Rep*, *Cancer Med* and *World J Gastroenterol* (Figure 4D), although we note that there was a wide variation between articles of the same journal.
+Then, we assessed for an association between mean analysis score and the SJR, for journals with 5 or more analyses (Figure 4E).
+Again, there was no association between analysis score and SJR (Pearson r=-0.008, p=0.95).
 Next we assessed whether there was any association between analysis scores and the number of citations received by the articles.
-After log transforming the citation data, there was no association between citations and analysis scores (Pearson r=0.002, p=0.94).
+After log transforming the citation data, there was no association between citations and analysis scores (Pearson r=0.002, p=0.95) (Figure 4F).
+Taken together, these findings suggest that these methodological issues are not limited to lower ranking or poorly cited articles.
 
@@
 
-![Figure 4E. Association between SJR and mean analysis score for journals with >=5 analyses.](images/mean_sjr1.png "Mean score vs SJR")
+![Figure 4E. Association between SJR and mean analysis score for journals with 5 or more analyses.](images/mean_sjr1.png "Mean score vs SJR")
+
+![Figure 4F. Association between SJR and citations.](images/cites1.png "Mean score vs citations")
 
@@
 
-While there have been several articles evaluating the performance of functional enrichment tools, there has been relatively little interest in how existing tools are (mis)used. 
-A recent study looked at the the popularity of frequently used tools and their performance (Xie et al, 2021), but this assumes that the tools were properly used.
+While there have been several articles benchmarking the performance of functional enrichment tools, there has been relatively little interest in how existing tools are (mis)used. 
+A recent study looked at the the popularity of frequently used tools and their performance (Xie et al, 2021), but this assumes that the tools were properly used and sufficiently reported.
 In this sample of open access research articles, there appears to be a bias toward tools that are easy to use.
@@
 
-The lack of methodological details regarding enrichment analysis in some articles was disturbing.
-To illustrate this, many articles mention prominently in the abstract the functional enrichment results but do not detail in their methods sections anything about what type of enrichment analysis was actually conducted.
-This is not limited to a few obscure journals; there were 88 journals that published articles where the tool used for enrichment analysis was not defined.
-
-As we screened these articles, we observed other types of details lacking.
-For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles.
-For ORA, in some instances it was unclear whether up- and down-regulated gene lists were combined before analysis.
-When applying ORA to differential expression, combining up- and down-regulated gene lists is considered poor practice (Hong et al, 2014).
-Our own tests indicate that combining up- and down-regulated gene lists prior to ORA results in a dramatic loss of sensitivity when the proportion of differentially expressed genes is high.
-Using an inappropriate background list resulted in substantially altered results (Figure 1B).
+Failing to properly describe the background gene list was the most common methodological issue (Figure 3J).
+This error can lead to substantially distorted results (Figure 1B).
 In the example here, using the inappropriate background resulted in 2.3 fold more differentially regulated gene sets.
 Thus if there are few or no significant results using a correct background list, there may be a temptation to use an inappropriate background.
-Results here indicate most ORA articles do not report using a background gene list, suggesting that inappropriate ORA use is common.
 
-We scored each of the analyses included in the screen which revealed only 30 articles scored a satisfactory score of zero or higher.
-Surprisingly, the analysis scores we generated did not correlate with a journal level bibliographic metric, nor accrued citations of individual articles.
-This suggests no association between methodological rigour and traditional research impact metrics in this sample.
+The proportion of ORA studies that combined up- and down-regulated gene set was also surprisingly high, at 70% (Figure 3K).
+As exemplified in Figure 1C, combining up- and down-regulated genes can severely reduce the number of significant  gene sets (from 293 pathways to just 22 in this example).
+This is consistent with a previous report indicating that combining up- and down-regulated gene lists is considered poor practice because it reduces sensitivity (Hong et al, 2014).
+Results shown here indicate that a large fraction of published works suffer from a high rate of false negatives, in effect missing a large fraction of truly differentially regulated gene sets.
+This is exacerbated by using an inappropriate background gene list, which leads to a high rate of false positives.
+
+Although analyses involving GSEA scored better overall, they were not free of issues.
+For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Figure 3L).
+
+We scored a total of 1630 analyses, revealing only a small fraction of articles obtained a satisfactory score of zero or higher.
+Surprisingly, the analysis scores we generated did not correlate with journal level bibliographic metric, nor accrued citations of individual articles, showing no association between methodological rigour and traditional research impact metrics in this sample.
+This suggests that methodological and reporting problems are not limited lower ranked journals, but are a more general problem.
 
@@
 3. If making claims about the regulation of a gene set, then results of a statistical test must be shown (including measures of significance and effect size).
+The number of genes belonging to a pathway on its own is insufficient to infer enrichment.
 "Data not shown" is not acceptable when it is possible to attach supplementary files.
@@
 This will reduce the chance of false positives when performing many tests simultaneously.
-Report what method was used for p-value adjustment; Bonferonni, FDR and FWER are examples.
+Report what method was used for p-value adjustment.
+Bonferonni, FDR and FWER are examples.
 
@@
 
+9. Functional enrichment analysis results shouldn't be considered proof of biological plausibility nor validity of omics data analysis (Timmons et al, 2015). 
+These tools are best used generating new hypotheses; informing subsequent biochemical or signaling experimentation.
+
 ### Additional recommendations for enhanced reproducibility
@@
 
-9. Preference FCS and pathway topology tools over ORA tools.
+10. Preference FCS and pathway topology tools over ORA tools.
 ORA tools have a lower sensitivity. 
@@
 
-10. Include the gene profile data (including any background lists) in the supplementary data in TSV or CSV formats.
+11. Include the gene profile data (including any background lists) in the supplementary data in TSV or CSV formats.
 Excel spreadsheets are not recommended for genomics work (Ziemann et al, 2016)
 
-11.
+12.
 Scripted analysis workflows are preferred over analysis involving graphical user interfaces because they can attain a higher degree of computational reproducibility, so consider investing in upgrading your analysis.
@@
 
-12. If using a scripted analysis, provide code by depositing it to a repository like GitHub or Zenodo.
+13. If using a scripted analysis, provide code by depositing it to a repository like GitHub or Zenodo.
 Link the data to the code, so anyone can download and reproduce your analysis (Peng 2011).
@@
 <hr>
-
-## Conclusions
-
-We echo the sentiments of Timmons et al (2015) that functional enrichment analysis results should be interpreted with extreme caution, and that they shouldn't be considered proof of biological plausibility nor validity of omics data analysis.
-Functional enrichment analysis is best used as a hypothesis generating procedure to inform subsequent biochemical or signaling experimentation.
 

### 2021-08-30 QC_analysis2.Rmd
minor edits to manuscript

@@
 res <- res[order(res)]
-sum(res[which(names(res)!="Yes")])/sum(res)*100
+res[which(names(res)!="Yes")]/sum(res)*100
 res

### 2021-08-30 manuscript.Rmd
minor edits to manuscript

@@
 Despite this popularity, there are concerns that these methods are being applied incorrectly and the results of some peer-reviewed publications are unreliable.
-These methodological problems include the use of inappropriate background gene lists, lack of false discovery rate correction and combining up-and downregulated gene sets before enrichment analysis.
+These methodological problems include the use of inappropriate background gene lists, lack of false discovery rate correction and combining up-and down-regulated gene sets before enrichment analysis.
 An example analysis of public RNA-seq reveals that these methodological errors alter enrichment results dramatically.
-To ascertain the frequency of these errors in the literature, we performed a screen of 1363 open access research articles describing functional enrichment results.
+To ascertain the frequency of these errors in the literature, we performed a screen of 186 open access research articles describing functional enrichment results.
 We find that 95% of analyses using over-representation tests did not implement an appropriate reference gene list or did not describe this in the methods.
-Failure to perform p-value correction for multiple tests was identified in 53% of analyses.
+Failure to perform p-value correction for multiple tests was identified in 43% of analyses.
 Many studies lacked detail in the methods section about the tools and gene sets used.
-Only 10% of studies achieved an overall satisfactory standard, which highlights the poor state of functional enrichment rigour and reporting in the contemporary literature.
+Only 15% of studies avoided major flaws, which highlights the poor state of functional enrichment rigour and reporting in the contemporary literature.
 We provide a set of minimum standards that should act as a checklist for researchers and peer-reviewers.
@@
 ORA tools can be stand alone software packages or web services, and they use one or more statistical tests (eg: Fisher’s Exact test, hypergeometric test, binomial test) (Draghici et al, 2003; Hosack et al, 2003).
+
 FCS tools involve giving each detected gene a differential expression score and then evaluating whether the scores are more positive or negative than expected by chance for each set of genes.
 The popular Gene Set Enrichment Analysis (GSEA) tool uses permutation approaches to establish whether a gene set is significantly associated with higher or lower scores, either by permuting sample labels or by permuting genes in the differential expression profile (Subramanian et al, 2005).
+
 From a user's perspective, ORA is easier to conduct because it is as simple as pasting a list of gene names into a text box on a website.
-On the other hand FCS tools are more difficult to use but are more sensitive at detecting subtle associations (Kaspi and Ziemann 2020; Maleki et al, 2020; Xie et al, 2021).
+FCS tools are more difficult to use, but they are more sensitive at detecting subtle associations (Kaspi and Ziemann 2020; Maleki et al, 2020; Xie et al, 2021).
 
 Although these are powerful tools to summarise complex genomics data, there are concerns that they are not being correctly used.
-Timmons et al (2015) highlight instances where a failure to account for sampling bias in ORA leads to biased results and invalidated conclusions in published work.
-Many times as peer reviewers, we have had to request authors include details about what reference or background gene list was used to account for sampling bias.
-In other cases we have requested authors to correct their p-values to account for the possibility of false positives when performing hundreds to thousands of gene set tests in parallel.
-The purpose of this work is to demonstrate the effect of misusing enrichment tools and determine how frequent methodological issues in functional enrichment are, specifically 
+Timmons et al (2015) highlight published articles where failure to account for sampling bias invalidated enrichment analysis and conclusions.
+
+The purpose of this work is to survey the frequency of methodological and reporting flaws in the literature, in particular;
 (i) lack of a background gene set, 
-(ii) lack of adjustment for multiple comparisons and 
+(ii) lack of p-value adjustment for multiple comparisons and 
 (ii) lack of essential methodological details.
-By performing a screen of published articles we will identify the most frequent error modes, and use this to inform us to develop a set of minimum standards for functional enrichment analysis.
+We use the results of this survey to inform a set of minimum standards for functional enrichment analysis.
 
@@
 
-* Whether assumptions was correctly applied
+* Whether the analysis had any major flaws that might invalidate the results.
+This includes, (i) background gene set not stated or inappropriate background set used,
+(ii) lack of FDR correction,
+(iii) no enrichment data shown,
+(iv) inference without performing any statistical test, and
+(v) misinterpreting p-values by stating results were significant when FDR values indicate they aren't.
 
@@
 These articles were cross checked by another team member and any discrepancies were resolved.
+
 For analyses using GSEA, we scanned the articles to identify whether key methodological steps were described, including (i) the gene weighting parameter, (ii) test type, ie: permuting sample labels or genes, and (iii) method used for ranking genes.
+
 For ORA of differential gene expression analyses, we inspected the articles to see evidence that researchers were conducting enrichment of up and down regulated genes separately or combined.
+
 Cleaned data were then loaded into R v4.1 for downstream analysis.
@@
 Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to glucose exposure.
+
 FCS revealed 95 up and 316 downregulated pathways (FDR<0.05).
@@
 There was a strong overlap between FCS and ORA results with a Jaccard statistic of 0.66 (Figure 1A).
+
 We then performed ORA using the whole genome gene list (indicated as ORA\*), which resulted in 147 up and 530 down regulated gene sets (FDR<0.05),
@@
 Interestingly, 26 gene sets were simultaneously up and downregulated with this approach.
+
 Then we performed ORA analysis after combining up and downregulated genes as sometimes observed in the literature (indicated as ORA comb) with the standard background and again with a whole genome background (indicated as ORA\* comb).
@@
 A search of PubMed Central showed 2941 articles published in 2019 with the keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
-From these, 1500 articles were screened for methodological errors; data supplied in Supplementary Table 1.
-We excluded 133 articles from the screen because they did not present any enrichment analysis. 
+From these, 200 articles were screened for methodological errors.
+We excluded 14 articles from the screen because they did not present any enrichment analysis. 
 Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
@@
       [10]: '118 articles excluded'
-      [11]: '1395 articles screened'
+      [11]: '1395 analyses screened'
       [12]: 'data entry to spreadsheet'
@@
 There were 31 different species under study, but human was the most common with 157 analyses (Figure S1C).
+
 We recorded the use of 26 different gene set libraries, with GO and KEGG being the most frequently used (Figure 3A).
@@
 Out of the 225 analyses that performed a statistical test, only 119 (53%) described correction of p-values for multiple testing (Figure 3D).
+
 There were 50 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 15 analyses (6.4%) did not state what tool was used (Figure 3E).
@@
 To assess this in a systematic way we assessed differential gene expression studies to determine how common it was to combine up- and down-regulated gene lists prior to ORA (Figure 3K).
-From 109 analyses, just 28 studies performed stated they performed separate analysis of up- and down-regulated gene sets.
+From 109 analyses, just 28 studies performed separate analysis of up- and down-regulated gene sets.
 
@@
 These are (i) the gene weighting parameter, (ii) test type, ie: permuting sample labels or genes, and (iii) method used for ranking genes.
-These parameters were not stated in more than half of the analyses.
+These parameters were not stated in more than half of the analyses (Figure 3L).
 
@@
 
-![Figure S1a. Representation of journals in the data set.](images/journals2.png "Journals")
+![Figure S1A. Representation of journals in the sample.](images/journals2.png "Journals")
 
-![Figure S1B. Representation of different omics types in the data set.](images/omics2.png "Omics")
+![Figure S1B. Representation of different omics types.](images/omics2.png "Omics")
 
-![Figure S1C. Representation of different organisms in the data set.](images/organisms2.png "Organisms")
+![Figure S1C. Representation of different organisms.](images/organisms2.png "Organisms")
 
-![Figure 3A. Representation of different gene set libraries in the data set.](images/genesetlib2.png "Geneset libraries")
+![Figure 3A. Representation of different gene set libraries.](images/genesetlib2.png "Geneset libraries")
 
-![Figure 3B. Proportion of analyses that reported version information for gene sets used.](images/genesetvers2.png "Geneset version")
+![Figure 3B. Proportion of analyses reporting gene set library version information.](images/genesetvers2.png "Geneset version")
 
-![Figure 3C. Representation of different statistical tests in the data set.](images/stattest2.png "Geneset libraries")
+![Figure 3C. Representation of different statistical tests.](images/stattest2.png "Geneset libraries")
 
-![Figure 3D. Recognition of p-value correction for multiple testing in the data set.](images/fdr2.png "FDR")
+![Figure 3D. Proportion of analysis conducting correction of p-values for multiple comparisons.](images/fdr2.png "FDR")
 
-![Figure 3E. Representation of different software tools for enrichment analysis in the data set.](images/app2.png "Software used")
+![Figure 3E. Representation of different software tools for enrichment analysis.](images/app2.png "Software used")
 
@@
 
-![Figure 3G. Background gene set usage and reporting in the dataset.](images/bg2.png "Background gene list used")
+![Figure 3G. Background gene set usage and reporting.](images/bg2.png "Background gene list used")
 
@@
 
-![Figure 3K. Representation of journals in the data set.](images/comb2.png "Journals")
+![Figure 3K. Whether up- and down-regulated genes were analysed separately or combined. Limited to ORA of differential expression only.](images/comb2.png "Journals")
 
@@
 There was no statistically significant association of analysis score and SJR (Pearson r=0.03, p=0.23).
+
 Next we wanted to know which journals had the highest and lowest scores.
@@
 The best scoring journals were *Transl Psychiatry*, *Metabolites* and *J Exp Clin Cancer Res* (Figure 4C), while the poorest were *RNA Biol*, *Mol Med Rep*, *Cancer Med* and *World J Gastroenterol* (Figure 4D), although we note that there was a wide variation between articles of the same journal.
-Then, we assessed for an association between mean analysis score and the SJR, for journals with 5 or more analyses (Figure 4E).
+
+Then we assessed for an association between mean analysis score and the SJR, for journals with 5 or more analyses (Figure 4E).
 Again, there was no association between analysis score and SJR (Pearson r=-0.008, p=0.95).
@@
 
-While there have been several articles benchmarking the performance of functional enrichment tools, there has been relatively little interest in how existing tools are (mis)used. 
-A recent study looked at the the popularity of frequently used tools and their performance (Xie et al, 2021), but this assumes that the tools were properly used and sufficiently reported.
+While there have been several articles benchmarking the performance of functional enrichment tools, there has been relatively little written on how existing tools are used (or misused). 
+A recent study looked at the the popularity of frequently used tools and benchmarked their performance (Xie et al, 2021), but this assumes that the tools were properly used and methods sufficiently reported.
 In this sample of open access research articles, there appears to be a bias toward tools that are easy to use.
@@
 In the example here, using the inappropriate background resulted in 2.3 fold more differentially regulated gene sets.
-Thus if there are few or no significant results using a correct background list, there may be a temptation to use an inappropriate background.
+Thus if there are few or no significant results using a correct background list, there may be a temptation to use an inappropriate background in order to avoid null results.
 
@@
 This is consistent with a previous report indicating that combining up- and down-regulated gene lists is considered poor practice because it reduces sensitivity (Hong et al, 2014).
-Results shown here indicate that a large fraction of published works suffer from a high rate of false negatives, in effect missing a large fraction of truly differentially regulated gene sets.
+These results suggest that a large fraction of published works suffer from a high rate of false negatives, in effect missing a large fraction of truly differentially regulated gene sets.
 This is exacerbated by using an inappropriate background gene list, which leads to a high rate of false positives.
@@
 
-8. If using ORA for gene expression analysis, refrain from combining up- and down-regulated genes before analysis.
+8. If using ORA for gene expression analysis, be sure to conduct ORA tests separately for up- and down-regulated genes before analysis.
+Combining up- and down-regulated genes into a single list will limit sensitivity.
 
 9. Functional enrichment analysis results shouldn't be considered proof of biological plausibility nor validity of omics data analysis (Timmons et al, 2015). 
-These tools are best used generating new hypotheses; informing subsequent biochemical or signaling experimentation.
+Such tools are best used to generate new hypotheses; informing subsequent biochemical or signaling experimentation.
 

### 2021-08-30 QC_analysis2.Rmd
manuscript updates

@@
 mylm1 <- lm (xm$scores ~ xm$SJR)
-plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
+plot(xm$SJR,xm$scores, bty="n",
+  xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
 abline(mylm1,col="red")
@@
 
-png("images/score_sjr1.png")
+png("images/score_sjr2.png")
 par(mar=c(5,5,3,1))
-plot(xm$SJR,xm$scores,xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
+plot(xm$SJR,xm$scores, bty="n",
+  xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
 abline(mylm1,col="red")
@@
 
-png("images/score_highest1.png")
+png("images/score_highest2.png")
 par(mar=c(5,12,3,1))
@@
 
-png("images/score_lowest1.png")
+png("images/score_lowest2.png")
 par(mar=c(5,12,3,1))

### 2021-08-30 analysis.Rmd
manuscript updates

@@
 mylm1 <- lm (xm$scores ~ xm$SJR)
-plot(xm$SJR,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2), pch=19,
+plot(xm$SJR,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2), pch=19, bty="n",
   xlab="SJR",ylab="score",main="Analysis scores vs. journal metrics")
@@
 par(mar=c(5,5,3,1))
-plot(xm$SJR,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2), pch=19,
+plot(xm$SJR,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2), pch=19,bty="n",
   xlab="SJR",ylab="score",main="Analysis scores vs. journal metrics")
@@
 mylm2 <- lm (jjres$scores ~ jjres$SJR)
-plot(jjres$SJR,jjres$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+plot(jjres$SJR,jjres$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="SJR",ylab="mean score",main="Mean analysis score vs. journal metrics")
@@
 par(mar=c(5,5,3,1))
-plot(jjres$SJR,jjres$scores,  col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+plot(jjres$SJR,jjres$scores,  col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="SJR",ylab="mean score",main="Mean analysis score vs. journal metrics")
@@
 mylm2 <- lm (xm$scores ~ xm$cites)
-plot(xm$cites,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+plot(xm$cites,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="Num cites",ylab="score",main="Association of analysis scores with number of citations")
@@
 mylm2 <- lm (xm$scores ~ log2(xm$cites))
-plot(log2(xm$cites),xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+plot(log2(xm$cites),xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
@@
 par(mar=c(5,5,3,1))
-plot(log2(xm$cites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,
+plot(log2(xm$cites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")

### 2021-08-31 QC_analysis2.Rmd
manuscript editing

@@
 
-png("images/score_sjr2.png")
-par(mar=c(5,5,3,1))
-plot(xm$SJR,xm$scores, bty="n",
-  xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
-abline(mylm1,col="red")
-grid()
-dev.off()
-
 ```
@@
 
-png("images/score_highest2.png")
-par(mar=c(5,12,3,1))
-barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
-        main = "Highest scoring journals",xlim=c(-4,0))
-grid()
-dev.off()
-
 par(mar=c(5,12,3,1))
@@
 grid()
-
-png("images/score_lowest2.png")
-par(mar=c(5,12,3,1))
-barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
-        main = "Lowest scoring journals",xlim=c(-5,0))
-grid()
-dev.off()
 

### 2021-08-31 manuscript.Rmd
manuscript updates
manuscript editing

@@
 FCS was performed using the mitch R package v1.4.0 with default settings, which uses a rank-ANOVA statistical test (Kaspi and Ziemann 2020).
-ORA analysis was performed using the clusterProfiler R package (v4.0.2) enricher function that implements a hypergeometric test (Yu et al, 2012).
+Genes with FDR<0.05 were used for ORA analysis using the clusterProfiler R package (v4.0.2) enricher function that implements a hypergeometric test (Yu et al, 2012).
+No fold-change threshold was used to select genes for ORA.
 For ORA, two types of background gene sets were used (i) all detected genes, or (ii) all genes in the annotation set.
@@
 
-We have collated a list of 2941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
+We collated 2941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
 We sampled 200 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.
@@
 * Whether the analysis had any major flaws that might invalidate the results.
-This includes, (i) background gene set not stated or inappropriate background set used,
+This includes:
+(i) background gene set not stated or inappropriate background set used,
 (ii) lack of FDR correction,
@@
 For assessment of enrichment analysis quality with journal metrics and citations, we required a larger set of studies so we sampled and analysed a further 1300 articles from PMC.
-Results from this sample were not double-checked and so may have a small number of inaccuracies.
+Results from this sample were not double-checked and so may contain a small number of inaccuracies.
 We rated each analysis with a simple approach that deducted points for missing methodological details and awarding points for including extra information (Table 1).
@@
 Using NCBI’s Eutils API, we collected the number of citations each article accrued since publication.
-We used Spearman and Pearson correlation tests to determine any association with the analysis scores we generated.
+We used Pearson correlation tests to assess the association with the analysis scores we generated.
 All data analyses were conducted in R v4.1 and the analysis scripts are available at GitHub (https://github.com/markziemann/SurveyEnrichmentMethods).
@@
 Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
-As some articles included more than one enrichment analysis, the dataset included 235 analyses from 186 articles (Supplementary Table 1).
+As some articles included more than one enrichment analysis, the dataset included 235 analyses from 186 articles; this data is available in Table S1.
 A flow diagram of the survey is provided in Figure 2.
@@
 In order to make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
-We therefore analysed a further 1300 articles, bringing the total number of analyses described up to 1365. 
-This larger sample did not undergo double-checking so may contain a small number of errors.
+We therefore analysed a further 1300 articles, bringing the total number of analyses described up to 1365; this dataset is available in Table S2.
 
@@
 
-![Figure 4F. Association between SJR and citations.](images/cites1.png "Mean score vs citations")
+![Figure 4F. Association between citations and analysis scores.](images/cites1.png "Mean score vs citations")
 
@@
 
+## Supporting information
+
+[**Table S1**](http://ziemann-lab.net/public/kaumadi/TableS1.tsv "A survey of 186 articles describing functional enrichment results.") A survey of 186 articles describing functional enrichment results.
+
+[**Table S2**](http://ziemann-lab.net/public/kaumadi/TableS2.tsv "A survey of 1365 articles describing functional enrichment results.") A survey of 1365 articles describing functional enrichment results.
+
+## Acknowledgements
+
+We thank Dr Nicholas C. Wong (Monash University) and Dr Antony Kaspi (Walter and Eliza Hall Institute) for comments on the manuscript.
+This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+
 ## References

### 2021-08-31 manuscript.Rmd
minor edits

@@
 
-Scimago Journal Ranks were downloaded from the Scimago website (https://www.scimagojr.com/journalrank.php) and used to rank journals by their citation metrics.
+Scimago Journal Rank (SJR) data were downloaded from the Scimago website (https://www.scimagojr.com/journalrank.php) and used to score journals by their citation metrics.
 Using NCBI’s Eutils API, we collected the number of citations each article accrued since publication.
@@
 In order to make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
-We therefore analysed a further 1300 articles, bringing the total number of analyses described up to 1365; this dataset is available in Table S2.
+We therefore analysed a further 1300 articles, bringing the total number of analyses described up to 1630; this dataset is available in Table S2.
 
-We then scored each analysis based on the inclusion or lack of methodological issues and included details.
+We then scored each analysis based on the presence or absence of methodological issues and included details.
 The median score was -4, with a mean of -3.5 and standard deviation of 1.4 (Figure 4A).
@@
 In this sample of open access research articles, there appears to be a bias toward tools that are easy to use.
-ORA tools that only require pasting lists of gene identifiers into a webpage (ie: DAVID, IPA, KOBAS and PANTHER) are collectively more popular than other solutions like GSEA (a standalone graphical user interface software for FCS) or any command line tool.
+ORA tools that only require pasting lists of gene identifiers into a webpage (ie: DAVID, KOBAS and PANTHER) are collectively more popular than other solutions like GSEA (a standalone graphical user interface software for FCS) or any command line tool.
 This is despite the fact that ORA tools perform worse than FCS and new generation pathway topology tools (Kaspi & Ziemann 2020; Maleki et al, 2020; Xie et al, 2021).
@@
 Surprisingly, the analysis scores we generated did not correlate with journal level bibliographic metric, nor accrued citations of individual articles, showing no association between methodological rigour and traditional research impact metrics in this sample.
-This suggests that methodological and reporting problems are not limited lower ranked journals, but are a more general problem.
+This suggests that methodological and reporting problems are not limited to lower ranked journals, but are a more general problem.
 
@@
 Many open access articles examined here are from journals not considered to be highly prestigious, and the distribution of analysis scores might be different for closed access articles.
-The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in more prestigious journals contain such details in the abstract at lower rates.
+The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in higher ranked journals contain such details in the abstract at lower rates.
 

### 2021-08-31 analysis.Rmd
manuscript editing
minor edits
fix journal names and generate pdfs

@@
 
-```
-
-## Filter for studies included in QC2
-
-```{r,xy}
+write.table(x,file="TableS2.tsv",quote=FALSE,sep="\t",col.names = TRUE,row.names = FALSE)
 
@@
     colors=brewer.pal(8, "Dark2"))
+
+names(res) <- gsub("RNA array","Gene expression array",names(res))
 
@@
 
+pdf("images/hist1.pdf",width=4,height=3)
+par(mar=c(5,5,3,1))
+barplot(table(scores), xlab="analysis score",ylab="frequency")
+dev.off()
+
 ```
@@
 
+pdf("images/score_sjr1.pdf",width=4,height=3)
+par(mar=c(4,4,3,2))
+plot(xm$SJR,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2), pch=19,bty="n",
+  xlab="SJR",ylab="score",main="Analysis scores vs. journal metrics")
+abline(mylm1,col="red")
+grid()
+dev.off()
+
 ```
@@
 
+pdf("images/score_highest1.pdf",width=3,height=4.5)
+par(mar=c(5,8,3,1))
+barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Highest scoring",xlim=c(-4,0))
+grid()
+dev.off()
+
+pdf("images/score_lowest1.pdf",width=3,height=4.5)
+par(mar=c(5,8,3,1))
+barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
+        main = "Lowest scoring",xlim=c(-5,0))
+grid()
+dev.off()
+
 j <- unique(xmm$Journal)
@@
 cor.test(jjres$scores,jjres$SJR,method="spearman")
+
+pdf("images/mean_sjr1.pdf",width=3.5,height=3)
+par(mar=c(5,5,3,1))
+plot(jjres$SJR,jjres$scores,  col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
+  xlab="SJR",ylab="mean score",main="Mean analysis score vs. journal metrics")
+abline(mylm2,col="red")
+dev.off()
 
@@
 cor.test(xm$scores , xm$cites, method="spearman")
+
+pdf("images/cites1.pdf",width=3.5,height=3)
+par(mar=c(5,5,3,1))
+plot(log2(xm$cites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
+  xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
+abline(mylm2,col="red")
+dev.off()
+
 ```

### 2021-08-31 manuscript.Rmd
fix journal names and generate pdfs

@@
 
-There were articles from 100 journals in the sample, with *PLoS One*, *Scientific Reports*, and *PeerJ* being the biggest contributors (Figure S1A).
+There were articles from 96 journals in the sample, with *PLoS One*, *Scientific Reports*, and *PeerJ* being the biggest contributors (Figure S1A).
 There were 18 different omics types, with gene expression array and RNA-seq being the most popular (Figure S1B).
@@
 We scored a total of 1630 analyses, revealing only a small fraction of articles obtained a satisfactory score of zero or higher.
-Surprisingly, the analysis scores we generated did not correlate with journal level bibliographic metric, nor accrued citations of individual articles, showing no association between methodological rigour and traditional research impact metrics in this sample.
+The analysis scores we generated did not correlate with journal or article metrics.
 This suggests that methodological and reporting problems are not limited to lower ranked journals, but are a more general problem.
@@
 5. Always report FDR or q-values (p-values that have been corrected for multiple tests).
-This will reduce the chance of false positives when performing many tests simultaneously.
+This will reduce the chance of false positives when performing many tests simultaneously (Hung et al, 2012).
 Report what method was used for p-value adjustment.
@@
 
-We thank Dr Nicholas C. Wong (Monash University) and Dr Antony Kaspi (Walter and Eliza Hall Institute) for comments on the manuscript.
+We thank XX for comments on the manuscript.
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
@@
 Huang, D. W., Sherman, B. T., & Lempicki, R. A. (2009). Bioinformatics enrichment tools: paths toward the comprehensive functional analysis of large gene lists. Nucleic Acids Research, 37(1), 1–13.
+
+Hung, J.-H., Yang, T.-H., Hu, Z., Weng, Z., & DeLisi, C. (2012). Gene set enrichment analysis: performance evaluation and usage guidelines. Briefings in Bioinformatics, 13(3), 281–291.
 

### 2021-09-01 example.Rmd
assembling the figs

@@
 })
+
+dir.create("images")
 
@@
 
+pdf("images/fcs_ora1.pdf",width=4,height=4)
+plot(euler(v1),quantities = TRUE, main="FCS vs ORA")
+dev.off()
+
+pdf("images/orabg1.pdf",width=4,height=4)
+plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")
+dev.off()
+
+pdf("images/oracomb1.pdf",width=4,height=4)
+plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
+dev.off()
+
 ```

### 2021-09-01 manuscript_draft.Rmd
assembling the figs

--- 2021-08-31 manuscript.Rmd
@@
 
+library(tidyverse)
 library(DiagrammeR)
+library(DiagrammeRsvg)
+library(rsvg)
+
 grViz("digraph flowchart {
@@
 width = 700)
+
+tmp <- grViz("digraph flowchart {
+      # node definitions with substituted label text
+      node [fontname = Helvetica, shape = rectangle]        
+      tab1 [label = '@@1']
+      tab2 [label = '@@2']
+      tab3 [label = '@@3']
+      tab4 [label = '@@4']
+      tab5 [label = '@@5']
+      tab6 [label = '@@6']
+      tab7 [label = '@@7']
+      tab8 [label = '@@8']
+      tab9 [label = '@@9']
+      tab10 [label = '@@10']
+      tab11 [label = '@@11']
+      tab12 [label = '@@12']
+      tab13 [label = '@@13']
+
+      # edge definitions with the node IDs
+      tab1 -> tab2  -> tab4 -> tab5 -> tab6 -> tab7;
+      tab2 -> tab3 ;
+      tab1 -> tab9 -> tab10;
+      tab5 -> tab13   ;
+      tab9 -> tab11 -> tab12 -> tab13 -> tab8 ; 
+      }
+      [1]: 'PMC keyword search (2019): 2941 results'
+      [2]: '200 PMC articles selected'
+      [3]: '14 articles excluded'
+      [4]: '235 analyses screened and cross-checked'
+      [5]: 'data entry to spreadsheet'
+      [6]: 'data analysis in R'
+      [7]: 'summary charts'
+      [8]: 'integration with journal and article metrics'
+      [9]: 'additional 1300 PMC articles selected'
+      [10]: '118 articles excluded'
+      [11]: '1395 analyses screened'
+      [12]: 'data entry to spreadsheet'
+      [13]: 'data analysis in R'
+")
+
+tmp2 <- capture.output(rsvg_pdf(charToRaw(export_svg(tmp)),'images/flowchart.pdf'))
 

### 2021-09-01 manuscript.Rmd
readme update

@@
 Although these are powerful tools to summarise complex genomics data, there are concerns that they are not being correctly used.
+Tipney and Hunter (2010) warn that inappropriate background set selection heavily influences enrichment results.
 Timmons et al (2015) highlight published articles where failure to account for sampling bias invalidated enrichment analysis and conclusions.
@@
 
-![Figure 1A. Comparison of FCS and ORA methods.](images/fcs_ora1.png "FCS vs ORA")
-
-![Figure 1B. ORA with standard and whole genome background.](images/orabg1.png "ORA with different background sets")
-
-![Figure 1C. Effect of combining up and downregulated genes on ORA results.](images/oracomb1.png "ORA combining up and downregulated genes")
+![Figure 1.
+Example of enrichment analysis misuse.
+(A) Comparison of FCS and ORA methods.
+(B) ORA with standard and whole genome background.
+(C) Effect of combining up and downregulated genes on ORA results.
+](images/Figure1.png "Figure 1")
 
@@
 
-```{r,diagram1,echo=FALSE,fig.width=8,fig.height=4}
-
-library(DiagrammeR)
-grViz("digraph flowchart {
-      # node definitions with substituted label text
-      node [fontname = Helvetica, shape = rectangle]        
-      tab1 [label = '@@1']
-      tab2 [label = '@@2']
-      tab3 [label = '@@3']
-      tab4 [label = '@@4']
-      tab5 [label = '@@5']
-      tab6 [label = '@@6']
-      tab7 [label = '@@7']
-      tab8 [label = '@@8']
-      tab9 [label = '@@9']
-      tab10 [label = '@@10']
-      tab11 [label = '@@11']
-      tab12 [label = '@@12']
-      tab13 [label = '@@13']
-
-      # edge definitions with the node IDs
-      tab1 -> tab2  -> tab4 -> tab5 -> tab6 -> tab7;
-      tab2 -> tab3 ;
-      tab1 -> tab9 -> tab10;
-      tab5 -> tab13   ;
-      tab9 -> tab11 -> tab12 -> tab13 -> tab8 ; 
-      }
-      [1]: 'PMC keyword search (2019): 2941 results'
-      [2]: '200 PMC articles selected'
-      [3]: '14 articles excluded'
-      [4]: '235 analyses screened and cross-checked'
-      [5]: 'data entry to spreadsheet'
-      [6]: 'data analysis in R'
-      [7]: 'summary charts'
-      [8]: 'integration with journal and article metrics'
-      [9]: 'additional 1300 PMC articles selected'
-      [10]: '118 articles excluded'
-      [11]: '1395 analyses screened'
-      [12]: 'data entry to spreadsheet'
-      [13]: 'data analysis in R'
-",
-width = 700)
-
-```
-*Figure 2. A summary of the survey of functional enrichment analyses.*
+![Figure 2. A summary of the survey of functional enrichment analyses. ](images/Figure2.png "Figure 2")
 
@@
 
-![Figure S1A. Representation of journals in the sample.](images/journals2.png "Journals")
+![Figure S1. General make up of the sample. (A) Journals. (B) Types of omics analysis applied. (C) Organisms under study.](images/FigureS1.png "Figure S1")
 
-![Figure S1B. Representation of different omics types.](images/omics2.png "Omics")
-
-![Figure S1C. Representation of different organisms.](images/organisms2.png "Organisms")
-
-![Figure 3A. Representation of different gene set libraries.](images/genesetlib2.png "Geneset libraries")
-
-![Figure 3B. Proportion of analyses reporting gene set library version information.](images/genesetvers2.png "Geneset version")
-
-![Figure 3C. Representation of different statistical tests.](images/stattest2.png "Geneset libraries")
-
-![Figure 3D. Proportion of analysis conducting correction of p-values for multiple comparisons.](images/fdr2.png "FDR")
-
-![Figure 3E. Representation of different software tools for enrichment analysis.](images/app2.png "Software used")
-
-![Figure 3F. Proportion of analyses that reported version information for software used.](images/appvers2.png "Software version")
-
-![Figure 3G. Background gene set usage and reporting.](images/bg2.png "Background gene list used")
-
-![Figure 3H. Proportion of scripted analyses that provided software code.](images/code2.png "Code available")
-
-![Figure 3I. Proportion of scripted analyses that provided gene profiles.](images/genelists2.png "Gene lists provided")
-
-![Figure 3J. Proportion of analyses with different methodological problems.](images/assumptions2.png "Assumptions violated")
-
-![Figure 3K. Whether up- and down-regulated genes were analysed separately or combined. Limited to ORA of differential expression only.](images/comb2.png "Journals")
-
-![Figure 3L. Reporting of GSEA parameters.](images/gseaparam2.png "Journals")
+![Figure 3.
+Findings from the survey of published enrichment analyses.
+(A) Representation of different gene set libraries.
+(B) Proportion of analyses reporting gene set library version information.
+(C) Representation of different statistical tests.
+(D) Proportion of analysis conducting correction of p-values for multiple comparisons.
+(E) Representation of different software tools for enrichment analysis.
+(F) Proportion of analyses that reported version information for software used.
+(G) Background gene set usage and reporting.
+(H) Proportion of scripted analyses that provided software code.
+(I) Proportion of analyses that provided gene profiles.
+(J)  Proportion of analyses with different methodological problems.
+(K) Whether up- and down-regulated genes were analysed separately or combined.
+Limited to ORA of differential expression only.
+(L) Reporting of GSEA parameters.](images/Figure3.png "Figure 3")
 
@@
 
-![Figure 4A. Distribution of analysis scores.](images/hist1.png "Analysis scores")
-
-![Figure 4B. Association of analysis scores with journal rank.](images/score_sjr1.png "Analysis scores versus journal rank")
-
-![Figure 4C. Journals with highest analysis scores.](images/score_highest1.png "Highest analysis scores")
-
-![Figure 4D. Journals with Lowest analysis scores.](images/score_lowest1.png "Lowest analysis scores")
-
-![Figure 4E. Association between SJR and mean analysis score for journals with 5 or more analyses.](images/mean_sjr1.png "Mean score vs SJR")
-
-![Figure 4F. Association between citations and analysis scores.](images/cites1.png "Mean score vs citations")
+![Figure 4.
+Comparison of Analysis scores to bibliometrics.
+(A) Distribution of analysis scores.
+(B) Association of analysis scores with Scimago Journal Rank (SJR).
+(C) Journals with highest analysis scores.
+(D) Journals with lowest analysis scores.
+(E) Association of mean analysis score with SJR for journals with 5 or more analyses.
+(F) Association of analysis scores with accrued citations.](images/Figure4.png "Figure 4")
 
@@
 
+Tipney, H., & Hunter, L. (2010). An introduction to effective use of enrichment analysis software. Human Genomics, 4(3), 202–206.
+
 Xie, C., Jauhari, S., & Mora, A. (2021). Popularity and performance of bioinformatics software: the case of gene set analysis. BMC Bioinformatics, 22(1), 191.

### 2021-09-01 manuscript.Rmd
readme update

@@
 
-### An example of gene set analysis misuse
+### An example of functional enrichment analysis misuse
 
-In order to demonstrate the effect of gene set test misuse, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes.
+In order to demonstrate the effect of functional enrichment analysis misuse, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes.
 Out of 39,297 genes in the annotation set, 15,635 were above the detection threshold (>10 reads per sample on average).
@@
 
+![Figure S1. General make up of the sample. (A) Journals. (B) Types of omics analysis applied. (C) Organisms under study.](images/FigureS1.png "Figure S1")
+
 We recorded the use of 26 different gene set libraries, with GO and KEGG being the most frequently used (Figure 3A).
@@
 Taken together, these data suggest methodological and reporting deficiencies are widespread in published functional enrichment analyses.
-
-![Figure S1. General make up of the sample. (A) Journals. (B) Types of omics analysis applied. (C) Organisms under study.](images/FigureS1.png "Figure S1")
 

### 2021-09-02 manuscript.Rmd
editing

@@
 ---
-
-Source: https://github.com/markziemann/SurveyEnrichmentMethods
-
-Target Journal: Briefings in Bioinformatics (JIF=9.0)
 
@@
 (ii) lack of p-value adjustment for multiple comparisons and 
-(ii) lack of essential methodological details.
+(iii) lack of essential methodological details.
 We use the results of this survey to inform a set of minimum standards for functional enrichment analysis.
@@
 
-We collated 2941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
+We collated 2,941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
 We sampled 200 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.
@@
 
+*Table 1. Scoring schema.*
 | 1 point deducted | 1 point awarded |
@@
 | Gene set library origin not stated | Code made available |
-| Gene set library version not stated | Gene lists provided |
-| Stat test not stated | |
-| No stat test conducted | |
+| Gene set library version not stated | Gene profile data provided |
+| Statistical test not stated | |
+| No statistical test conducted | |
 | No FDR correction conducted | |
@@
 
-Scimago Journal Rank (SJR) data were downloaded from the Scimago website (https://www.scimagojr.com/journalrank.php) and used to score journals by their citation metrics.
+Scimago Journal Rank (SJR) data for 2020 were downloaded from the Scimago website (https://www.scimagojr.com/journalrank.php) and used to score journals by their citation metrics.
 Using NCBI’s Eutils API, we collected the number of citations each article accrued since publication.
@@
 Out of 39,297 genes in the annotation set, 15,635 were above the detection threshold (>10 reads per sample on average).
-Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to glucose exposure.
+Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to elevated glucose exposure.
 
@@
 Example of enrichment analysis misuse.
+Euler diagrams show the overlap between sets of differentially regulated gene sets with different methods (FDR<0.05).
 (A) Comparison of FCS and ORA methods.
 (B) ORA with standard and whole genome background.
-(C) Effect of combining up and downregulated genes on ORA results.
+Whole genome background analysis is indicated with a \* (C) Effect of combining up and downregulated genes on ORA results.
+Analysis using combined gene list is indicated by *ORA comb*, while *ORA\* comb* indicates combined gene list analysis with a whole genome background.
 ](images/Figure1.png "Figure 1")
@@
 
-A search of PubMed Central showed 2941 articles published in 2019 with the keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
-From these, 200 articles were screened for methodological errors.
+A search of PubMed Central showed 2,941 articles published in 2019 with the keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
+From these, 200 articles were selected for methodological analysis.
 We excluded 14 articles from the screen because they did not present any enrichment analysis. 
@@
 For analyses using ORA methods, we studied what background gene set was used (Figure 3G). 
-This revealed that in most cases, the background list was not stated or it was clear from the article methods section that no background list was used.
-In a few cases, background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
+This revealed that in most cases, the background list was not defined or it was clear from the article methods section that no background list was used.
+In a few cases, a background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
 In only 8/197 of cases (4.1%), the appropriate background list was described in the article.
@@
 In order to make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
-We therefore analysed a further 1300 articles, bringing the total number of analyses described up to 1630; this dataset is available in Table S2.
+We therefore analysed a further 1,300 articles, bringing the total number of analyses described up to 1630; this dataset is available in Table S2.
 
@@
 
-The proportion of ORA studies that combined up- and down-regulated gene set was also surprisingly high, at 70% (Figure 3K).
+The proportion of ORA studies that combined up- and down-regulated gene list prior to analysis was also surprisingly high, at 70% (Figure 3K).
 As exemplified in Figure 1C, combining up- and down-regulated genes can severely reduce the number of significant  gene sets (from 293 pathways to just 22 in this example).
@@
 Although analyses involving GSEA scored better overall, they were not free of issues.
-For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Figure 3L).
+For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Figure 3L), limiting reproducibility.
 
@@
 Still there are some limitations that need to be recognised.
-Many open access articles examined here are from journals not considered to be highly prestigious, and the distribution of analysis scores might be different for closed access articles.
+Many open access articles examined here are from lower-ranked journals, and might not be representative of articles in paywalled journals.
 The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in higher ranked journals contain such details in the abstract at lower rates.
@@
 
-3. If making claims about the regulation of a gene set, then results of a statistical test must be shown (including measures of significance and effect size).
+3. If making claims about the regulation of a gene set, then results of a statistical test must be shown including measures of significance and effect size.
 The number of genes belonging to a pathway on its own is insufficient to infer enrichment.
-"Data not shown" is not acceptable when it is possible to attach supplementary files.
 
@@
 If using GSEA or another FCS tool, parameters around ranking, gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels).
-Any parameters that are vary from the defaults should be reported.
+Any parameters that vary from the defaults should be reported.
 
@@
 
-### Additional recommendations for enhanced reproducibility
+### Additional recommendations for enhanced reliability and reproducibility
 

### 2021-09-03 manuscript.Rmd
text edits

@@
 A statistical test is performed to ascertain whether the number of DEGs belonging to a particular pathway (gene set) is higher than that expected by random chance.
-ORA tools can be stand alone software packages or web services, and they use one or more statistical tests (eg: Fisher’s Exact test, hypergeometric test, binomial test) (Draghici et al, 2003; Hosack et al, 2003).
+ORA tools can be stand-alone software packages or web services, and they use one or more statistical tests (eg: Fisher’s Exact test, hypergeometric test, binomial test) (Draghici et al, 2003; Hosack et al, 2003).
 
@@
 
-For assessment of enrichment analysis quality with journal metrics and citations, we required a larger set of studies so we sampled and analysed a further 1300 articles from PMC.
+For assessment of enrichment analysis quality with journal metrics and citations, we required a larger sample, so we selected a further 1300 articles from PMC for analysis.
 Results from this sample were not double-checked and so may contain a small number of inaccuracies.
@@
 *Table 1. Scoring schema.*
+
 | 1 point deducted | 1 point awarded |
@@
 
-![Figure S1. General make up of the sample. (A) Journals. (B) Types of omics analysis applied. (C) Organisms under study.](images/FigureS1.png "Figure S1")
+![Figure S1. General make-up of the sample. (A) Journals. (B) Types of omics analysis applied. (C) Organisms under study.](images/FigureS1.png "Figure S1")
 
@@
 For analyses using ORA methods, we studied what background gene set was used (Figure 3G). 
-This revealed that in most cases, the background list was not defined or it was clear from the article methods section that no background list was used.
+This revealed that in most cases, the background list was not defined, or it was clear from the article methods section that no background list was used.
 In a few cases, a background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
@@
 
-During the course of this survey, we also noticed some studies that grouped up- and down-regulated gene lists together prior to ORA, a practice we were not expecting.
+During this survey, we noticed some studies grouped up- and down-regulated gene lists together prior to ORA, a practice we were not expecting.
 To assess this in a systematic way we assessed differential gene expression studies to determine how common it was to combine up- and down-regulated gene lists prior to ORA (Figure 3K).
@@
 
-In order to make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
+To make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
 We therefore analysed a further 1,300 articles, bringing the total number of analyses described up to 1630; this dataset is available in Table S2.
@@
 
-Next we wanted to know which journals had the highest and lowest scores.
+Next, we wanted to know which journals had the highest and lowest scores.
 Only journals with five or more analyses were included.
@@
 Again, there was no association between analysis score and SJR (Pearson r=-0.008, p=0.95).
-Next we assessed whether there was any association between analysis scores and the number of citations received by the articles.
+Next, we assessed whether there was any association between analysis scores and the number of citations received by the articles.
 After log transforming the citation data, there was no association between citations and analysis scores (Pearson r=0.002, p=0.95) (Figure 4F).
@@
 While there have been several articles benchmarking the performance of functional enrichment tools, there has been relatively little written on how existing tools are used (or misused).
-A recent study looked at the the popularity of frequently used tools and benchmarked their performance (Xie et al, 2021), but this assumes that the tools were properly used and methods sufficiently reported.
+A recent study looked at the popularity of frequently used tools and benchmarked their performance (Xie et al, 2021), but this assumes that the tools were properly used, and methods sufficiently reported.
 In this sample of open access research articles, there appears to be a bias toward tools that are easy to use.
-ORA tools that only require pasting lists of gene identifiers into a webpage (ie: DAVID, KOBAS and PANTHER) are collectively more popular than other solutions like GSEA (a standalone graphical user interface software for FCS) or any command line tool.
+ORA tools that only require pasting lists of gene identifiers into a webpage (ie: DAVID, KOBAS and PANTHER) are collectively more popular than other solutions like GSEA (a stand-alone graphical user interface software for FCS) or any command line tool.
 This is despite the fact that ORA tools perform worse than FCS and new generation pathway topology tools (Kaspi & Ziemann 2020; Maleki et al, 2020; Xie et al, 2021).
@@
 This error can lead to substantially distorted results (Figure 1B).
-In the example here, using the inappropriate background resulted in 2.3 fold more differentially regulated gene sets.
-Thus if there are few or no significant results using a correct background list, there may be a temptation to use an inappropriate background in order to avoid null results.
+In the example here, using the inappropriate background resulted in 2.3-fold more differentially regulated gene sets.
+Thus, if there are few or no significant results using a correct background list, there may be a temptation to use an inappropriate background to avoid null results.
 
@@
 Although analyses involving GSEA scored better overall, they were not free of issues.
-For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Figure 3L), limiting reproducibility.
+For example, GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Figure 3L), limiting reproducibility.
 
@@
 Still there are some limitations that need to be recognised.
-Many open access articles examined here are from lower-ranked journals, and might not be representative of articles in paywalled journals.
+Many open access articles examined here are from lower-ranked journals, that might not be representative of articles in paywalled journals.
 The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in higher ranked journals contain such details in the abstract at lower rates.
 
-Nevertheless, these results are a wake up call for reproducibility and highlight the urgent need for minimum standards for functional enrichment analysis. 
+Nevertheless, these results are a wake-up call for reproducibility and highlight the urgent need for minimum standards for functional enrichment analysis. 
 To address this, we provide a set of minimum standards for functional enrichment analysis, as well as additional recommendations to achieve gold standard reliability and reproducibility (Box 1).
@@
 Scripted analysis workflows are preferred over analysis involving graphical user interfaces because they can attain a higher degree of computational reproducibility, so consider investing in upgrading your analysis.
-Similarly, free and open source software tools are preferred over proprietary software.
+Similarly, free and open-source software tools are preferred over proprietary software.
 

### 2021-09-03 example.Rmd
Figure 1 edits

--- 2021-08-13 example.Rmd
@@
 
-plot(euler(v0),quantities = TRUE,main="effect of combining up and down regulated genes")
+plot(euler(v0),quantities = TRUE, edges = "gray", main="effect of combining up and down regulated genes")
 
@@
 
-plot(euler(v1),quantities = TRUE, main="FCS compared to ORA")
+plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS compared to ORA")
 
@@
 
-plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")
+plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
 
@@
 
-plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
+plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
 
@@
 
-plot(euler(v3),quantities = TRUE, main="all sets")
+plot(euler(v3),quantities = TRUE, edges = "gray", main="all sets")
 
 png("images/fcs_ora1.png")
-plot(euler(v1),quantities = TRUE, main="FCS vs ORA")
+plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
 dev.off()
@@
 png("images/orabg1.png")
-plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")
+plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
 dev.off()
@@
 plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
+dev.off()
+
+pdf("images/fcs_ora1.pdf",width=4,height=4)
+plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
+dev.off()
+
+pdf("images/orabg1.pdf",width=4,height=4)
+plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
+dev.off()
+
+pdf("images/oracomb1.pdf",width=4,height=4)
+plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
 dev.off()

### 2021-09-04 manuscript.Rmd
Figure 1 edits
edits thanks to feedback

@@
 To ascertain the frequency of these errors in the literature, we performed a screen of 186 open access research articles describing functional enrichment results.
-We find that 95% of analyses using over-representation tests did not implement an appropriate reference gene list or did not describe this in the methods.
+We find that 95% of analyses using over-representation tests did not implement an appropriate background gene list or did not describe this in the methods.
 Failure to perform p-value correction for multiple tests was identified in 43% of analyses.
@@
 In ORA, differentially expressed genes (DEGs) meeting a significance and/or fold change threshold are queried against curated pathways (gene sets).
-A statistical test is performed to ascertain whether the number of DEGs belonging to a particular pathway (gene set) is higher than that expected by random chance.
-ORA tools can be stand-alone software packages or web services, and they use one or more statistical tests (eg: Fisher’s Exact test, hypergeometric test, binomial test) (Draghici et al, 2003; Hosack et al, 2003).
+A statistical test is performed to ascertain whether the number of DEGs belonging to a particular pathway (gene set) is higher than that expected by random chance, as determined by comparison to a background gene list.
+These ORA tools can be stand-alone software packages or web services, and they use one or more statistical tests (eg: Fisher’s Exact test, hypergeometric test) (Draghici et al, 2003; Hosack et al, 2003).
 
-FCS tools involve giving each detected gene a differential expression score and then evaluating whether the scores are more positive or negative than expected by chance for each set of genes.
+In the case of ORA for differential expression (eg: RNA-seq), a whole genome background is inappropriate because in any tissue, most genes are not expressed and therefore have no chance of being classified as DEGs.
+A good rule of thumb is to use a background gene list consisting of genes detected in the assay at a level where they have a chance of being classified as DEG (Tipney and Hunter 2010).
+
+FCS tools involve giving each detected gene a differential expression score and then evaluating whether the scores are more positive or negative than expected by chance for each gene set.
 The popular Gene Set Enrichment Analysis (GSEA) tool uses permutation approaches to establish whether a gene set is significantly associated with higher or lower scores, either by permuting sample labels or by permuting genes in the differential expression profile (Subramanian et al, 2005).
@@
 Tipney and Hunter (2010) warn that inappropriate background set selection heavily influences enrichment results.
-Timmons et al (2015) highlight published articles where failure to account for sampling bias invalidated enrichment analysis and conclusions.
+Timmons et al (2015) highlight two cases where an inappropriate background list led to invalid enrichment results in published articles.
 
 The purpose of this work is to survey the frequency of methodological and reporting flaws in the literature, in particular;
-(i) lack of a background gene set, 
+(i) inappropriate background gene set, 
 (ii) lack of p-value adjustment for multiple comparisons and 
@@
 Transcript level counts were aggregated to genes using the getDEE2 R package v1.2.0. 
-Next, genes with an average of 10 reads per sample were discarded.
+Next, genes with an average of less than 10 reads per sample were omitted from downstream analysis.
 Differential expression statistical analysis was conducted with DESeq2 v1.32.0 (Love et al, 2014) to identify genes altered by high glucose exposure.
@@
 No fold-change threshold was used to select genes for ORA.
-For ORA, two types of background gene sets were used (i) all detected genes, or (ii) all genes in the annotation set.
+For ORA, two types of background gene sets were used (i) detected genes, or (ii) all genes in the genome.
 For genes and gene sets, a false discovery rate adjusted p-value (FDR) of 0.05 was considered significant.
@@
 
-* Whether a background gene set was used
+* Whether an appropriate background gene set was used
 
@@
 In order to demonstrate the effect of functional enrichment analysis misuse, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes.
-Out of 39,297 genes in the annotation set, 15,635 were above the detection threshold (>10 reads per sample on average).
+Out of 39,297 genes in the whole genome, 15,635 were above the detection threshold (≥10 reads per sample on average).
 Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to elevated glucose exposure.
@@
 ORA with a background gene list consisting of detected genes revealed 55 upregulated and 238 downregulated gene sets (FDR<0.05).
-There was a strong overlap between FCS and ORA results with a Jaccard statistic of 0.66 (Figure 1A).
+The overlap between FCS and ORA shown in Figure 1A is reflected by a Jaccard statistic of 0.66, indicating moderate concordance between these methods.
 
 We then performed ORA using the whole genome gene list (indicated as ORA\*), which resulted in 147 up and 530 down regulated gene sets (FDR<0.05),
-The overlap of ORA with ORA\* was relatively smaller, with a Jaccard statistic of 0.41 (Figure 1B).
+The overlap of ORA with ORA\* was relatively small, with a Jaccard statistic of 0.41 (Figure 1B).
 Interestingly, 26 gene sets were simultaneously up and downregulated with this approach.
@@
 ORA\* comb yielded 64 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.08.
-These results demonstrate that commonly observed practices of combining the up and downregulated gene lists and using the whole genome background gene list severely distort ORA results.
+These results suggest that when used properly, FCS and ORA yield similar results, but when used improperly, ORA results can be severely distorted.
 
@@
 
-For analyses using ORA methods, we studied what background gene set was used (Figure 3G). 
+For analyses using ORA methods, we examined what background gene set was used (Figure 3G). 
 This revealed that in most cases, the background list was not defined, or it was clear from the article methods section that no background list was used.
@@
 This error can lead to substantially distorted results (Figure 1B).
-In the example here, using the inappropriate background resulted in 2.3-fold more differentially regulated gene sets.
+In the example here, using the inappropriate whole genome background resulted in 2.3-fold more differentially regulated gene sets.
 Thus, if there are few or no significant results using a correct background list, there may be a temptation to use an inappropriate background to avoid null results.
@@
 
-1. Report the origin of the gene sets and the version used or date downloaded.
+1. Before starting the analysis, read the tool's user manual.
+
+2. Report the origin of the gene sets and the version used or date downloaded.
 These databases are regularly upgraded so this is important for reproducibility.
 
-2. Report the tool used and its software version.
+3. Report the tool used and its software version.
 As these tools are regularly updated, this will aid reproducibility.
+For online tools, record the date the tool was used.
 
-3. If making claims about the regulation of a gene set, then results of a statistical test must be shown including measures of significance and effect size.
+4. If making claims about the regulation of a gene set, then results of a statistical test must be shown including measures of significance and effect size.
 The number of genes belonging to a pathway on its own is insufficient to infer enrichment.
 
-4. Report which statistical test was used.
+5. Report which statistical test was used.
 This will help long term reproducibility, for example if in future the tool is no longer available.
@@
 
-5. Always report FDR or q-values (p-values that have been corrected for multiple tests).
+6. Always report FDR or q-values (p-values that have been corrected for multiple tests).
 This will reduce the chance of false positives when performing many tests simultaneously (Hung et al, 2012).
@@
 
-6. If ORA is used, it is vital to use a background list consisting of all detected genes, and report this in the methods section.
+7. If ORA is used, it is vital to use a background list consisting of all detected genes, and report this in the methods section.
 Avoid tools that don't accept a background list.
 
-7. Report selection criteria and parameters.
+8. Report selection criteria and parameters.
 If performing ORA, then the inclusion thresholds need to be mentioned.
@@
 
-8. If using ORA for gene expression analysis, be sure to conduct ORA tests separately for up- and down-regulated genes before analysis.
+9. If using ORA for gene expression analysis, be sure to conduct ORA tests separately for up- and down-regulated genes before analysis.
 Combining up- and down-regulated genes into a single list will limit sensitivity.
 
-9. Functional enrichment analysis results shouldn't be considered proof of biological plausibility nor validity of omics data analysis (Timmons et al, 2015). 
+10. Functional enrichment analysis results shouldn't be considered proof of biological plausibility nor validity of omics data analysis (Timmons et al, 2015). 
 Such tools are best used to generate new hypotheses; informing subsequent biochemical or signaling experimentation.
@@
 
-10. Preference FCS and pathway topology tools over ORA tools.
+11. Preference FCS and pathway topology tools over ORA tools.
 ORA tools have a lower sensitivity. 
@@
 
-11. Include the gene profile data (including any background lists) in the supplementary data in TSV or CSV formats.
+12. Include the gene profile data (including any background lists) in the supplementary data in TSV or CSV formats.
 Excel spreadsheets are not recommended for genomics work (Ziemann et al, 2016)
 
-12.
+13.
 Scripted analysis workflows are preferred over analysis involving graphical user interfaces because they can attain a higher degree of computational reproducibility, so consider investing in upgrading your analysis.
@@
 
-13. If using a scripted analysis, provide code by depositing it to a repository like GitHub or Zenodo.
+14. If using a scripted analysis, provide code by depositing it to a repository like GitHub or Zenodo.
 Link the data to the code, so anyone can download and reproduce your analysis (Peng 2011).
@@
 
-We thank XX for comments on the manuscript.
+We thank Antony Kaspi (Walter and Elisa Hall Institute), Nick Wong and Anup Shah (Monash University) for comments on the manuscript.
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).

### 2021-09-17 analysis.Rmd
"histogram fix"

@@
 
+table(scores)
+
 png("images/hist1.png",width=300,height=200)
-par(mar=c(5,5,3,1))
-barplot(table(scores), xlab="analysis score",ylab="frequency")
+par(mar=c(5,5,2,1))
+barplot(table(scores), xlab="analysis score",ylab="frequency",ylim=c(0,600))
+text(x= 1:length(table(scores))* 1.2 -0.5, y = table(scores), label = table(scores), pos = 3, cex = 1, col = "black")
 dev.off()
 
-pdf("images/hist1.pdf",width=4,height=3)
-par(mar=c(5,5,3,1))
-barplot(table(scores), xlab="analysis score",ylab="frequency")
+pdf("images/hist1.pdf",width=5,height=4)
+par(mar=c(5,5,2,1))
+barplot(table(scores), xlab="analysis score",ylab="frequency",ylim=c(0,600))
+text(x= 1:length(table(scores))* 1.2 -0.5, y = table(scores), label = table(scores), pos = 3, cex = 1, col = "black")
+dev.off()
+
+scores2 <- round(table(scores)/sum(table(scores))*100,1)
+scores2
+
+png("images/hist2.png",width=300,height=200)
+par(mar=c(5,5,2,1))
+barplot(scores2, xlab="analysis score",ylab="frequency",ylim=c(0,50))
+text(x= 1:length(scores2)* 1.2 -0.5, y = scores2, label = scores2, pos = 3, cex = 0.9, col = "black")
+dev.off()
+
+pdf("images/hist2.pdf",width=5,height=4)
+par(mar=c(5,5,2,1))
+barplot(scores2, xlab="analysis score",ylab="frequency",ylim=c(0,40))
+text(x= 1:length(scores2)* 1.2 -0.5, y = scores2, label = scores2, pos = 3, cex = 0.9, col = "black")
 dev.off()

### 2021-11-23 example1.Rmd
updated ORA/FCS examples

--- 2021-09-03 example.Rmd
@@
     toc_float: true
+    code_folding: hide
     fig_width: 7
@@
 Here we are performing an analysis of some gene expression data to demonstrate the difference between ORA and FCS methods and to highlight the differences caused by improper background gene set use.
+
+The dataset being used is SRP128998 and we are comparing the cells grown in normal glucose condition (control) to the high glucose condition (case).
+
+Data are obtained from http://dee2.io/
 
@@
 
-## Get expression data
+## Get expression data and make an MDS plot
 
@@
 
-```{r,getdata,fig.width=7,fig.height=7}
+```{r,getdata}
 
+name="SRP128998"
 mdat<-getDEE2Metadata("hsapiens")
@@
 samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$HG<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
+samplesheet$trt<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
 samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
@@
 gt$accession<-rownames(gt)
-head(gt)
 
@@
 x1<-x[,which(colnames(x) %in% s1$SRR_accession)]
+
+```
+
+Here show the number of genes in the annotation set, and those detected above the detection threshold.
+
+```{r,filter}
 
@@
 
-# MDS plot
-plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$HG, main="MDS")
+```
 
-#run DESeq2 for LG vs HG
-y <- DESeqDataSetFromMatrix(countData = round(x1), colData = s1, design = ~ HG)
-y <- DESeq(y)
-LGvHG <- results(y)
-LGvHG<-as.data.frame(LGvHG[order(LGvHG$pvalue),])
-rownames(LGvHG)<-sapply(strsplit(rownames(LGvHG),"\\."),"[[",1)
-head(LGvHG)
+Now multidimensional scaling (MDS) plot to show the correlation between the datasets.
+If the control and case datasets are clustered separately, then it is likely that there will be many differentially expressed genes with FDR<0.05.
+
+```{r,mds}
+
+plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$trt, main="MDS")
 
 ```
+
+## Differential expression
+
+Now run DESeq2 for control vs case.
+
+```{r,deseq2}
+
+y <- DESeqDataSetFromMatrix(countData = round(x1), colData = s1, design = ~ trt)
+y <- DESeq(y)
+de <- results(y)
+de<-as.data.frame(de[order(de$pvalue),])
+rownames(de)<-sapply(strsplit(rownames(de),"\\."),"[[",1)
+head(de) %>% kbl() %>% kable_paper("hover", full_width = F)
+
+```
+
+Now let's have a look at some of the charts showing differential expression.
+In particular, an MA plot and volcano plot.
 
@@
 
-maplot(LGvHG,"LG v HG")
-make_volcano(LGvHG,"LG v HG")
+maplot(de,name)
+make_volcano(de,name)
 
@@
 ## Gene sets from Reactome
+
+In order to perform gene set analysis, we need some gene sets.
 
@@
 
+Mitch uses rank-ANOVA statistics for enrichment detection.
+
 ```{r,mitch1}
 
-m <- mitch_import(LGvHG,DEtype = "DEseq2", geneTable = gt)
-head(m)
+m <- mitch_import(de,DEtype = "DEseq2", geneTable = gt)
 mres <- mitch_calc(m,genesets = genesets)
-head(mres$enrichment_result) 
 m_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
 m_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
-length(m_up)
-length(m_dn)
+message(paste("Number of up-regulated pathways:",length(m_up) ))
+message(paste("Number of down-regulated pathways:",length(m_dn) ))
+head(mres$enrichment_result,10)  %>% kbl() %>% kable_paper("hover", full_width = F)
 
@@
 
+Clusterprofiler uses a hypergeometric test.
+Firstly I will conduct the analysis separately for up and down regulated genes and with the correct backgound (as intended by the developers).
+
 ```{r,cp1}
@@
 genesets2 <- read.gmt("ReactomePathways.gmt")
-head(genesets2)
 
-LGvHG_up <- rownames(subset(LGvHG,log2FoldChange>0,padj<0.05))
-LGvHG_up <- unique(gt[which(rownames(gt) %in% LGvHG_up),1])
-head(LGvHG_up)
+de_up <- rownames(subset(de,log2FoldChange>0,padj<0.05))
+de_up <- unique(gt[which(rownames(gt) %in% de_up),1])
 
-LGvHG_dn <- rownames(subset(LGvHG,log2FoldChange<0,padj<0.05))
-LGvHG_dn <- unique(gt[which(rownames(gt) %in% LGvHG_dn),1])
-head(LGvHG_dn)
+de_dn <- rownames(subset(de,log2FoldChange<0,padj<0.05))
+de_dn <- unique(gt[which(rownames(gt) %in% de_dn),1])
 
-LGvHG_bg <- rownames(LGvHG)
-LGvHG_bg <- unique(gt[which(rownames(gt) %in% LGvHG_bg),1])
-length(LGvHG_bg)
-head(LGvHG_bg)
+de_bg <- rownames(de)
+de_bg <- unique(gt[which(rownames(gt) %in% de_bg),1])
 
-c_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
-c_up <- head(c_up,1000)
-head(c_up)
+c_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
 c_up <- rownames(subset(c_up, p.adjust < 0.05))
 
-c_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
-c_dn <- head(c_dn,1000)
-head(c_dn)
+c_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
 c_dn <- rownames(subset(c_dn, p.adjust < 0.05))
@@
 
-## ORA with clusterprofiler combining up and down
+Now performing ORA with clusterprofiler combining up and down.
 
@@
 
-LGvHG_de <- rownames(subset(LGvHG,padj<0.05))
-LGvHG_de <- unique(gt[which(rownames(gt) %in% LGvHG_de),1])
-head(LGvHG_de)
+de_de <- rownames(subset(de,padj<0.05))
+de_de <- unique(gt[which(rownames(gt) %in% de_de),1])
 
-d_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
-d_de <- head(d_de,1000)
+d_de <- as.data.frame(enricher(gene = de_de, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
 d_de <- rownames(subset(d_de, p.adjust < 0.05))
-head(d_de)
 
@@
 
-## ORA with clusterprofiler with whole genome background list
+Now performing ORA with clusterprofiler with whole genome background list
 
@@
 
-LGvHG_bg <- w$GeneInfo$GeneSymbol
-length(LGvHG_bg)
-head(LGvHG_bg)
+de_bg <- w$GeneInfo$GeneSymbol
 
-f_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg,  maxGSSize = 5000, TERM2GENE = genesets2)
-f_up <- head(f_up,1000)
-head(f_up)
+f_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
 f_up <- rownames(subset(f_up, p.adjust < 0.05))
 
-f_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, maxGSSize = 5000, TERM2GENE = genesets2)
-f_dn <- head(f_dn,1000)
-head(f_dn)
+f_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2))
 f_dn <- rownames(subset(f_dn, p.adjust < 0.05))
@@
 
-## ORA combined with clusterprofiler with whole genome background list
+Now performing ORA (combining up and down gene lists) with clusterprofiler with whole genome background list
 
@@
 
-e_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, maxGSSize = 5000, TERM2GENE = genesets2)
-e_de <- head(e_de,1000)
-head(e_de)
+e_de <- as.data.frame(enricher(gene = de_de, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2))
 e_de <- rownames(subset(e_de, p.adjust < 0.05))
@@
 ## Venn diagram comparison
+
+The Venn (or Euler to be more correct) diagram is useful to visualise the overlaps between sets.
 
@@
            "FCS up"=m_up, "FCS dn"=m_dn)
-
-plot(euler(v3),quantities = TRUE, edges = "gray", main="all sets")
 
@@
 dc <- length(intersect(d_de, c(c_up,c_dn))) / length(union(d_de, c(c_up,c_dn)))
-dc
 
@@
 ec <- length(intersect(e_de, c(c_up,c_dn))) / length(union(e_de, c(c_up,c_dn)))
-ec
 
@@
 cm <- length(intersect(c(c_up,c_dn), c(m_up,m_dn))) / length(union(c(c_up,c_dn), c(m_up,m_dn)))
-cm
 
@@
 cf <- length(intersect(c_de, f_de )) / length(union(c_de, f_de))
-cf
 
@@
 mf <- length(intersect(m_de, f_de )) / length(union(m_de, f_de))
-mf
 

### 2021-11-23 example3.Rmd
updated ORA/FCS examples

--- 2021-11-23 example1.Rmd
@@
 ---
-title: "Example gene set analysis"
+title: "Example gene set analysis: The case of HDAC inhibitor SAHA"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
-The dataset being used is SRP128998 and we are comparing the cells grown in normal glucose condition (control) to the high glucose condition (case).
+The dataset being used is SRP096178 and we are comparing the cells grown in normal condition (control) to those grown with addition of SAHA (case).
 
@@
 
-dir.create("images")
-
 ```
 
-## Get expression data and make an MDS plot
+## Get expression data
 
-I'm using some RNA-seq data looking at the effect of hyperglycemia on hepatocytes.
+I'm using some RNA-seq data looking at the effect of SAHA on HAEC cells.
 
-```{r,getdata}
+```{r,getdata,fig.width=7,fig.height=7}
 
-name="SRP128998"
+name="SRP037718"
 mdat<-getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP037718",mdat$SRP_accession),]
 samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$trt<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
-samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
-s1 <- subset(samplesheet,VPA==0)
 
-w<-getDEE2("hsapiens",samplesheet$SRR_accession,metadata=mdat,legacy = TRUE)
+samplesheet$trt<-as.factor(c(1,1,1,0,0,0))
+s1 <- samplesheet
+
+w<-getDEE2("hsapiens",s1$SRR_accession,metadata=mdat,legacy = TRUE)
 x<-Tx2Gene(w)
@@
 Clusterprofiler uses a hypergeometric test.
-Firstly I will conduct the analysis separately for up and down regulated genes and with the correct backgound (as intended by the developers).
+Firstly I will conduct the analysis separately for up and down regulated genes and with the correct background (as intended by the developers).
 

### 2021-11-23 example4.Rmd
updated ORA/FCS examples

--- 2021-11-23 example1.Rmd
@@
 ---
-title: "Example gene set analysis"
+title: "Example gene set analysis: The case of Set7 knockdown"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
-The dataset being used is SRP128998 and we are comparing the cells grown in normal glucose condition (control) to the high glucose condition (case).
+The dataset being used is SRP038101 and we are comparing the cells expressing a set7kd shRNA construct (case) compared to the scrambled construct (control).
 
@@
 
-dir.create("images")
-
 ```
 
-## Get expression data and make an MDS plot
+## Get expression data
 
-I'm using some RNA-seq data looking at the effect of hyperglycemia on hepatocytes.
+I'm using some RNA-seq data looking at the effect of Set7 knockdown on HMEC cells.
 
-```{r,getdata}
+```{r,getdata,fig.width=7,fig.height=7}
 
-name="SRP128998"
+name="SRP096177"
 mdat<-getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP096177",mdat$SRP_accession),]
 samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$trt<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
-samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
-s1 <- subset(samplesheet,VPA==0)
+samplesheet$trt<-as.factor(c(1,1,1,0,0,0))
+s1 <- samplesheet
 
@@
 # counts 
-x1<-x[,which(colnames(x) %in% s1$SRR_accession)]
+x1<-x[,which(colnames(x) %in% samplesheet$SRR_accession)]
 
+getDEE2::getDEE2_bundle()
 ```

### 2021-11-23 example5.Rmd
updated ORA/FCS examples

--- 2021-11-23 example4.Rmd
@@
 ---
-title: "Example gene set analysis: The case of Set7 knockdown"
+title: "Example gene set analysis: The case of LHON causing mutations"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
-The dataset being used is SRP038101 and we are comparing the cells expressing a set7kd shRNA construct (case) compared to the scrambled construct (control).
+The dataset being used is SRP247621 and we are comparing the fibroblast cells from LHON patients (case) versus healthy controls.
 
@@
 
-I'm using some RNA-seq data looking at the effect of Set7 knockdown on HMEC cells.
+I'm using some RNA-seq data looking at the difference in fibroblast gene expression between control and LHON patients.
 
@@
 
-name="SRP096177"
+name="SRP247621"
 mdat<-getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP096177",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP247621",mdat$SRP_accession),]
 samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$trt<-as.factor(c(1,1,1,0,0,0))
+samplesheet$trt<-as.factor(c(1,1,1,2,2,2,0,0,0)) # exclude carriers for simplicity
+samplesheet <- samplesheet[which(samplesheet$trt!=2),]
 s1 <- samplesheet
@@
 
-getDEE2::getDEE2_bundle()
 ```

### 2021-11-23 example2.Rmd
updates

--- 2021-08-13 example.Rmd
@@
 Here we are performing an analysis of some gene expression data to demonstrate the difference between ORA and FCS methods and to highlight the differences caused by improper background gene set use.
+
+The dataset being used is SRP038101 and we are comparing the cells grown in normal condition (control) to those grown with addition of Azacitidine (case).
+
+Data are obtained from http://dee2.io/
 
@@
 
-I'm using some RNA-seq data looking at the effect of hyperglycemia on hepatocytes.
+I'm using some RNA-seq data looking at the effect of Azacytidine on AML3 cells.
 
@@
 
+name="SRP038101"
 mdat<-getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
+samplesheet <- mdat[grep("SRP038101",mdat$SRP_accession),]
 samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$HG<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
-samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
-s1 <- subset(samplesheet,VPA==0)
+samplesheet$trt<-as.factor(c(1,1,1,0,0,0))
+s1 <- samplesheet
 
@@
 gt$accession<-rownames(gt)
-head(gt)
 
 # counts 
-x1<-x[,which(colnames(x) %in% s1$SRR_accession)]
+x1<-x[,which(colnames(x) %in% samplesheet$SRR_accession)]
+
+```
+
+Here show the number of genes in the annotation set, and those detected above the detection threshold.
+
+```{r,filter}
 
@@
 
-plot(euler(v3),quantities = TRUE, main="all sets")
-
 png("images/fcs_ora1.png")
-plot(euler(v1),quantities = TRUE, main="FCS vs ORA")
+plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
 dev.off()

### 2021-11-23 example2.Rmd
expansion of the evaluation of misuse

@@
 
-png("images/fcs_ora1.png")
+png("images/fcs_ora2.png")
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-png("images/orabg1.png")
+png("images/orabg2.png")
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-pdf("images/fcs_ora1.pdf",width=4,height=4)
+pdf("images/fcs_ora2.pdf",width=4,height=4)
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-pdf("images/orabg1.pdf",width=4,height=4)
+pdf("images/orabg2.pdf",width=4,height=4)
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-pdf("images/oracomb1.pdf",width=4,height=4)
+pdf("images/oracomb2.pdf",width=4,height=4)
 plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
@@
 
+saveRDS(dat,file = "ex2dat.rds")
+
 ```

### 2021-11-23 example3.Rmd
expansion of the evaluation of misuse

@@
 
-png("images/fcs_ora1.png")
+png("images/fcs_ora3.png")
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-png("images/orabg1.png")
+png("images/orabg3.png")
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-png("images/oracomb1.png")
+png("images/oracomb3.png")
 plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
@@
 
-pdf("images/fcs_ora1.pdf",width=4,height=4)
+pdf("images/fcs_ora3.pdf",width=4,height=4)
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-pdf("images/orabg1.pdf",width=4,height=4)
+pdf("images/orabg3.pdf",width=4,height=4)
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-pdf("images/oracomb1.pdf",width=4,height=4)
+pdf("images/oracomb3.pdf",width=4,height=4)
 plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
@@
 
+saveRDS(dat,file = "ex3dat.rds")
+
 ```

### 2021-11-23 example4.Rmd
expansion of the evaluation of misuse

@@
 
-png("images/fcs_ora1.png")
+png("images/fcs_ora4.png")
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-png("images/orabg1.png")
+png("images/orabg4.png")
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-png("images/oracomb1.png")
+png("images/oracomb4.png")
 plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
@@
 
-pdf("images/fcs_ora1.pdf",width=4,height=4)
+pdf("images/fcs_ora4.pdf",width=4,height=4)
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-pdf("images/orabg1.pdf",width=4,height=4)
+pdf("images/orabg4.pdf",width=4,height=4)
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-pdf("images/oracomb1.pdf",width=4,height=4)
+pdf("images/oracomb4.pdf",width=4,height=4)
 plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
@@
 
+saveRDS(dat,file = "ex4dat.rds")
+
 ```

### 2021-11-23 example5.Rmd
expansion of the evaluation of misuse

@@
 
-png("images/fcs_ora1.png")
+png("images/fcs_ora5.png")
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-png("images/orabg1.png")
+png("images/orabg5.png")
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-png("images/oracomb1.png")
+png("images/oracomb5.png")
 plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
@@
 
-pdf("images/fcs_ora1.pdf",width=4,height=4)
+pdf("images/fcs_ora5.pdf",width=4,height=4)
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-pdf("images/orabg1.pdf",width=4,height=4)
+pdf("images/orabg5.pdf",width=4,height=4)
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-pdf("images/oracomb1.pdf",width=4,height=4)
+pdf("images/oracomb5.pdf",width=4,height=4)
 plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
@@
 
+saveRDS(dat,file = "ex5dat.rds")
+
 ```

### 2021-11-23 examples.Rmd
expansion of the evaluation of misuse

---
title: "Example gene set analysis: bringing it together"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichmentMethods

## Intro

Previously we analysed five datasets below and looked at the differences between FCS and ORA methods for gene set enrichment.

| Dataset | SRA accesion | genes in annotation set | genes detected | genes differentially expressed |
| :---: | :---: | :---: | :---: | :---: | 
| 1. | SRP128998 | 39297 | 15635 | 3472 |
| 2. | SRP038101 | 39297 | 13926 | 3589 | 
| 3. | SRP096178 | 39297 | 15477 | 9488 |
| 4. | SRP038101 | 39297 | 15607 | 5150 | 
| 5. | SRP247621 | 39297 | 14288 | 230 |

```{r,begin}

suppressPackageStartupMessages({
library("kableExtra")
library("beeswarm")
})

```

## Summarise

Here I'm summarising the results of the five example data sets.

```{r,getdata,fig.width=7,fig.height=5}

exdat=NULL
ex1dat <- readRDS("ex1dat.rds")
ex2dat <- readRDS("ex2dat.rds")
ex3dat <- readRDS("ex3dat.rds")
ex4dat <- readRDS("ex4dat.rds")
ex5dat <- readRDS("ex5dat.rds")

exdat <- rbind(ex1dat,ex2dat,ex3dat,ex4dat,ex5dat)
rownames(exdat) <- c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621")

mysd <- apply(exdat,2,sd)

par(mar=c(5,10,3,1))
barplot(rev(colMeans(exdat)),xlab="mean Jaccard index",horiz = TRUE, las =1, xlim=c(0,.7) )

pdf("images/jacbar1.pdf",width=4,height=4)
par(mar=c(5,10,3,1))
barplot(rev(colMeans(exdat)),xlab="mean Jaccard index",horiz = TRUE, las =1, xlim=c(0,.7) )
dev.off()

png("images/jacbar1.png")
par(mar=c(5,10,3,1))
barplot(rev(colMeans(exdat)),xlab="mean Jaccard index",horiz = TRUE, las =1, xlim=c(0,.7) )
dev.off()

exdat <- rbind(exdat,"mean"=colMeans(exdat))

cols <- c("#CC79A7","#56B4E9","#E69F00","#F0E442","#009E73","black")
coll <- list(cols, cols, cols, cols, cols)
par(mar=c(5,10,3,1))
beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=2,pch=19, xlab = "Jaccard index")
legend("bottomright", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","mean"), pch=19,  col=cols, cex=1,pt.cex = 1.5)

pdf("images/jacswarm1.pdf",width=6,height=5)
par(mar=c(5,9,3,1))
beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=2,pch=19, xlab = "Jaccard index")
legend("bottomright", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","mean"), pch=19,  col=cols, cex=1.0, pt.cex = 1.7)
dev.off()

png("images/jacswarm1.png")
par(mar=c(5,10,3,1))
beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=2,pch=19, xlab = "Jaccard index")
legend("bottomright", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","mean"), pch=19,  col=cols, cex=1,pt.cex = 1.5)
dev.off()

rbind(exdat,mysd) %>% kbl() %>% kable_paper("hover", full_width = F)

```

## Session information

```{r,session}

sessionInfo()

```

### 2021-11-25 examples.Rmd
more work on examples

@@
 | 5. | SRP247621 | 39297 | 14288 | 230 |
+| 6. | SRP253951 | 39297 | 15182 | 8588 |
+| 7. | SRP068733 | 39297 | 14255 | 7365 |
 
@@
 ex5dat <- readRDS("ex5dat.rds")
+ex6dat <- readRDS("ex6dat.rds")
+ex7dat <- readRDS("ex7dat.rds")
 
-exdat <- rbind(ex1dat,ex2dat,ex3dat,ex4dat,ex5dat)
-rownames(exdat) <- c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621")
+exdat <- rbind(ex1dat,ex2dat,ex3dat,ex4dat,ex5dat,ex6dat,ex7dat)
+rownames(exdat) <- c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","SRP253951","SRP068733")
 
+exdat %>% kbl(caption="jaccard index values across seven RNA-seq studies") %>% kable_paper("hover", full_width = F)
+
+```
+
+```{r,summary_stats}
+
+means <- colMeans(exdat)
+medians <- colMedians(exdat)
+sds <- sd(exdat)
+summstats <- rbind(means,medians,sds)
+summstats %>% kbl(caption="summary statistics of different enrichment analysis comparisons") %>% kable_paper("hover", full_width = F)
+
+```
+
+Here let's plot the jaccard scores.
+First as a boxplot and then overlay the dataset studies over the top.
+
+```{r,jacbox}
+
+cols <- c("#CC79A7","#56B4E9","#E69F00","#F0E442","#009E73","#0072B2","#D55E00")
+coll <- list(cols, cols, cols, cols, cols)
+par(mar=c(5,10,3,1))
+boxplot(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),names = rev(colnames(exdat)),horizontal = TRUE, las=1 , col="white", cex=1.5,pch=19, xlab = "Jaccard index")
+beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=1.5,pch=19, xlab = "Jaccard index",add = TRUE)
+legend("bottomright", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","SRP253951","SRP068733"), pch=19,  col=cols, cex=1,pt.cex = 1.5)
+
+pdf("images/jacbox1.pdf",width=5,height=4)
+par(mar=c(5,9,3,2))
+boxplot(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),names = rev(colnames(exdat)),horizontal = TRUE, las=1 , col="white", cex=1.5,pch=19, xlab = "Jaccard index")
+beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=1.5,pch=19, xlab = "Jaccard index",add = TRUE)
+legend("topleft", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","SRP253951","SRP068733"), pch=19,  col=cols, cex=1,pt.cex = 1.5)
+dev.off()
+
+png("images/jacbox1.png")
+par(mar=c(5,9,3,1))
+boxplot(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),names = rev(colnames(exdat)),horizontal = TRUE, las=1 , col="white", cex=1.5,pch=19, xlab = "Jaccard index")
+beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=1.5,pch=19, xlab = "Jaccard index",add = TRUE)
+legend("bottomright", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","SRP253951","SRP068733"), pch=19,  col=cols, cex=1.2,pt.cex = 1.5)
+dev.off()
+
+```
+
+Now plot as a simple bargraph
+
+```{r,jacbar}
+
+mymean <- colMeans(exdat)
 mysd <- apply(exdat,2,sd)
@@
 dev.off()
+
+```
+
+Now plot as a beeswarm.
+
+```{r,jacswarm}
 

### 2021-11-25 examples.Rmd
bugfix

@@
 medians <- colMedians(exdat)
-sds <- sd(exdat)
+sds <- apply(exdat,2,sd)
 summstats <- rbind(means,medians,sds)

### 2021-11-29 replication/PMC6349697.Rmd
replication study 1

---
title: "PMC6349697: A replication study"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichmentMethods

## Intro

```{r,begin}

suppressPackageStartupMessages({
library("getDEE2")
library("DESeq2")
library("clusterProfiler")
library("org.Hs.eg.db")
library("mitch")
library("kableExtra")
library("eulerr")
})

```

In [Kaumadi et al 2021](https://www.biorxiv.org/content/10.1101/2021.09.06.459114v1) we
showed that major problems plague functional enrichment analysis in many journal articles.

Here I will do my best to reproduce and replicate the findings of some studies.
These were selected on the following criteria:

* Included in the 2019 group that was studies in Kaumadi et al 2021.

* Human study

* RNA-seq gene expression

* Performed DAVID gene set analysis

* Provided gene list

From this subset, four studies were selected

| Study | Analytical issues |
| --- | --- |
| PMC6349697 | Background |
| PMC6425008 | Background, FDR |
| PMC6463127 | Background, FDR |
| PMC6535219 | Background |

In this document I will be seeing if I can reproduce the same findings by using the same
tool with the gene list provided in the article.
Also I will correct the method wherever possible.
This means using the correct background gene list, using FDR control and performing 
separate analysis of up and down-regulated gene lists.

The study examined in this document is PMC6349697: 
[Liu et al, 2019](https://pubmed.ncbi.nlm.nih.gov/30723491/).

In this study, LINC00941 is the focus.
From TGCA expression data, a coexpression network was built and then genes associated with
LINC00941 were obtained.
These genes were then subject to enrichment analysis of GO terms and KEGG pathways using 
DAVID v6.8, with pathways FDR<0.05 considered significant.
No information is provided in the article about a background gene list.

From the article, they present in [Supplementary Table S5](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6349697/#SM5)
a list of 123 co-expressed genes.

In [Supplementary Table S6](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6349697/#SM6) 
they show the 3180 genes that underwent clustering, along with the cluster
assignment.
There are 124 genes in the "yellow" cluster.

In [Supplementary Table S7](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6349697/#SM7),
authors show the significant results of the enrichment tests.

| Category | Term | Count | FDR |
| --- | --- | --- | --- |
| GOTERM_BP_DIRECT | GO:0030574~collagen catabolic process | 15 | 2.73E-15 |
| GOTERM_BP_DIRECT | GO:0030199~collagen fibril organization | 13 | 8.48E-15 |
| GOTERM_BP_DIRECT | GO:0030198~extracellular matrix organization | 20 | 1.43E-14 |
| GOTERM_BP_DIRECT | GO:0007155~cell adhesion | 21 | 8.40E-09 |
| GOTERM_BP_DIRECT | GO:0001501~skeletal system development | 10 | 2.67E-04 |
| GOTERM_BP_DIRECT | GO:0001503~ossification | 7 | 0.0154 |
| KEGG | hsa04512:ECM-receptor interaction | 10 | 1.81E-06 |
| KEGG | hsa04974:Protein digestion and absorption | 9 | 4.53E-05 |
| KEGG | hsa04510:Focal adhesion | 10 | 0.0032 |
| KEGG | hsa04151:PI3K-Akt signaling pathway | 11 | 0.032 |

## Try to reproduce

Here I'm going to reproduce the findings with DAVID as best I can.

Here is the list of genes used as the foreground.
The gene IDs are saved as a txt file on this repository called "PMC6349697_fg.txt"

```{r,fg}

fg <- readLines("PMC6349697_fg.txt")

fg

```

These genes were submitted to DAVID v6.8 on 29th November 2021 without any particular 
background gene set, which gave the following results.

```{r,res1}

res1 <- read.table("PMC6349697_res1.tsv",header=TRUE,sep="\t")

res1 %>% kbl(caption="DAVID results obtained using article gene list without background") %>% kable_paper("hover", full_width = F)

```

This result means that most pathways could be replicated using the gene list as provided 
by the authors without the use of a custom background gene list.

## Try to reproduce with corrected background gene list

Now I will try to use a corrected gene list.
I see that the Table S6 has all the other clustered genes sets, totalling 3180 genes, 
however this is likely not all the  genes that were detected in the dataset.
Therefore, I downloaded the TGCA data used in the study and identified all the genes with
an average of 10 reads per sample or more.

There are 25947 genes in the background.

```{r,bg1}

bg <- readLines("PMC6349697_bg.txt")

length(bg)

```

These genes are stored as "PMC6349697_bg.txt" in the repo.

Next I repeated DAVID analysis, this time using the correct background gene list.
Here is the result I obtained.

```{r,res2}

res2 <- read.table("PMC6349697_res2.tsv",header=TRUE,sep="\t")

res2 %>% kbl(caption="DAVID results obtained using the correct background") %>% kable_paper("hover", full_width = F)

```

In contrast to the six GOs and four KEGGs identified with FDR<0.05 originally, with
the corrected background there's eight GOs and six KEGGs now.

## What about the conclusions of the study?

Some interesting statements about the enrichment analysis:

* Abstract:

>"... functional enrichment analysis of LINC00941 co-expression network demonstrated that LINC00941 might be an essential regulator of tumor metastasis and cancer cell proliferation. "

* Introduction final paragraph:

> "... through the functional enrichment analysis, we found that LINC00941 might be an essential regulator of tumor metastasis and cancer cell proliferation. "

* Results:

> "Some GO terms, such as extracellular matrix organization (GO: 0030198) and cell adhesion (GO: 0007155) are cell migration-related GO processes, which are associated with tumor metastasis. We detected that ECM-receptor interaction (KEGG: hsa04512) and Focal adhesion (KEGG: hsa04510) are metastasis-related pathways. PI3K-Akt signaling pathway (KEGG: hsa04151) is cell proliferation-related pathway. Our findings demonstrated that LINC00941 could be a potential regulator of tumor metastasis and cancer cell proliferation."

* Discussion:

> "The results of GO terms and KEGG pathways were mainly enriched in cell proliferation, cell migration, and tumor metastasis."

Indeed the results indicate LINC00941 is co-regulated with genes involved in ECM structure and metabolism.
This pathway might be involved with tissue remodelling involved with tumour maturation.
This looks to all be okay, apart from one statement about proliferation.
None of these ontologies/pathways are related to cell cycle or proliferation.
The statement in the discussion is also suspect, because none of these pathways are closely linked to proliferation, migration or 
metastasis.
There are likely other gene sets available which can be used to test these associations more thoroughly.
For example there are 58 gene sets containing the keyword "metastasis" in the MSigDB v7.4 collection (29/11/2021).

## What about a replication using an R script

Here I'm using the enrichGO function to analyse the data.
The algorithm is slightly different and the gene sets might be a different version.
First GO biological processes.

```{r,repl2_go}

go_bp <- enrichGO(gene = fg,  keyType = "ENSEMBL", universe = bg,
    OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", pvalueCutoff  = 0.01,
    qvalueCutoff  = 0.05, readable = TRUE)

go_bp <- data.frame(go_bp)

go_bp <- subset(go_bp,qvalue<0.05)

nrow(go_bp)

writeLines(go_bp$Description,"PMC6349697_gobp.txt")

go_bp[,c(2:7,9)] %>% kbl(caption="Clusterprofiler GO results obtained using the correct background") %>% kable_paper("hover", full_width = F)

```

Now KEGG analysis but need to convert to entrez first

```{r,repl2_kegg}

fg_entrez <- unlist(mget(fg, org.Hs.egENSEMBL2EG, ifnotfound = NA))

bg_entrez <- unlist(mget(bg, org.Hs.egENSEMBL2EG, ifnotfound = NA))

kegg <- enrichKEGG(gene = fg_entrez, universe = bg_entrez,
    organism = "hsa", pAdjustMethod = "BH", pvalueCutoff  = 0.01,
    qvalueCutoff  = 0.05)

kegg <- as.data.frame(kegg)

writeLines(kegg$Description,"PMC6349697_kegg.txt")

kegg[,c(2:7,9)] %>% kbl(caption="Clusterprofiler KEGG results obtained using the correct background") %>% kable_paper("hover", full_width = F)

```

The clusterProfiler GO:BP results gave 69 significant terms.
Prominent similar terms to the article were ECM, collagen, skeletal system and ossification, whereas clusterprofiler 
also gave some interesting gene sets such as angiogenesis, smooth muscle cell migration, response to transforming 
growth factor beta.

With KEGG analysis, only two sets were consistent, while focal adhesion and PI3K-Akt were not significant.

In conclusion, it appears that the lack of a background gene set did not have much of an effect on the results,
however the conclusions drawn in the article regarding proliferation and metastasis based on enrichment analysis
are unfounded.

## Session information

```{r,session}

sessionInfo()

```

### 2021-11-30 replication/PMC6425008.Rmd
adding replication data

---
title: "PMC6425008: A replication study"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichmentMethods

## Intro

```{r,begin}

suppressPackageStartupMessages({
library("getDEE2")
library("DESeq2")
library("clusterProfiler")
library("org.Hs.eg.db")
library("mitch")
library("kableExtra")
library("eulerr")
})

```

In [Kaumadi et al 2021](https://www.biorxiv.org/content/10.1101/2021.09.06.459114v1) we
showed that major problems plague functional enrichment analysis in many journal articles.

Here I will do my best to reproduce and replicate the findings of some studies.
These were selected on the following criteria:

* Included in the 2019 group that was studies in Kaumadi et al 2021.

* Human study

* RNA-seq gene expression

* Performed DAVID gene set analysis

* Provided gene list

From this subset, four studies were selected

| Study | Analytical issues |
| --- | --- |
| PMC6349697 | Background |
| PMC6425008 | Background, FDR |
| PMC6463127 | Background, FDR |
| PMC6535219 | Background |

In this document I will be seeing if I can reproduce the same findings by using the same
tool with the gene list provided in the article.
Also I will correct the method wherever possible.
This means using the correct background gene list, using FDR control and performing 
separate analysis of up and down-regulated gene lists.

The study examined in this document is PMC6425008: 
[Balusamy et al, 2019](https://pubmed.ncbi.nlm.nih.gov/30890753/).

In this study, the effect of citral (3,7-dimethyl-2,6-octadienal) on cells is investigated.
In the paper, they isolate the compound with chromatography, followed by in vitro 
treatment of AGL cell line with different levels of the compound.
Treated cells were characterised in terms of morphology and viability.
Citral caused shrunken cells, amorphous shape and lower cell number.
To understand the pattern of gene expression, RNA-seq of n=1 control and n=1 treated
samples was performed yielding 75 and 65 M read pairs respectively which were deposited to
SRA (SRP150561).

Some type of differential analysis was performed with Cuffdiff, but with no replicates,
this might not be reliable.
Anyway, the up and down regulated gene lists are found in (Supplementary File S6)[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6425008/bin/41598_2019_41406_MOESM7_ESM.xlsx].
It contains 612 genes; 216 up and 396 downregulated genes.

There is a network analysis which looks a bit like a hairball and isn't described in any
detail.

Figures 5-7 show the results of the enrichment analysis.
A number of gene sets are presented as significant, despite lack of FDR correction.
There was no mention of a background gene list in the article.

Here I have summarised the "significant" findings presented in the supplements.
There are some problems.
Firstly, only 7 of these have an FDR (BH method) < 0.05.
Secondly, some of these show that only 1 gene was found in a gene set.
Surely one gene on its own can't constitute a valid enrichment.

```{r,orig_res1}

orig <- read.table("PMC6425008_orig.tsv",sep="\t",header=TRUE)

orig %>% kbl(caption="Enrichment results presented in PMC6425008 supplementary files") %>% kable_paper("hover", full_width = F)

```

In this work, I will see if I can replicate the results for the 7 FDR significant gene sets.

```{r,orig_res3}

subset(orig,BH.method < 0.05)$GO.Term

```

The article also shows some KEGG pathway analysis in Table 2 but shows results for individual genes and doesn't present any results
of enrichment tests for gene sets.

In Table 3 of the article, GO enrichment table is shown for apoptosis pathways.
It seems only two of the sets are significant: "Response to unfolded protein" and "Regulation of cell cycle"

```{r,orig_res2}

orig2 <- read.table("PMC6425008_orig2.tsv",sep="\t",header=TRUE)

orig2 %>% kbl(caption="Enrichment results presented in PMC6425008 Table 3") %>% kable_paper("hover", full_width = F)

```

In addition, the results of Table 3 are inconsistent with the GO results shown in the supplements.
For example "Response to unfolded protein" and "Regulation of cell cycle" are missing from the supplementary tables.

## Try to reproduce

I submitted the 216 upregulated genes to DAVID and these were mapped to 213 DAVID gene IDs.
This analysis without a background list gave 31 BP, 17 CC and 10 MF significant GO terms (see table below).
This is drastically different to what is shown in the article.

```{r,repl1}

repl1 <- read.table("PMC6425008_repl1.tsv", header=TRUE, sep="\t")

repl1 %>% kbl(caption="DAVID Enrichment results when using the genes provided in the supplement and a corrected background gene list.") %>% kable_paper("hover", full_width = F)

```

## Try to reproduce with corrected background gene list

Now I'll try this again, but using a correct background gene set.
As the data is found on the DEE2 database, I can see which genes are detected and I can use those ones as a background.
As the results below show, there are 14209 genes in the background.

```{r,bg}

m <- getDEE2::getDEE2Metadata(species="hsapiens")
ss <- m[grep("SRP150561",m$SRP_accession),]
x <- getDEE2(species="hsapiens",SRRvec=ss$SRR_accession,legacy=TRUE)
mx <- x$GeneCounts
mxf <- mx[which(rowMeans(mx)>=10),]
nrow(mx) #58302
nrow(mxf) #14119
bg <- rownames(mxf)

geneinfo <- x$GeneInfo
bgg <- unique(geneinfo[which(rownames(geneinfo) %in%bg),"GeneSymbol"])
length(bgg)
writeLines(bgg,"PMC6425008_bg.txt")

```
So I tried DAVID with the new background, but was greeted with "Due to the ambiguous nature of Gene Symbol, DAVID does not allow it to be used for a background."
This poses a problem for many researchers.

Anyway so I will use Ensembl IDs instead (14218 IDs).

```{r,bg2}

writeLines(bg,"PMC6425008_bg2.txt")

```

Here are the results when using all detected genes as the background.

```{r,repl2}

repl2 <- read.table("PMC6425008_repl2.tsv", header=TRUE, sep="\t")

repl2 %>% kbl(caption="DAVID Enrichment results when using the genes provided in the supplement.") %>% kable_paper("hover", full_width = F)

```

## Do the results reproduce?

Considering the results shown in the supplement, there were 7 FDR significant sets.

When the analysis was rerun without a specific background there were mixed results with most of the findings not reproducing.

* Did reproduce: "Proteasome complex", "Inclusion body"

* Did not reproduce: "Protein metabolism", "Heat shock protein activity", "Ubiquitin-specific protease activity", "Exosomes", "Lysosome"

When considering the re-analysis with the corrected background set:

* Did reproduce: "Proteasome complex", "Inclusion body"

* Did not reproduce: "Protein metabolism", "Heat shock protein activity", "Ubiquitin-specific protease activity", "Exosomes", "Lysosome"

This suggests that the main problem here is that the results shown in the supplement are incorrect and are inconsistent with the gene list provided
in the supplement.

Regarding the results shown in Table 3, "Response to unfolded protein" was reproduced, while "Regulation of cell cycle" wasn't reproduced.

## What about the conclusions of the study?

Some interesting statements about the enrichment analysis:

* Abstract:

>"... Furthermore, the enrichment and KEGG pathway results suggested that there are several genes involved to induce apoptosis pathway. "

>".. The enrichment analysis identified DEGs genes from transcriptome libraries including cell death, cell cycle, apoptosis and cell growth. "

>"The present study showed the significant inhibition effect upon citral by regulating various genes involved in signaling pathways, inhibits metastasis, colony formation and induced apoptosis both in silico and in vitro."

* Results and Discussion:

> "KEGG analysis revealed that MAPK signaling pathway, NF-kappa B signaling pathway, p53 signaling pathway, PI3-K-Akt signaling pathway, pathways in cancer, and prostate cancer pathways are mostly enriched in up and down-regulated genes which identified between normal and citral treated samples in AGS. Among them, cell cycle, pathways in cancer, and MAPK signaling pathways were observed as mostly enriched pathways. In addition, several common genes were identified in these pathways were listed in Table 2. Among them, cell cycle (hsa04110) and cancer (hsa05200) occupied several down-regulated DEG genes. The genes involved in cell cycle synthesis are very crucial for the inhibition of cancer cell progression19. For example, cyclin-dependent kinases (CKD2) are the most important regulators for the transition and cell cycle progression and play a significant role in regulating cell cycle division including centromere duplication, DNA synthesis, G1-S transition, and modulation of G2 progression20–22. Furthermore, the enrichment analysis and KEGG pathway results indicated that, there are several genes involved to initiate apoptosis pathway that are directly/indirectly related to the treatment of citral. In addition, enrichment analysis identified DEGs genes from transcriptome libraries were mainly involved in cell death, cell cycle, apoptotic process, and cell growth (Table 3)."

To summarise the conclusions, authors allude to enrichment analysis identifying apoptosis pathways.
They speak about the KEGG classifications but there was no statistical enrichment test performed so this is not a quantitative enrichment.
Specifically they mention table 3 supports enrichment analysis that indicates "cell death, cell cycle, apoptotic process, and cell growth"

The only hint of apoptosis pathways identified was the "Ubiquitin-specific protease activity" found in the original study (FDR=3.1e-04), and this was 
partially reproduced with DAVID without a background set (GO:0051436~negative regulation of ubiquitin-protein ligase activity involved in mitotic cell cycle; 
FDR=3.154072e-09), however when a correct background gene set was used, no ubiquitin-related pathways were identified as significant.

## My conclusion

So to conclude, it appears that most pathway results weren't reproduced regarding cell cycle, signaling or apoptosis.

What I did find, was that the upregulated genes were enriched in functions related to protein misfolding, eg: chaperones, proteasome, etc.

## Session information

```{r,session}

sessionInfo()

```

### 2021-12-01 replication/PMC6463127.Rmd
PMC6463127.Rmd

---
title: "PMC6463127: A replication study"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichmentMethods

## Intro

```{r,begin}

suppressPackageStartupMessages({
library("getDEE2")
library("DESeq2")
library("clusterProfiler")
library("org.Hs.eg.db")
library("mitch")
library("kableExtra")
library("eulerr")
})

```

In [Kaumadi et al 2021](https://www.biorxiv.org/content/10.1101/2021.09.06.459114v1) we
showed that major problems plague functional enrichment analysis in many journal articles.

Here I will do my best to reproduce and replicate the findings of some studies.
These were selected on the following criteria:

* Included in the 2019 group that was studies in Kaumadi et al 2021.

* Human study

* RNA-seq gene expression

* Performed DAVID gene set analysis

* Provided gene list

From this subset, four studies were selected

| Study | Analytical issues |
| --- | --- |
| PMC6349697 | Background |
| PMC6425008 | Background, FDR |
| PMC6463127 | Background, FDR |
| PMC6535219 | Background |

In this document I will be seeing if I can reproduce the same findings by using the same
tool with the gene list provided in the article.
Also I will correct the method wherever possible.
This means using the correct background gene list, using FDR control and performing 
separate analysis of up and down-regulated gene lists.

The study examined in this document is PMC6463127: 
[Chai et al, 2019](https://pubmed.ncbi.nlm.nih.gov/30832348/).

The focus of this study is to compare the RNA expression profiles of parathyroid adenomas with normal parathyroid gland tissue.
They found 247 differentially expressed genes; 45 up-regulated, 202 down-regulated.

They used DAVID for GO/KEGG analysis.
They did not perform FDR correction of enrichment results.

Here are the upregulated and downregulated GO terms and the downregulated KEGG pathways. 
(Presuming there were no significantly upregulated KEGG pathways.)

```{r,orig1}

orig1 <- read.table("PMC6463127_orig_up.tsv",header=TRUE,sep="\t")

orig1 %>% kbl(caption="upregulated GO terms shown in the article") %>% kable_paper("hover", full_width = F)

orig2 <- read.table("PMC6463127_orig_dn.tsv",header=TRUE,sep="\t")

orig2 %>% kbl(caption="down-regulated GO terms shown in the article") %>% kable_paper("hover", full_width = F)

orig3 <- read.table("PMC6463127_orig_keggdn.tsv",header=TRUE,sep="\t")

orig3 %>% kbl(caption="down-regulated KEGG pathways shown in the article") %>% kable_paper("hover", full_width = F)

```

## Try to reproduce

So let's see if I can reproduce their result using DAVID without a background gene list.
The study provided the differentially regulated gene lists in Table S2

Here are the significant GO terms for the up and down directions, along with the downregulated KEGG pathways.

```{r,repl1}

repl1up <- read.table("PMC6463127_repl_up.tsv",header=TRUE,sep="\t")

repl1up %>% kbl(caption="Upregulated GO terms obtained using article gene list without background") %>% kable_paper("hover", full_width = F)

repl1dn <- read.table("PMC6463127_repl_dn.tsv",header=TRUE,sep="\t")

repl1dn %>% kbl(caption="Downregulated GO terms obtained using article gene list without background") %>% kable_paper("hover", full_width = F)

repl1kegg <- read.table("PMC6463127_repl_kegg.tsv",header=TRUE,sep="\t")

repl1kegg %>% kbl(caption="Downregulated KEGG pathways obtained using article gene list without background") %>% kable_paper("hover", full_width = F)

```

There is a supplement with all expressed genes (Supplementary Table S2) which I will use as a background.
To get this to work, I will convert the gene symbols to Ensembl identifiers.
There were only ~8000 expressed genes which is a bit low.

```{r,bg}

bgg <- readLines("PMC6463127_bgg.txt")
head(bgg)
length(bgg)

geneinfo <- read.table("http://dee2.io/data/hsapiens/hsa_gene_info.tsv",header=TRUE)

bg <- geneinfo[which(geneinfo$GeneSymbol %in% bgg),"GeneID"]
bg <- bg[which(!is.na(bg))]
bg <- unique(bg)
length(bg)
writeLines(bg,"PMC6463127_bge.txt")

```

## Try to reproduce with corrected background gene list

Then I ran DAVID again using the correct background. 
Here is the result of the upregulated genes.

```{r,replbg1up}

repl2up <- read.table("PMC6463127_repl_up2.tsv",header=TRUE,sep="\t")

repl2up %>% kbl(caption="Upregulated GO terms obtained using article gene list with a background") %>% kable_paper("hover", full_width = F)

repl2dn <- read.table("PMC6463127_repl_dn2.tsv",header=TRUE,sep="\t")

repl2dn %>% kbl(caption="Downregulated GO terms obtained using article gene list with a background") %>% kable_paper("hover", full_width = F)

repl2kegg <- read.table("PMC6463127_repl_kegg2.tsv",header=TRUE,sep="\t")

repl2kegg %>% kbl(caption="Downregulated KEGG pathways obtained using article gene list with a background") %>% kable_paper("hover", full_width = F)

```

## What about a replication using an R script

Here I'm using the enrichGO function to analyse the data.
The algorithm is slightly different and the gene sets might be a different version.
First GO biological processes.

```{r,repl2_go}

# from supp table 2
upg <- readLines("PMC6463127_up.txt")
dng <- readLines("PMC6463127_dn.txt")

upe <- geneinfo[which(geneinfo$GeneSymbol %in% upg),"GeneID"]
dne <- geneinfo[which(geneinfo$GeneSymbol %in% dng),"GeneID"]

# up
go_up <- enrichGO(gene = upe,  keyType = "ENSEMBL", universe = bg,
    OrgDb = org.Hs.eg.db, ont = "ALL", pAdjustMethod = "BH", pvalueCutoff  = 0.05,
    qvalueCutoff  = 1, readable = TRUE)

go_up <- data.frame(go_up)
nrow(go_up)

go_up <- subset(go_up,qvalue<0.05)
nrow(go_up)

go_up[,c(1,3:7)] %>% kbl(caption="Upregulated GO terms obtained using clusterProfiler with the correct background") %>% kable_paper("hover", full_width = F)

# dn
go_dn <- enrichGO(gene = dne,  keyType = "ENSEMBL", universe = bg,
    OrgDb = org.Hs.eg.db, ont = "ALL", pAdjustMethod = "BH", pvalueCutoff  = 0.05,
    qvalueCutoff  = 1, readable = TRUE)

go_dn <- data.frame(go_dn)
nrow(go_dn)

go_dn <- subset(go_dn,qvalue<0.05)
nrow(go_dn)

go_dn[,c(1,3:7)] %>% kbl(caption="Downregulated GO terms obtained using clusterProfiler with the correct background") %>% kable_paper("hover", full_width = F)

```

Now KEGG analysis but need to convert to entrez first

```{r,repl2_kegg}

dn_entrez <- unlist(mget(dne, org.Hs.egENSEMBL2EG, ifnotfound = NA))

bg_entrez <- unlist(mget(bg, org.Hs.egENSEMBL2EG, ifnotfound = NA))

kegg <- enrichKEGG(gene = dn_entrez, universe = bg_entrez,
    organism = "hsa", pAdjustMethod = "BH", pvalueCutoff  = 0.05,
    qvalueCutoff  = 1)

kegg <- as.data.frame(kegg)

kegg[,c(2:7)] %>% kbl(caption="Clusterprofiler KEGG results obtained using the correct background") %>% kable_paper("hover", full_width = F)

```

## Are the conclusions of the study supported?

In the article, they obtained 45 upregulated and 78 downregulated GO terms, along with 5 downregulated KEGG pathways.

The authors don't specifically draw any conclusions from the enrichment analysis, but let the reader come to their
own conclusions on the basis of the results tables presented.

Anyway, we can look at whether the top results were confirmed with a corrected method.

I repeated the analysis without a background and received the same upregulated GO terms as the article 
(16 terms with p<0.05), but of these, only 5 gave an FDR<0.05.

When a correct background gene list was used, only 2 GOs were upregulated with FDR<0.05.
Basically, there were no significantly upregulated GO BP or MF terms when using a correct background and FDR control. 
Using a clusterProfiler script, I got 9 upregulated GOs, 8 of which were different to the DAVID results.

It was the same story with the downregulated GOs, but after using a background gene list
the p-values were larger and the ranks of the terms were changed.

I was able to repeat the KEGG results without a specialised background.
When using the appropriate background, the KEGG results were similar.

In summary the results shown in the article were described as significant, but as
authors did not perform FDR correction, they are not reliable.

Incidentally, the hypergeometric test used in clusterProfiler found many more ontologies
and KEGG pathways that were statistically significant.

## Session information

```{r,session}

sessionInfo()

```

### 2021-12-02 citation_fetching.Rmd
solved bug

@@
   N_CITES <- length(CITES[1])
-  PUBDATE <- get_pub_date(PMCID)
+  #PUBDATE <- get_pub_date(PMCID)
+  PUBDATE <- get_pub_date(PMID)
   return(c(PMCID,PMID,N_CITES,PUBDATE))
@@
 
-## Get cites  and pu
+## Get cites for small group
 
-Now to get a bunch of these.
+<!-- Now to get results for a few PMC IDs to test the functions are working OK. -->
 
-```{r,reutils3}
+```{r,test1}
 
@@
 
+## Get cites for all
+
+Now to get a bunch of these.
+
+```{r,getcites}
+
+x <- read.table("data/PMC_2019-analysis.tsv",sep="\t",header=TRUE)
+head(x)
+PMCIDS <- unique(x$Pubmed.Central.ID)
+str(PMCIDS)
+
+PMCIDS <- as.numeric(gsub("PMC","",PMCIDS))
+
+res <- lapply(X = PMCIDS, FUN = get_cites)
+mx <- do.call(rbind,res)
+mx <- apply(X = mx, MARGIN = 2, FUN = as.numeric)
+df <- as.data.frame(mx)
+colnames(df) <- c("PMCID","PMID","NumCites","Pubdate")
+
+df %>% kbl %>% kable_paper("hover", full_width = F)
+
+plot(df$Pubdate,df$NumCites)
+mylm <- lm(df$NumCites ~ df$Pubdate)
+
+mylm
+
+saveRDS(df,"cites_data.rds")
+
+```
+
 ## Session information

### 2021-12-03 citation_fetching.Rmd
using JSON to retrieve results

--- 2021-04-29 citation_fetching.Rmd
@@
 
-```{r,reutils2}
+4. Pubdate
+
+```{r,library}
 
 library("reutils")
-
-get_cites <- function(PMCID) {
-  mylink <- elink(uid=PMCID , dbFrom="pmc", dbTo = "pubmed")
-  PMID <- mylink[1]
-  CITES <- elink(uid = PMID , dbFrom = "pubmed"  , linkname = "pubmed_pubmed_citedin" )
-  N_CITES <- length(CITES[1])
-  return(c(PMCID,PMID,N_CITES))
-}
-
-PMCID=3155154
-
-get_cites(PMCID = PMCID)
+library("XML")
+library("kableExtra")
+library("anytime")
 
@@
 
-## Try the following
+Here are the functions.
 
-Now to get a bunch of these.
+You can see that there are tryCatch expressions used to manage the occasional error that are raised by 
+eutils queries which I can't reproduce nor debug. 
+So for those queries I just attempt again.
 
-```{r,reutils3}
+```{r,functions}
+
+cites <- function(PMID) {
+  tryCatch(
+    expr = {
+        CITES <- elink(uid=PMID , dbFrom="pubmed", linkname = "pubmed_pubmed_citedin" )
+    },
+    error = function(e){ 
+        CITES <- elink(uid=PMID , dbFrom="pubmed", linkname = "pubmed_pubmed_citedin" )
+    },
+    finally = {
+      return(CITES)
+    }
+)}
+
+get_cites <- function(PMID) {
+  CITES <- cites(PMID)
+  N_CITES <- length(CITES[1])
+  return(c(PMID,N_CITES))
+}
+
+# test the function
+PMID=30763910
+res <- get_cites(PMID = PMID)
+res
+
+```
+
+## Get cites for small group
+
+Now to get results for a few PMC IDs to test the functions are working OK.
+
+```{r,test1}
 
@@
 
-PMCIDS <- as.numeric(gsub("PMC","",PMCIDS))
+PMCIDS <- gsub("PMC","",PMCIDS)
+es <- esummary(PMCIDS, "pmc",retmode = "json")
 
-res <- sapply(X = PMCIDS, FUN = get_cites)
+es <- content(es, "parsed")
+es <- es[2]
+es <- es[[1]]
+es <- es[2:length(es)]
 
-res
+pmid <- unlist(lapply(es,function(x){
+  x$articleids[1,2]
+}))
+
+pubdate <- unlist(lapply(es,function(x){
+  x$pubdate
+}))
+
+df <- data.frame(PMCIDS,pmid,pubdate)
+
+ncites <- t(sapply(df[,2],get_cites))
+ncites
+
+df$ncites <- ncites[,2]
+head(df)
+df <- df[,c(1,2,4,3)]
+colnames(df) <- c("PMCID","PMID","NumCites","Pubdate")
+df$PMCID <- gsub("^","PMC",df$PMCID)
+df$Pubdate <- anydate(df$Pubdate)
+df$NumCites <- as.numeric(df$NumCites)
+str(df)
+
+df %>% kbl %>% kable_paper("hover", full_width = F)
+
+plot(df$Pubdate,df$NumCites)
+mylm <- lm(df$NumCites ~ df$Pubdate)
+
+mylm
+
 ```
@@
 Looks like it works :)
+
+Except reutils fails for articles with any special characters in there.
+
+## Get cites for all
+
+Now to get a bunch of these.
+
+```{r,getcites}
+
+x <- read.table("data/PMC_2019-analysis.tsv",sep="\t",header=TRUE)
+head(x)
+PMCIDS <- unique(x$Pubmed.Central.ID)
+
+mycitesfunc <- function(PMCIDS) {
+  PMCIDS <- gsub("PMC","",PMCIDS)
+  es <- esummary(PMCIDS, "pmc",retmode = "json")
+  es <- content(es, "parsed")
+  es <- es[2]
+  es <- es[[1]]
+  es <- es[2:length(es)]
+  pmid <- unlist(lapply(es,function(x){
+    x$articleids[1,2]
+  }))
+  pubdate <- unlist(lapply(es,function(x){
+    x$pubdate
+  }))
+  df <- data.frame(PMCIDS,pmid,pubdate)
+  ncites <- t(sapply(df[,2],get_cites))
+  df$ncites <- ncites[,2]
+  df <- df[,c(1,2,4,3)]
+  colnames(df) <- c("PMCID","PMID","NumCites","Pubdate")
+  df$PMCID <- gsub("^","PMC",df$PMCID)
+  df$Pubdate <- anydate(df$Pubdate)
+  df$NumCites <- as.numeric(df$NumCites)
+  return(df)
+}
+
+df1 <- mycitesfunc(PMCIDS[1:500])
+df2 <- mycitesfunc(PMCIDS[501:1000])
+df3 <- mycitesfunc(PMCIDS[1001:1500])
+
+df <- rbind(df1,df2,df3)
+
+df %>% kbl %>% kable_paper("hover", full_width = F)
+
+plot(df$Pubdate,df$NumCites)
+mylm <- lm(df$NumCites ~ df$Pubdate)
+
+mylm
+
+saveRDS(df,"cites_data.rds")
+
+```
 

### 2021-12-07 researcher_guidelines_for_enrichment_analysis.Rmd
guidelines

# Guidelines for reliable and reproducible functional enrichment analysis

2021-12-07

Kaumadi Wijesooriya, Sameer A Jadaan, Kaushalya L Perera, Tanuveer Kaur, Mark Ziemann

<hr>

Based upon literature and our study called "Urgent need for consistent standards in functional enrichment analysis", 
we have drafted recommended best practice guidelines for conducting functional enrichment analysis.

Links to literature articles are listed at the bottom of this document.

We have broken this list of best practices into two sections, minimum standards and gold standards.
Minimum standards are designed to avoid the most severe methodological issues while requiring minimal effort to address.
These minimum reporting standards will increase the chance that the analysis can be replicated.
Gold standards are aimed towards total transparency, code sharing and long term reproducibility.

## Minimum standards

### Statistics

1. Before starting the analysis, read the tool’s user manual.
Learn what type of statistical test is being conducted and what the null hypothesis is.
Learn whether the test being conducted is a self contained or competitive test and what the advantages/disadvantages are for each test type [1].

2. If making claims about the regulation of a gene set, then results of a statistical test must be shown including measures of significance and effect size.
The number of genes belonging to a pathway on its own is insufficient to infer enrichment [2].

3. Report which statistical test was used.
This will help long term reproducibility, for example if in future the tool is no longer available.
This is especially relevant for tools that report the results of different statistical tests.

4. Always report FDR or q-values (p-values that have been corrected for multiple tests).
This will reduce the chance of false positives when performing many tests simultaneously [2,3].
Report what method was used for p-value adjustment. Bonferroni, FDR and FWER are examples.

5. If ORA is used, it is vital to use a background list consisting of all detected genes, and report this in the methods section [2,4,5].
Avoid tools that don’t accept a background list.

6. If using ORA for gene expression analysis, be sure to conduct ORA tests separately for up- and down-regulated genes.
Combining up- and down-regulated genes into a single list can limit sensitivity [6].

### Reporting

7. Report the origin of the gene sets and the version used or date downloaded.
These databases are regularly upgraded so this is important for reproducibility.

8. Report the tool used and its software version.
As these tools are regularly updated, this will aid reproducibility.
For online tools, record the date the tool was used.

9. Report selection criteria and parameters.
If performing ORA, then the inclusion thresholds need to be mentioned.
If using GSEA or another FCS tool, parameters around ranking, gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels). 
Any parameters that vary from the defaults should be reported.

10. Functional enrichment analysis results shouldn’t be considered proof of biological plausibility nor validity of omics data analysis [5].
Such tools are best used to generate new hypotheses; informing subsequent biochemical or signaling experimentation.

## Gold standards

1. Preference FCS and pathway topology tools over ORA tools.
ORA tools are reported to have a lower sensitivity, especially when the number of "significant" genes is low (<100) or high (>3000) [7-9]

2. Include the gene profile data (including any background lists) in the supplementary data to facilitate reproducibility.
TSV or CSV formats are recommended as Excel spreadsheets can be problematic for gene names [10].

3. Scripted analysis workflows are preferred over analysis involving graphical user interfaces because they can attain a higher degree of computational reproducibility [11,12]
Similarly, free and open-source software tools are preferred over proprietary software as it lowers the bar for reproducibility.

4. Foster computational reproducibility by depositing code to a repository like GitHub or Zenodo.
Link the data to the code, so anyone can download and reproduce your results [12].

## Final comments

We drafted these best practices based on our reading of the literature (selected references below),
along with our experience in conducting enrichment analysis and 
examining hundreds of research articles presenting enrichment results.

These are not authoritative, as we are consulting the bioinformatics community for additional input and discussion.
We welcome your contribution; to make a suggestion, please raise an issue to this repository and include any supporing information including links to published articles.

## References

1. Khatri P, Sirota M, Butte AJ. Ten years of pathway analysis: current approaches and outstanding challenges. PLoS Comput Biol. 2012;8(2):e1002375. doi:10.1371/journal.pcbi.1002375

2. Tilford CA, Siemers NO. Gene set enrichment analysis. Methods Mol Biol. 2009;563: 99-121.

3. Hung JH, Yang TH, Hu Z, Weng Z, DeLisi C. Gene set enrichment analysis: performance evaluation and usage guidelines. Brief Bioinform. 2012;13: 281-291.

4. Tipney H, Hunter L. An introduction to effective use of enrichment analysis software. Hum Genomics. 2010;4: 202-206.

5. Timmons JA, Szkop KJ, Gallagher IJ. Multiple sources of bias confound functional enrichment analysis of global -omics data. Genome Biol. 2015;16: 186.

6. Hong G, Zhang W, Li H, Shen X, Guo Z. Separate enrichment analysis of pathways for up- and downregulated genes. J R Soc Interface. 2014;11: 20130950.

7. Kaspi A, Ziemann M. Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data. BMC Genomics. 2020;21: 447.

8. Maleki F, Ovens K, Hogan DJ, Kusalik AJ. Gene set analysis: Challenges, opportunities, and future research. Front Genet. 2020;11: 654.

9. Xie C, Jauhari S, Mora A. Popularity and performance of bioinformatics software: the case of gene set analysis. BMC Bioinformatics. 2021;22: 191.

10. Ziemann M, Eren Y, El-Osta A. Gene name errors are widespread in the scientific literature. Genome Biol. 2016;17(1):177.

11. Peng RD. Reproducible research in computational science. Science. 2011;334: 1226-1227.

12. Tan TW, Tong JC, Khan AM, de Silva M, Lim KS, Ranganathan S. Advancing standards for bioinformatics activities: persistence, reproducibility, disambiguation and Minimum Information About a Bioinformatics investigation (MIABi). BMC Genomics. 2010;11: S27.

### 2021-12-07 examples/example1.Rmd
tidy1

--- 2021-11-23 example1.Rmd
@@
 s1 <- subset(samplesheet,VPA==0)
+
+s1 %>% kbl(caption = "sample sheet") %>% kable_paper("hover", full_width = F)
 
@@
 dat
+saveRDS(dat,file = "ex1dat.rds")
 
-barplot(dat,ylab="jaccard metric")
+par(mar=c(5,10,3,1))
+barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
+text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
+
+pdf("images/jacbar.pdf",width=4,height=3)
+par(mar=c(5,9,3,1))
+barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.8) , main=name)
+text( x=rev(dat)+0.08 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
+dev.off()
+
+png("images/jacbar.png")
+par(mar=c(5,10,3,1))
+barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
+text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
+dev.off()
 

### 2021-12-07 examples/example6.Rmd
tidy1

--- 2021-11-23 example4.Rmd
@@
 ---
-title: "Example gene set analysis: The case of Set7 knockdown"
+title: "Example gene set analysis: The case of SARS-CoV2 infection"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
-The dataset being used is SRP038101 and we are comparing the cells expressing a set7kd shRNA construct (case) compared to the scrambled construct (control).
+The dataset being used is SRP253951 and we are comparing A549 cells with and without infection with SARS-CoV-2.
 
@@
 
-I'm using some RNA-seq data looking at the effect of Set7 knockdown on HMEC cells.
+I'm using some RNA-seq data looking at the difference in transcriptome caused by SARS-CoV-2 infection.
 
@@
 
-name="SRP096177"
+name="SRP253951"
 mdat<-getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP096177",mdat$SRP_accession),]
-samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
-samplesheet$trt<-as.factor(c(1,1,1,0,0,0))
+samplesheet <- mdat[grep("SRP253951",mdat$SRP_accession),]
+
+samplesheet <- mdat[which(mdat$SRX_accession %in% c("SRX8089264","SRX8089265","SRX8089266","SRX8089267","SRX8089268","SRX8089269")),]
+
+samplesheet$trt <- factor(c(0,0,0,1,1,1))
 s1 <- samplesheet
+
+s1 %>% kbl(caption = "sample sheet") %>% kable_paper("hover", full_width = F)
 
@@
 x<-Tx2Gene(w)
-x<-x$Tx2Gene
+x <- x$Tx2Gene
 
@@
 
-png("images/fcs_ora1.png")
+png("images/fcs_ora6.png")
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-png("images/orabg1.png")
+png("images/orabg6.png")
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-png("images/oracomb1.png")
+png("images/oracomb6.png")
 plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
@@
 
-pdf("images/fcs_ora1.pdf",width=4,height=4)
+pdf("images/fcs_ora6.pdf",width=4,height=4)
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-pdf("images/orabg1.pdf",width=4,height=4)
+pdf("images/orabg6.pdf",width=4,height=4)
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-pdf("images/oracomb1.pdf",width=4,height=4)
+pdf("images/oracomb6.pdf",width=4,height=4)
 plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
@@
 
+saveRDS(dat,file = "ex6dat.rds")
+
 ```

### 2021-12-07 examples/example7.Rmd
tidy1

--- 2021-12-07 examples/example6.Rmd
@@
 ---
-title: "Example gene set analysis: The case of SARS-CoV2 infection"
+title: "Example gene set analysis: The case of P300 knockdown"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
-The dataset being used is SRP253951 and we are comparing A549 cells with and without infection with SARS-CoV-2.
+The dataset being used is SRP068733 and we are comparing the healthy endothelial cells with a scrambled siRNA to cells treated with a p300 targeting siRNA.
 
@@
 
-I'm using some RNA-seq data looking at the difference in transcriptome caused by SARS-CoV-2 infection.
+I'm using some RNA-seq data looking at the difference in healthy endothelial cell gene expression between vehicle and C646 samples.
 
@@
 
-name="SRP253951"
+name="SRP068733"
 mdat<-getDEE2Metadata("hsapiens")
-samplesheet <- mdat[grep("SRP253951",mdat$SRP_accession),]
-
-samplesheet <- mdat[which(mdat$SRX_accession %in% c("SRX8089264","SRX8089265","SRX8089266","SRX8089267","SRX8089268","SRX8089269")),]
-
-samplesheet$trt <- factor(c(0,0,0,1,1,1))
+samplesheet <- mdat[grep("SRP068733",mdat$SRP_accession),]
+samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
+SRRvec <- c("SRR3112216","SRR3112217","SRR3112218","SRR3112219","SRR3112220","SRR3112221")
+samplesheet <- samplesheet[which(samplesheet$SRR_accession %in% SRRvec),]
+samplesheet$trt<-as.factor(c(0,0,0,1,1,1)) 
 s1 <- samplesheet
@@
 
-w<-getDEE2("hsapiens",samplesheet$SRR_accession,metadata=mdat,legacy = TRUE)
+w<-getDEE2("hsapiens",SRRvec,metadata=mdat,legacy = TRUE)
 x<-Tx2Gene(w)
-x <- x$Tx2Gene
+x<-x$Tx2Gene
 
@@
 
-png("images/fcs_ora6.png")
+png("images/fcs_ora7.png")
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-png("images/orabg6.png")
+png("images/orabg7.png")
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-png("images/oracomb6.png")
+png("images/oracomb7.png")
 plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
@@
 
-pdf("images/fcs_ora6.pdf",width=4,height=4)
+pdf("images/fcs_ora7.pdf",width=4,height=4)
 plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
@@
 
-pdf("images/orabg6.pdf",width=4,height=4)
+pdf("images/orabg7.pdf",width=4,height=4)
 plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
@@
 
-pdf("images/oracomb6.pdf",width=4,height=4)
+pdf("images/oracomb7.pdf",width=4,height=4)
 plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
@@
 
-saveRDS(dat,file = "ex6dat.rds")
+saveRDS(dat,file = "ex7dat.rds")
 

### 2021-12-07 examples/examples.Rmd
tidy1

--- 2021-11-25 examples.Rmd
@@
 means <- colMeans(exdat)
-medians <- colMedians(exdat)
-sds <- sd(exdat)
+medians <- apply(exdat,2,median)
+sds <- apply(exdat,2,sd)
 summstats <- rbind(means,medians,sds)
@@
 
-Now plot as a beeswarm.
-
-```{r,jacswarm}
-
-exdat <- rbind(exdat,"mean"=colMeans(exdat))
-
-cols <- c("#CC79A7","#56B4E9","#E69F00","#F0E442","#009E73","black")
-coll <- list(cols, cols, cols, cols, cols)
-par(mar=c(5,10,3,1))
-beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=2,pch=19, xlab = "Jaccard index")
-legend("bottomright", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","mean"), pch=19,  col=cols, cex=1,pt.cex = 1.5)
-
-pdf("images/jacswarm1.pdf",width=6,height=5)
-par(mar=c(5,9,3,1))
-beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=2,pch=19, xlab = "Jaccard index")
-legend("bottomright", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","mean"), pch=19,  col=cols, cex=1.0, pt.cex = 1.7)
-dev.off()
-
-png("images/jacswarm1.png")
-par(mar=c(5,10,3,1))
-beeswarm(list(exdat[,5],exdat[,4],exdat[,3],exdat[,2],exdat[,1]),labels = rev(colnames(exdat)),horizontal = TRUE, las=1 , pwcol = coll,cex=2,pch=19, xlab = "Jaccard index")
-legend("bottomright", legend=c("SRP128998", "SRP038101", "SRP096178","SRP038101","SRP247621","mean"), pch=19,  col=cols, cex=1,pt.cex = 1.5)
-dev.off()
-
-rbind(exdat,mysd) %>% kbl() %>% kable_paper("hover", full_width = F)
-
-```
-
 ## Session information

### 2021-12-07 analysis_code/QC_analysis2.Rmd
tidy3

--- 2021-08-31 QC_analysis2.Rmd
@@
 dev.off()
+
+pdf("images/journals2.pdf",width=4,height=4)
+par(mar=c(5,9,3,2))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Journal", xlim=c(0,180))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 15, label = res2, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 
+names(res) <- gsub("RNA array","Gene expression array",names(res))
+
 other <- sum(res[1:(nrow(res)-10)])
@@
 text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 6, label = res2, pos = 3, cex = 1, col = "black")
+dev.off()
+
+pdf("images/omics2.pdf",width=4,height=4)
+par(mar=c(5,10,3,2))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Omics type", xlim=c(0,110))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 10, label = res2, pos = 3, cex = 1, col = "black")
 dev.off()
@@
 
+pdf("images/organisms2.pdf",width=4,height=4)
+par(mar=c(5,11,3,2))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Organism", xlim=c(0,200))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 25, label = res2, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 
+pdf("images/genesetlib2.pdf",width=4,height=4)
+par(mar=c(5,11,3,2))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Gene set library", xlim=c(0,150))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 16, label = res2, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 text(y = (1:length(res)*1.2)-1.2  , x = res + 20, label = res, pos = 3, cex = 1, col = "black")
+dev.off()
+
+pdf("images/genesetvers2.pdf",width=3,height=2)
+par(mar=c(5,4,3,2))
+barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Gene set version defined", xlim=c(0,260))
+text(y = (1:length(res)*1.2)-1.2  , x = res + 25, label = res, pos = 3, cex = 1, col = "black")
 dev.off()
@@
 
+pdf("images/stattest2.pdf",width=4,height=4)
+par(mar=c(5,11,3,2))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Test used", xlim=c(0,170))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 20, label = res2, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 dev.off()
+
+pdf("images/fdr2.pdf",width=3,height=3)
+par(mar=c(5,5,3,2))
+barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "FDR correction performed", xlim=c(0,160))
+text(y = (1:length(res)*1.2)-1.0  , x = res + 15, label = res, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 
+pdf("images/app2.pdf",width=4,height=4)
+par(mar=c(5,12,3,2))
+barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "App used", xlim=c(0,80))
+text(y = (1:length(res2)*1.2)-1.2  , x = res2 + 10, label = res2, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 
+pdf("images/appvers2.pdf",width=3,height=2)
+par(mar=c(5,4,3,2))
+barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "App version defined", xlim=c(0,200))
+text(y = (1:length(res)*1.2)-1.2  , x = res + 19, label = res, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 text(y = (1:length(res)*1.2)-1.2  , x = res + 3, label = res, pos = 3, cex = 1, col = "black")
+dev.off()
+
+pdf("images/code2.pdf",width=3,height=2)
+par(mar=c(5,4,3,2))
+barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Code availability", xlim=c(0,55))
+text(y = (1:length(res)*1.2)-1.2  , x = res + 5, label = res, pos = 3, cex = 1, col = "black")
 dev.off()
@@
 
+pdf("images/bg2.pdf",width=4,height=3)
+par(mar=c(5,10,3,2))
+barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Background list specified", xlim=c(0,220))
+text(y = (1:length(res)*1.2)-1.1  , x = res + 20, label = res, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 text(y = (1:length(res)*1.2)-1.2  , x = res + 9, label = res, pos = 3, cex = 1, col = "black")
+dev.off()
+
+pdf("images/genelists2.pdf",width=3,height=2)
+par(mar=c(5,4,3,2))
+barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Gene lists provided", xlim=c(0,170))
+text(y = (1:length(res)*1.2)-1.25  , x = res + 15, label = res, pos = 3, cex = 1, col = "black")
 dev.off()
@@
 
+pdf("images/assumptions2.pdf",width=4,height=3)
+par(mar=c(5,11,3,2))
+barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Methodological flaws", xlim=c(0,230))
+text(y = (1:length(res)*1.2)-1.2  , x = res + 25, label = res, pos = 3, cex = 1, col = "black")
+dev.off()
+
 ```
@@
 text(y = (1:length(res)*1.2)-1.1  , x = res + 5, label = res, pos = 3, cex = 1, col = "black")
+dev.off()
+
+pdf("images/comb2.pdf",width=3,height=2.5)
+par(mar=c(5,5,3,2))
+barplot(res,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
+        main = "Separated or combined ORA", xlim=c(0,100))
+text(y = (1:length(res)*1.2)-1.1  , x = res + 10, label = res, pos = 3, cex = 1, col = "black")
 dev.off()
@@
 text(y = (1:nrow(params)*1.2) - 0.7 , x = params[,1]+2, label = params[,2], pos = 3, cex = 1, col = "black")
+dev.off()
+
+pdf("images/gseaparam2.pdf",width=4.5,height=3)
+par(mar=c(5,6,3,2))
+barplot(t(params),horiz=TRUE,las=1,xlim=c(0,30),legend = colnames(params),
+        main="GSEA parameter reporting", xlab="no. analyses",
+        args.legend = list(x = "topleft", bg="white",inset = c(0.05, 0.1)))
+text(y = (1:nrow(params)*1.2) - 0.8 , x = params[,1]-3, label = params[,1], pos = 3, cex = 1, col = "white")
+text(y = (1:nrow(params)*1.2) - 0.8 , x = params[,1]+2, label = params[,2], pos = 3, cex = 1, col = "black")
 dev.off()

### 2021-12-07 analysis_code/analysis.Rmd
tidy3

--- 2021-09-17 analysis.Rmd
@@
   cites <- readRDS("cites_data.rds")
-} else {
-  PMCIDS <- as.numeric(gsub("PMC","",x$Pubmed.Central.ID ))
-  mylink <- elink(uid=PMCIDS , dbFrom="pmc", dbTo = "pubmed")
-  PMID <- mylink[1]
-  #writeLines(PMID,"pubmedids.txt")
-  res <- sapply(X=PMID,FUN=function(x){
-      Sys.sleep(1)
-      CITES <- elink(uid = x , dbFrom = "pubmed"  , linkname = "pubmed_pubmed_citedin" ) 
-      CITES=length(CITES[1])
-      return(CITES)
-   })
-  mylink2 <- elink(uid=PMID , dbFrom="pubmed", dbTo = "pmc")
-  cites <- data.frame(mylink2[1],PMID,res)
-  colnames(cites) <- c("PMCID","PMID","cites")
-  saveRDS(cites,file="cites_data.rds")
 } 
-
-cites$PMCID <- gsub("^","PMC",cites$PMCID)
 cites$PMID <- gsub("^","PMID",cites$PMID)
@@
 dim(xm)
+head(xm)
 
@@
 
-mylm2 <- lm (xm$scores ~ xm$cites)
-plot(xm$cites,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
+mylm2 <- lm (xm$scores ~ xm$NumCites)
+plot(xm$NumCites,xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="Num cites",ylab="score",main="Association of analysis scores with number of citations")
@@
 
-cor.test(xm$scores , xm$cites, method="pearson")
+cor.test(xm$scores , xm$NumCites, method="pearson")
 
-cor.test(xm$scores , xm$cites, method="spearman")
+cor.test(xm$scores , xm$NumCites, method="spearman")
 
-mylm2 <- lm (xm$scores ~ log2(xm$cites))
-plot(log2(xm$cites),xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
+mylm2 <- lm (xm$scores ~ log2(xm$NumCites+1))
+plot(log2(xm$NumCites),xm$scores, col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
@@
 par(mar=c(5,5,3,1))
-plot(log2(xm$cites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
+plot(log2(xm$NumCites),xm$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="log2(citations)",ylab="score",main="Analysis scores vs. log2(citations)")
@@
 
-cor.test(xm$scores , xm$cites, method="pearson")
+cor.test(xm$scores , xm$NumCites, method="pearson")
 
-cor.test(xm$scores , xm$cites, method="spearman")
+cor.test(xm$scores , xm$NumCites, method="spearman")
 

### 2021-12-07 analysis_code/citation_fetching.Rmd
tidy3

--- 2021-12-03 citation_fetching.Rmd
@@
   CITES <- cites(PMID)
-  N_CITES <- length(CITES[1])
+  N_CITES <- length(linkset(CITES)[1]$pubmed_pubmed_citedin)
   return(c(PMID,N_CITES))
@@
 PMID=30763910
+res <- get_cites(PMID = PMID)
+res
+
+# test a paper with 0 cites
+PMID=31148949
 res <- get_cites(PMID = PMID)
@@
 
-Except reutils fails for articles with any special characters in there.
+Except reutils fails for articles with any special characters in there, looks like it is causedby improper handling of non
+ASCII characters by the XML parser.
+Solution is to use JSON format.
 
@@
 mylm <- lm(df$NumCites ~ df$Pubdate)
-
-mylm
-
+abline(mylm,col="red")
+summary(mylm)
+cor.test(df$NumCites , as.numeric(df$Pubdate))
 saveRDS(df,"cites_data.rds")

### 2021-12-07 docs/manuscript.Rmd
tidy3

--- 2021-09-04 manuscript.Rmd
@@
 
-* Gene set library used, and whether a version was reported
+* Gene set library used, and whether a version or date was reported
 
@@
 
-* Software package used, and whether a version was reported
+* Software package used, and whether a version or date was reported
 
@@
 
-* Whether gene profile was provided in the supplement
+* Whether gene profile was provided (in the supplement or main text).
 

### 2021-12-07 analysis_code/analysis1.Rmd
tidy4

--- 2021-12-07 analysis_code/QC_analysis2.Rmd
@@
 ---
-title: "PMC Survey QC Analysis"
+title: "PMC Survey Analysis Part 1"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 Here we are performing an analysis of 200 articles which is randomly selected from 1500 PMC articles.
+These articles were examined independently by two team members and the results were compared.
+Any inconsistencies were resolved.
+
+The code shown here was used to generate Figure 2 of the manuscript.
 
@@
 
-x <- read.table("data/QC-analysis2.tsv",header=TRUE,fill=TRUE,sep="\t")
+x <- read.table("../data/QC-analysis2.tsv",header=TRUE,fill=TRUE,sep="\t")
 head(x)
@@
 
-## Score
-
-Here I am proposing a scoring scheme.
-If there is information missing from the article, or basic mistakes are made, then a point is deducted.
-If the article goes over and above the basic reproducibility, then they are awarded a point.
-
-Gene set library origin not stated = -1
-
-Gene set library version not stated = -1
-
-Stat test not stated = -1
-
-No stat test conducted = -1
-
-No FDR correction conducted = -1
-
-App used not stated = -1
-
-App version not stated = -1
-
-Background list not defined = -1
-
-Inappropriate background list used = -1
-
-Code availability = +1
-
-Gene lists provided = +1
-
-```{r,score}
-
-score <- function(r){
-  r[is.na(r)] <- 0
-  SCORE=0
-  # gene set lib
-  if(r["Gene.set.library"]=="Not stated"){
-    SCORE=SCORE-1
-  }
-  # GS version
-  if(r["GS.version"]=="No"){
-    SCORE=SCORE-1
-  }
-  # stat test
-  if(r["Statistical.test.used"]=="No test"){
-    SCORE=SCORE-1
-  }
-  if(r["Statistical.test.used"]=="Not stated"){
-    SCORE=SCORE-1
-  }
-  # FDR
-  if(r["FDR.Correction"]!="Yes"){
-    SCORE=SCORE-1
-  }
-  # app used
-  if(r["App.used"]=="Not stated"){
-    SCORE=SCORE-1
-  }
-  # App version
-  if(r["App.Version"]=="No"){
-    SCORE=SCORE-1
-  }
-  # Code availability
-  if(r["Code.availability"]=="Yes"){
-    SCORE=SCORE+1
-  }
-  # Background
-  if(r["Background.gene.set"]=="Not stated"){
-    SCORE=SCORE-1
-  }
-  if(r["Background.gene.set"]=="No"){
-    SCORE=SCORE-1
-  }
-  if(r["Background.gene.set"]=="Stated, but incorrect"){
-    SCORE=SCORE-1
-  }
-  # gene list
-  if(r["Gene.lists.provided"]=="Yes"){
-    SCORE=SCORE+1
-  }
-  return(SCORE)
-}
-
-scores <- apply(X = x,MARGIN = 1, FUN = score)
-barplot(table(scores), xlab="analysis score",ylab="frequency")
-x$scores <- scores
-length(which(scores>0))
-
-mean(x$scores)
-sd(x$scores)
-median(x$scores)
-
-```
-
-## Journal rank
-
-```{r,journal rank}
-
-jmetrics <- read.table("scimagojr_2020.csv",sep=";",header=TRUE,dec = ",")
-jmetrics$Title <- toupper(jmetrics$Title)
-head(jmetrics,3)
-dim(jmetrics)
-
-nlmcat <- readLines("nlmcatalog_result.txt")
-journaltitle <- nlmcat[grep("Title\\(s\\):",nlmcat)]
-journaltitle <- gsub("Title\\(s\\): ","",journaltitle)
-journaltitle <- gsub("\\.$","",journaltitle)
-journaltitle <- toupper(journaltitle) 
-journalabbrev <- nlmcat[grep("Title Abbreviation:",nlmcat)]
-journalabbrev <- sapply(strsplit(journalabbrev,":"),"[[",2)
-journalabbrev <- gsub(" $","",journalabbrev)
-journalabbrev <- gsub("^ ","",journalabbrev)
-jdf <- data.frame(journalabbrev,journaltitle)
-
-mjdf <- merge(jdf,jmetrics,by.x="journaltitle",by.y="Title")
-
-xm <- merge(x,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)
-
-```
-
-## Analysis scores vs Journal rank
-
-```{r,scoresvsjrank1}
-
-tail(xm$Journal)
-
-mylm1 <- lm (xm$scores ~ xm$SJR)
-plot(xm$SJR,xm$scores, bty="n",
-  xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
-abline(mylm1,col="red")
-
-cor.test(xm$scores,xm$SJR,method="pearson")
-
-cor.test(xm$scores,xm$SJR,method="spearman")
-
-```
-
-## Best and worst journals
-
-Minimum of 5 analyses.
-
-```{r,goodbadugly}
-
-xmm <-xm[,c("Journal","scores")]
-tab <- table(xmm$Journal)
-tab <- tab[which(tab>3)]
-xmm <- xmm[which(xmm$Journal %in% names(tab)),]
-dim(xmm)
-jres <- aggregate(. ~ Journal, xmm ,mean)
-jsd <- aggregate(. ~ Journal, xmm ,sd)
-jres$sd <- jsd$scores
-jres <- jres[order(jres$scores),]
-dim(jres)
-head(jres)
-
-par(mar=c(5,12,3,1))
-barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
-        main = "Highest scoring journals",xlim=c(-4,0))
-grid()
-
-par(mar=c(5,12,3,1))
-barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
-        main = "Lowest scoring journals",xlim=c(-5,0))
-grid()
-
-j <- unique(xmm$Journal)
-jscores <- sapply(j,function(jj) {
- xmm[which(xmm$Journal == jj ),2]  
-})
-
-jscores <- jscores[order(unlist(lapply(jscores,mean)))]
-
-par(mar=c(5,12,3,1))
-vioplot(tail(jscores,20),horizontal = TRUE,las=1,cex.axis=0.75,main="Highest scoring journals")
-
-par(mar=c(5,12,3,1))
-vioplot(head(jscores,20),horizontal = TRUE,las=1,cex.axis=0.75,main="Lowest scoring journals")
-
-```
-
 ## Session information

### 2021-12-07 analysis_code/analysis2.Rmd
tidy4

--- 2021-12-07 analysis_code/analysis.Rmd
@@
 ---
-title: "PMC Survey Data Analysis"
+title: "PMC Survey Analysis Part 2"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
-Here we are performing an analysis of the ways enrichment analysis has been done in a sample of 1500 PMC articles from 2019.
+Here we are performing the analysis that is described in Figure 3 of the manuscript.
+This involves calculating an analysis score for a 1500 articles from 2019.
+We then identify the best/worst journals, and correlate analysis scores with journal and article level metrics.
 
@@
 write.table(x,file="TableS2.tsv",quote=FALSE,sep="\t",col.names = TRUE,row.names = FALSE)
-
-```
-
-## Journal
-
-```{r,Journal1}
-
-journal <- x$Journal
-journal_split <- strsplit(journal,", ")
-journal <- unlist(journal_split)
-res <- table(journal)
-res <- res[order(res)]
-length(res)
-res
-
-par(mar=c(1,1,1,1))
-#names(res) <- gsub("Gene expression array","RNA array",names(res))
-wordcloud(words = names(res), freq = res, min.freq = 1, 
-    max.words=200, random.order=FALSE, rot.per=0.35, 
-    colors=brewer.pal(8, "Dark2"))
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Journal", xlim=c(0,100))
-grid()
-
-other <- sum(res[1:(nrow(res)-10)])
-res2 <- c(other,tail(res,9))
-names(res2)[1] <- "Other"
-par(mar=c(5,12,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Journal", xlim=c(0,1200))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2+50, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-
-dir.create("images")
-png("images/journals1.png")
-par(mar=c(5,8,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Journal", xlim=c(0,1200))
-text(y = (1:length(res2)*1.2)-0.8  , x = res2+50, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-```
-
-## Omics type
-
-```{r,omics1}
-
-omics <- x$Omics.type
-omics_split <- strsplit(omics,", ")
-omics <- unlist(omics_split)
-res <- table(omics)
-res <- res[order(res)]
-length(res)
-res
-
-par(mar=c(1,1,1,1))
-names(res) <- gsub("Gene expression array","RNA array",names(res))
-wordcloud(words = names(res), freq = res, min.freq = 1, 
-    max.words=200, random.order=FALSE, rot.per=0.35, 
-    colors=brewer.pal(8, "Dark2"))
-
-names(res) <- gsub("RNA array","Gene expression array",names(res))
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Omics type", xlim=c(0,600))
-grid()
-
-other <- sum(res[1:(nrow(res)-10)])
-res2 <- c(other,tail(res,9))
-names(res2)[1] <- "Other"
-par(mar=c(5,12,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Omics type", xlim=c(0,600))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/omics1.png")
-par(mar=c(5,10,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Omics type", xlim=c(0,600))
-text(y = (1:length(res2)*1.2)-0.8  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-
-```
-
-## Organism
-
-```{r,org1}
-
-org <- x$Organism
-org_split <- strsplit(org,", ")
-org <- unlist(org_split)
-res <- table(org)
-res <- res[order(res)]
-length(res)
-res
-
-par(mar=c(1,1,1,1))
-names(res) <- gsub("Homo sapiens","human",names(res))
-wordcloud(words = names(res), freq = res, min.freq = 1, 
-    max.words=200, random.order=FALSE, rot.per=0.35, 
-    colors=brewer.pal(8, "Dark2"), scale=c(4,.5))
-
-par(mar=c(5,12,3,1))
-names(res) <- gsub("human","Homo sapiens",names(res))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Organism", xlim=c(0,1200))
-grid()
-
-other <- sum(res[1:(nrow(res)-10)])
-res2 <- c(other,tail(res,9))
-names(res2)[1] <- "Other"
-par(mar=c(5,12,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Organism", xlim=c(0,1300))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/organisms1.png")
-par(mar=c(5,10,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Organism", xlim=c(0,1300))
-text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-
-```
-
-## Gene set library
-
-```{r,gsl1}
-
-GSL <-x$Gene.set.library
-GSL_split <- strsplit(GSL,", ")
-GSL <- unlist(GSL_split)
-res <- table(GSL)
-res <- res[order(res)]
-length(res)
-which(names(res)=="Not stated")/sum(res)*100
-res
-
-par(mar=c(1,1,1,1))
-names(res) <- gsub("Homo sapiens","human",names(res))
-wordcloud(words = names(res), freq = res, min.freq = 1, 
-    max.words=200, random.order=FALSE, rot.per=0.35, 
-    colors=brewer.pal(8, "Dark2"))
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Gene set library", xlim=c(0,1000))
-grid()
-
-other <- sum(res[1:(nrow(res)-10)])
-res2 <- c(other,tail(res,9))
-names(res2)[1] <- "Other"
-par(mar=c(5,12,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Gene set library", xlim=c(0,1100))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/genesetlib1.png")
-par(mar=c(5,12,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Gene set library", xlim=c(0,1100))
-text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-```
-
-## Gene set version
-
-```{r,gsv1}
-
-GSV <-x$GS.version
-res <- table(GSV)
-res
-res[1]/sum(res)*100
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Gene set version defined", xlim=c(0,1600))
-text(y = (1:length(res)*1.2) - 0.6 , x = res-65, label = res, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/genesetvers1.png",width = 260, height = 260)
-par(mar=c(5,7,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Gene set version defined", xlim=c(0,1900))
-text(y = (1:length(res)*1.2) - 0.6 , x = res+200, label = res, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-```
-
-## Statistical test used
-
-```{r,test1}
-
-test <-x$Statistical.test.used
-test <- strsplit(test,", ")
-test <- unlist(test)
-res <- table(test)
-res <- res[order(res)]
-res[which(names(res)=="Not stated")] / sum(res) * 100
-res
-
-par(mar=c(1,1,1,1))
-wordcloud(words = names(res), freq = res, min.freq = 1, 
-    max.words=200, random.order=FALSE, rot.per=0.35, 
-    colors=brewer.pal(8, "Dark2"))
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Test used", xlim=c(0,1200))
-grid()
-
-other <- sum(res[1:(nrow(res)-10)])
-res2 <- c(other,tail(res,9))
-names(res2)[1] <- "Other"
-par(mar=c(5,12,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "Test used", xlim=c(0,1200))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/stattest1.png")
-par(mar=c(5,10,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Test used", xlim=c(0,1200))
-text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-```
-
-## FDR Correction
-
-```{r,fdr1}
-
-fdr <-x$FDR.Correction
-fdr <- strsplit(fdr,", ")
-fdr <- unlist(fdr)
-res <- table(fdr)
-res <- res[order(res)]
-sum(res[which(names(res)!="Yes")])/sum(res)*100
-res
-
-par(mar=c(1,1,1,1))
-wordcloud(words = names(res), freq = res, min.freq = 1, 
-    max.words=200, random.order=FALSE, rot.per=0.35, 
-    colors=brewer.pal(8, "Dark2"))
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "FDR", xlim=c(0,900))
-text(y = (1:length(res)*1.2)-0.7  , x = res-30, label = res, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/fdr1.png",width = 260, height = 260)
-par(mar=c(5,7,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "FDR", xlim=c(0,1000))
-text(y = (1:length(res)*1.2)-0.7  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-
-```
-
-## App used
-
-```{r,app1}
-
-App <-x$App.used
-App_split <- strsplit(App,", ")
-App <- unlist(App_split)
-res <- table(App)
-res <- res[order(res)]
-res[which(names(res)=="Not stated")]/sum(res)*100
-length(unique(subset(x,App.used=="Not stated")$Journal))
-unique(subset(x,App.used=="Not stated")$Journal)
-length(res)
-res
-
-par(mar=c(1,1,1,1))
-wordcloud(words = names(res), freq = res, min.freq = 1, 
-    max.words=200, random.order=FALSE, rot.per=0.35, 
-    colors=brewer.pal(8, "Dark2"))
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "App used", xlim=c(0,400))
-grid()
-
-other <- sum(res[1:(nrow(res)-10)])
-res2 <- c(other,tail(res,9))
-names(res2)[1] <- "Other"
-par(mar=c(5,12,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
-        main = "App used", xlim=c(0,500))
-text(y = (1:length(res2)*1.2)-1.2  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/app1.png")
-par(mar=c(5,12,3,1))
-barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "App used", xlim=c(0,500))
-text(y = (1:length(res2)*1.2)-0.8  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-
-length(unique(subset(x,App.used=="Not stated")$Journal))
-
-```
-
-## App version
-
-```{r,appv1}
-
-APV <-x$App.Version
-res <- table(APV)
-res
-res[1]/sum(res)*100
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "App version defined", xlim=c(0,1600))
-text(y = (1:length(res)*1.2) - 0.6 , x = res-70, label = res, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/appvers1.png",width = 260, height = 260)
-par(mar=c(5,7,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "App version defined", xlim=c(0,1600))
-text(y = (1:length(res)*1.2) - 0.6 , x = res-200, label = res, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-
-```
-
-## Code available
-
-```{r,code1}
-
-code <-x$Code.availability
-res <- table(code)
-res
-res[1]/sum(res)*100
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Code availability", xlim=c(0,330))
-text(y = (1:length(res)*1.2) - 0.6 , x = res+15, label = res, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/code1.png",width = 260, height = 260)
-par(mar=c(5,7,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Code availability", xlim=c(0,350))
-text(y = (1:length(res)*1.2) - 0.6 , x = res+30, label = res, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-
-```
-
-## Background gene set
-
-```{r,background1}
-
-BG <-x$Background.gene.set
-res <- table(BG)
-res
-sum(res[which(names(res)!="Yes")])/sum(res)*100
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Background list defined", xlim=c(0,1300))
-text(y = (1:length(res)*1.2)-0.7  , x = res+60, label = res, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/bg1.png")
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Background list defined", xlim=c(0,1400))
-text(y = (1:length(res)*1.2)-0.7  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-```
-
-## Gene lists provided
-
-```{r,genelists1}
-
-GL <-x$Gene.lists.provided
-res <- table(GL)
-res
-sum(res[which(names(res)!="Yes")])/sum(res)*100
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Gene lists provided", xlim=c(0,1200))
-text(y = (1:length(res)*1.2) - 0.6 , x = res-70, label = res, pos = 3, cex = 1, col = "black")
-grid()
-colnames(x)
-
-png("images/genelists1.png",width = 260, height = 260)
-par(mar=c(5,7,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Gene lists provided", xlim=c(0,1200))
-text(y = (1:length(res)*1.2) - 0.6 , x = res-120, label = res, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
-
-```
-
-## Assumptions violated
-
-```{r,assumptions1}
-
-ok <- nrow(subset(x,Assumptions.violated=="No"))
-ok
-bad <- nrow(subset(x,Assumptions.violated!="No"))
-bad
-ok/sum(bad,ok)*100
-
-ass <-x$Assumptions.violated
-ass <- strsplit(ass,", ")
-ass <- unlist(ass)
-res <- table(ass)
-res <- res[order(res)]
-res
-
-par(mar=c(1,1,1,1))
-wordcloud(words = names(res), freq = res, min.freq = 1, 
-    max.words=200, random.order=FALSE, rot.per=0.35, 
-    colors=brewer.pal(8, "Dark2"))
-
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Assumptions violated", xlim=c(0,1500))
-text(y = (1:length(res)*1.2)-0.75  , x = res+60, label = res, pos = 3, cex = 1, col = "black")
-grid()
-
-png("images/assumptions1.png")
-par(mar=c(5,12,3,1))
-barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
-        main = "Assumptions violated", xlim=c(0,1500))
-text(y = (1:length(res)*1.2)-0.75  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
-grid()
-dev.off()
 

### 2021-12-07 analysis_code/analysis2.Rmd
tidy5
tidy6

@@
 library("wordcloud")
-library("RColorBrewer")
-library("wordcloud2")
-library("reutils")
-library("XML")
-library("kableExtra")
 library("vioplot")
@@
 
-Here we are also using the QC2 data where possible.
-
 ```{r,exclude1}
 
-x <- read.table("data/PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
-y <- read.table("data/QC-analysis2.tsv",header=TRUE,fill=TRUE,sep="\t")
+x <- read.table("../data/PMC_2019-analysis.tsv",header=TRUE,fill=TRUE,sep="\t")
+y <- read.table("../data/QC-analysis2.tsv",header=TRUE,fill=TRUE,sep="\t")
 y <- y[,1:16]
@@
 
+dir.create("images")
 png("images/hist1.png",width=300,height=200)
@@
 
-jmetrics <- read.table("scimagojr_2020.csv",sep=";",header=TRUE,dec = ",")
+jmetrics <- read.table("../data/scimagojr_2020.csv",sep=";",header=TRUE,dec = ",")
 jmetrics$Title <- toupper(jmetrics$Title)
@@
 
-nlmcat <- readLines("nlmcatalog_result.txt")
+nlmcat <- readLines("../data/nlmcatalog_result.txt")
 journaltitle <- nlmcat[grep("Title\\(s\\):",nlmcat)]
@@
 
-if (file.exists("cites_data.rds") ) {
-  cites <- readRDS("cites_data.rds")
+if (file.exists("../data/cites_data.rds") ) {
+  cites <- readRDS("../data/cites_data.rds")
+} else {
+  stop("citation data not found. Please run 'citation_fetching.Rmd' to obtain updated citation info")
 }

### 2021-12-07 analysis_code/citation_fetching.Rmd
tidy6
one command execute

@@
 
-x <- read.table("data/PMC_2019-analysis.tsv",sep="\t",header=TRUE)
+x <- read.table("../data/PMC_2019-analysis.tsv",sep="\t",header=TRUE)
 head(x)
@@
 cor.test(df$NumCites , as.numeric(df$Pubdate))
-saveRDS(df,"cites_data.rds")
+saveRDS(df,"../data/cites_data.rds")
 

### 2021-12-07 docs/manuscript.Rmd
one command execute

--- 2021-09-04 manuscript.Rmd
@@
 ---
-title: "Guidelines for reliable and reproducible functional enrichment analysis"
+title: "Urgent need for consistent standards in functional enrichment analysis"
 author: "Kaumadi Wijesooriya, Sameer A Jadaan, Tanuveer Kaur, Kaushalya L Perera, Mark Ziemann"
@@
 
-Kaumadi Wijesooriya<sup>1</sup>, Sameer A Jadaan<sup>2</sup>, Tanuveer Kaur<sup>1</sup>, Kaushalya L Perera<sup>1</sup>, Mark Ziemann<sup>1</sup>
+Kaumadi Wijesooriya<sup>1</sup>, Sameer A Jadaan<sup>2</sup>, Tanuveer Kaur<sup>1</sup>, Kaushalya L Perera<sup>1</sup>, Mark Ziemann<sup>1</sup>*
 
@@
 
+(*) Corresponding author: m.ziemann@deakin.edu.au
+
 ## Abstract
@@
 Despite this popularity, there are concerns that these methods are being applied incorrectly and the results of some peer-reviewed publications are unreliable.
-These methodological problems include the use of inappropriate background gene lists, lack of false discovery rate correction and combining up-and down-regulated gene sets before enrichment analysis.
-An example analysis of public RNA-seq reveals that these methodological errors alter enrichment results dramatically.
+These problems include the use of inappropriate background gene lists, lack of false discovery rate correction and lack of methodological detail.
 To ascertain the frequency of these errors in the literature, we performed a screen of 186 open access research articles describing functional enrichment results.
@@
 Many studies lacked detail in the methods section about the tools and gene sets used.
-Only 15% of studies avoided major flaws, which highlights the poor state of functional enrichment rigour and reporting in the contemporary literature.
-We provide a set of minimum standards that should act as a checklist for researchers and peer-reviewers.
+Using seven independent RNA-seq datasets, we show misuse of enrichment tools changes results substantially.
+In conclusion, only 15% of functional enrichment studies examined here avoided major flaws, which highlights the poor state of functional enrichment rigour and reporting in the contemporary literature.
+This highlights the urgent need for stronger standards for enrichment analysis
+
+## Author summary
+
+Functional enrichment analysis is a commonly used technique to identify trends in large scale biological datasets.
+In biomedicine, functional enrichment analysis of gene expression data is frequently applied to identify disease and drug mechanisms.
+While enrichment tests were once primarily conducted using complicated scripting languages, web based tools are becoming more widely used.
+Users can paste a list of genes into a web-page and receive enrichment results in a matter of seconds.
+Despite the popularity of these tools, there are concerns that statistical problems and incomplete reporting could be undermining research quality.
+In this article, we conducted a systematic examination of published enrichment analyses and assessed whether (i) any statistical flaws were present and (ii) sufficient methodological detail is provided such that the study could be replicated.
+We found that lack of methodological detail and errors in statistical analysis were common, which undermines the reliability and reproducibility of these research articles.
+A set of best practices is urgently needed to raise the quality of published work.
 
@@
 
-Since the turn of the millennium, high throughput "omics" techniques like microarrays and next generation sequencing have brought with them a deluge of data.
+Since the turn of the millennium, high throughput “omics” techniques like microarrays and high throughput sequencing have brought with them a deluge of data.
 These experiments involve the measurement of thousands of genes simultaneously, and can identify hundreds or even thousands of significant associations with developmental stages or diseases.
@@
 In ORA, differentially expressed genes (DEGs) meeting a significance and/or fold change threshold are queried against curated pathways (gene sets).
-A statistical test is performed to ascertain whether the number of DEGs belonging to a particular pathway (gene set) is higher than that expected by random chance, as determined by comparison to a background gene list.
-These ORA tools can be stand-alone software packages or web services, and they use one or more statistical tests (eg: Fisher’s Exact test, hypergeometric test) (Draghici et al, 2003; Hosack et al, 2003).
+A statistical test is performed to ascertain whether the number of DEGs belonging to a particular gene set is higher than that expected by random chance, as determined by comparison to a background gene list.
+These ORA tools can be stand-alone software packages or web services, and they use one or more statistical tests (eg: Fisher’s exact test, hypergeometric test) [1,2].
 
 In the case of ORA for differential expression (eg: RNA-seq), a whole genome background is inappropriate because in any tissue, most genes are not expressed and therefore have no chance of being classified as DEGs.
-A good rule of thumb is to use a background gene list consisting of genes detected in the assay at a level where they have a chance of being classified as DEG (Tipney and Hunter 2010).
+A good rule of thumb is to use a background gene list consisting of genes detected in the assay at a level where they have a chance of being classified as DEG [3,4].
+Using the whole genome background gene list may be suitable for cases where all genes have the capacity of being detected, for example in studies of genetic variation (eg [5]).
 
 FCS tools involve giving each detected gene a differential expression score and then evaluating whether the scores are more positive or negative than expected by chance for each gene set.
-The popular Gene Set Enrichment Analysis (GSEA) tool uses permutation approaches to establish whether a gene set is significantly associated with higher or lower scores, either by permuting sample labels or by permuting genes in the differential expression profile (Subramanian et al, 2005).
+The popular Gene Set Enrichment Analysis (GSEA) tool uses permutation approaches to establish whether a gene set is significantly associated with higher or lower scores, either by permuting sample labels or by permuting genes in the differential expression profile [6].
 
-From a user's perspective, ORA is easier to conduct because it is as simple as pasting a list of gene names into a text box on a website.
-FCS tools are more difficult to use, but they are more sensitive at detecting subtle associations (Kaspi and Ziemann 2020; Maleki et al, 2020; Xie et al, 2021).
+From a user’s perspective, ORA is easier to conduct because it is as simple as pasting a list of gene names into a text box on a website.
+FCS tools are more difficult to use, but are reported to have superior sensitivity in detecting subtle associations [7-9].
 
 Although these are powerful tools to summarise complex genomics data, there are concerns that they are not being correctly used.
-Tipney and Hunter (2010) warn that inappropriate background set selection heavily influences enrichment results.
-Timmons et al (2015) highlight two cases where an inappropriate background list led to invalid enrichment results in published articles.
+Tipney and Hunter [3] warn that inappropriate background set selection heavily influences enrichment results.
+Timmons et al [10] highlight two cases where an inappropriate background list led to invalid enrichment results in published articles.
 
-The purpose of this work is to survey the frequency of methodological and reporting flaws in the literature, in particular;
-(i) inappropriate background gene set, 
-(ii) lack of p-value adjustment for multiple comparisons and 
-(iii) lack of essential methodological details.
-We use the results of this survey to inform a set of minimum standards for functional enrichment analysis.
+While the nominal p-value from the gene set test is appropriate when a single set is examined, functional enrichment analysis typically involves hundreds to thousands of parallel tests, one for each gene set in the library (eg MSigDB v7.4 library contains 32,284 sets [11]).
+False discovery rate (FDR) correction of enrichment p-values is therefore required to limit the number of false positives when performing so many concurrent tests [4,12].
+
+Lack of methodological detail can severely weaken reproducibility.
+In 2001, minimum information about a microarray experiment (MIAME) guidelines were established, resulting in higher reporting standards in publications describing microarray data [13].
+Reporting standards for computational biology have been proposed [14,15], but are not widely adopted.
+At a minimum, functional enrichment analysis reports should describe the methods used in such detail that the analysis could be replicated.
+
+The main purpose of this work is to survey the frequency of methodological and reporting flaws in the literature, in particular; (i) inappropriate background gene set, (ii) lack of p-value adjustment for multiple comparisons and (iii) lack of essential methodological details.
+Secondly, we examine several RNA-seq datasets to evaluate the effect of such issues on the functional enrichment results obtained.
+
+## Results
+
+### A screen of functional enrichment analyses in open access journals
+
+A search of PubMed Central showed 2,941 articles published in 2019 with the keywords “enrichment analysis”, “pathway analysis” or “ontology analysis”.
+From these, we initially selected 200 articles for detailed methodological analysis.
+We excluded 14 articles from the screen because they did not present any enrichment analysis.
+Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
+As some articles included more than one enrichment analysis, the dataset included 235 analyses from 186 articles; this data is available in S1 Table.
+A flow diagram of the survey is provided in Fig. 1.
+
+```{r,diagram1,echo=FALSE,fig.width=8,fig.height=4}
+
+library(tidyverse)
+library(DiagrammeR)
+library(DiagrammeRsvg)
+library(rsvg)
+suppressWarnings(dir.create("images"))
+grViz("digraph flowchart {
+      # node definitions with substituted label text
+      node [fontname = Helvetica, shape = rectangle]        
+      tab1 [label = '@@1']
+      tab2 [label = '@@2']
+      tab3 [label = '@@3']
+      tab4 [label = '@@4']
+      tab5 [label = '@@5']
+      tab6 [label = '@@6']
+      tab7 [label = '@@7']
+      tab8 [label = '@@8']
+      tab9 [label = '@@9']
+      tab10 [label = '@@10']
+      tab11 [label = '@@11']
+      tab12 [label = '@@12']
+      tab13 [label = '@@13']
+
+      # edge definitions with the node IDs
+      tab1 -> tab2  -> tab4 -> tab5 -> tab6 -> tab7;
+      tab2 -> tab3 ;
+      tab1 -> tab9 -> tab10;
+      tab5 -> tab13   ;
+      tab9 -> tab11 -> tab12 -> tab13 -> tab8 ; 
+      }
+      [1]: 'PMC keyword search (2019): 2941 results'
+      [2]: '200 PMC articles selected'
+      [3]: '14 articles excluded'
+      [4]: '235 analyses screened and cross-checked'
+      [5]: 'data entry to spreadsheet'
+      [6]: 'data analysis in R'
+      [7]: 'summary charts'
+      [8]: 'integration with journal and article metrics'
+      [9]: 'additional 1300 PMC articles selected'
+      [10]: '118 articles excluded'
+      [11]: '1395 analyses screened'
+      [12]: 'data entry to spreadsheet'
+      [13]: 'data analysis in R'
+",
+width = 700)
+
+tmp <- grViz("digraph flowchart {
+      # node definitions with substituted label text
+      node [fontname = Helvetica, shape = rectangle]        
+      tab1 [label = '@@1']
+      tab2 [label = '@@2']
+      tab3 [label = '@@3']
+      tab4 [label = '@@4']
+      tab5 [label = '@@5']
+      tab6 [label = '@@6']
+      tab7 [label = '@@7']
+      tab8 [label = '@@8']
+      tab9 [label = '@@9']
+      tab10 [label = '@@10']
+      tab11 [label = '@@11']
+      tab12 [label = '@@12']
+      tab13 [label = '@@13']
+
+      # edge definitions with the node IDs
+      tab1 -> tab2  -> tab4 -> tab5 -> tab6 -> tab7;
+      tab2 -> tab3 ;
+      tab1 -> tab9 -> tab10;
+      tab5 -> tab13   ;
+      tab9 -> tab11 -> tab12 -> tab13 -> tab8 ; 
+      }
+      [1]: 'PMC keyword search (2019): 2941 results'
+      [2]: '200 PMC articles selected'
+      [3]: '14 articles excluded'
+      [4]: '235 analyses screened and cross-checked'
+      [5]: 'data entry to spreadsheet'
+      [6]: 'data analysis in R'
+      [7]: 'summary charts'
+      [8]: 'integration with journal and article metrics'
+      [9]: 'additional 1300 PMC articles selected'
+      [10]: '118 articles excluded'
+      [11]: '1395 analyses screened'
+      [12]: 'data entry to spreadsheet'
+      [13]: 'data analysis in R'
+")
+
+tmp2 <- capture.output(rsvg_pdf(charToRaw(export_svg(tmp)),'images/flowchart.pdf'))
+
+```
+*Fig 1. A summary of the survey of functional enrichment analyses.
+The survey consists of two parts, with 200 and 1300 PMC articles considered respectively.*
+
+There were articles from 96 journals in the sample, with PLoS One, Scientific Reports, and PeerJ being the biggest contributors (S1 Fig).
+There were 18 different omics types, with gene expression array and RNA-seq being the most popular (S1 Fig).
+There were 31 different species under study, but Homo sapiens was the most common with 157 analyses (S1 Fig).
+
+![Figure S1A. Representation of journals in the sample.](../analysis_code/images/journals2.png "Journals")
+
+![Figure S1B. Representation of different omics types.](../analysis_code/images/omics2.png "Omics")
+
+![Figure S1C. Representation of different organisms.](../analysis_code/images/organisms2.png "Organisms")
+
+We recorded the use of 26 different gene set libraries, with GO and KEGG being the most frequently used (Fig 2A).
+There were 14 analyses where the gene set libraries used were not defined in the article.
+Only 18 analyses reported the version of the gene set library used (Fig 2B).
+There were 12 different statistical tests used, and the most common reported tests were Fisher’s exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Fig 2C).
+Fourteen analyses did not conduct any statistical test, as they only showed the number of genes belonging to different sets.
+Out of the 225 analyses that performed a statistical test, only 119 (53%) described correcting p-values for multiple testing (Fig 2D).
+
+There were 50 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 15 analyses (6.4%) did not state what tool was used (Fig 2E).
+The version of the software used was provided in only 68 of 235 analyses (29%) (Fig 2F).
+
+For analyses using ORA methods, we examined what background gene set was used (Fig 2G).
+This revealed that in most cases, the background list was not defined, or it was clear from the article methods section that no specific background list was used.
+In a few cases, a background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
+In only 8/197 of cases (4.1%), the appropriate background list was described in the article.
+
+Of the 47 analyses which used a scripted computer language, only 3 provided links to the code used for enrichment analysis (6.4%) (Fig 2H).
+For 93 of 235 analyses (40%), the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Fig 2I).
+
+Next, we quantified the frequency of methodological issues and reporting that would undermine the conclusions (Fig 2J).
+Lack of appropriate background was the most common issue (179 cases), followed by FDR (94 cases), then lack of data shown (13), inference without test (11 cases), and misinterpreted FDR values (2 cases).
+Only 35 analyses (15%) did not exhibit any of these major methodological issues.
+
+During this survey, we noticed some studies grouped up- and down-regulated gene lists together prior to ORA, a practice we were not expecting.
+To assess this in a systematic way we assessed differential gene expression studies to determine how common it was to combine up- and down-regulated gene lists prior to ORA (Fig 2K).
+From 109 analyses, just 28 studies performed separate analysis of up- and down-regulated gene sets.
+
+We also looked at studies performing GSEA, and whether three important analytical choices were described in the methods.
+These are (i) the gene weighting parameter, (ii) test type, ie: permuting sample labels or genes, and (iii) method used for ranking genes.
+These parameters were not stated in more than half of the analyses (Fig 2L).
+
+Taken together, these data suggest methodological and reporting deficiencies are widespread in published functional enrichment analyses.
+
+![Figure 2A. Representation of different gene set libraries.](../analysis_code/images/genesetlib2.png "Geneset libraries")
+
+![Figure 2B. Proportion of analyses reporting gene set library version information.](../analysis_code/images/genesetvers2.png "Geneset version")
+
+![Figure 2C. Representation of different statistical tests.](../analysis_code/images/stattest2.png "Geneset libraries")
+
+![Figure 2D. Proportion of analysis conducting correction of p-values for multiple comparisons.](../analysis_code/images/fdr2.png "FDR")
+
+![Figure 2E. Representation of different software tools for enrichment analysis.](../analysis_code/images/app2.png "Software used")
+
+![Figure 2F. Proportion of analyses that reported version information for software used.](../analysis_code/images/appvers2.png "Software version")
+
+![Figure 2G. Background gene set usage and reporting.](../analysis_code/images/bg2.png "Background gene list used")
+
+![Figure 2H. Proportion of scripted analyses that provided software code.](../analysis_code/images/code2.png "Code available")
+
+![Figure 2I. Proportion of analyses that provided gene profiles.](../analysis_code/images/genelists2.png "Gene lists provided")
+
+![Figure 2J. Proportion of analyses with different methodological problems.](../analysis_code/images/assumptions2.png "Assumptions violated")
+
+![Figure 2K. Whether up- and down-regulated genes were analysed separately or combined. Limited to ORA of differential expression only.](../analysis_code/images/comb2.png "Journals")
+
+![Figure 2L. Reporting of GSEA parameters.](../analysis_code/images/gseaparam2.png "Journals")
+
+### Analysis scores
+
+To make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
+We therefore analysed a further 1,300 articles, bringing the total number of analyses described up to 1630; this dataset is available in S2 Table.
+
+We then scored each analysis based on the presence or absence of methodological issues and included details.
+The median score was -4, with a mean of -3.5 and standard deviation of 1.4 (Fig 3A).
+Next, we assessed whether these analysis scores are associated with Scimago Journal Rank (SJR), a journal-level citation metric (Fig 3B).
+There was no statistically significant association of analysis score and SJR (Pearson r=0.03, p=0.23).
+
+Next, we wanted to know which journals had the highest and lowest scores.
+Only journals with five or more analyses were included.
+The best scoring journals were Transl Psychiatry, Metabolites and J Exp Clin Cancer Res (Fig 3C), while the poorest were RNA Biol, Mol Med Rep, Cancer Med and World J Gastroenterol (Fig 3D), although we note that there was a wide variation between articles of the same journal.
+
+Then we assessed for an association between mean analysis score and the SJR, for journals with 5 or more analyses (Fig 3E).
+Again, there was no association between analysis score and SJR (Pearson r=-0.008, p=0.95).
+Next, we assessed whether there was any association between analysis scores and the number of citations received by the articles.
+After log transforming the citation data, there was no association between citations and analysis scores (Pearson r=0.02, p=0.39) (Fig 3F).
+Taken together, these findings suggest that these methodological issues are not limited to lower ranking or poorly cited articles.
+
+![Figure 3A. Distribution of analysis scores.](../analysis_code/images/hist1.png "Analysis scores")
+
+![Figure 3B. Association of analysis scores with Scimago Journal Rank (SJR).](../analysis_code/images/score_sjr1.png "Analysis scores versus journal rank")
+
+![Figure 3C. Journals with highest mean analysis scores.](../analysis_code/images/score_highest1.png "Highest analysis scores")
+
+![Figure 3D. Journals with Lowest mean analysis scores.](../analysis_code/images/score_lowest1.png "Lowest analysis scores")
+
+![Figure 3E. Association of mean analysis score with SJR for journals with 5 or more analyses.](../analysis_code/images/mean_sjr1.png "Mean score vs SJR")
+
+![Figure 3F. Association of analysis scores with accrued citations.](../analysis_code/images/cites1.png "Mean score vs citations")
+
+### An example of functional enrichment analysis misuse
+
+To demonstrate whether functional enrichment analysis misuse affects results and downstream interpretation, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes (SRA accession SRP128998).
+Out of 39,297 genes in the annotation set, 15,635 were above the detection threshold (≥10 reads per sample on average).
+Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to high glucose exposure.
+
+FCS revealed 95 up and 316 down regulated Reactome gene sets (FDR<0.05).
+ORA with a background gene list consisting of detected genes revealed 55 upregulated and 238 downregulated gene sets (FDR<0.05).
+The overlap between results derived by FCS and ORA methods are shown in Fig 4A, and is reflected by a Jaccard statistic of 0.66, indicating moderate concordance between these methods.
+
+We then performed ORA using a background list consisting of all genes in the annotation set, not just those detected by the assay (indicated as ORA*), which resulted in 147 up and 530 down regulated gene sets (FDR<0.05).
+The overlap of ORA with ORA* was relatively small, with a Jaccard statistic of 0.41 (Fig 4B).
+Interestingly, 26 gene sets were simultaneously up and downregulated with this approach.
+
+Then we performed ORA analysis after combining up and downregulated genes as sometimes observed in the literature (indicated as ORA comb) with the standard background and again the whole annotation set background (indicated as ORA* comb).
+ORA comb revealed only 22 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.04 (Fig 4C).
+ORA* comb yielded 64 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.08.
+Jaccard values from the above comparisons are summarised in Fig 4D.
+
+This analysis was conducted with an additional six independent RNA-seq experiments for confirmation (Fig 4E).
+The Jaccard values obtained confirm a consistently high degree of similarity between FCS and ORA (mean Jaccard = 0.59).
+ORA with a whole genome background consistently gave mean Jaccard values <0.45 as compared to ORA with the recommended background.
+
+These findings suggest that when used properly, FCS and ORA yield similar results, but when used improperly, ORA results can be severely distorted, with less than 50% of the results being the same.
+
+![Figure 4A. Euler diagram comparing FCS and ORA methods.](../examples/images/fcs_ora1.png "FCS vs ORA")
+
+![Figure 4B.
+Euler diagram comparing ORA with standard and whole genome background.
+Whole genome background analysis is indicated with a *.](../examples/images/orabg1.png "ORA with different background sets")
+
+![Figure 4C.
+Euler diagram demonstrating the effect of combining up and downregulated genes on ORA results.
+Analysis using combined gene list is indicated by “ORA comb”, while “ORA* comb” indicates combined gene list analysis with a whole genome background.](../examples/images/oracomb1.png "ORA combining up and downregulated genes")
+
+![Figure 4D. Jaccard index values comparing enrichment analysis results when conducted in different ways for SRP128998.](../examples/images/jacbar.png "Jaccard index values")
+
+![Figure 4E. Jaccard index values comparing enrichment analysis results when conducted in different ways seven independent RNA-seq datasets.](../examples/images/jacbox1.png "Jaccard index values")
+
+## Discussion
+
+There have been several articles benchmarking the performance of functional enrichment tools, but there has been relatively little written on how enrichment tools are used (or misused).
+A recent study looked at the popularity of frequently used tools and benchmarked their performance [9], but this assumes that the tools were properly used, and methods sufficiently reported.
+In this sample of open-access research articles, there appears to be a bias toward tools that are easy to use.
+ORA tools that only require pasting lists of gene identifiers into a webpage (ie: DAVID, KOBAS and PANTHER) are collectively more popular than other solutions like GSEA (a stand-alone graphical user interface software for FCS) or any command line tool.
+This is despite ORA tools being reported to lack sensitivity when compared to FCS according to previous benchmarking studies [7-9].
+
+Failing to properly describe the background gene list was the most common methodological issue (Fig 2J).
+In the seven RNA-seq examples examined here, using the inappropriate whole genome background gave results that were on average only 32% similar to results obtained using the correct background (Fig 4E).
+The proportion of ORA studies that combined up- and down-regulated gene lists prior to analysis was high, at 70% (Fig 2K).
+As exemplified in Fig 4C, combining up- and down-regulated genes can severely reduce the number of significant gene sets (from 293 gene sets to just 22 in this example).
+This is consistent with a previous report indicating that combining up- and down-regulated gene lists is considered poor practice because it reduces sensitivity [16].
+
+Although analyses involving GSEA scored better overall, they were not free of issues.
+For example, GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Fig 2L), limiting reproducibility.
+
+We scored a total of 1630 analyses, revealing only a small fraction of articles obtained a satisfactory score of zero or higher.
+The analysis scores we generated did not correlate with journal or article metrics.
+This suggests that methodological and reporting problems are not limited to lower ranked journals, but are a more general problem.
+
+There are some limitations of this study that need to be recognised.
+Many open access articles examined here are from lower-ranked journals that might not be representative of articles in paywalled journals.
+The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in higher ranked journals contain such details in the abstract at lower rates.
+
+Further, it is difficult to ascertain whether these methodological issues invalidate the conclusions of these studies.
+We are currently working on a systematic large scale replication study to determine the reliability of these studies with corrected methods.
+
+Nevertheless, these results are a wake-up call for reproducibility and highlight the urgent need for minimum standards for functional enrichment analysis.
 
@@
 
-### An example gene set analysis
-
-To demonstrate the effect of misusing gene set analysis, a publicly available RNA-seq dataset (accession SRP128998) was downloaded from DEE2 on 13th August 2021 (Ziemann et al, 2019).
-This data consists of immortalised human hepatocytes cultured in standard or high glucose, first described by Felisbino et al (2021).
-Transcript level counts were aggregated to genes using the getDEE2 R package v1.2.0. 
-Next, genes with an average of less than 10 reads per sample were omitted from downstream analysis.
-Differential expression statistical analysis was conducted with DESeq2 v1.32.0 (Love et al, 2014) to identify genes altered by high glucose exposure.
-For gene set analysis, Reactome gene sets (Jassal et al, 2020) were downloaded (URL: https://reactome.org/download/current/ReactomePathways.gmt.zip , accessed 13th August 2021).
-FCS was performed using the mitch R package v1.4.0 with default settings, which uses a rank-ANOVA statistical test (Kaspi and Ziemann 2020).
-Genes with FDR<0.05 were used for ORA analysis using the clusterProfiler R package (v4.0.2) enricher function that implements a hypergeometric test (Yu et al, 2012).
-No fold-change threshold was used to select genes for ORA.
-For ORA, two types of background gene sets were used (i) detected genes, or (ii) all genes in the genome.
-For genes and gene sets, a false discovery rate adjusted p-value (FDR) of 0.05 was considered significant.
-Analyses were conducted in R version 4.1.0.
-
 ### Survey of enrichment analysis
 
-We collated 2,941 articles in PubMed Central published in 2019 that have keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
-We sampled 200 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.
+We collated 2,941 articles in PubMed Central published in 2019 that have keywords “enrichment analysis”, “pathway analysis” or “ontology analysis”.
+We initially sampled 200 of these articles and collected the following information from the article, searching the methods sections and other parts of the article including the supplement.
 
@@
 
-* Whether the analysis had any major flaws that might invalidate the results.
-This includes:
+* Whether the analysis had any major flaws that might invalidate the results. This includes:
 (i) background gene set not stated or inappropriate background set used,
@@
 (iv) inference without performing any statistical test, and
-(v) misinterpreting p-values by stating results were significant when FDR values indicate they aren't.
+(v) misinterpreting p-values by stating results were significant when FDR values indicate they weren’t.
 
 We excluded articles describing novel enrichment analysis techniques/tools, review articles and conference abstracts.
-Some articles presented the results of >1 enrichment analysis, so additional rows were added to accommodate them. 
+Some articles presented the results of >1 enrichment analysis, so additional rows were added to the table to accommodate them.
 These data were entered into a Google Spreadsheet by a team of five researchers.
@@
 For analyses using GSEA, we scanned the articles to identify whether key methodological steps were described, including (i) the gene weighting parameter, (ii) test type, ie: permuting sample labels or genes, and (iii) method used for ranking genes.
-
 For ORA of differential gene expression analyses, we inspected the articles to see evidence that researchers were conducting enrichment of up and down regulated genes separately or combined.
-
-Cleaned data were then loaded into R v4.1 for downstream analysis.
+Cleaned data were then loaded into R for downstream analysis.
 
@@
 
-Scimago Journal Rank (SJR) data for 2020 were downloaded from the Scimago website (https://www.scimagojr.com/journalrank.php) and used to score journals by their citation metrics.
-Using NCBI’s Eutils API, we collected the number of citations each article accrued since publication.
-We used Pearson correlation tests to assess the association with the analysis scores we generated.
-All data analyses were conducted in R v4.1 and the analysis scripts are available at GitHub (https://github.com/markziemann/SurveyEnrichmentMethods).
+SJR data for 2020 were downloaded from the Scimago website (accessed 5th August 2021) and used to score journals by their citation metrics.
+Using NCBI’s Eutils API, we collected the number of citations each article accrued since publication (accessed 3rd December 2021).
+Citation data were log2 transformed prior to regression analysis.
+In R, we used Pearson correlation tests to assess the association with the analysis scores we generated.
 
-## Results
+### Exemplifying functional enrichment analysis misuse
 
-### An example of functional enrichment analysis misuse
+To demonstrate the effect of misusing functional enrichment analysis, a publicly available RNA-seq dataset (SRA accession SRP128998) was downloaded from DEE2 on 13th August 2021 [17].
+This data consists of immortalised human hepatocytes cultured in standard (n=3) or high glucose (n=3), first described by Felisbino et al [18].
+Transcript level counts were aggregated to genes using the getDEE2 R package v1.2.0.
+Next, genes with an average of less than 10 reads per sample were omitted from downstream analysis.
+Differential expression statistical analysis was conducted with DESeq2 v1.32.0 [19] to identify genes altered by high glucose exposure.
+For gene set analysis, human Reactome gene sets [20] were downloaded in GMT format from the Reactome website (accessed 13th August 2021).
+FCS was performed using the mitch R package v1.4.0 with default settings, which uses a rank-ANOVA statistical test [7].
+Genes with FDR<0.05 were used for ORA analysis using the clusterProfiler R package (v4.0.2) enricher function that implements a hypergeometric test [21].
+No fold-change threshold was used to select genes for ORA. For ORA, two types of background gene sets were used: (i) detected genes, or (ii) all genes in the genome annotation set.
+For genes and gene sets, a false discovery rate adjusted p-value (FDR) of 0.05 was considered significant.
+Analyses were conducted in R version 4.1.0.
+To understand whether these results are consistent across other experiments, we repeated this analysis for an additional six independent published RNA-seq studies [22-27].
+Details of the contrasts examined are shown in Table 2.
 
-In order to demonstrate the effect of functional enrichment analysis misuse, we used an example RNA-seq dataset examining the effect of high glucose exposure on hepatocytes.
-Out of 39,297 genes in the whole genome, 15,635 were above the detection threshold (≥10 reads per sample on average).
-Statistical analysis revealed 3,472 differentially expressed genes with 1,560 up-regulated and 1,912 down-regulated (FDR<0.05) due to elevated glucose exposure.
+*Table 2.
+Seven independent RNA-seq experiments used for functional enrichment analysis.
+Detection threshold is an average of 10 reads per sample.
+Differentially expressed genes are defined as FDR<0.05 using DESeq2.*
 
-FCS revealed 95 up and 316 downregulated pathways (FDR<0.05).
-ORA with a background gene list consisting of detected genes revealed 55 upregulated and 238 downregulated gene sets (FDR<0.05).
-The overlap between FCS and ORA shown in Figure 1A is reflected by a Jaccard statistic of 0.66, indicating moderate concordance between these methods.
-
-We then performed ORA using the whole genome gene list (indicated as ORA\*), which resulted in 147 up and 530 down regulated gene sets (FDR<0.05),
-The overlap of ORA with ORA\* was relatively small, with a Jaccard statistic of 0.41 (Figure 1B).
-Interestingly, 26 gene sets were simultaneously up and downregulated with this approach.
-
-Then we performed ORA analysis after combining up and downregulated genes as sometimes observed in the literature (indicated as ORA comb) with the standard background and again with a whole genome background (indicated as ORA\* comb).
-ORA comb revealed only 22 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.04 (Figure 1C). 
-ORA\* comb yielded 64 differentially regulated gene sets with a Jaccard similarity to ORA of only 0.08.
-These results suggest that when used properly, FCS and ORA yield similar results, but when used improperly, ORA results can be severely distorted.
-
-![Figure 1.
-Example of enrichment analysis misuse.
-Euler diagrams show the overlap between sets of differentially regulated gene sets with different methods (FDR<0.05).
-(A) Comparison of FCS and ORA methods.
-(B) ORA with standard and whole genome background.
-Whole genome background analysis is indicated with a \* (C) Effect of combining up and downregulated genes on ORA results.
-Analysis using combined gene list is indicated by *ORA comb*, while *ORA\* comb* indicates combined gene list analysis with a whole genome background.
-](images/Figure1.png "Figure 1")
-
-### A screen of functional enrichment analyses in open access journals
-
-A search of PubMed Central showed 2,941 articles published in 2019 with the keywords "enrichment analysis", "pathway analysis" or "ontology analysis".
-From these, 200 articles were selected for methodological analysis.
-We excluded 14 articles from the screen because they did not present any enrichment analysis. 
-Those excluded articles included articles describing novel enrichment analysis techniques or tools, review articles or conference abstracts.
-As some articles included more than one enrichment analysis, the dataset included 235 analyses from 186 articles; this data is available in Table S1.
-A flow diagram of the survey is provided in Figure 2.
-
-![Figure 2. A summary of the survey of functional enrichment analyses. ](images/Figure2.png "Figure 2")
-
-There were articles from 96 journals in the sample, with *PLoS One*, *Scientific Reports*, and *PeerJ* being the biggest contributors (Figure S1A).
-There were 18 different omics types, with gene expression array and RNA-seq being the most popular (Figure S1B).
-There were 31 different species under study, but human was the most common with 157 analyses (Figure S1C).
-
-![Figure S1. General make-up of the sample. (A) Journals. (B) Types of omics analysis applied. (C) Organisms under study.](images/FigureS1.png "Figure S1")
-
-We recorded the use of 26 different gene set libraries, with GO and KEGG being the most frequently used (Figure 3A).
-There were 14 analyses where the gene set libraries used were not defined in the article.
-Only 18 analyses reported the version of the gene set library used (Figure 3B).
-There were 12 different statistical tests used, and the most common reported tests were Fisher's Exact, GSEA and hypergeometric tests; but the statistical test used was not reported for the majority of analyses (Figure 3C).
-Out of the 225 analyses that performed a statistical test, only 119 (53%) described correction of p-values for multiple testing (Figure 3D).
-
-There were 50 different tools used to perform enrichment analysis, with DAVID and GSEA being the most common; while 15 analyses (6.4%) did not state what tool was used (Figure 3E).
-The version of the software used was provided in only 68 of 235 analyses (29%) (Figure 3F).
-
-For analyses using ORA methods, we examined what background gene set was used (Figure 3G). 
-This revealed that in most cases, the background list was not defined, or it was clear from the article methods section that no background list was used.
-In a few cases, a background list was mentioned but was inappropriate, for example using a whole genome background for an assay like RNA-seq.
-In only 8/197 of cases (4.1%), the appropriate background list was described in the article.
-
-Of the 47 analyses which used a scripted computer language, only 3 provided links to the code used for enrichment analysis (6.4%) (Figure 3H).
-For 93 of 235 analyses (40%), the corresponding gene lists/profiles were provided either in the supplement or in the article itself (Figure 3I).
-
-Next, we quantified the frequency of methodological issues and reporting that would undermine the conclusions (Figure 3J).
-Lack of appropriate background was the most common issue (179 cases), followed by FDR (94 cases), then lack of data shown (13), inference without test (11 cases), and misinterpreted FDR values (2 cases).
-Only 35 analyses (15%) did not exhibit any of these major methodological issues.
-
-During this survey, we noticed some studies grouped up- and down-regulated gene lists together prior to ORA, a practice we were not expecting.
-To assess this in a systematic way we assessed differential gene expression studies to determine how common it was to combine up- and down-regulated gene lists prior to ORA (Figure 3K).
-From 109 analyses, just 28 studies performed separate analysis of up- and down-regulated gene sets.
-
-We also looked at studies performing GSEA, and whether three important analytical choices were described in the methods.
-These are (i) the gene weighting parameter, (ii) test type, ie: permuting sample labels or genes, and (iii) method used for ranking genes.
-These parameters were not stated in more than half of the analyses (Figure 3L).
-
-Taken together, these data suggest methodological and reporting deficiencies are widespread in published functional enrichment analyses.
-
-![Figure 3.
-Findings from the survey of published enrichment analyses.
-(A) Representation of different gene set libraries.
-(B) Proportion of analyses reporting gene set library version information.
-(C) Representation of different statistical tests.
-(D) Proportion of analysis conducting correction of p-values for multiple comparisons.
-(E) Representation of different software tools for enrichment analysis.
-(F) Proportion of analyses that reported version information for software used.
-(G) Background gene set usage and reporting.
-(H) Proportion of scripted analyses that provided software code.
-(I) Proportion of analyses that provided gene profiles.
-(J)  Proportion of analyses with different methodological problems.
-(K) Whether up- and down-regulated genes were analysed separately or combined.
-Limited to ORA of differential expression only.
-(L) Reporting of GSEA parameters.](images/Figure3.png "Figure 3")
-
-### Analysis scores
-
-To make an accurate assessment of association between analysis quality and bibliometrics, we required a larger sample of articles.
-We therefore analysed a further 1,300 articles, bringing the total number of analyses described up to 1630; this dataset is available in Table S2.
-
-We then scored each analysis based on the presence or absence of methodological issues and included details.
-The median score was -4, with a mean of -3.5 and standard deviation of 1.4 (Figure 4A).
-Next, we assessed whether these analysis scores associate with journal citation metrics (Figure 4B).
-There was no statistically significant association of analysis score and SJR (Pearson r=0.03, p=0.23).
-
-Next, we wanted to know which journals had the highest and lowest scores.
-Only journals with five or more analyses were included.
-The best scoring journals were *Transl Psychiatry*, *Metabolites* and *J Exp Clin Cancer Res* (Figure 4C), while the poorest were *RNA Biol*, *Mol Med Rep*, *Cancer Med* and *World J Gastroenterol* (Figure 4D), although we note that there was a wide variation between articles of the same journal.
-
-Then we assessed for an association between mean analysis score and the SJR, for journals with 5 or more analyses (Figure 4E).
-Again, there was no association between analysis score and SJR (Pearson r=-0.008, p=0.95).
-Next, we assessed whether there was any association between analysis scores and the number of citations received by the articles.
-After log transforming the citation data, there was no association between citations and analysis scores (Pearson r=0.002, p=0.95) (Figure 4F).
-Taken together, these findings suggest that these methodological issues are not limited to lower ranking or poorly cited articles.
-
-![Figure 4.
-Comparison of Analysis scores to bibliometrics.
-(A) Distribution of analysis scores.
-(B) Association of analysis scores with Scimago Journal Rank (SJR).
-(C) Journals with highest analysis scores.
-(D) Journals with lowest analysis scores.
-(E) Association of mean analysis score with SJR for journals with 5 or more analyses.
-(F) Association of analysis scores with accrued citations.](images/Figure4.png "Figure 4")
-
-## Discussion
-
-While there have been several articles benchmarking the performance of functional enrichment tools, there has been relatively little written on how existing tools are used (or misused).
-A recent study looked at the popularity of frequently used tools and benchmarked their performance (Xie et al, 2021), but this assumes that the tools were properly used, and methods sufficiently reported.
-In this sample of open access research articles, there appears to be a bias toward tools that are easy to use.
-ORA tools that only require pasting lists of gene identifiers into a webpage (ie: DAVID, KOBAS and PANTHER) are collectively more popular than other solutions like GSEA (a stand-alone graphical user interface software for FCS) or any command line tool.
-This is despite the fact that ORA tools perform worse than FCS and new generation pathway topology tools (Kaspi & Ziemann 2020; Maleki et al, 2020; Xie et al, 2021).
-
-Failing to properly describe the background gene list was the most common methodological issue (Figure 3J).
-This error can lead to substantially distorted results (Figure 1B).
-In the example here, using the inappropriate whole genome background resulted in 2.3-fold more differentially regulated gene sets.
-Thus, if there are few or no significant results using a correct background list, there may be a temptation to use an inappropriate background to avoid null results.
-
-The proportion of ORA studies that combined up- and down-regulated gene list prior to analysis was also surprisingly high, at 70% (Figure 3K).
-As exemplified in Figure 1C, combining up- and down-regulated genes can severely reduce the number of significant  gene sets (from 293 pathways to just 22 in this example).
-This is consistent with a previous report indicating that combining up- and down-regulated gene lists is considered poor practice because it reduces sensitivity (Hong et al, 2014).
-These results suggest that a large fraction of published works suffer from a high rate of false negatives, in effect missing a large fraction of truly differentially regulated gene sets.
-This is exacerbated by using an inappropriate background gene list, which leads to a high rate of false positives.
-
-Although analyses involving GSEA scored better overall, they were not free of issues.
-For example, GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles (Figure 3L), limiting reproducibility.
-
-We scored a total of 1630 analyses, revealing only a small fraction of articles obtained a satisfactory score of zero or higher.
-The analysis scores we generated did not correlate with journal or article metrics.
-This suggests that methodological and reporting problems are not limited to lower ranked journals, but are a more general problem.
-
-Still there are some limitations that need to be recognised.
-Many open access articles examined here are from lower-ranked journals, that might not be representative of articles in paywalled journals.
-The articles included in this study contained keywords related to functional enrichment in the abstract, and it is plausible that articles in higher ranked journals contain such details in the abstract at lower rates.
-
-Nevertheless, these results are a wake-up call for reproducibility and highlight the urgent need for minimum standards for functional enrichment analysis. 
-To address this, we provide a set of minimum standards for functional enrichment analysis, as well as additional recommendations to achieve gold standard reliability and reproducibility (Box 1).
-
-<hr>
-
-### Essential minimum standards for functional enrichment analysis
-
-The following guidelines are selected to avoid the most severe methodological issues while requiring minimal effort to address.
-
-1. Before starting the analysis, read the tool's user manual.
-
-2. Report the origin of the gene sets and the version used or date downloaded.
-These databases are regularly upgraded so this is important for reproducibility.
-
-3. Report the tool used and its software version.
-As these tools are regularly updated, this will aid reproducibility.
-For online tools, record the date the tool was used.
-
-4. If making claims about the regulation of a gene set, then results of a statistical test must be shown including measures of significance and effect size.
-The number of genes belonging to a pathway on its own is insufficient to infer enrichment.
-
-5. Report which statistical test was used.
-This will help long term reproducibility, for example if in future the tool is no longer available.
-This is especially relevant for tools that report the results of different statistical tests.
-
-6. Always report FDR or q-values (p-values that have been corrected for multiple tests).
-This will reduce the chance of false positives when performing many tests simultaneously (Hung et al, 2012).
-Report what method was used for p-value adjustment.
-Bonferonni, FDR and FWER are examples.
-
-7. If ORA is used, it is vital to use a background list consisting of all detected genes, and report this in the methods section.
-Avoid tools that don't accept a background list.
-
-8. Report selection criteria and parameters.
-If performing ORA, then the inclusion thresholds need to be mentioned.
-If using GSEA or another FCS tool, parameters around ranking, gene weighting and type of test need to be disclosed (eg: permuting sample labels or gene labels).
-Any parameters that vary from the defaults should be reported.
-
-9. If using ORA for gene expression analysis, be sure to conduct ORA tests separately for up- and down-regulated genes before analysis.
-Combining up- and down-regulated genes into a single list will limit sensitivity.
-
-10. Functional enrichment analysis results shouldn't be considered proof of biological plausibility nor validity of omics data analysis (Timmons et al, 2015). 
-Such tools are best used to generate new hypotheses; informing subsequent biochemical or signaling experimentation.
-
-### Additional recommendations for enhanced reliability and reproducibility
-
-The following guidelines are directed at going over and above the minimum standards towards a gold standard of reliability and reproducibility.
-
-11. Preference FCS and pathway topology tools over ORA tools.
-ORA tools have a lower sensitivity. 
-In practice, they are rarely used properly.
-
-12. Include the gene profile data (including any background lists) in the supplementary data in TSV or CSV formats.
-Excel spreadsheets are not recommended for genomics work (Ziemann et al, 2016)
-
-13.
-Scripted analysis workflows are preferred over analysis involving graphical user interfaces because they can attain a higher degree of computational reproducibility, so consider investing in upgrading your analysis.
-Similarly, free and open-source software tools are preferred over proprietary software.
-
-14. If using a scripted analysis, provide code by depositing it to a repository like GitHub or Zenodo.
-Link the data to the code, so anyone can download and reproduce your analysis (Peng 2011).
-
-<hr>
+| SRA accession and citation | Control datasets | Case datasets | Genes detected | Genes differentially expressed | 
+| --- | --- | --- | --- | --- | 
+| SRP128998 [18] | GSM2932797 GSM2932798 GSM2932799 | GSM2932791 GSM2932792 GSM2932793 | 15635 | 3472 | 
+| SRP038101 [22] | GSM1329862 GSM1329863 GSM1329864 | GSM1329859 GSM1329860 GSM1329861 | 13926 | 3589 |
+| SRP037718 [23] | GSM1326472 GSM1326473 GSM1326474 | GSM1326469 GSM1326470 GSM1326471 | 15477 | 9488 |
+| SRP096177 [24] | GSM2448985 GSM2448986 GSM2448987 | GSM2448982 GSM2448983 GSM2448984 | 15607 | 5150 |
+| SRP247621 [25] | GSM4300737 GSM4300738 GSM4300739 | GSM4300731 GSM4300732 GSM4300733 | 14288 | 230 |
+| SRP253951 [26] | GSM4462339 GSM4462340 GSM4462341 | GSM4462336 GSM4462337 GSM4462338 | 15182 | 8588 |
+| SRP068733 [27] | GSM2044431 GSM2044432 GSM2044433 | GSM2044428 GSM2044429 GSM2044430 | 14255 | 7365 |
 
@@
 
-[**Table S1**](http://ziemann-lab.net/public/kaumadi/TableS1.tsv "A survey of 186 articles describing functional enrichment results.") A survey of 186 articles describing functional enrichment results.
+[**S1 Table**](http://ziemann-lab.net/public/kaumadi/TableS1.tsv "A survey of 186 articles describing functional enrichment results.") A survey of 186 articles describing functional enrichment results.
 
-[**Table S2**](http://ziemann-lab.net/public/kaumadi/TableS2.tsv "A survey of 1365 articles describing functional enrichment results.") A survey of 1365 articles describing functional enrichment results.
+[**S2 Table**](http://ziemann-lab.net/public/kaumadi/TableS2.tsv "A survey of 1365 articles describing functional enrichment results.") A survey of 1365 articles describing functional enrichment results.
+
+Code and data supporting this study are deposited to GitHub (https://github.com/markziemann/SurveyEnrichmentMethods).
 
@@
 
-Draghici, S., Khatri, P., Martins, R. P., Ostermeier, G. C., & Krawetz, S. A. (2003). Global functional profiling of gene expression. Genomics, 81(2), 98–104.
+1. Hosack DA, Dennis G Jr, Sherman BT, Lane HC, Lempicki RA. Identifying biological themes within lists of genes with EASE. Genome Biol. 2003;4: R70.
 
-Felisbino, M.
-B., Ziemann, M., Khurana, I., Okabe, J., Al-Hasani, K., Maxwell, S., … El-Osta, A.
-(2021).
+2. Draghici S, Khatri P, Martins RP, Ostermeier GC, Krawetz SA. Global functional profiling of gene expression. Genomics. 2003;81: 98-104.
+
+3. Tipney H, Hunter L. An introduction to effective use of enrichment analysis software. Hum Genomics. 2010;4: 202-206.
+
+4. Tilford CA, Siemers NO. Gene set enrichment analysis. Methods Mol Biol. 2009;563: 99-121.
+
+5.
+Cirillo E, Kutmon M, Gonzalez Hernandez M, Hooimeijer T, Adriaens ME, Eijssen LMT, et al.
+From SNPs to pathways: Biological interpretation of type 2 diabetes (T2DM) genome wide association study (GWAS) results.
+PLoS One.
+2018;13: e0193515.
+
+6.
+Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et al.
+Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles.
+Proc Natl Acad Sci U S A.
+2005;102: 15545-15550.
+
+7. Kaspi A, Ziemann M. Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data. BMC Genomics. 2020;21: 447.
+
+8. Maleki F, Ovens K, Hogan DJ, Kusalik AJ. Gene set analysis: Challenges, opportunities, and future research. Front Genet. 2020;11: 654.
+
+9. Xie C, Jauhari S, Mora A. Popularity and performance of bioinformatics software: the case of gene set analysis. BMC Bioinformatics. 2021;22: 191.
+
+10. Timmons JA, Szkop KJ, Gallagher IJ. Multiple sources of bias confound functional enrichment analysis of global -omics data. Genome Biol. 2015;16: 186.
+
+11. Liberzon A, Birger C, Thorvaldsdóttir H, Ghandi M, Mesirov JP, Tamayo P. The Molecular Signatures Database (MSigDB) hallmark gene set collection. Cell Syst. 2015;1:417-425.
+
+12. Hung JH, Yang TH, Hu Z, Weng Z, DeLisi C. Gene set enrichment analysis: performance evaluation and usage guidelines. Brief Bioinform. 2012;13: 281-291.
+
+13.
+Brazma A, Hingamp P, Quackenbush J, Sherlock G, Spellman P, Stoeckert C, et al.
+Minimum information about a microarray experiment (MIAME)-toward standards for microarray data.
+Nat Genet.
+2001;29:365-371.
+
+14.
+Tan TW, Tong JC, Khan AM, de Silva M, Lim KS, Ranganathan S.
+Advancing standards for bioinformatics activities: persistence, reproducibility, disambiguation and Minimum Information About a Bioinformatics investigation (MIABi).
+BMC Genomics.
+2010;11: S27.
+
+15. Peng RD. Reproducible research in computational science. Science. 2011;334: 1226-1227.
+
+16. Hong G, Zhang W, Li H, Shen X, Guo Z. Separate enrichment analysis of pathways for up- and downregulated genes. J R Soc Interface. 2014;11: 20130950.
+
+17. Ziemann M, Kaspi A, El-Osta A. Digital expression explorer 2: a repository of uniformly processed RNA sequencing data. Gigascience. 2019;8: giz022.
+
+18.
+Felisbino MB, Ziemann M, Khurana I, Okabe J, Al-Hasani K, Maxwell S, et al.
 Valproic acid influences the expression of genes implicated with hyperglycaemia-induced complement and coagulation pathways.
-Scientific Reports, 11(1), 2163.
+Sci Rep.
+2021;11: 2163.
 
-Hong, G., Zhang, W., Li, H., Shen, X., & Guo, Z. (2014). Separate enrichment analysis of pathways for up- and downregulated genes. Journal of the Royal Society, Interface, 11(92), 20130950.
+19. Love MI, Huber W, Anders S. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biol. 2014;15: 550.
 
-Hosack, D. A., Dennis, G., Jr, Sherman, B. T., Lane, H. C., & Lempicki, R. A. (2003). Identifying biological themes within lists of genes with EASE. Genome Biology, 4(10), R70.
+20. Jassal B, Matthews L, Viteri G, Gong C, Lorente P, Fabregat A, et al. The reactome pathway knowledgebase. Nucleic Acids Res. 2020;48: D498–D503.
 
-Huang, D. W., Sherman, B. T., & Lempicki, R. A. (2009). Bioinformatics enrichment tools: paths toward the comprehensive functional analysis of large gene lists. Nucleic Acids Research, 37(1), 1–13.
+21. Yu G, Wang L-G, Han Y, He Q-Y. clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS. 2012;16: 284-287.
 
-Hung, J.-H., Yang, T.-H., Hu, Z., Weng, Z., & DeLisi, C. (2012). Gene set enrichment analysis: performance evaluation and usage guidelines. Briefings in Bioinformatics, 13(3), 281–291.
+22.
+Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al.
+DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML.
+Genome Biol.
+2014;15: 406.
 
-Jassal, B., Matthews, L., Viteri, G., Gong, C., Lorente, P., Fabregat, A., … D’Eustachio, P. (2020). The reactome pathway knowledgebase. Nucleic Acids Research, 48(D1), D498–D503.
+23. Rafehi H, Balcerczyk A, Lunke S, Kaspi A, Ziemann M, Kn H, et al. Vascular histone deacetylation by pharmacological HDAC inhibition. Genome Res. 2014;24: 1271–1284.
 
-Kaspi, A., & Ziemann, M. (2020). Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data. BMC Genomics, 21(1), 447.
+24. Keating ST, Ziemann M, Okabe J, Khan AW, Balcerczyk A, El-Osta A. Deep sequencing reveals novel Set7 networks. Cell Mol Life Sci. 2014;71: 4471–486.
 
-Love, M. I., Huber, W., & Anders, S. (2014). Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biology, 15(12), 550.
+25.
+Lopez Sanchez MIG, Van Bergen NJ, Kearns LS, Ziemann M, Liang H, Hewitt AW, et al.
+OXPHOS bioenergetic compensation does not explain disease penetrance in Leber hereditary optic neuropathy.
+Mitochondrion.
+2020;54: 113–121.
 
-Maleki, F., Ovens, K., Hogan, D. J., & Kusalik, A. J. (2020). Gene set analysis: Challenges, opportunities, and future research. Frontiers in Genetics, 11, 654.
+26. Blanco-Melo D, Nilsson-Payant BE, Liu WC, Uhl S, Hoagland D, Møller R, et al. Imbalanced Host Response to SARS-CoV-2 Drives Development of COVID-19. Cell. 2020;181: 1036–1045.e9.
 
-Peng, R. D. (2011). Reproducible research in computational science. Science (New York, N.Y.), 334(6060), 1226–1227.
-
-Subramanian, A., Tamayo, P., Mootha, V.
-K., Mukherjee, S., Ebert, B.
-L., Gillette, M.
-A., … Mesirov, J.
-P.
-(2005).
-Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles.
-Proceedings of the National Academy of Sciences of the United States of America, 102(43), 15545–15550.
-
-Timmons, J. A., Szkop, K. J., & Gallagher, I. J. (2015). Multiple sources of bias confound functional enrichment analysis of global -omics data. Genome Biology, 16(1), 186.
-
-Tipney, H., & Hunter, L. (2010). An introduction to effective use of enrichment analysis software. Human Genomics, 4(3), 202–206.
-
-Xie, C., Jauhari, S., & Mora, A. (2021). Popularity and performance of bioinformatics software: the case of gene set analysis. BMC Bioinformatics, 22(1), 191.
-
-Yu, G., Wang, L.-G., Han, Y., & He, Q.-Y. (2012). clusterProfiler: an R package for comparing biological themes among gene clusters. Omics: A Journal of Integrative Biology, 16(5), 284–287.
-
-Ziemann, M., Eren, Y., & El-Osta, A. (2016). Gene name errors are widespread in the scientific literature. Genome Biology, 17(1), 177.
-
-Ziemann, M., Kaspi, A., & El-Osta, A. (2019). Digital expression explorer 2: a repository of uniformly processed RNA sequencing data. GigaScience, 8(4). doi:10.1093/gigascience/giz022
+27.
+Rafehi H, Kaspi A, Ziemann M, Okabe J, Karagiannis TC, El-Osta A.
+Systems approach to the pharmacological actions of HDAC inhibitors reveals EP300 activities and convergent mechanisms of regulation in diabetes.
+Epigenetics.
+2017;12: 991–1003.

### 2021-12-07 examples/examples.Rmd
one command execute

@@
 ---
-title: "Example gene set analysis: bringing it together"
+title: "Gene set analysis misuse examples"
 author: "Mark Ziemann & Kaumadi Wijesooriya"
@@
 
-Previously we analysed five datasets below and looked at the differences between FCS and ORA methods for gene set enrichment.
+This script is designed to generate the results of the manuscript figure 4.
+
+Panels A-D are generated with "example1.Rmd" and "example1-7.Rmd" are required 
+for panel E.
+
+Here are the datasets considered.
 
@@
 
-```{r,begin}
+```{r,begin0}
 
@@
 
+## Execute the examples
+
+Here I'm running the scripts to analyse each of the example datasets shown in the above
+table.
+
+```{r,execute0}
+
+rmarkdown::render("example1.Rmd")
+rmarkdown::render("example2.Rmd")
+rmarkdown::render("example3.Rmd")
+rmarkdown::render("example4.Rmd")
+rmarkdown::render("example5.Rmd")
+rmarkdown::render("example6.Rmd")
+rmarkdown::render("example7.Rmd")
+
+```
+
 ## Summarise
@@
 
-```{r,getdata,fig.width=7,fig.height=5}
+```{r,getdata0,fig.width=7,fig.height=5}
 
@@
 
-```{r,summary_stats}
+```{r,summary_stats0}
 
@@
 
-```{r,jacbox}
+```{r,jacbox0}
 
@@
 
-```{r,jacbar}
+```{r,jacbar0}
 
@@
 
-```{r,session}
+```{r,session0}
 

### 2021-12-07 reproduce_analysis_and_manuscript.Rmd
one command execute

---
title: "Reproduce the analysis and regenerate the manuscript"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichmentMethods

## Intro

This script is designed to enable the complete analysis described in the article 
to be replicated with a single command.

By knitting this Rmd, it will reanalyse the data and regenerate all figures of
the manuscript.

First of all, it is a good idea to check whether all of the requisite 
dependancies are installed.

Ubuntu users may need to install `librsvg2-dev` and `libv8-dev` to get `DiagrammeR` to work.

```{r,libraries}

# for figure 2 and 3
library("wordcloud")
library("vioplot")
library("reutils")
library("XML")
library("kableExtra")
library("anytime")

# for figure 4
library("getDEE2") 
library("DESeq2")
library("clusterProfiler")
library("mitch")
library("kableExtra")
library("eulerr")
library("kableExtra")
library("beeswarm")

# for figure 1 / manuscript draft
library(rmdformats)
library(tidyverse)
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

```

## Execute the scripts

Here will re-run all of the analysis and re-build the manuscript draft.

```{r,run_all}

# Figure 2
rmarkdown::render("analysis_code/analysis1.Rmd")

# Figure 3
rmarkdown::render("analysis_code/analysis2.Rmd")

# uncomment the below line in order to update the number of accumulated
# citations for each article considered
#rmarkdown::render("analysis_code/citation_fetching.Rmd")

# Figure 4
rmarkdown::render("examples/examples.Rmd")

# Figure 1 and assemble a manuscript draft
rmarkdown::render("docs/manuscript.Rmd")

```

### 2021-12-07 reproduce_analysis_and_manuscript.Rmd
finishing touches

@@
 library("mitch")
-library("kableExtra")
 library("eulerr")
-library("kableExtra")
 library("beeswarm")

### 2021-12-07 docs/manuscript.Rmd
finishing touches

@@
 
-library(tidyverse)
-library(DiagrammeR)
-library(DiagrammeRsvg)
-library(rsvg)
+library("DiagrammeR")
+library("DiagrammeRsvg")
+library("rsvg")
 suppressWarnings(dir.create("images"))

### 2022-01-12 analysis_code/analysis2.Rmd
fixed minor issue with journal name mismatches betwen scimago and nlm

@@
 library("vioplot")
+library("ggplot2")
+library("ggExtra")
+library("europepmc")
 
@@
 
+There is a problem that the journal names from NLM/PUBMED CENTRAL do not exactly match those given by SciMago.
+This results in some journals being dropped from the SJR correlation analysis.
+To avoid this, I will use the ISSN numbers which should be unique and non-ambiguous identifiers.
+This is obtained with a query of Europe PMC with the europepmc package.
+
 ```{r,journal rank}
@@
 
-nlmcat <- readLines("../data/nlmcatalog_result.txt")
-journaltitle <- nlmcat[grep("Title\\(s\\):",nlmcat)]
-journaltitle <- gsub("Title\\(s\\): ","",journaltitle)
-journaltitle <- gsub("\\.$","",journaltitle)
-journaltitle <- toupper(journaltitle) 
-journalabbrev <- nlmcat[grep("Title Abbreviation:",nlmcat)]
-journalabbrev <- sapply(strsplit(journalabbrev,":"),"[[",2)
-journalabbrev <- gsub(" $","",journalabbrev)
-journalabbrev <- gsub("^ ","",journalabbrev)
-jdf <- data.frame(journalabbrev,journaltitle)
+getissn <- function(pmc){
+  Sys.sleep(1)
+  pmcres <- europepmc::epmc_search(pmc)
+  pmcissn <- pmcres$journalIssn
+  pmcissn <- gsub("-","",pmcissn)
+  return(pmcissn)
+}
 
-mjdf <- merge(jdf,jmetrics,by.x="journaltitle",by.y="Title")
+issn <- sapply(x$Pubmed.Central.ID,getissn)
+issn2 <- lapply(issn,function(x) { unlist(strsplit(x,"\\; ")) } )
+l <- lapply(issn2,length)
+l <- unlist(l)
+lnames <- names(l)
+pmcnames <- unlist(sapply(1:length(l), function(i) { rep(lnames[i],l[i]) }  ))
+issn2 <- unname(unlist(issn2))
+issn2 <- gsub("-","",issn2)
+pmcissn <- data.frame(pmcnames,issn2)
+xissn <- merge(x,pmcissn,by.x="Pubmed.Central.ID",by.y="pmcnames")
 
-xm <- merge(x,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)
+xm <- merge(xissn,jmetrics,by.x="issn2",by.y="Issn")
 
@@
 ## Analysis scores vs Journal rank
+
+Here I will be making a scatter plot of the analysis scores vs the SJR.
+
+I will also attempt to make the same scatterplot with marginal histograms according to (https://stackoverflow.com/questions/8545035/scatterplot-with-marginal-histograms-in-ggplot2)
 
@@
 
+df <- data.frame("SJR" = xm$SJR, "score" = xm$scores )
+p <- ggplot(df, aes(SJR, score)) + 
+  geom_point(alpha=0.1, size = 3) + 
+  stat_smooth(method = lm,fullrange=TRUE) +
+  theme_classic()
+ggExtra::ggMarginal(p, type = "histogram")
+
+pdf("images/score_sjr2.pdf",width=4,height=3)
+p <- ggplot(df, aes(SJR, score)) + 
+  geom_point(alpha=0.1, size = 3) + 
+  stat_smooth(method = lm,fullrange=TRUE) +
+  theme_classic()
+ggExtra::ggMarginal(p, type = "histogram")
+dev.off()
+
 ```
@@
 
-xmm <-xm[,c("Journal","scores")]
-tab <- table(xmm$Journal)
+x2 <-x[,c("Journal","scores")]
+tab <- table(x2$Journal)
 tab <- tab[which(tab>=5)]
-xmm <- xmm[which(xmm$Journal %in% names(tab)),]
-dim(xmm)
-jres <- aggregate(. ~ Journal, xmm ,mean)
-jsd <- aggregate(. ~ Journal, xmm ,sd)
+x2 <- x2[which(x2$Journal %in% names(tab)),]
+dim(x2)
+jres <- aggregate(scores ~ Journal, x2 ,mean)
+jsd <- aggregate(scores ~ Journal, x2 ,sd)
 jres$sd <- jsd$scores
@@
 
-j <- unique(xmm$Journal)
+j <- unique(x2$Journal)
 jscores <- sapply(j,function(jj) {
- xmm[which(xmm$Journal == jj ),2]  
+ x2[which(x2$Journal == jj ),2]  
 })
@@
 
-jjres <- merge(jres,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)
+x2 <-xm[,c("Journal","SJR","scores")]
+tab <- table(x2$Journal)
+tab <- tab[which(tab>=5)]
+x2 <- x2[which(x2$Journal %in% names(tab)),]
+dim(x2)
+jres <- aggregate(scores ~ Journal, x2 ,mean)
+jsjr <- aggregate(SJR ~ Journal, x2 ,mean)
+jres$SJR <- jsjr$SJR
+jres <- jres[order(jres$scores),]
+dim(jres)
+head(jres)
 
-mylm2 <- lm (jjres$scores ~ jjres$SJR)
-plot(jjres$SJR,jjres$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
+mylm2 <- lm (jres$scores ~ jres$SJR)
+plot(jres$SJR,jres$scores,col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="SJR",ylab="mean score",main="Mean analysis score vs. journal metrics")
@@
 par(mar=c(5,5,3,1))
-plot(jjres$SJR,jjres$scores,  col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
+plot(jres$SJR,jres$scores,  col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="SJR",ylab="mean score",main="Mean analysis score vs. journal metrics")
@@
 
-cor.test(jjres$scores,jjres$SJR,method="pearson")
+cor.test(jres$scores,jres$SJR,method="pearson")
 
-cor.test(jjres$scores,jjres$SJR,method="spearman")
+cor.test(jres$scores,jres$SJR,method="spearman")
 
@@
 par(mar=c(5,5,3,1))
-plot(jjres$SJR,jjres$scores,  col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
+plot(jres$SJR,jres$scores,  col = rgb(red=0,blue=0,green=0,alpha=0.2) , pch=19,bty="n",
   xlab="SJR",ylab="mean score",main="Mean analysis score vs. journal metrics")
 abline(mylm2,col="red")
+dev.off()
+
+df <- data.frame("SJR" = jres$SJR, "mean score" = jres$scores )
+p <- ggplot(df, aes(SJR, mean.score)) +
+  geom_point(alpha=0.3, size = 3) +
+  stat_smooth(method = lm,fullrange=TRUE) +
+  theme_classic()
+ggExtra::ggMarginal(p, type = "histogram")
+
+pdf("images/mean_sjr2.pdf",width=4,height=3)
+p <- ggplot(df, aes(SJR, mean.score)) +
+  geom_point(alpha=0.3, size = 3) +
+  stat_smooth(method = lm,fullrange=TRUE) +
+  theme_classic()
+ggExtra::ggMarginal(p, type = "histogram")
 dev.off()
@@
 
-cor.test(xm$scores , xm$NumCites, method="pearson")
+cor.test(xm$scores , log2(xm$NumCites+1), method="pearson")
 
-cor.test(xm$scores , xm$NumCites, method="spearman")
+cor.test(xm$scores , log2(xm$NumCites+1), method="spearman")
 
@@
 abline(mylm2,col="red")
+dev.off()
+
+df <- data.frame("log2.citations" = log2(xm$NumCites+1), "score" = xm$scores )
+
+p <- ggplot(df, aes(log2.citations, score)) +
+  geom_point(alpha=0.1, size = 3) +
+  stat_smooth(method = lm,fullrange=TRUE) +
+  theme_classic()
+ggExtra::ggMarginal(p, type = "histogram")
+
+pdf("images/cites2.pdf",width=4,height=3)
+p <- ggplot(df, aes(log2.citations,score)) +
+  geom_point(alpha=0.1, size = 3) +
+  stat_smooth(method = lm,fullrange=TRUE) +
+  theme_classic()
+ggExtra::ggMarginal(p, type = "histogram")
 dev.off()

### 2022-01-17 examples/example1.Rmd
remove mention of comb

@@
 
+m_up_nom <- subset(mres$enrichment_result,pANOVA<0.05 & s.dist > 0)[,1]
+m_dn_nom <- subset(mres$enrichment_result,pANOVA<0.05 & s.dist < 0)[,1]
+
 ```
@@
 
-c_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
+c_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
 c_up <- rownames(subset(c_up, p.adjust < 0.05))
 
-c_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
+c_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
 c_dn <- rownames(subset(c_dn, p.adjust < 0.05))
+
+c_up_nom <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none" ))
+c_up_nom <- rownames(subset(c_up_nom, pvalue < 0.05))
+
+c_dn_nom <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
+c_dn_nom <- rownames(subset(c_dn_nom, pvalue < 0.05))
 
@@
 
-Now performing ORA with clusterprofiler combining up and down.
+Now performing ORA with clusterprofiler combining up and down. (Not shown in final article)
 
@@
 
-d_de <- as.data.frame(enricher(gene = de_de, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
+d_de <- as.data.frame(enricher(gene = de_de, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
 d_de <- rownames(subset(d_de, p.adjust < 0.05))
@@
 
-f_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
+f_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
 f_up <- rownames(subset(f_up, p.adjust < 0.05))
 
-f_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2))
+f_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
 f_dn <- rownames(subset(f_dn, p.adjust < 0.05))
+
+f_up_nom <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
+f_up_nom <- rownames(subset(f_up_nom, p.adjust < 0.05))
+
+f_dn_nom <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
+f_dn_nom <- rownames(subset(f_dn_nom, p.adjust < 0.05))
 
@@
 e_de <- rownames(subset(e_de, p.adjust < 0.05))
+
+```
+
+## Bar chart of significance
+
+Here the idea is to classify the pathways as 1 not significant, 2 nominally significant, or 3 FDR significant
+
+```{r,stacked1}
+
+c_up_df <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, 
+  pAdjustMethod="fdr" ,qvalueCutoff=1,pvalueCutoff=1))
+
+n_c_up_ns <- nrow( subset(c_up_df,pvalue>0.05) )
+n_c_up_nom <- nrow( subset(c_up_df,pvalue<0.05 ) )
+n_c_up_fdr <- nrow( subset(c_up_df,p.adjust<0.05) )
+
+c_dn_df <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, 
+  pAdjustMethod="fdr" ,qvalueCutoff=1,pvalueCutoff=1))
+
+n_c_dn_ns <- nrow( subset(c_dn_df,pvalue>0.05) )
+n_c_dn_nom <- nrow( subset(c_dn_df,pvalue<0.05 ) )
+n_c_dn_fdr <- nrow( subset(c_dn_df,p.adjust<0.05) )
+
+n_m_up <- length(m_up)
+n_m_dn <- length(m_dn)
+
+n_m_up_nom <- length(m_up_nom)
+n_m_dn_nom <- length(m_dn_nom)
+
+par(mar=c(5,10,5,2))
+
+ngenes <- c("ORA up fdr"=n_c_up_fdr,"ORA up nom"=n_c_up_nom,
+  "ORA dn fdr"=n_c_dn_fdr,"ORA dn nom"=n_c_dn_nom,
+  "FCS up fdr"=n_m_up,"FCS up nom"=n_m_up_nom,
+  "FCS dn fdr"=n_m_dn,"FCS dn nom"=n_m_dn_nom )
+
+barplot(ngenes, horiz=TRUE,las=1)
+
+n_c_fdr <- length(union(c_dn,c_up))
+n_c_nom <- length(union(c_dn_nom,c_up_nom))
+n_m_fdr <- length(subset(mres$enrichment_result,p.adjustANOVA<0.05 )[,1])
+n_m_nom <- length(subset(mres$enrichment_result,pANOVA<0.05 )[,1])
+
+par(mar=c(5,5,5,2))
+
+ngenes <- c("ORA FDR<0.05"=n_c_fdr,"ORA p<0.05"=n_c_nom,"FCS FDR<0.05"=n_m_fdr,"FCS p<0.05"=n_m_nom)
+barplot(ngenes,ylab="no. gene sets")
+text((0:3*1.2)+0.7,ngenes-50,labels=ngenes,cex=1.2)
+
+pdf("images/fdr1.pdf",width=4,height=4)
+par(mar=c(5,7,2,2))
+barplot(ngenes,xlab="no. gene sets",horiz=TRUE,las=1)
+text( ngenes - 70, (0:3*1.2)+0.7,labels=ngenes,cex=1)
+dev.off()
 
@@
 
+In the next Venn I will omit combined analysis and include nominal p-value sets.
+
+```{r,venn1}
+
+par(cex.main=0.5)
+
+par(mar=c(2,2,2,2))
+
+v1 <- list("FCS up"=m_up, "FCS dn"=m_dn,
+           "ORA up"=c_up,"ORA dn"=c_dn)
+
+plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS compared to ORA")
+
+v0 <- list("FDR up"=m_up, "FDR dn"=m_dn,
+           "Nom up"=m_up_nom,"Nom dn"=m_dn_nom)
+
+plot(euler(v0),quantities = TRUE, edges = "gray", main="Effect of FDR correction on FCS results")
+
+v0 <- list("FDR up"=c_up, "FDR dn"=c_dn,
+           "Nom up"=c_up_nom,"Nom dn"=c_dn_nom)
+
+plot(euler(v0),quantities = TRUE, edges = "gray", main="Effect of FDR correction on ORA results")
+
+ora_nom <- union(c_up_nom,c_dn_nom)
+ora_fdr <- union(c_up,c_dn)
+fcs_nom <- union(m_up_nom,m_dn_nom)
+fcs_fdr <- union(m_up,m_dn)
+
+v3 <- list("ORA nom"=ora_nom, "ORA FDR"=ora_fdr,
+           "FCS nom"=fcs_nom,"FCS FDR"=fcs_fdr)
+
+plot(euler(v3),quantities = TRUE, edges = "gray", main="Effect of FDR correction")
+
+v2 <- list("ORA up"=c_up,"ORA dn"=c_dn,
+           "ORA* up"=f_up,"ORA* dn"=f_dn )
+
+plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
+
+png("images/fcs_ora1.png")
+plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
+dev.off()
+
+png("images/orabg1.png")
+plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
+dev.off()
+
+pdf("images/fcs_ora1.pdf",width=4,height=4)
+plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
+dev.off()
+
+pdf("images/orabg1.pdf",width=4,height=4)
+plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
+dev.off()
+
+```
+
 ## Jaccard calculation
@@
 
+As per reviewer suggestion, remove the comparison of combined and separated ORA.
+
+I will include using nominal p-value.
+
+```{r,jacc2}
+
+# FCS vs ORA
+cm <- length(intersect(c(c_up,c_dn), c(m_up,m_dn))) / length(union(c(c_up,c_dn), c(m_up,m_dn)))
+
+m_up <- gsub("^","up ",m_up)
+m_dn <- gsub("^","dn ",m_dn)
+m_de <- union(m_up,m_dn)
+
+c_up <- gsub("^","up ",c_up)
+c_dn <- gsub("^","dn ",c_dn)
+c_de <- union(c_up,c_dn)
+
+f_up <- gsub("^","up ",f_up)
+f_dn <- gsub("^","dn ",f_dn)
+f_de <- union(f_up,f_dn)
+
+# ORA vs ORA*
+cf <- length(intersect(c_de, f_de )) / length(union(c_de, f_de))
+
+# FCS vs ORA*
+mf <- length(intersect(m_de, f_de )) / length(union(m_de, f_de))
+
+dat <- c("FCS vs ORA"=cm,"ORA vs ORA*"=cf,"FCS vs ORA*"=mf)
+
+dat
+saveRDS(dat,file = "ex1dat.rds")
+
+par(mar=c(5,10,3,1))
+barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
+text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
+
+pdf("images/jacbar2.pdf",width=4,height=3)
+par(mar=c(5,9,3,1))
+barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.8) , main=name)
+text( x=rev(dat)+0.08 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
+dev.off()
+
+png("images/jacbar2.png")
+par(mar=c(5,10,3,1))
+barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
+text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
+dev.off()
+
+```
+
 ## Session information
